### 2016-11-15 greedy_hamiltonian/chapters/1.introduction.tex
fist commit

\section{Introduction}
Parameterized partial differential equations are very often used as tools to model many problems in engineering and the applied sciences. While the need for more accuracy has led to the development of exceedingly complex models, the computational and storage limitations often make direct approaches impossible. Hence, we must seek alternative methods that allow us to approximate the desired output under variation of the input parameters while keeping the computational costs to a minimum.

Reduced basis methods have emerged as a powerful approach for the reduction of the intrinsic complexity of such models \cite{Ito:1998up,Ito:1998ch,Ito:2001ev,Peterson:1989ki}. These methods contain two stages: the offline and the online. In the offline stage, one explors the parameter space to construct a low-dimensional basis that accurately represents the parametrized solution to the partial differential equation. In this stage, evaluation of the solution of the original model for multiple parameter values is required. The online stage comprises a Galerkin projection onto the span of the reduced basis, which allows exploration of the parameter space at significantly reduced complexity \cite{Antoulas:2005:ALD:1088857,Anonymous:2016wl}.

Convectional reduced basis techniques e.g. proper orthogonal decomposition (POD) \cite{Kunisch:2002er,Atwell:2001ja,Ravindran:2002hn}, require the exploration of the entire parameter space. This leads to a very expensive and often impractical offline stages when dealing with multi-dimensional paramtere domains. On the other hand, sampling techniques, usually of a greedy nature, search through the parameter space selectively often guided by an error estimate to certify the accuracy of the basis. This approach, accompanied with an efficient sampling procedure, trades considerable amount of computation with overall accuracy of the reduced-basis \cite{Cuong:2005gd,Rozza:2005ie,Anonymous:2016wl}.

Beside computational complexity, another aspect of reduced order modeling is the preservation of structures and, in particular, stability of the original model. In general, reduced order models do not guarantee that such properties are preserved \cite{Anonymous:pMn0O0Q4}.

In the context of Hamiltonian and Lagrangian systems, recent work suggest modifications of POD to preserve geometric structure. Lall et al. \cite{Lall:2003iy} and Carlberg et al. \cite{Carlberg:2014ky} suggests that the reduced-order system should be identified by a Lagrangian function on a low-ordered configuration space. In this way the geometric structure of the original system is inherited by the reduced system. For Hamiltonian systems Peng et al. \cite{Peng:2014di}, using a symplectic transformation, constructs a reduced Hamiltonian, as an approximation to the Hamiltonian of the original system. As a result the reduced system preserves the symplectic structure. Although these methods preserve geometric structure, they use a POD-like approach for constructing the reduced basis, and are not well suitable for problems with a high-dimensional parameter domain.

In this paper, we present a greedy approach for the construction of a reduced system that preserves the geometric structure of Hamiltonian systems, as an iterative routine that successively enriches the reduced basis at each iteration. On one hand this technique results in a reduced Hamiltonian system that mimics symplectic properties of the original system and preserves the Hamiltonian structure and stability over the course of time integration. On the other hand, since time integration of the original system is only required once per iteration, the proposed method saves substantial computational cost during the offline stage when compared to alternative POD-like approaches. Moreover, we demonstrate that assumptions, natural for the set of all solutions of the original Hamiltonian system under variation of parameters, lead to exponentially fast convergence of the greedy algorithm. For nonlinear Hamiltonian systems, we show how the basis can be combined with the discrete empirical interpolation method (DEIM) to allow a fast evaluation of nonlinear terms while maintaining the symplectic structure.

This paper is organized as follows. Section \ref{chap:MoOr:1} presents a brief overview of model order reduction, POD and DEIM. In Section \ref{chap:Hasy:1} we cover the required topics from symplectic geometry and Hamiltonian systems. Section \ref{chap:SyMo:1} discusses the greedy generation of a symplectic reduced basis as well as other SVD-based symplectic model reduction techniques. Accuracy, stability, and efficiency of the greedy method compared to other SVD based methods are discussed in Section \ref{chap:NuRe:1}. Finally we offer the conclusion in Section \ref{chap:Con:1}.

### 2016-11-15 greedy_hamiltonian/chapters/2.mor.tex
fist commit

\section{Model Order Reduction} \label{chap:MoOr:1}
Consider a parameterized, finite dimensional dynamical system that can be described by a set of first order ordinary differential equations on the form
\begin{equation} \label{eq:MoOr:1}
\left\{
\begin{split}
& \frac{d}{dt} \mathbf{x}(t,\omega) = \mathbf f (t,\mathbf x,\omega), \\
& \mathbf x(0,\omega) = \mathbf x_0(\omega).
\end{split}
\right.
\end{equation}
Here $\mathbf x \in \mathbb R^n$ is the state vector, $\omega \in \Gamma$ is a vector containing all the parameters of the system that belongs to a compact set $\Gamma$ ($\subset \mathbb R^d$) and $\mathbf f : \mathbb R \times \mathbb R^n \times \Gamma \to \mathbb R^n$ is a general function of the state variables and parameters.

We define the solution manifold as the set of all solutions to (\ref{eq:MoOr:1}) under variation of parameter $\omega$
\begin{equation} \label{eq:MoOr:2}
	\mathcal M = \{ \mathbf x(\omega) | \omega \in \Gamma \} \subset \mathbb R^n.
\end{equation}
Note that the exact solution and solution manifold is not always available; we assume that we have a numerical integrator in hand that can approximate the solution to (\ref{eq:MoOr:1}) for any realization of $\omega$ with arbitrary accuracy. By abuse of notation we refer to $\mathbf x$ and $\mathcal M$ as the exact solution and the exact solution manifold, respectively, rather than the discrete solution and discrete solution manifold.

Model order reduction is based on the assumption that $\mathcal M$ is of low dimension \cite{Anonymous:2016wl,Antoulas:2005:ALD:1088857}, and that the span of appropriately chosen basis vectors $\{v_i\}_{i=1}^k$ covers most of the solution manifold with a small error. The set $\{v_i\}_{i=1}^k$ is denoted as the reduced basis and its span as the reduced space. Assuming that a $k$-dimensional $(k\ll n)$ reduced basis is given, the approximated solution can be represented as
\begin{equation} \label{eq:MoOr:3}
	\mathbf x \approx V \mathbf y,
\end{equation}
where $V$ is a matrix containing the reduced basis vectors as its columns and $\mathbf y$ is the coordinates of the approximation in this basis. By substituting (\ref{eq:MoOr:3}) into (\ref{eq:MoOr:1}) we obtain the overdetermined system
\begin{equation} \label{eq:MoOr:4}
	V \frac{d}{dt} \mathbf y = \mathbf f (t , V \mathbf y , \omega) + \mathbf r(t,\omega).
\end{equation}
Here we added the residual $\mathbf r$ to emphasise that (\ref{eq:MoOr:4}) is an approximation of (\ref{eq:MoOr:1}). Taking the Petrov-Galerkin projection \cite{Antoulas:2005:ALD:1088857} we construct a basis $W$ of size $n-k$ that is orthogonal to the residual $\mathbf r$ and assume $W^T V$ is invertible. This yields
\begin{equation} \label{eq:MoOr:5}
	\frac{d}{dt} \mathbf y = (W^TV)^{-1} \mathbf f(t,V\mathbf y,\omega).
\end{equation}
Equation (\ref{eq:MoOr:5}) consists of $k$ equations and is called the reduced system. Solving the reduced system instead of the original system can reduce the computational costs since $k$ is significantly smaller than $n$. For nonlinear systems, the evaluation of $\mathbf f$ may still have computational complexity that depends on $n$. We return to this question in detail in Section \ref{chap:MoOr.DEIM:1}.

\subsection{Proper Orthogonal Decomposition} \label{chap:MoOr.PrOr:1}
Let $\mathbf x (t_i,\omega_j)$ with $i=1,\dots,m$ and $j=1,\dots,n$ be a finite number of samples, referred to as snapshots, from the solution manifold (\ref{eq:MoOr:2}). If we Suppose that a reduced basis $V$ is provided, the projection operator from $\mathbb R^n$ onto the reduced space can be constructed as $VV^T$. The proper orthogonal decomposition (POD) requires the total error of projecting all the snapshots onto the reduced space to be minimum. The POD basis of size $k$ is thus the solution of the optimization problem
\begin{equation} \label{eq:MoOr:6}
\begin{aligned}
& \underset{V\in \mathbb R^{n\times k}}{\text{minimize}}
& & \| S - VV^TS\|_F \\
& \text{subject to}
& & V^TV = I_k
\end{aligned}
\end{equation}
Here $S$ is the snapshot matrix,, containing snapshots $\mathbf x(t_i,\omega_j)$ in its columns, $\|\cdot \|_F$ is the Frobenius norm of a matrix and $I_k$ is the identity matrix of size $k$. According to Schmidt-Mirsky-Eckard-Young theorem \cite{Antoulas:2005:ALD:1088857}, the solution to the optimization (\ref{eq:MoOr:6}) is equivalent to the truncated singular value decomposition (SVD) of the snapshot matrix $S$ given by
\begin{equation} \label{eq:MoOr:7}
	V = \sigma_1 u_1 v^T_1 + \cdots + \sigma_k u_k v^T_k.
\end{equation}
Here $\sigma_i, u_i$ and $v_i$ are the singular values, left singular vectors and right singular vectors of $S$, respectively.

\subsection{Discrete Empirical Interpolation Method (DEIM)} \label{chap:MoOr.DEIM:1} \nocite{Chaturantabut:2010cz}
Let us now discuss the inefficiency of evaluating nonlinearities when using projection based reduced model. Suppose that the right hand side function in (\ref{eq:MoOr:1}) is of the form $\mathbf f(t,\mathbf x , \omega) = L\mathbf x + \mathbf g(t,\mathbf x ,\omega)$, where $L\in \mathbb R^{n\times n}$ reflects the linear part, and $\mathbf g$ is a nonlinear function. Now suppose a $k$-dimensional reduced basis $V$ is provided, the reduced system takes the form
\begin{equation} \label{eq:MoOr:8}
	\frac{d}{dt} \mathbf y = \underbrace{(WV)^{-1} L V}_{\tilde L} \mathbf{y} + \underbrace{(WV)^{-1} \mathbf g(t,V\mathbf y,\omega)}_{\tilde N (\mathbf y)}.
\end{equation}
Here, $\tilde L$ is a $k\times k$ matrix which can be computed before time integration of the reduced system. However, evaluation of $\tilde N (\mathbf y)$ has a complexity that depends on $n$, the size of the original system. Suppose that evaluation of $\mathbf g$ with $n$ components has complexity $\alpha(n)$, for some function $\alpha$. Then the complexity of evaluating $\tilde N(\mathbf y)$ is $\mathcal{O}(\alpha(n) + 4nk)$ which consists of 2 matrix-vector operations and the evaluation of the nonlinear function. This means that estimation of the nonlinear terms can be as expensive as solving the original system.

To overcome this bottleneck we take an approach similar to what we did above. Assume that the manifold $\mathcal M_{\mathbf g} = \{ \mathbf g(t,\mathbf x , \omega)| t\in \mathbb R, \mathbf x \in \mathbb R , \omega \in \Gamma\}$ is of a low dimension and that $\mathbf g$ can be approximated by a linear subspace of dimension $m\ll n$, spanned by the basis $\{ u_1 , \dots , u_m \}$, i.e.
\begin{equation} \label{eq:MoOr:10}
	\mathbf g(t,\mathbf x,\omega) \approx U \mathbf c(t,\mathbf x,\omega).
\end{equation}
Here $U$ contains basis vectors $u_i$ and $\mathbf c$ is the corresponding vector of coefficients. Now suppose $p_1,\dots,p_m$ are $m$ indices from $\{1,\dots,n\}$ and define an $n\times m$ matrix
\begin{equation} \label{eq:MoOr:11}
	P = [e_{p_1},\dots,e_{p_m}],
\end{equation}
where $e_{p_i}$ is the $p_i$-th column of the identity matrix $I_n$. Hence, multiplying $P$ on $\mathbf g$ selects components $p_1,\dots,p_m$ of $\mathbf g$. If we assume that $P^TU$ is non-singular, the coefficient vector $\mathbf c$ can be uniquely determined from
\begin{equation} \label{eq:MoOr:12}
	P^T \mathbf g = (P^TU)\mathbf c.
\end{equation}
Finally the approximation of $\mathbf g$ is determined by
\begin{equation} \label{eq:MoOr:13}
	\mathbf g(t,\mathbf x,\omega) \approx U \mathbf c(t,\mathbf x,\omega) = U (P^TU)^{-1} P^T \mathbf g(t,\mathbf x,\omega),
\end{equation}
which is referred to as the \emph{Discrete Empirical Interpolation} (DEIM) approximation \cite{Chaturantabut:2010cz}. Applying DEIM on the reduced system (\ref{eq:MoOr:5}) yields
\begin{equation} \label{eq:MoOr:14}
	\frac{d}{dt} \mathbf y = \tilde L \mathbf y + (WV)^{-1} U(P^TU)^{-1}P^T \mathbf g(t,V\mathbf y , \omega).
\end{equation}
Note that the matrix $(WV)^{-1} U(P^TU)^{-1}$ can be computed offline and since $\mathbf g$ is evaluated only at $m$ of its components, the evaluation of the nonlinear term in (\ref{eq:MoOr:14}) does not depend on $n$.

To obtain the projection basis $U$, the POD can be applied to the ensemble of the discrete evaluation of $\mathbf g$. There is no additional cost for computing the nonlinear snapshots since they are generated when computing the trajectory snapshot matrix $S$. The interpolating indices $p_1,\dots,p_m$ can be constructed as follows. Given the projection basis $U = \{u_1,\dots,u_m\}$, the first interpolation index $p_1$ is chosen according to the component of $u_1$ with the largest magnitude. The rest of the interpolation indices, $p_2,\dots,p_m$ correspond to the component of the largest magnitude of the residual vector $\mathbf r = u_l - U \mathbf c$. It is shown in \cite{Chaturantabut:2010cz} that if the residual vector is a nonzero vector in each iteration then $P^TU$ is non-singular and (\ref{eq:MoOr:13}) is well defined. We summarize the algorithm of generating interpolating indices in Algorithm \ref{alg:MoOr:1}.

\begin{algorithm} 
\caption{Discrete Empirical Interpolation Method} \label{alg:MoOr:1}
{\bf Input:}  Basis vectors $\{u_1,\dots , u_m\}\subset \mathbb R^n$
\begin{enumerate}
\item pick $p_1$ to be the index of the largest component of $u_1$.
\item $U \leftarrow [u_1]$
\item $P \leftarrow [p_1]$
\item \textbf{for} $i\leftarrow 2$ \textbf{to} $m$
\item \hspace{0.5cm} solve $(P^TU)\mathbf c = P^T u_i$ for $\mathbf c$
\item \hspace{0.5cm} $\mathbf r \leftarrow u_i - U\mathbf c$
\item \hspace{0.5cm} pick $p_i$ to be the index of the largest component of $\mathbf r$
\item \hspace{0.5cm} $U \leftarrow [u_1,\dots,u_i]$
\item \hspace{0.5cm} $P \leftarrow [p_1,\dots,p_i]$
\item \textbf{end for}
\end{enumerate}
\vspace{0.5cm}
{\bf Output:} Interpolating indices $\{p_1,\dots,p_m\}$
\end{algorithm}

If we use an implicit time integration scheme, a similar inefficiency occurs for the computation of Newton iteration. In addition to the nonlinear term, the Jacobian of $\mathbf g(\mathbf x)$ with respect to the reduced state vector must be computed at each iteration. The Jacobian takes the form
\begin{equation} \label{eq:MoOr:9}
	\mathbf J_{\mathbf y}(\mathbf g) = (WV)^{-1} \mathbf J_{\mathbf x}(\mathbf g) V.
\end{equation}
The complexity of computing (\ref{eq:MoOr:9}) is $\mathcal{O}(\alpha(n) +2n^2k+2nk^2+2nk)$ comprising several matrix-vector multiplications and an evaluation of the Jacobian which depends on the size of the original system. Approximating the Jacobian in (\ref{eq:MoOr:9}) is usually both problem and discretization dependent. Very often the nonlinear function $\mathbf g$ is evaluated component-wise i.e.
\begin{equation} \label{eq:MoOr:15}
	\mathbf g(\mathbf x) =
	\begin{pmatrix}
		g_1(x_1,\dots,x_n) \\
		g_2(x_1,\dots,x_n) \\
		\vdots \\
		g_n(x_1,\dots,x_n)
	\end{pmatrix}
	=
	\begin{pmatrix}
		g_1(x_1) \\
		g_2(x_2) \\
		\vdots \\
		g_n(x_n)
	\end{pmatrix}.
\end{equation}
with $\mathbf x$ the configuration vector. In such cases the interpolating index matrix $P$ and the nonlinear function $\mathbf g$ commute. I.e., we have
\begin{equation} \label{eq:MoOr:16}
	\tilde N(\mathbf y) \approx (WV)^{-1} U(P^TU)^{-1}P^T \mathbf g(V\mathbf y) = (WV)^{-1} U(P^TU)^{-1}\mathbf g(P^TV\mathbf y)
\end{equation}
If we now take the Jacobian of the approximate function we recover
\begin{equation}
	\mathbf J_{\mathbf y}(\mathbf g) = \underbrace{ (WV)^{-1} U(P^TU)^{-1} }_{k\times m} \underbrace{ \mathbf J_{\mathbf x}(\mathbf g(P^T V \mathbf y) ) }_{m\times m} \underbrace{P^T V}_{k\times k}.
\end{equation}
The matrix $(WV)^{-1} U(P^TU)^{-1}$ can be computed offline and the Jacobian is evaluated only at $m\times m$ components. Hence the overall complexity of computing the Jacobian is now independent of $n$.

### 2016-11-15 greedy_hamiltonian/chapters/3.ham.tex
fist commit

\section{Hamiltonian Systems} \label{chap:Hasy:1}
Let $P$ be a manifold and $\Omega:P\times P \to \mathbb R$ be a closed, nondegenerate two-form on $P$. The pair $(P,\Omega)$ is called a \emph{symplectic manifold}. The vector field $X_H:P \to TP$ is called a \emph{Hamiltonian vector field} if there exists a function $H:P\to \mathbb R$ such that for every vector $v\in T_zP$ we have
\begin{equation} \label{eq:Hasy:1}
	\Omega_z(X_H(z),v) = \mathbf d H(z) \cdot v.
\end{equation}
where $\mathbf d H(z) \cdot v$ is the directional derivative of $H$ along $v$. Note that when $P$ belongs to a Euclidean space then $\mathbf d H = \nabla_z H$. The equations of evolution are then defined by
\begin{equation} \label{eq:Hasy:2}
	\dot z = X_H(z)
\end{equation}
and known as the \emph{Hamilton's equation} \cite{Marsden:1999ck}. A fundamental feature of Hamiltonian systems is the conservation of the Hamiltonian along integral curves on $P$. To emphasise the importance of this property we state

\begin{theorem} \label{theorem:Hasy:1}
Suppose that $X_H$ is a Hamiltonian vector field with flow $\phi_t$ on a symplectic manifold $P$. Then $H\circ \phi_t = H$.
\end{theorem}

\begin{proof}
We show that $H$ is constant along integral curves
\begin{equation} \label{eq:Hasy:3}
\begin{aligned}
	\frac{d}{dt}(H\circ \phi_t)(z) &= \mathbf d H(\phi_t(z)) \cdot( \frac{d}{dt} \phi_t(z) ) \\
	&= \mathbf d H (\phi_t(z))\cdot X_H(\phi_t(z)) \\
	&= \Omega_z( X_H(\phi_t(z)), X_H(\phi_t(z)) ) = 0,
\end{aligned}
\end{equation}
where we used the chain rule in the first line and bilinearity of $\Omega$ in the argument.
\end{proof}

For the case where the symplectic manifold is also a linear vector space, the pair $(P,\Omega)$ is also referred to as a \emph{symplectic vector space}. We state the following theorems regarding symplectic vector spaces without proof. We refer the reader to \cite{Marsden:1999ck,de2006symplectic,Silva01lectureson} for detailed proofs.

\begin{itemize}
\item If $P$ is a vector space then $\Omega$ is a constant form, that is $\Omega_z$ is independent of $z\in P$. 
\item If $(P,\Omega)$ is a finite-dimensional symplectic manifold then $P$ is even dimensional.
\item If $(P,\Omega)$ is a $2n$-dimensional symplectic vector space, then there is a basis $e_1,\dots e_n,f_1, \dots , f_n$ of $P$ such that
\begin{equation} \label{eq:Hasy:4}
\begin{aligned}
	& \Omega(e_i,e_j) = 0 = \Omega(f_i,f_j), \quad & i\neq j,\\
	& \Omega(e_i,f_j) = \delta_{ij}, & i\leq i,j \leq n.
\end{aligned}
\end{equation}
where $\delta$ is the Kronecker's delta function. Moreover if $P = \mathbb{R}^{2n}$ we can choose the basis vectors $\{e_i,f_i\}_{i=1}^n$ such that
\begin{equation} \label{eq:Hasy:5}
	\Omega(v_1,v_2) = v_1^T \mathbb J_{2n} v_2, \qquad v_1,v_2\in \mathbb R^n,
\end{equation}
with $\mathbb J_{2n}$ being the symplectic matrix, defined as
\begin{equation} \label{eq:Hasy:6}
	\mathbb{J}_{2n} = 
	\begin{pmatrix}
		0_n & I_n \\
		-I_n & 0_n
	\end{pmatrix},
\end{equation}
Here $I_n$ and $0_n$ is the identity matrix and the zero square matrix of size $n$, respectively.
\item On a finite-dimensional symplectic vector space the relationship (\ref{eq:Hasy:1}) becomes
\begin{equation} \label{eq:Hasy:7}
\left\{
\begin{aligned}
	&\dot {\mathbf z} = \mathbb{J}_{2n} \nabla_{\mathbf z} H(\mathbf z), \\
	& \mathbf z(0) = \mathbf z_0.
\end{aligned}
\right.
\end{equation}
or, by introducing the canonical coordinates $\mathbf z = (\mathbf q^T, \mathbf p^T)^T$,
\begin{equation} \label{eq:Hasy:8}
\left\{
\begin{aligned}
	&\dot {\mathbf q} = \nabla_{\mathbf p} H(\mathbf q,\mathbf p),\\
	&\dot {\mathbf p} = - \nabla_{\mathbf q} H(\mathbf q,\mathbf p).
\end{aligned}
\right.
\end{equation}
\end{itemize}
We now introduce \emph{symplectic transformations}, i.e. mappings between symplectic manifolds which preserve the two-form $\Omega$. Accurate numerical treatment of Hamiltonian systems often requires preservation of the symmetry expressed in Theorem \ref{theorem:Hasy:1}. Symplectic transformations can be used to construct such symmetry preserving numerical methods.

\begin{definition}
Let $(P,\Omega)$ and $(Q,\Pi)$ be two symplectic manifolds. A $C^\infty$-mapping $\phi:P\to Q$ is called symplectic or canonical if, for any $u,v \in T_zP$ where $z\in P$, we have
\begin{equation} \label{eq:Hasy:9}
	\Omega_z(u,v) = \Pi_{\phi(z)}(T_z\phi\cdot u,T_z \phi \cdot v)
\end{equation}
where $T_z \phi$ is the derivative of $\phi$ at $z$.
\end{definition}

If $(P,\Omega)$ and $(Q,\Pi)$ are symplectic linear vector spaces of size $2k$ and $2n$, respectively, then we can represent a linear mapping $\phi:P\to Q$ on matrix form
\begin{equation} \label{eq:Hasy:10}
	\phi(x) = Ax, \quad x\in P\text{ and }A\in \mathbb{R}^{2n\times 2k}.
\end{equation}
As tangent spaces $T_zP$ and $T_zQ$ coincide with $P$ and $Q$, we conclude that $T_z\phi(u) = Au$ for all $u\in T_zP$. Hence condition (\ref{eq:Hasy:9}) will be equivalent to
\begin{equation} \label{eq:Hasy:11}
	A^T \mathbb{J}_{2n}A = \mathbb{J}_{2k}.
\end{equation}
A matrix of size $2n\times 2k$ satisfying (\ref{eq:Hasy:11}) is called a \emph{symplectic matrix}.

\begin{definition}
	The \emph{symplectic inverse} of a matrix $A\in \mathbb{R}^{2n\times 2k}$ is denoted by $A^+$ and defined by
\begin{equation}\label{eq:Hasy:12}
	A^+ := \mathbb{J}_{2k}^T A^T \mathbb{J}_{2n}.
\end{equation}
\end{definition}
We point out the properties of the symplectic inverse and refer the reader to \cite{Peng:2014di} for detailed proof.
\begin{lemma} \label{lemma:Hasy:1}
Let $A\in \mathbb{R}^{2n\times 2k}$ be a symplectic matrix and $A^+$ its symplectic inverse as defined in (\ref{eq:Hasy:12}). Then ${(A^+)}^T$ is a symplectic matrix and $A^+A = I_{2k}$.
\end{lemma}

It is straight forward to verify that $AA^+$ is idempotent, i.e. a symplectic projection onto the column span of $A$.

Common numerical integrators e.g., Runge-Kutta methods, do not preserve the symplectic structure, i.e. The Liouville measure, of Hamiltonian systems \cite{Hairer:1250576,Marsden:1999ck}. This is often observed in the decay or the blow up of the Hamiltonian. Symplectic transformations can be used to construct a numerical scheme which does conserve symplectic structure, resulting a stable system under long time integration \cite{Hairer:1250576}. The Str\"omer-Verlet time stepping scheme is an example of symplectic integrators and is given by
\begin{equation} \label{eq:Hasy:13}
\begin{aligned}
	q_{n+1/2} &= q_n + \frac{\Delta t}{2} \nabla_pH(q_{n+1/2},p_n), \\
	p_{n+1} &= p_n - \frac{\Delta t}{2} \left( \nabla_qH(q_{n+1/2},p_n) + \nabla_qH(q_{n+1/2},p_{n+1}) \right),\\
	q_{n+1} &= q_{n+1/2} + \frac{\Delta t}{2} \nabla_pH(q_{n+1/2},p_{n+1}),
\end{aligned}
\end{equation}
and
\begin{equation} \label{eq:Hasy:14}
\begin{aligned}
	p_{n+1/2} &= p_n - \frac{\Delta t}{2} \nabla_qH(q_{n},p_{n+1/2}), \\
	q_{n+1} &= q_n + \frac{\Delta t}{2} \left( \nabla_pH(q_{n},p_{n+1/2}) + \nabla_pH(q_{n+1},p_{n+1/2}) \right),\\
	p_{n+1} &= p_{n+1/2} - \frac{\Delta t}{2} \nabla_qH(q_{n+1},p_{n+1/2}).
\end{aligned}
\end{equation}
For a general Hamiltonian system, Str\"omer-Verlet scheme is implicit. However, for separable Hamiltonians, i.e. $H(q,p) = K(p) + U(q)$, this scheme become explicit. We refer the reader to \cite{Hairer:1250576} for more information about the construction and applications of symplectic and geometric numerical integrators.

### 2016-11-15 greedy_hamiltonian/chapters/4.sympmor.tex
fist commit

\section{Symplectic Model Reduction} \label{chap:SyMo:1}
We now propose how to modify reduced order modeling to ensure that they preserve the symplectic structure of Hamiltonian system.

Consider a Hamiltonian system (\ref{eq:Hasy:7}) on a $2n$-dimensional symplectic vector space $(P,\Omega)$. Suppose that the solution manifold $\mathcal M_H$ is well approximated, by a low dimensional symplectic subspace $(Q,\Omega)$ of dimension $2k$ $(k\ll n)$. We can construct a symplectic basis $A$ for $Q$ and approximate the solution to (\ref{eq:Hasy:7}) as
\begin{equation} \label{eq:SyMo:1}
	\mathbf z \approx A\mathbf y.
\end{equation}
Substituting this into (\ref{eq:Hasy:7}) we obtain
\begin{equation} \label{eq:SyMo:2}
	A\mathbf y = \mathbb{J}_{2n} \nabla_{\mathbf z} H(A \mathbf y). 
\end{equation}
Multiplying both sides with the symplectic inverse of $A$ and using the chain rule we have,
\begin{equation} \label{eq:SyMo:3}
	\mathbf y = A^+ \mathbb J_{2n} (A^+)^T \nabla_{\mathbf y} H(A\mathbf y).
\end{equation}
Since $A$ is a symplectic basis, Lemma \ref{lemma:Hasy:1} ensures that $(A^+)^T$ is a symplectic matrix i.e. $A^+ \mathbb J_{2n} (A^+)^T = \mathbb{J}_{2k}$. By defining the reduced Hamiltonian $\tilde H:\mathbb R^{2k} \to \mathbb R$ as $\tilde H (y) = H(Ay)$ we obtain the reduced system
\begin{equation} \label{eq:SyMo:4}
\left\{
\begin{aligned}
	 \frac{d}{dt} \mathbf y &= \mathbb J_{2k} \nabla_{\mathbf y} \tilde H(\mathbf y), \\
	 \mathbf y_0 &= A^+ \mathbf z_0.
\end{aligned}
\right.
\end{equation}
The system obtained from the Petrov-Galerkin projection in (\ref{eq:MoOr:5}) is not a Hamiltonian system and therefore does not guarantee conservation of the symplectic structure. On the other hand we observe that the reduced system in (\ref{eq:SyMo:4}) is of the form (\ref{eq:Hasy:7}) and, hence, is a Hamiltonian system, i.e. the symplectic structure will be conserved along integral curves of (\ref{eq:SyMo:4}). Note that the original and the reduced systems are derived from different Hamiltonians. In the next proposition we show that the loss in the Hamiltonian is constant in time.

\begin{proposition}
Let $\mathbf{z} (t)$ be the solution of (\ref{eq:Hasy:7}) at time $t$. Further suppose that $\tilde{\mathbf{z}} (t)$ is the approximate solution of the reduced system (\ref{eq:SyMo:4}) in the original coordinate system. Then the error in the Hamiltonian defined by
\begin{equation} \label{eq:SyMo:5}
	\Delta H(t)  = |H(\mathbf z(t)) - H(\tilde{\mathbf z}(t))|,
\end{equation}
is constant for all $t\in \mathbb R$.
\end{proposition}

\begin{proof}
	Let $\phi_t$ and $\psi_t$ be the Hamiltonian flow of the original and the reduced system respectively. By definition $\mathbf z(t) = \phi_t(\mathbf z_0)$ and $\mathbf y(t) = \psi_t(\mathbf y_0)$. Using Theorem \ref{theorem:Hasy:1} we have
\begin{equation} \label{eq:SyMo:6}
\begin{aligned}
	H(\tilde{\mathbf{z}} (t)) = H( A\mathbf y (t) ) = \tilde H(\mathbf y (t)) = \tilde H(\psi_t(\mathbf y_0)) = \tilde H(\mathbf y_0) = \tilde H(A^+ \mathbf z_0) = H(AA^+\mathbf z_0).
\end{aligned}
\end{equation}
Also $H(\phi_t(\mathbf z_0)) = H(\mathbf z_0)$. The error in the Hamiltonian can be written in terms of $\mathbf z_0$ and the symplectic basis $A$ as
\begin{equation} \label{eq:SyMo:7}
	\Delta H(t) = |H(\mathbf z_0) - H(AA^+\mathbf z_0)|
\end{equation}
\end{proof}

The following theorems guarantees stability of the original and the reduced Hamiltonian systems. We refer the reader to \cite{Peng:2014di} for the proofs.

\begin{theorem} \label{theorem:SyMo:1}
Consider the initial value problem (\ref{eq:Hasy:7}) with its symplectic reduced system (\ref{eq:SyMo:4}). If we can find a bounded neighborhood $N$ of $z_0$ in $P$ such that $H(z) < H(z_0)$ or $H(z) > H(z_0)$, for all $z\in \partial N$, then the original system and the reduced system are bounded for all $t\in \mathbb R$.
\end{theorem}

\begin{theorem} \label{theorem:SyMo:2}
If the original system (\ref{eq:Hasy:7}) is a linear Hamiltonian system, i.e. $H(x) = \frac 1 2 x^T L x$ with $L$ a $2n\times 2n$ real symmetric matrix, then both the original system and the reduced system (\ref{eq:SyMo:4}) are stable.
\end{theorem}

While the symplectic structure is not guaranteed to be preserved in the reduced systems obtained by the Petrov-Galerkin projection, the reduced system obtained by the symplectic projection guarantees the preservation of the energy up to the error in the Hamiltonian (\ref{eq:SyMo:5}). In the next section we discuss  different methods for obtaining a symplectic basis.

\subsection{Proper Symplectic Decomposition (PSD)} \label{chap:SyMo.PrSy:1}

Similar to Section \ref{chap:MoOr.PrOr:1} we gather snapshots $\mathbf z_i = [q_i^T , p_i^T]^T$ in the snapshot matrix $S$. Suppose that a symplectic basis $A$ of size $2n\times2k$ and its symplectic inverse $A^+$ is provided. The proper symplectic decomposition requires that the error of the symplectic projection onto the symplectic subspace to be minimum. Analogous to the minimization (\ref{eq:MoOr:6}), the PSD symplectic basis of size $2k$ is the solution of the optimization problem

\begin{equation} \label{eq:SyMo:8}
\begin{aligned}
& \underset{V\in \mathbb R^{2n\times 2k}}{\text{minimize}}
& & \| S - AA^+S\|_F \\
& \text{subject to}
& & A^T \mathbb{J}_{2n}A = \mathbb{J}_{2k}
\end{aligned}
\end{equation}
Note $A$ is not necessarily an orthogonal matrix. Compared to POD, in (\ref{eq:SyMo:8}) the orthogonal projection is replaced with a symplectic projection $AA^+$. At first, the minimization looks similar to the one obtained by POD. However, the lack of an SVD-like decomposition for symplectic subspaces, makes the minimization (\ref{eq:SyMo:8}) a harder problem than (\ref{eq:MoOr:6}).

As the optimization problem (\ref{eq:SyMo:8}) is nonlinear, direct solution usually expensive. A simplified version of the optimization (\ref{eq:SyMo:8}) can be found in \cite{Peng:2014di}, but there is no guarantee that the method provides a near optimal basis. In the following we outline non-direct methods for finding solutions to (\ref{eq:SyMo:8}), proposed by \cite{Peng:2014di}, while assuming a specific structure for $A$. Finally, we introduce a greedy approach for symplectic basis generation.

\subsubsection{SVD Based Methods for Symplectic Basis Generation} \label{chap:SyMo.PrSy:2}

\paragraph{\bf Cotangent lift} Suppose that $A$ is of the form
\begin{equation} \label{eq:SyMo:9}
	A = 
	\begin{pmatrix}
		\Phi & 0 \\
		0 & \Phi
	\end{pmatrix},
\end{equation}
where $\Phi \in \mathbb{R}^{n\times k}$ is an orthonormal matrix. It is easy to check that $A$ is a symplectic matrix, i.e., $A^T \mathbb J_{2n} A = \mathbb J_{2k}$. The construction of $A$ suggests that the range of $\Phi$ should cover both the potential and the momentum spaces. Hence, we can construct $A$ by forming the combined snapshot matrix
\begin{equation} \label{eq:SyMo:10}
	S_{\text{combined}} = [q_1,\dots,q_n,p_1,\dots,p_n], \qquad \mathbf z_i = (q_i^T,p_i^T)^T,
\end{equation}
and define $\Phi=[u_i,\dots,u_k]$, where $u_i$ is the $i$-th left singular vector of $S_{\text{combined}}$. It is shown in \cite{Peng:2014di} that among all symplectic bases of the form (\ref{eq:SyMo:9}) cotangent lift minimizes the projection error.

%Algorithm \ref{alg:SyMo:1} summarizes the cotengent lift method for finding a symplectic basis.
%
%\begin{algorithm} 
%\caption{Real SVD method for symplectic basis generation} \label{alg:SyMo:1}
%{\bf Input:} Snapshots $\{ q_i , p_i \}_{i=1}^N$ of the solution manifold and $2k$ the size of the symplectic basis.
%\begin{enumerate}
%\item Construct the combined snapshot matrix $S_{\text{combined}}$ in (\ref{eq:SyMo:10}).
%\item Compute the SVD decomposition of $S_{\text{combined}}=U\Sigma V^T$.
%\item Define $\Phi = [u_1,\dots,u_k]$ with $u_i$ the $i$-th left singular vector.
%\item Construct $A$ according to (\ref{eq:SyMo:9}).
%\end{enumerate}
%\vspace{0.5cm}
%{\bf Output:} Symplectic matrix $A$ of the size $2n\times 2k$.
%\end{algorithm}
%

\paragraph{\bf Complex SVD} We now consider another form of symplectic basis $A$. Suppose that $A$ takes the form
\begin{equation} \label{eq:SyMo:11}
	A = 
	\begin{pmatrix}
		\Phi & -\Psi \\
		\Psi & \Phi
	\end{pmatrix},
\end{equation}
such that $\Phi$ and $\Psi$ are real matrices of the size $n\times k$ satisfying conditions
\begin{equation} \label{eq:SyMo:12}
\Phi^T \Phi + \Psi^T \Psi = I_k,\quad \Phi^T \Psi = \Psi^T \Phi.
\end{equation}
It can be checked that $A$ forms a symplectic matrix. To construct $A$ we first define the complex snapshot matrix
\begin{equation} \label{eq:SyMo:13}
	S_{\text{complex}} = [ q_1 + i p_1, \dots , q_N + i p_N ].
\end{equation}
The SVD decomposition of $S_{\text{complex}}$ can be written as $S_{\text{complex}} = U \Sigma V^H$ with the superscript $H$ indicating the conjugate transpose of a complex valued matrix. Each left singular vector of $S_{\text{complex}}$ takes the form $u_m = r_m + i s_m$. We define
\begin{equation} \label{eq:SyMo:14}
	 \Phi = [r_1,\dots, r_k], \quad \Psi = [s_1,\dots, s_k].
\end{equation}
Then we can easily check that conditions (\ref{eq:SyMo:12}) is satisfied since $U$ is unitary. It is shown in \cite{Peng:2014di} that among all symplectic bases of the form (\ref{eq:SyMo:11}) the complex SVD minimizes the projection error.

%Algorithm \ref{alg:SyMo:2} summarizes the complex SVD method for finding a symplectic basis.
%
%\begin{algorithm} 
%\caption{Complex SVD method for symplectic basis generation} \label{alg:SyMo:2}
%{\bf Input:} Snapshots $\{ q_i , p_i \}_{i=1}^N$ of the solution manifold and $2k$ the size of the symplectic basis.
%\begin{enumerate}
%\item Construct the complex snapshot matrix $S_{\text{complex}}$ in (\ref{eq:SyMo:13}).
%\item Compute the complex SVD decomposition $S_{\text{complex}} = U \Sigma V^H$.
%\item Define $\Phi$ and $\Psi$ as in (\ref{eq:SyMo:14}).
%\item Construct $A$ according to (\ref{eq:SyMo:11}).
%\end{enumerate}
%\vspace{0.5cm}
%{\bf Output:} Symplectic matrix $A$ of the size $2n\times 2k$.
%\end{algorithm}

\subsubsection{\bf The Greedy Approach to Symplectic Basis Generation} The greedy generation of the reduced basis is an iterative procedure which, in each iteration, adds the two best possible basis vectors to the symplectic basis to enhance overall accuracy. In contrast to the cotangent lift and the complex SVD methods, the greedy approach does not require the symplectic basis to have a specific structure. This results in more compact bases and/or more accurate reduced systems. In addition, for parametric problems, the greedy approach only requires one numerical solution to be computed per iteration which saves substantial computational cost in the offline stage.

{\bf Symplectic Gram-Schmidt:} As discussed in Section \ref{chap:Hasy:1}, any finite dimensional linear symplectic vector space has a symplectic basis that satisfy conditions (\ref{eq:Hasy:4}). Suppose that $S=\{ e_1 , \dots , e_k \} \bigcup \{ f_1 , \dots , f_k \}$ is a symplectic basis and $a$ and $b$ are two vectors such that $a,b \not \in \text{span}(S)$ and $a \nparallel b$. We seek to grow the size of $S$ by the vectors $a$ and $b$. First we modify $a$ and $b$ to ensure they are symplectic (symplectic orthogonal) to $S$, i.e., we add a linear combination of elements of $S$ to $a$ and $b$, so the resulting vectors are symplectic to $S$. This can be achieved since we assumed $a,b\not \in \text{span}(S)$. So for $a$ we seek $\alpha_1,\dots,\alpha_k,\beta_1,\dots,\beta_k \in \mathbb R$ such that
\begin{equation}
	\Omega\left(a + \sum_{i=1}^k \alpha_i e_i + \sum_{i=1}^k \beta_i f_i , \sum_{i=1}^k \bar{\alpha}_i e_i + \sum_{i=1}^k \bar{\beta}_i f_i \right) = 0,
\end{equation}
for all possible $\bar{\alpha}_1,\dots,\bar{\alpha}_k,\bar{\beta}_1,\dots,\bar{\beta}_k \in \mathbb R$. It is easily seen that the unique solution is as 
\begin{equation}
	\alpha_i = - \Omega(a,f_i), \quad \beta_i = \Omega(a,e_i),
\end{equation}
for $i=1,\dots,k$. Now we define the modified vectors as
\begin{equation}
\begin{aligned}
	& \tilde a = a - \sum_{i=1}^k \Omega(a,f_i) e_i + \sum_{i=1}^k \Omega(a,e_i) f_i, \\
	& \tilde b = b - \sum_{i=1}^k \Omega(b,f_i) e_i + \sum_{i=1}^k \Omega(b,e_i) f_i.
\end{aligned}
\end{equation}
Finally we construct the new basis vectors as
\begin{equation}
\begin{aligned}
	& e_{k+1} = \text{sign}(\Omega(\tilde a , \tilde b)) \cdot \frac{\tilde a}{ \sqrt{ |\Omega(\tilde a , \tilde b)| } }, \quad f_{k+1} = \frac{\tilde b}{ \sqrt{ |\Omega(\tilde a , \tilde b)| } }.
\end{aligned}
\end{equation}
One easily checks that the new basis vectors $e_1,\dots,e_{k+1},f_1,\dots,f_{k+1}$ satisfy all the conditions (\ref{eq:Hasy:4}). Similar to the QR decomposition, the symplectic Gram-Schmidt process leads to a matrix decomposition for any full ranked matrix of size $2n\times 2k$. More details about this matrix decomposition are given in the Appendix.

In the case of finite symplectic vector fields, there is an alternative way to update the symplectic basis $S$. To illustrate, we first note that if $e_i$ or $f_i$ is a basis vector in $S$, we have $\mathbb J_{2n}e_i = -f_i$ and $\mathbb J_{2n}f_i = e_i$. Since
\begin{equation}
	e_i^T\mathbb J_{2n} f_i = \Omega(e_i,f_i) = 1 \Longleftrightarrow \mathbb J_{2n} e_i = - f_i \text{ and } \mathbb J_{2n} f_i = e_i.
\end{equation}
Thus if we consider a vector $v\not \in \text{span}(S)$ then $\mathbb J_{2n}^T v $ also does not belong to span$(S)$. Since, if it did, we could write the vector as a linear combination of basis vectors of $S$ as
\begin{equation}
	\mathbb J_{2n}^T v = \sum_{i=1}^k \alpha_i e_i + \sum_{i=1}^k \beta_i f_i.
\end{equation}
Multiplying both sides by $\mathbb J_{2n}$ yields
\begin{equation}
	v = \sum_{i=1}^m \alpha_i \mathbb J_{2n} e_i + \sum_{i=1}^k \beta_i \mathbb J_{2n} f_i,
\end{equation}
But all vectors on the right hand side belong to span$(S)$, hence reaching a contradiction. For updating $S$ using $v$, we first symplectify (symplectic orthogonalize) $v$ with respect to $S$ to recover
\begin{equation}
	\tilde v = v - \sum_{i=1}^k \Omega(v,f_i)e_i + \sum_{i=1}^k \Omega (v,e_i)f_i.
\end{equation}
We can write the update for the symplectic basis as
\begin{equation}
	e_{k+1} = \frac{\tilde v}{\|\tilde v\|},\quad f_{k+1} = \mathbb J_{2n}^T \tilde v.
\end{equation}
It is easily checked that the new set of basis vectors $\{e_1,\dots,e_{k+1},f_1,\dots,f_{k+1} \}$ forms a symplectic basis. This update, not only provides a symplectic basis but also an orthonormal basis.

The key element of the greedy algorithm is the availability of an error function which evaluates the error associated with the model reduction \cite{Anonymous:2016wl}. In the framework of symplectic model reduction, one possible candidate is the error in the Hamiltonian (\ref{eq:SyMo:5}). Correctly approximating symplectic systems relies on preservation of the Hamiltonian, hence the error in the Hamiltonian appears to be a natural choice. Moreover, since the error in the Hamiltonian depends on the initial condition and the reduced symplectic basis, evaluation of error does not require time integration of the full system.

Suppose that a $2k$ dimensional symplectic basis $A_{2k} = \{e_1,\dots ,e_k,f_1,\dots ,f_k\}$ is generated at the $k$-th step of the greedy method and we seek to enrich it by two additional vectors. Using the error in the Hamiltonian (\ref{eq:SyMo:7}) we search the parameter space to find the value that maximises the error in the Hamiltonian
\begin{equation}
	\omega_{k+1} := \underset{\omega\in \Gamma}{\text{argmax }}\Delta H(\omega).
\end{equation}
The goal is to approximate the Hamiltonian function as well as possible.

We then propagate (\ref{eq:MoOr:1}) in time to produce trajectory snapshots $S=\{ \mathbf z(t_i,\omega_{k+1}) | i = 1,\dots,M \}$. The next basis vector is the snapshot that maximises the projection error (\ref{eq:SyMo:8})
\begin{equation}
	z := \underset{s\in S}{\text{argmax }} \| s - AA^+s \|.
\end{equation}
Finally, we update the basis as
\begin{equation}
	e_{k+1} = \tilde z, \quad f_{k+1} = \mathbb J_{2n}^T \tilde z,
\end{equation}
where $\tilde z$ is the vector obtained from applying the symplectic Gram-Schmidt process to $z$.

Since the maximization over the entire parameter space $\Gamma$ is impossible, we discretize the parameter set into a grid with $N$ points: $\Gamma_N = \{ \omega_1,\dots,\omega_N\}$. However, since the selection of parameters only require the evaluation of the error in the Hamiltonian and not time integration of the original system, then $\Gamma_N$ can be chosen very rich. Other sampling techniques can be found in \cite{Quarteroni:2016wi}.

We summarized the greedy algorithm for generation of a symplectic basis in Algorithm \ref{alg:SyMo:3}.

\begin{algorithm} 
\caption{The greedy algorithm for generation of a symplectic basis} \label{alg:SyMo:3}
{\bf Input:} Tolerated loss in the Hamiltonian $\delta$, parameter set $\Omega_N = \{\omega_1,\dots,\omega_N\}$, initial condition $\mathbf z_0(\omega)$
\begin{enumerate}
\item $\omega^* \leftarrow \omega_1$
\item $e_1 \leftarrow \mathbf z_0(\omega^*)$
\item $f_1 \leftarrow \mathbb J^T_{2n} \mathbf z_0(\omega^*)$
\item $A \leftarrow [e_1,f_1]$
\item $k \leftarrow 1$
\item \textbf{while} $\Delta H(\omega) > \delta$ for all $\omega \in \Omega_N$
\item \hspace{0.5cm} $w^* \leftarrow$ $\underset{\omega\in \Omega_N}{\text{argmax }}\Delta H(\omega)$
\item \hspace{0.5cm} Compute trajectory snapshots $S=\{ \mathbf z(t_i,\omega^*) | i = 1,\dots,M \}$
\item \hspace{0.5cm} $\mathbf z^* \leftarrow$ $\underset{s\in S}{\text{argmax }} \| s - AA^+s \|$
\item \hspace{0.5cm} Apply symplectic Gram-Schmidt on $\mathbf z^*$
\item \hspace{0.5cm} $e_{k+1} \leftarrow \mathbf z^*/ \| z^*\|$
\item \hspace{0.5cm}$f_{k+1} \leftarrow \mathbb J^T_{2n} \mathbf z^*$
\item \hspace{0.5cm} $A \leftarrow [e_1,\dots ,e_{k+1} , f_1,\dots,f_{k+1}]$
\item \hspace{0.5cm} $k \leftarrow k+1$
\item \textbf{end while}
\end{enumerate}
\vspace{0.5cm}
{\bf Output:} Symplectic basis $A$.
\end{algorithm}

The convergence and convergence rate of the greedy algorithm with respect to Kolmogorov $n$-width is investigated in Section \ref{chap:SyMo.PrSy:3}. We finish this section by stating the advantage of using the greedy algorithm over the POD-based methods:
\begin{itemize}
\item Lack of a symplectic linear algebra toolbox for solving the minimization in (\ref{eq:SyMo:8}) requires a possibly very expensive nonlinear optimizations to recover near optimal symplectic bases. Alternative approaches for finding solutions to (\ref{eq:SyMo:8}) requires assumptions on the structure of the symplectic basis. The greedy algorithm on the other hand, is a cheap alternative to the nonlinear optimization and it allows flexibility in terms of the structure.
\item The greedy method reduces the total cost of the offline stage by solving the original system (\ref{eq:Hasy:7}) only $k$ times to obtain a reduced system of size $2k$, unlike POD based methods that require time integration of (\ref{eq:Hasy:7}) for the entire parameter space.
\item If enrichment of the symplectic basis is required, the greedy algorithm can add new basis vectors. The POD-based methods, are bound to re-compute the basis for the refined parameter set.
\item POD-based methods, e.g., the cotangent lift and the complex SVD, require performing SVD decompositions on possibly very large matrices. The greedy algorithm avoids this.
\end{itemize}

\subsubsection{Convergence of the Greedy Method} \label{chap:SyMo.PrSy:3}

To show convergence of the greedy method we define a slightly different version based on the projection error. The error in the Hamiltonian is then introduced as a cheap surrogate to the projection error to accelerate parameter selection.

Suppose that we are given with a compact subset $S$ of $\mathbb R^{2n}$. Our intention is to find a set of vectors $A=\{e_1,\dots,e_k,f_1,\dots,f_k\}$, such that $A$ forms a symplectic basis and any $s\in S$ is well approximated by elements of the subspace span$(A)$. The modified greedy method for generating basis vectors $e_i$ and $f_i$ is as follows. In the initial step we pick $e_1$ such that $\Omega(e_1,-\mathbb{J}_{2n}^T e_1) = \|e_1\|_2 = \max_{s\in S} \|s\|_2$, where $\|\cdot\|_2$ is the Euclidean norm. Then define $f_1 = -\mathbb{J}_{2n}^T e_1$. It is easy to check that the span of $A_2 = \{e_1,f_1\}$ is symplectic, so $A_2$ is the first subspace that approximates elements of $S$. In the $k$-th step of the greedy method, suppose we have a basis $A_{2k} = \{ e_1,\dots, e_k , f_1,\dots ,f_k \}$. We define $P_{2k}$ to be a symplectic projection operator that projects elements of $S$ onto span$(A_{2k})$ and define
\begin{equation} \label{eq:new1}
	\sigma_{2k}(s) := \|s-P_{2k}(s)\|_2,
\end{equation}
as the projection error. Moreover we denote by $\sigma_{2k}$ the maximum approximation error of $S$ using elements in span$(A_{2k})$ as
\begin{equation} \label{eq:new2}
	\sigma_{2k} := \max_{s\in S} \sigma_{2k}(s).
\end{equation}
The next set of basis vectors in the greedy selection are
\begin{equation} \label{eq:new3}
\begin{aligned}
	& e_{k+1} := \underset{s\in S}{\text{argmax }}\sigma_{2k}(s), \\
	& f_{k+1} := \mathbb{J}^T e_{k+1}.
\end{aligned}
\end{equation}
We emphasise that the sequence of basis vectors generated by the greedy is generally not unique.

To estimate the quality of the reduced subspace, it is natural to compare it with the best possible $2k$-dimensional subspace in the sense of minimum projection (not necessary symplectic) error. For this we need to introduce the Kolmogorov $n$-width \cite{Kolmogoroff:1936fj,Pinkus:1985vy}.

\begin{definition}
Let $S$ be a subset of $\mathbb R^{2n}$ and $Y_n$ be a general $n$-dimensional subspace of $\mathbb R^{2n}$. The angle between $S$ and $Y_n$ is given by
\begin{equation} \label{eq:new4}
	E(S,Y_n) := \sup_{s\in S} \inf_{y\in Y_n} \|s-y\|_2.
\end{equation}
The Kolmogorov $n$-width of $S$ in $\mathbb R^n$ is given by
\begin{equation} \label{eq:new5}
	d_{n}(S,\mathbb{R}^n) := \inf_{Y_n} E(S,Y_n) = \inf_{Y_n} \sup_{s\in S} \inf_{y\in Y_n} \|s-y\|_2
\end{equation}
\end{definition}

For a given subspace $Y_n$, the angle between $S$ and $Y_n$ measures the worst possible projection error of elements in $S$ onto $Y_n$. Hence the Kolmogorov $n$-width quantifies how well $S$ can be approximated by any $n$-dimensional subspace.

We seek to show that the decay of $\sigma_{2k}$, obtained by the greedy algorithm, has the same rate as of $d_{2k}(S)$, i.e., the greedy method provides the best possible accuracy attained by a $2k$-dimensional subspace.

We start by symplectifying the vectors provided by the greedy algorithm as
\begin{equation} \label{eq:new6}
\begin{aligned}
	& \xi_1 = e_i, & \bar{\xi}_1 = \mathbb{J}_{2n}^T \xi_1, &\\
	& \xi_i = e_i - P_{2(i-1)} (e_i), & \bar{\xi}_i = \mathbb{J}_{2n}^T, \xi_i &\quad i = 2,3,\dots
\end{aligned}
\end{equation}
Now the projection of a vector $s\in S$ onto span$(A_{2k})$ can be written using the symplectic basis as
\begin{equation} \label{eq:new7}
	P_{2k}(s) = \sum_{i=1}^k \left( \alpha_i(s) \xi_i + \bar{\alpha}_i(s) \bar{\xi}_i \right),
\end{equation}
where $\alpha_i(s)$ and $\bar{\alpha}_i(s)$ for $i=1,\dots,k$ are the expansion coefficients
\begin{equation} \label{eq:new8}
	\alpha_i(s) = - \frac{\Omega(\bar{\xi}_i,s)}{\Omega(\xi_i,\bar{\xi}_i)}, \quad \bar{\alpha}_i(s) = \frac{\Omega(\xi_i,s)}{\Omega(\xi_i,\bar{\xi}_i)},
\end{equation}
for any $s\in S$. Since $\bar{\xi}_i$ is symplectic to span$(A_{2(k-1)})$ we have
\begin{equation} \label{eq:new9}
\begin{aligned}
	|\alpha_i(s)| = \frac{|\Omega(\bar{\xi}_i,s)|}{|\Omega(\xi_i,\bar{\xi}_i)|} &= \frac{\Omega( \bar{\xi}_i, s - P_{2(k-1)}(s))}{|\Omega(\xi_i,\bar{\xi}_i)|} \\
	&\leq \frac{\|\bar{\xi}_i\|_2 \| s - P_{2(k-1)}(s) \|_2}{ \|\xi_i\|_2 \|\bar{\xi}_i\|_2 } \\
	&= \frac{\| s - P_{2(k-1)}(s) \|_2}{\| e_i - P_{2(k-1)}(e_i) \|_2} \leq 1.
\end{aligned}
\end{equation}
Here, we use the fact that $|\Omega(\xi_i,\bar{\xi}_i)| = |\xi_i^T \mathbb{J}_{2n}\bar{\xi}_i| = \| \xi_i \|^2 = \|\bar{\xi}_i\|^2$ and the last inequality follows from the greedy algorithm which maximizes $e_i$. Similarly we deduce that $|\bar{\alpha}_i(s)|\leq 1$.

We write
\begin{equation} \label{eq:new10}
\begin{aligned}
	\xi_j = \sum_{i=1}^j \left( \mu_i^j e_i + \gamma_i^j f_i \right), \quad \bar{\xi}_j = \sum_{i=1}^j \left( \lambda_i^j e_i + \eta_i^j f_i,  \right), \quad j=1,2,\dots
\end{aligned}
\end{equation}
with
\begin{equation} \label{eq:new11}
\begin{aligned}
	&\mu^j_j = 1, \quad \gamma^j_j = 0, \\
	&\mu_i^j = \sum_{l=i}^{j-1}\left( - \alpha_l(f_j) \mu_i^l  + \bar{\alpha}_l(f_j) \gamma_i^l \right), \quad\gamma_i^j = \sum_{l=i}^{j-1}\left( - \alpha_l(f_j) \gamma_i^l  + \bar{\alpha}_l(f_j) \mu_i^l \right), \\
	&\lambda^j_i = - \gamma ^j_i, \quad \eta^j_i = \mu^j_i.
\end{aligned}
\end{equation}
for $j=2,3,\dots$. By induction and using the bound in (\ref{eq:new9}) we deduce that
\begin{equation} \label{eq:new12}
	\mu^j_i,\gamma^j_i,\lambda^j_i,\eta^j_i \leq 3^{j-i}, \quad \text{for } j\geq i.
\end{equation}
Now let $2k$ be the dimension of the desired reduced space. Looking at the definition of Kolmogorov $n$-width we observe that for any $\theta > 1$ we can find a subspace $Y_{2k}$ such that $E(S,Y_{2k}) \leq \theta d_{2k}(S,\mathbb R^n)$. Hence can find vectors $v_1,\dots,v_k,u_1,\dots,u_k\in Y_{2k}$ such that
\begin{equation} \label{eq:new13}
\begin{aligned}
	& \|e_i - v_i\|_2 \leq \theta d_{2k}(S,\mathbb R^n), \\
	& \|f_i - u_i\|_2 \leq \theta d_{2k}(S,\mathbb R^n).
\end{aligned}
\end{equation}
Now we construct a set of $2(k+1)$ new vectors
\begin{equation} \label{eq:new14}
\begin{aligned}
	& \zeta_j = \sum_{i=1}^{k+1} \mu_i^j v_i + \gamma^j_i u_i,\quad \bar{\zeta}_j = \sum_{i=1}^{k+1} \lambda_i^j v_i + \eta^j_i u_i.
\end{aligned}
\end{equation}
for $j = 1,\dots,k+1$. Note that since all $u_i$ and $v_i$ belong to $Y_{2k}$ so does their linear combination including all $\zeta_j$ and $\bar{\zeta}_j$. We can use the inequality (\ref{eq:new12}) to write
\begin{equation}
	\| \xi_i - \zeta_i \|_2 \leq 3^i \theta d_{2k}(S,\mathbb R^n),\quad \| \bar{\xi}_i - \bar{\zeta}_i \|_2 \leq 3^i \theta d_{2k}(S,\mathbb R^n).
\end{equation}
Moreover since $Y_{2k}$ is of dimension $2k$ we can find $\kappa_i$, $i=1,\dots,2(k+1)$ such that
\begin{equation} \label{eq:new15}
	\sum_{i=1}^{2(k+1)} \kappa_i^2 = 1,
\end{equation}
and
\begin{equation} \label{eq:new16} 
	\sum_{i=1}^{k+1} \kappa_i \zeta_i + \sum_{i=1}^{k+1} \kappa_{i+k+1} \bar{\zeta}_i = 0.
\end{equation}
We have
\begin{equation} \label{eq:new17}
\begin{aligned}
	\left\| \sum_{i=1}^{k+1} \kappa_i \xi_i + \sum_{i=1}^{k+1} \kappa_{i+k+1} \bar{\xi}_i \right\|_2 &= \left\| \sum_{i=1}^{k+1} \kappa_i (\xi_i - \zeta_i) + \sum_{i=1}^{k+1} \kappa_{i+k+1} (\bar{\xi}_i-\bar{\zeta}_i) \right\|_2 \\
	&\leq 2\cdot 3^{k+1} \sqrt{2(k+1)} \theta d_{2k}(S,\mathbb R^n).
\end{aligned}
\end{equation}
We know there exists $1 \leq j\leq 2k+2$ such that $\kappa_j > 1/\sqrt{2(k+1)}$. Without loss of generality let us assume that $j\leq k+1$. This yields
\begin{equation} \label{eq:new18}
	\left\| \xi_j +  \kappa_j^{-1} \sum_{i=1,i\neq j}^{k+1} \kappa_i \xi_i + \kappa_j^{-1}\sum_{i=1}^{k+1} \kappa_{i+k+1} \bar{\xi}_i \right\|_2 \leq 4\cdot 3^{k+1} (k+1) \theta d_{2k}(S,\mathbb R^n).
\end{equation}
Define $c = \kappa_j^{-1} \sum_{i=1,i\neq j}^{k+1} \kappa_i \xi_i + \kappa_j^{-1}\sum_{i=1}^{k+1} \kappa_{i+k+1} \bar{\xi}_i$. And note that $\mathbb{J}_{2n}^T c$ is symplectic to $\xi_j$. From this we deduce
\begin{equation} \label{eq:new19}
\begin{aligned}
	\| \xi_j \|_2 &\leq \| \xi_j \|_2 + \| c \|_2 \\
		&= \Omega(\xi_j,\mathbb{J}_{2n}^T \xi_j) + \Omega(c,\mathbb{J}_{2n}^T c) \\
	       &= \Omega(\xi_j,\mathbb{J}_{2n}^T \xi_j) + \Omega(c,\mathbb{J}_{2n}^T c) + \Omega(\xi_j,\mathbb{J}_{2n}^T c) + \Omega(c,\mathbb{J}_{2n}^T \xi_j) \\
       	&= \Omega(\xi_j + c, \mathbb{J}^T_{2n} (\xi_j + c)) \\
	&= \| \xi_j + c \|_2	
\end{aligned}
\end{equation}
Combining this with inequality (\ref{eq:new18}) yields
\begin{equation} \label{eq:new20}
	\| \xi_j \|_2 \leq 4\cdot 3^{k+1} (k+1) \theta d_{2k}(S,\mathbb R^n).
\end{equation}
Finally using the definition of $\xi_j$ for all $s\in S$ we have
\begin{equation} \label{eq:new21}
	\| s - P_{2(j-1)}(s) \|_2 \leq \| f_j - P_{2(j-1)}(f_j) \|_2 = \|\xi_j \|_2 \leq 4\cdot 3^{k+1} (k+1) \theta d_{2k}(S,\mathbb R^n)
\end{equation}
Hence, for any given $\lambda > 1$
\begin{equation} \label{eq:new22}
	\| s - P_{2k}(s) \|_2 \leq \| s - P_{2(j-1)}(s) \|_2 \leq 4\cdot 3^{k+1} (k+1) \theta d_{2k}(S,\mathbb R^n).
\end{equation}
This establishes the following theorem.
\begin{theorem}
	Let $S$ be a compact subset of $\mathbb{R}^{2n}$ with exponentially small Kolmogorov $n$-width $d_{n}\leq c\exp(-\alpha k)$ with $\alpha > \log3$. Then there exists $\beta>0$ such that the symplectic subspaces $A_{2k}$ generated by the greedy algorithm provides exponential approximation properties such that
\begin{equation} \label{eq:new23}
	\| s - P_{2k}(s) \|_2 \leq C \exp(-\beta k)
\end{equation}
for all $s\in S$ and some $C>0$.
\end{theorem}

\subsection{Symplectic Discrete Empirical Interpolation Method (SDEIM)} Consider the Hamiltonian system (\ref{eq:Hasy:7}) and its reduced system (\ref{eq:SyMo:4}) equipped with a symplectic transformation $A$. One can split the Hamiltonian function $H = H_1 + H_2$ such that $\nabla H_1 = L\mathbf z$ and $\nabla H_2 = \mathbf g(\mathbf z)$, where $L$ is a constant matrix in $\mathbb R^{n\times n}$ and $\mathbf g$ is a nonlinear function. The reduced system takes the form
\begin{equation} \label{eq:new24}
	\frac{d}{dt} \mathbf y = \underbrace{A^+ \mathbb J_{2n} L A}_{\tilde L} \mathbf y + A^+ \mathbb J_{2n} \mathbf g(A\mathbf y)
\end{equation}
As discussed in Section \ref{chap:MoOr.DEIM:1}, the complexity of evaluating the nonlinear term still depends on $n$, the size of the original system. To overcome this computational bottleneck we use the DEIM approximation for evaluating the nonlinear function $\mathbf g$ as
\begin{equation} \label{eq:new25}
	\frac{d}{dt} \mathbf y = \tilde L \mathbf y + \underbrace{ A^+ \mathbb J_{2n} V (P^TV)^{-1} P^T \mathbf g(A\mathbf y) }_{\tilde N(\mathbf y)}
\end{equation}
One may argue that if DEIM's projection basis $V$ is rich enough then the approximation of the nonlinear term does not affect the symplectic structure significantly since $\mathbf g$ is already a symplectic function. Nonetheless, for a general choice of $V$ the system (\ref{eq:new25}) is not guaranteed to be a Hamiltonian system. However, we can guarantee that (\ref{eq:new25}) is a Hamiltonian system by choosing $V=(A^+)^T$. To understand why, we note that the system (\ref{eq:new25}) is a Hamiltonian system if and only if $\tilde N(\mathbf y) = \mathbb J_{2k} \nabla_{\mathbf y} \mathbf g(\mathbf y)$. Also we have 
\begin{equation} \label{eq:new26}
	\mathbf g(A\mathbf y) = \nabla_{\mathbf z} H_2(\mathbf z) = (A^+)^T \nabla_{\mathbf y} H_2(A \mathbf y),
\end{equation}
where the chain rule is used for the second equality. Substituting this into $\tilde N$ we obtain
\begin{equation} \label{eq:new27}
	\tilde N(\mathbf y)= A^+ \mathbb J_{2n} V (P^TV)^{-1} P^T  (A^+)^T \nabla_{\mathbf y} H_2(A \mathbf y).
\end{equation}
Taking $V = (A^+)^T$ yields
\begin{equation} \label{eq:new28}
	\tilde N(\mathbf y) = A^+ \mathbb J_{2n}(A^+)^T H_2(A \mathbf y) = \mathbb J_{2k} H_2(A \mathbf y),
\end{equation}
using that $(A^+)^T$ is a symplectic matrix. Hence, $V = (A^+)^T$ is a sufficient condition for (\ref{eq:new25}) to be a Hamiltonian system.

Regarding the construction of the projection space, suppose that we have already constructed a symplectic basis $A=\{ e_1,\dots , e_k,f_1,\dots f_k \}$ using the greedy algorithm. Note that $(A^+)^T$ is a symplectic basis and $(A^+)^+=A$. Thus, we can move between these two symplectic bases by simply using the transpose operator and the symplectic inverse operator. Let $S_{\mathbf g} = \{ \mathbf g (\mathbf x(t_i,\omega_j)) \}$ with $i = 1,\dots,M$ and $ j = 1 ,\dots,N$ be the nonlinear snapshots that were gathered in the greedy algorithm. We then form $(A^+)^T = \{ e'_1,\dots, e'_k,f'_1,\dots,f'_k\}$ and use a greedy approach for adding new basis vectors to $(A^+)^T$. At the $i$-th iteration of the symplectic DEIM, we use $(A^+)^T$ to approximate elements in $S_{\mathbf g}$ and choose the vector that maximizes error as the next basis vector. 
\begin{equation}
	s^* := \underset{s \in S_{\mathbf g}}{\text{argmax }}\| s - (A^+)^T A^+ s \|_2.	
\end{equation}
After applying the symplectic Gram-Schmidt on $s^*$, we update $(A^+)^T$ as
\begin{equation}
\begin{aligned}
	e'_{k+i+1} = \frac{s^*}{\| s^* \|_2},\quad f'_{k+i+1} = \mathbb J_{2n}^T e'_{k+i+1}.
\end{aligned}
\end{equation}
Finally when $(A^+)^T$ approximates elements $S_{\mathbf g}$ with the desired accuracy, we transpose and symplecticaly inverse $(A^+)^T$ to obtain $A$, yielding a Hamiltonian reduced system. Note that the above update only enriches the span of $A$, containing the basis we started with. We summarized the symplectic DEIM algorithm in Algorithm \ref{alg:SyMo:4}.

\begin{algorithm} 
\caption{Symplectic Discrete Empirical Interpolation Method} \label{alg:SyMo:4}
{\bf Input:} Symplectic basis $A=\{ e_1,\dots,e_k,f_1,\dots,f_k \}$, nonlinear snapshots $S_{\mathbf g} = \{ \mathbf g(\mathbf x(t_i,\omega_j)) \}$ and tolerance $\delta$
\begin{enumerate}
\item Compute $(A^+)^T = \{ e'_1,\dots,e'_k,f'_1,\dots,f'_k \}$
\item $i \leftarrow 1$
\item \textbf{while} max$\| s - (A^+)^T A^+s \| > \delta$ for all $s\in S_{\mathbf g}$
\item \hspace{0.5cm} $s^* \leftarrow \underset{s \in S_{\mathbf g}}{\text{argmax }}\| s - (A^+)^T A^+ s \|$
\item \hspace{0.5cm} Apply symplectic Gram-Schmidt on $s^*$
\item \hspace{0.5cm} $e'_{k+i} = s^* / \| s^* \|$
\item \hspace{0.5cm} $f'_{k+i} = \mathbb J_{2n} e'_{k+i}$
\item \hspace{0.5cm} $(A^+)^T \leftarrow [e'_1,\dots,e'_{k+i},f'_1,\dots,f'_{k+i}]$
\item \hspace{0.5cm} $i\leftarrow i+1$
\item \textbf{end while}
\item take transpose and symplectic inverse of $(A^+)^T$
\end{enumerate}
\vspace{0.5cm}
{\bf Output:} Symplectic basis $A$ that guarantees a Hamiltonian reduced system.
\end{algorithm}

When using an implicit time integration scheme we face inefficiencies when evaluating the Jacobian of nonlinear terms, as discussed in Section \ref{chap:MoOr.DEIM:1}. We recall that the key to fast approximation of the Jacobian is that the interpolating index matrix $P$ obtained in the DEIM approximation, commutes with the nonlinear function. Nonlinear terms in Hamiltonian systems often take the from
\begin{equation}
	\mathbf g (\mathbf z) = \mathbf g (\mathbf q,\mathbf p) = 
	\begin{pmatrix}
		g_1(q_1,p_1) \\
		g_2(q_2,p_2) \\
		\vdots \\
		g_{2n}(q_{n},p_{n})
	\end{pmatrix}.
\end{equation}
Thus, the interpolating index matrix, obtained by Algorithm \ref{alg:MoOr:1} does not necessarily commute with the function $\mathbf g$. To overcome this, when index $\mathfrak p_i$ with $\mathfrak p_i\leq n$ or $\mathfrak p_i>n$ is chosen in Algorithm \ref{alg:MoOr:1} we also include $\mathfrak p_i + n$ or $\mathfrak p_i-n$, respectively. This guarantees that $\mathbf g$ and $P$ commute.

### 2016-11-15 greedy_hamiltonian/chapters/5.numres.tex
fist commit

\section{Numerical Results} \label{chap:NuRe:1}
In this section, we illustrate the performance of the greedy generation of a symplectic basis. First, the parametric linear wave equation is considered to compare SVD based methods with the greedy method. The nonlinear model order reduction using the combination of DIEM and the symplectic basis is then illustrated by considering the parametric nonlinear Schr\"odinger equation.

\subsection{Parametric Linear Wave equation} Consider the stochastic linear wave equation
\begin{equation} \label{eq:NuRe:1}
\left\{
\begin{aligned}
& u_{tt}(x,t,\omega) = \kappa(\omega) u_{xx}(x,t,\omega), \\
& u(x,0) = u^0(x),
\end{aligned}
\right.
\end{equation}
where $x$ belongs to a one-dimensional torus of length $L$, $\omega = (\omega_1,\dots,\omega_4)$ and
\begin{equation} \label{eq:NuRe:2}
	\kappa(\omega) = c^2\left( \sum_{l=1}^4 \frac{1}{l^2} \omega_l \right).
\end{equation}
Each $\omega_l$ is uniformly distributed in $[0,1]$ and $c\in \mathbb{R}$ is a constant number. By rewriting (\ref{eq:NuRe:1}) in canonical form, using the change of variable $q = u$ and $\partial q/ \partial t= p$, we obtain the symplectic form
\begin{equation} \label{eq:NuRe:3}
\left\{
\begin{aligned}
& q_t(x,t,\omega) = p(x,t,\omega), \\
& p_t(x,t,\omega) = \kappa(\omega) q_{xx}(x,t,\omega),
\end{aligned}
\right.
\end{equation}
with the associated Hamiltonian given by
\begin{equation} \label{eq:NuRe:4}
	H(q,p,\omega) = \frac 1 2 \int_0^L p^2 + \kappa(\omega) q_x^2 \ dx.
\end{equation}
We discretized the torus into $N$ equidistant points and define $\Delta x = L/N$, $x_i = i\Delta x$, $q_i=q(t,x_i,\omega)$ and $p_i=p(t,x_i,\omega)$ for $i = 1, \dots, N$. Furthermore, we discretize (\ref{eq:NuRe:3}) using a standard central finite differences scheme to obtain
\begin{equation} \label{eq:NuRe:5}
	\frac{d}{dt} \mathbf z = \mathbb{J}_{2N} L\mathbf z,
\end{equation}
where $\mathbf z=(q,\dots,q_N,p_q,\dots,p_n)^T$ and
\begin{equation} \label{eq:NuRe:6}
L = 
\begin{pmatrix}
	I_n & 0_N \\
	0_N & \kappa(\omega)D_{xx}
\end{pmatrix},\quad 
\end{equation}
with $D_{xx}$ the central finite differences matrix operator. The discrete Hamiltonian can finally be written as
\begin{equation} \label{eq:NuRe:7}
	H_{\Delta x}(\mathbf z) = \frac{\Delta x}2 \sum_{i=1}^{N} \left( p_i^2 + \kappa(\omega) \frac{(q_{i+1} - q_i)^2}{2\Delta x ^ 2} + \kappa(\omega) \frac{(q_{i} - q_{i-1})^2}{2\Delta x ^ 2} \right).
\end{equation}

The initial condition is given by
\begin{equation} \label{eq:NuRe:8}
	q_i(0) = h( 10\times|x_i - \frac{1}{2}| ), \quad p_i = 0, \quad i=1,\dots,N
\end{equation}
where $h(s)$ is the cubic spline function
\begin{equation} \label{eq:NuRe:9}
h(s) = 
\left\{
\begin{aligned}
& 1 - \frac{3}{2}s^2 + \frac{3}{4}s^3, \quad & 0\leq s \leq 1, \\
& \frac{1}{4}(2-s)^3, & 1< s \leq 2, \\
& 0, & s > 2.
\end{aligned}
\right.
\end{equation}
This will result in waves propagating in both directions on the torus.

For numerical time integration we used the Str\"omer-Verlet (\ref{eq:Hasy:13}) scheme, which becomes an explicit scheme since the Hamiltonian is separable for the linear wave-equation. The full model uses the following parameter set \vspace{0.5cm}
\begin{center}
\begin{tabular}{|l|l|}
\hline
Domain length & $L = 1$ \\
No. grid points & $N = 500$ \\
Space discretization size & $\Delta x = 0.002$ \\
Time discretization size & $\Delta t = 0.01$ \\
\hline
\end{tabular}
\end{center}
\vspace{0.5cm}

We compare the reduced system obtained by the greedy algorithm with the methods based on SVD. To generate snapshots, we discretize the parameter space $[0,1]^4$ into in total of $5^4$ equidistant grid points. For the SVD based methods and POD, snapshots are gathered in the snapshot matrices $S$, $S_{\text{combined}}$ and $S_{\text{complex}}$, respectively, and the SVD is performed to construct the reduced basis. The greedy method is applied according to Algorithm \ref{alg:SyMo:3}; as input, the tolerance for the error in the Hamiltonian is set to $\delta = 5 \times 10^{-3}$. All reduced systems are taken to have an identical size ($k=80$ for POD and $k=40$ for the symplectic methods). We use the Verlet-Str\"omer scheme for symplectic methods and a second order Runge-Kutta method for the POD. Finally we use transformation (\ref{eq:SyMo:1}) to transfer the solution of the reduced systems into the high-dimensional space for illustration purposes.

We reduced the cost by 50\% in the offline stage when using the greedy method as compared to SVD-based methods (cotangent lift and complex SVD method). This happens because the SVD-based methods require time integration of the full system for all discrete parameter points, while the greedy method picks a number of parameters from the parameter space.

%We summarized the duration of offline and online stages of different model reduction methods in the table below.
%
%\vspace{0.5cm}
%\begin{center}
%\begin{tabular}{c|cc}
% & offline (seconds) & online (seconds) \\
%\hline
%full system & - & 0.19 \\
%cotangent lift & 61.4 & 0.11 \\
%complex SVD & 61.4 & 0.11 \\
%greedy & 33.7 & 0.11 \\
%\end{tabular}
%\end{center}
%\vspace{0.5cm}
%
%In our implementations the offline stage of cotangent lift and the complex SVD are substantially longer compared to the greedy method. This is because SVD based methods need to explore the entire parameter space while the greedy method only hand pick a number of parameters from the parameter space. We see that the online stage of all reduced systems are faster than solving the full model. Of course efficiency of reduced systems depend on the implementation.

Figure \ref{fig:NuRe:3} shows the solution of the linear wave equation for parameter value $c=0.1019$, chosen to be different from training parameters, at $t=0$, $t = 1$ and $t=2$. While we see instability and divergence from the exact solution for the POD reduced system, the symplectic methods provide a good approximation of the full model.

The decay of the singular values for the POD, cotangent lift and the complex SVD method are presented in Figure \ref{fig:NuRe:4}(a). The fast decay of the singular values suggests that a low dimensional solution manifold indeed exists. However since the linear subspace constructed by the POD is not symplectic, we observe the blow up of the Hamiltonian function in Figure \ref{fig:NuRe:4}(b) and the instability of the solution in Figure \ref{fig:NuRe:3}. The symplectic methods (using a reduced basis of the same size as POD) preserve the Hamiltonian function as is seen in Figure \ref{fig:NuRe:4}(b).

Figure \ref{fig:NuRe:4}(c) shows the $L^2$-error between the solution of the full model and the reduced systems constructed by different methods. We note that the error for the POD reduced system rapidly increases, confirming that projection based reduced system does not yield a stable solution. Furthermore, the symplectic methods provide a better approximation since the geometric structure of the original system is preserved. Although the greedy method is almost twice faster than SVD-based method in the offline stage, its accuracy is comparable. For complex systems were the solution of the full system is expensive and for parameter domains with higher dimensions, we expect the greedy method to be superior in terms of computational cost.

\begin{figure*}[t]
\begin{center}
\begin{tabular}{cc}
	\includegraphics[width=0.4\textwidth]{./figs/wave/solution/solution_t0} &
	\includegraphics[width=0.4\textwidth]{./figs/wave/solution/solution_t1} \\
	$t=0$ & $t = 1$
\end{tabular}
\end{center}
\end{figure*}

\begin{figure}[t]
\centering
\begin{tabular}{cc}
\includegraphics[width=0.4\textwidth]{./figs/wave/solution/solution_t2} \\
	$t=2$
\end{tabular}
\caption{The solution $q$ at $t=0$, $t=1$ and $t=2$ of the linear wave equation for parameter value $c= 0.1019$ different from training parameters. Here the solution of full system together with the solution of the POD, cotangent lift, complex SVD and the greedy reduced system is presented.}\label{fig:NuRe:3}
\end{figure}

\begin{figure*}[t]
\begin{center}
\begin{tabular}{cc}
	\includegraphics[width=0.5\textwidth]{./figs/wave/singular} &
	\includegraphics[width=0.5\textwidth]{./figs/wave/hamiltonian} \\
	(a) & (b)
\end{tabular}
\end{center}
\end{figure*}

\begin{figure}[t] 
\begin{center}
\begin{tabular}{c}
	\includegraphics[width=0.5\textwidth]{./figs/wave/error} \\
	(c)
\end{tabular}
\end{center}
\caption{(a) The decay of singular values for POD, cotangent lift and complex SVD methods. (b) The $L^2$ error between the solution of th full system and the reduced system for different model reduction methods for $t \in [0,30]$. (c) Plot of the Hamiltonian function for $t \in [0,30]$.}\label{fig:NuRe:4}
\end{figure}

\subsection{Nonlinear Schr\"odinger equation} Let us consider the one-dimensional parametric Schr\"odinger equation
\begin{equation} \label{eq:NuRe:10}
\left\{
\begin{aligned}
	& i u_t(t,x,\epsilon) = - u_{xx}(t,x,\epsilon) - \epsilon |u(t,x,\epsilon)|^2 u(t,x,\epsilon),\\
	& u(0,x) = u_0(x),
\end{aligned}
\right.
\end{equation}
where $u$ is a complex valued wave function, $i$ is the imaginary unit, $|\cdot|$ is the modulus operator and $\epsilon$ is a parameter that belongs to the interval $\Gamma = [0.9,1.1]$. We consider periodic boundary conditions, i.e., $x$ belongs to a one dimensional torus of length $L$. We consider the initial condition
\begin{equation} \label{eq:NuRe:11}
	u_0(x) = \frac{\sqrt 2}{\cosh(x - x_0)} \exp(i\frac{c(x-x0)}{2}),
\end{equation}
for a positive constant $c$. In quantum mechanics, the quantity $|u(t,x)|^2$ represents the probability of finding the system in state $x$ at time $t$. For the choice of $\epsilon = 1$, $|u(x,t)|$ becomes a solitary wave, and the initial condition will be transported in the positive $x$ direction with a constant speed. For other choices of $\epsilon$, the solution comprises an ensemble of solitary waves, moving in either direction \cite{Faou:2012vh}.

By introducing the real and imaginary variables $u = p + iq$, we can rewrite (\ref{eq:NuRe:10}) in canonical form as
\begin{equation} \label{eq:NuRe:12}
\left\{
\begin{aligned}
 q_t &= p_{xx} + \epsilon (q^2+p^2)p, \\
 p_t &= -q_{xx} - \epsilon (q^2 + p^2)q,
\end{aligned}
\right.
\end{equation}
with the Hamiltonian function
\begin{equation} \label{eq:NuRe:13}
	H(q,p) = \int_{0}^{L} (q_x^2 + p_x^2) + \frac \epsilon 2 (q^2 + p^2)^2\ dx.
\end{equation}

We discretize the torus into $N$ equidistant points and take define $\Delta x = L/N$, $x_i = i\Delta x$, $q_i=q(t,x,\epsilon)$ and $p_i = p(t,x_i,\omega)$ for $i = 1 ,\dots,N$. A central finite differences scheme is used to discretize (\ref{eq:NuRe:11}) as
\begin{equation}  \label{eq:NuRe:14}
	\frac{d}{dt} \mathbf z = \mathbb J_{2N} L\mathbf z + \mathbb J_{2N} \mathbf g(\mathbf z).
\end{equation}
Here $\mathbf z = (q_1,\dots,q_N,p_1,\dots,p_n)^T$ and
\begin{equation}  \label{eq:NuRe:15}
	L = 
	\begin{pmatrix}
		D_{xx} & 0_N \\
		0_N & D_{xx}
	\end{pmatrix}
\end{equation}
with $D_{xx}$ being the central finite differences matrix operator for periodic boundary conditions. $\mathbf g$ is a vector valued nonlinear function defined as
\begin{equation}  \label{eq:NuRe:16}
	\mathbf g(\mathbf z) =
	\begin{pmatrix}
	(q_1^2 + p_1^2)q_1 \\
	\vdots \\
	(q_N^2 + p_N^2)q_N \\
	(q_1^2 + p_1^2)p_1 \\
	\vdots \\
	(q_N^2 + p_N^2)p_N
	\end{pmatrix}.
\end{equation}
We discretize the Hamiltonian to obtain
\begin{equation}  \label{eq:NuRe:17}
	H_{\Delta x}(\mathbf z) = {\Delta x}\sum_{i=1}^{N} \left( \frac{q_i q_{i-1} - q_i^2}{\Delta x ^2} + \frac{p_i p_{i-1} - p_i^2}{\Delta x ^2} + \frac \epsilon 4 (p_i^2 + q_i^2)^2  \right).
\end{equation}
We use a Str\"omer-Verlet (\ref{eq:Hasy:13}) scheme for time integration. Since the Hamiltonian function (\ref{eq:NuRe:17}) is non-separable, this scheme becomes implicit so in each time iteration, a system of nonlinear equations is solved using Newton's iteration. We summarize the physical and numerical parameters for the full model in the following table

\vspace{0.5cm}
\begin{center}
\begin{tabular}{|l|l|}
\hline
Domain length & $L = 2\pi /l$ \\
Domain scaling factor & $l = 0.11$ \\
wave speed & $c =1$\\
No. grid points & $N = 256$ \\
Space discretization size & $\Delta x = 0.2231$ \\
Time discretization size & $\Delta t = 0.01$ \\
\hline
\end{tabular}
\end{center}
\vspace{0.5cm}

Regarding computation of the nonlinear terms of reduced systems, We compare the DEIM with the symplectic DEIM. For generation of the DEIM reduced basis we apply Algorithm \ref{alg:MoOr:1} to the set of nonlinear snapshots. Algorithm \ref{alg:SyMo:4} is used to construct a reduced basis appropriate for the symplectic DEIM. As input, we provide the symplectic basis generated by Algorithm \ref{alg:SyMo:3} with the set of nonlinear snapshots and a tolerance for the error $\delta = 10^{-4}$.

We compared the reduced system obtained by the greedy algorithm with the cotangent lift, complex SVD, DEIM, symplectic DEIM and also POD. For the-SVD based methods, we discretize the parameter space $[0.9,1.1]$ into $M=500$ equidistant grid points to recover the discrete parameter space $\Gamma_M = \{\epsilon_1,\dots,\epsilon_M \}$, and gather trajectory snapshots for each $\epsilon_i$ for $i = 1,\dots,M$ in the snapshots matrix $S$. All reduced system are taken to have identical sizes ($k=90$ for symplectic methods and $k=180$ for the POD method). Following algorithm \ref{alg:SyMo:3} we construct the reduced system using the same discrete parameter space $\Gamma_M$. The tolerance for the error in the Hamiltonian is set to $\delta = 10^{-3}$. Moreover, for DEIM and symplectic DEIM, we construct bases of size $k'=80$. Note that in the reduced system, generated in the symplectic DEIM, will be of size $k+k'=170$.

The cost of the offline stage is reduced to 20\% when using the greedy method for constructing a symplectic basis of size $k=90$, as compared to the SVD-based methods. The online stage, i.e., time integration for a new parameter in $\Gamma$, is generally more than 3 times faster than the original system. We point out that the efficiency of reduced systems are implementation and platform dependent.

%The duration of the online and offline stages for different model reduction techniques are summarized in table below. The offline time for the cotangent lift and the complex SVD consist of time integration of the full model for all the parameters in the discrete parameter space $\Omega_M$. This is while the greedy method requires substantially shorter offline stage to construct a reduced basis of size $k = 90$. For a reduced system of size $k=90$, all methods almost have the same online duration. The efficiency of reduced systems compared to the full system are heavily implementation dependent, nevertheless solving the reduced systems with our implementation is more than 3 times faster than solving the original system.

%\vspace{0.5cm}
%\begin{center}
%\begin{tabular}{c|cc}
% & offline (hours) & online (seconds) \\
%\hline
%full system & - & 187 \\
%cotangent lift & 25.8 & 59$\pm 1$ \\
%complex SVD & 25.8 & 59$\pm 1$ \\
%greedy & 4.6 & 59$\pm 1$ \\
%greedy + SDEIM & 4.7 & ???
%\end{tabular}
%\end{center}
%\vspace{0.5cm}

\begin{figure*}[t]
\begin{center}
\begin{tabular}{cc}
	\includegraphics[width=0.5\textwidth]{./figs/schrodinger/solution/solution_t0} &
	\includegraphics[width=0.5\textwidth]{./figs/schrodinger/solution/solution_t10} \\
	$t = 0$ & $t = 10$
\end{tabular}
\end{center}
\end{figure*}

\begin{figure}[t] 
\begin{center}
\begin{tabular}{c}
	\includegraphics[width=0.5\textwidth]{./figs/schrodinger/solution/solution_t20}\\
	$$t = 20$$
\end{tabular}
\end{center}
\caption{The solution $|u(t,x)| = \sqrt{q^2 + p^2}$ at $t=0$, $t=10$ and $t=20$ of the Nonlinear Schr\"odinger equation for parameter value $\epsilon = 1.0932$. Here the solution of full system together with the solution of the POD, cotangent lift, complex SVD and the greedy reduced system is presented.}\label{fig:NuRe:1}
\end{figure}

\begin{figure*}[t]
\begin{center}
\begin{tabular}{cc}
	\includegraphics[width=0.5\textwidth]{./figs/schrodinger/singular} &
	\includegraphics[width=0.5\textwidth]{./figs/schrodinger/hamiltonian} \\
	(a) & (b)
\end{tabular}
\end{center}
\end{figure*}

\begin{figure}[t] 
\begin{center}
\begin{tabular}{c}
	\includegraphics[width=0.5\textwidth]{./figs/schrodinger/error} \\
	(c)
\end{tabular}
\end{center}
\caption{(a) The decay of singular values for POD, cotangent lift and complex SVD methods. (b) The $L^2$ error between the solution of th full system and the reduced system for different model reduction methods for $t \in [0,30]$. (c) Plot of the Hamiltonian function for $t \in [0,30]$.}\label{fig:NuRe:2}
\end{figure}

Figure \ref{fig:NuRe:1} shows the solution of the Schr\"odinger equation for parameter value $\epsilon = 1.0932$ at $t=0$, $t=10$ and $t=20$. We first compare the reduced system obtained by the greedy algorithm with POD, cotangent lift and the complex SVD method. The size of the reduced systems are taken identical for all methods ($k=180$ for POD and $k=90$ for the rest). Although the decay of the singular values in Figure \ref{fig:NuRe:2} suggests that the accuracy of the POD reduced system should be comparable to that of the other methods, we observe instabilities in the solution at $t=10$. The greedy, the cotangent lift and the complex SVD method, on the other hand, generate a stable reduced system that accurately approximates the solution of the full model.

The value of the Hamiltonian function for the reduced systems discussed above as a function of time is illustrated in Figure \ref{fig:NuRe:2}(b). We see that the symplectic methods preserve the Hamiltonian function. We like to emphasise that the combination of the reduced basis by the greedy and the DEIM does not preserve the symplectic structure as suggested in this figure.

Figure \ref{fig:NuRe:2}(c) illustrates the $L^2$-error between the solution of the full model with the reduced systems generated by different methods. We first observe that symplectic methods yield a lower computational error compared to non-symplectic methods. Secondly, we observe that although the reduced systems from the cotangent lift and the complex SVD are of the same size, their accuracy is different by an order of magnitude. We notice that the greedy algorithm is slightly less accurate than the cotangent lift method while its offline computational cost is less than 20\%. Lastly we notice that the combination of the greedy reduced basis and DEIM yields large errors in the solution while the solution using the symplectic DEIM is very accurate. We note that the symplectic DEIM is even more accurate than the greedy itself since it has been enriched by the nonlinear snapshots.

### 2016-11-15 greedy_hamiltonian/chapters/appendix.tex
fist commit

\section*{Appendix} \label{chap:Append:1}

Similar to the QR decomposition for constructing an orthogonal matrix, for any full rank matrix $M$ of size $2n\times 2n$ we may have a decomposition
\begin{equation}
	M = AR,
\end{equation}
where $A$ is a symplectic square matrix (and not necessarily orthogonal) and $R$ has the structure
\begin{equation}
	R =
	\begin{pmatrix}
		S & T \\
		U & V
	\end{pmatrix}
\end{equation}
where $S$, $T$, $U$ and $V$ are all upper triangular square matrices of size $n$. Also $T$ and $U$ have zeros along the diagonal. The algorithm for a symplectic QR (SQR) is shown in Algorithm \ref{alg:Append:1}.

\begin{algorithm} 
\caption{Symplectic QR decomposition (SQR)} \label{alg:Append:1}
{\bf Input:} Full rank matrix $M = [u_1,\dots , u_k,v_1,\dots,v_k]$ of size $2n \times 2k$.
\begin{enumerate}
\item $\alpha \leftarrow \Omega(u_1,v_1)$
\item $e_1 \leftarrow \text{sign}(\alpha)\cdot u_1 / \sqrt{|\alpha|} $
\item $f_1 \leftarrow v_1 / \sqrt{|\alpha|}$
\item \textbf{for} $i\leftarrow 2$ \textbf{to} $k$
\item \hspace{0.5cm} $q \leftarrow u_i$
\item \hspace{0.5cm} $p \leftarrow v_i$
\item \hspace{0.5cm} \textbf{for} $j\leftarrow1$ \textbf{to} $i-1$
\item \hspace{0.5cm} \hspace{0.5cm} $q \leftarrow q - \Omega(q,f_j)e_j + \Omega(q,e_j)f_j$
\item \hspace{0.5cm} \hspace{0.5cm} $p \leftarrow p - \Omega(p,f_j)e_j + \Omega(p,e_j)f_j$
\item \hspace{0.5cm} \textbf{end for}
\item \hspace{0.5cm} $\alpha \leftarrow \Omega(q,p)$
\item \hspace{0.5cm} $e_i \leftarrow \text{sign}(\alpha) \cdot q / \sqrt{|\alpha|} $
\item \hspace{0.5cm} $f_i \leftarrow p / \sqrt{|\alpha|}$
\item \textbf{end for}
\item $A \leftarrow [e_1, \dots , e_k , f_1, \dots , f_k]$
\item $R \leftarrow A^+M$
\end{enumerate}

\vspace{0.5cm}
{\bf Output:} Matrix decomposition $M = AR$.
\end{algorithm}

### 2016-11-15 greedy_hamiltonian/paper.tex
fist commit

%%
%% This is file `docultex.tex', 
%% Documentation for siam supplemental file macros for use with LaTeX 2e
%% 
%% December 19, 2013
%%
%% Supplemental Files Supported Version 1.0.0
%% 
%% You are not allowed to change this file. 
%% 
%% You are allowed to distribute this file under the condition that 
%% it is distributed together with all of the files in the siam macro 
%% distribution. These are:
%%
%%  siamltex1213.cls (main LaTeX macro for SIAM)
%%  siam10.clo   (size option for 10pt papers)
%%  subeqn.clo   (allows equation numbners with lettered subelements)
%%  siam.bst     (bibliographic style file for BibTeX)
%%  docultex.tex (this file)
%%
%% If you receive only some of these files from someone, complain! 
%% 
%% You are NOT ALLOWED to distribute this file alone. You are NOT 
%% ALLOWED to take money for the distribution or use of either this 
%% file or a changed version, except for a nominal charge for copying 
%% etc. 
%% \CharacterTable
%%  {Upper-case    \A\B\C\D\E\F\G\H\I\J\K\L\M\N\O\P\Q\R\S\T\U\V\W\X\Y\Z
%%   Lower-case    \a\b\c\d\e\f\g\h\i\j\k\l\m\n\o\p\q\r\s\t\u\v\w\x\y\z
%%   Digits        \0\1\2\3\4\5\6\7\8\9
%%   Exclamation   \!     Double quote  \"     Hash (number) \#
%%   Dollar        \$     Percent       \%     Ampersand     \&
%%   Acute accent  \'     Left paren    \(     Right paren   \)
%%   Asterisk      \*     Plus          \+     Comma         \,
%%   Minus         \-     Point         \.     Solidus       \/
%%   Colon         \:     Semicolon     \;     Less than     \<
%%   Equals        \=     Greater than  \>     Question mark \?
%%   Commercial at \@     Left bracket  \[     Backslash     \\
%%   Right bracket \]     Circumflex    \^     Underscore    \_
%%   Grave accent  \`     Left brace    \{     Vertical bar  \|
%%   Right brace   \}     Tilde         \~}

\documentclass[final,leqno,onefignum,onetabnum]{siamltex1213}

%-------------------------------------------------------------------------
%
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%
%-------------------------------------------------------------------------

\usepackage{./pkg/packages}

\title{Structure Preserving Model Reduction of Parametric Hamiltonian Systems}

\author{
	Babak Maboudi Afkham\thanks{Department of Mathematics, Chair of Computational Mathematics and Simulation Schience (MCSS), \'Ecole Polytechnique F\'ed\'erale de Lausanne, Switzerland (babak.maboudi@epfl.ch)} 
	\and 
	Jan S. Hesthaven\thanks{Department of Mathematics, Chair of Computational Mathematics and Simulation Schience (MCSS), \'Ecole Polytechnique F\'ed\'erale de Lausanne, Switzerland (jan.hesthaven@epfl.ch)}}

\begin{document}
\maketitle
\slugger{mms}{xxxx}{xx}{x}{x--x}%slugger should be set to mms, siap, sicomp, sicon, sidma, sima, simax, sinum, siopt, sisc, or sirev

\begin{abstract}
While reduced-order models (ROMs) are popular for approximately solving large systems of differential equations, the stability of reduced models over long-time integration remains an open question. We present a greedy approach for ROM generation of parametric Hamiltonian systems which captures the symplectic structure of Hamiltonian systems to ensure stability of the reduced model. Through the greedy selection of basis vectors, two new vectors are added at each iteration to the set of basis vectors to increase the overall accuracy of the reduce basis. We used the error in the Hamiltonian function due to model reduction, as an error indicator to search the parameter space and find the next best basis vectors. We show that the greedy algorithm converges with exponential rate, under natural assumptions on the set of all solutions of the Hamiltonian system under variation of the parameters. Moreover, we demonstrate that combining the greedy basis with the discrete empirical interpolation method also preserves the symplectic structure. This enables the reduction of computational cost for nonlinear Hamiltonian systems. The efficiency, accuracy, and stability of this model reduction technique is illustrated through simulations of the parametric wave equation and the parametric Schr\"odinger equation.

%Through the greedy selection of basis vectors, we used the Hamiltonian function as an error estimator to search for the

%to add basis vectors, two at a time, and increase the overall accuracy of the reduced basis. It therefor requires a single numerical solution to be computed at each iteration which can save computational costs in stage offline stage. Under

%We also illustrated that convergence rate of the greedy approach is proportional to the best possible convergence rate of any possible iterative update of the reduced basis. Moreover, for nonlinear Hamiltonian systems, we combined the reduced basis obtained from the greedy method with discrete empirical interpolation method and demonstrated the

\end{abstract}

\begin{keywords}Symplectic model reduction, Hamiltonian system, Greedy basis generation, Symplectic Discrete Empirical Interpolation (SDEIM) \end{keywords}

\begin{AMS}\end{AMS}

\pagestyle{myheadings}
\thispagestyle{plain}
\markboth{Babak Maboudi Afkham and Jan S. Hesthaven}{Symplectic Model reduction of Parametric Hamiltonian Systems}

\input{./chapters/1.introduction.tex}

\input{./chapters/2.mor.tex}

\input{./chapters/3.ham.tex}

\input{./chapters/4.sympmor.tex}

\input{./chapters/5.numres.tex}

\input{./chapters/6.conclusion.tex}

\input{./chapters/appendix.tex}

\bibliography{ref}
\bibliographystyle{plain}

\end{document}
%% end of file `docultex.tex'

### 2016-11-15 greedy_hamiltonian_new_template/chapters/1.introduction.tex
fist commit

--- 2016-11-15 greedy_hamiltonian/chapters/1.introduction.tex
@@
 
-In the context of Hamiltonian and Lagrangian systems, recent work suggest modifications of POD to preserve geometric structure.
+In the context of Hamiltonian and Lagrangian systems, recent work suggests modifications of POD to preserve geometric structure.
 Lall et al.
@@
 Section \ref{chap:SyMo:1} discusses the greedy generation of a symplectic reduced basis as well as other SVD-based symplectic model reduction techniques.
-Accuracy, stability, and efficiency of the greedy method compared to other SVD based methods are discussed in Section \ref{chap:NuRe:1}.
+Accuracy, stability, and efficiency of the greedy method compared to other SVD-based methods are discussed in Section \ref{chap:NuRe:1}.
 Finally we offer the conclusion in Section \ref{chap:Con:1}.

### 2016-11-15 greedy_hamiltonian_new_template/chapters/2.mor.tex
fist commit

--- 2016-11-15 greedy_hamiltonian/chapters/2.mor.tex
@@
 Model order reduction is based on the assumption that $\mathcal M$ is of low dimension \cite{Anonymous:2016wl,Antoulas:2005:ALD:1088857}, and that the span of appropriately chosen basis vectors $\{v_i\}_{i=1}^k$ covers most of the solution manifold with a small error.
-The set $\{v_i\}_{i=1}^k$ is denoted as the reduced basis and its span as the reduced space.
+The set $\{v_i\}_{i=1}^k$ is denoted as the \emph{reduced basis} and its span as the \emph{reduced space}.
 Assuming that a $k$-dimensional $(k\ll n)$ reduced basis is given, the approximated solution can be represented as
@@
 \subsection{Proper Orthogonal Decomposition} \label{chap:MoOr.PrOr:1}
-Let $\mathbf x (t_i,\omega_j)$ with $i=1,\dots,m$ and $j=1,\dots,n$ be a finite number of samples, referred to as snapshots, from the solution manifold (\ref{eq:MoOr:2}).
+Let $\mathbf x (t_i,\omega_j)$ with $i=1,\dots,m$ and $j=1,\dots,n$ be a finite number of samples, referred to as \emph{snapshots}, from the solution manifold (\ref{eq:MoOr:2}).
 If we Suppose that a reduced basis $V$ is provided, the projection operator from $\mathbb R^n$ onto the reduced space can be constructed as $VV^T$.
@@
 \end{equation}
-Here $S$ is the snapshot matrix,, containing snapshots $\mathbf x(t_i,\omega_j)$ in its columns, $\|\cdot \|_F$ is the Frobenius norm of a matrix and $I_k$ is the identity matrix of size $k$.
+Here $S$ is the snapshot matrix, containing snapshots $\mathbf x(t_i,\omega_j)$ in its columns, $\|\cdot \|_F$ is the Frobenius norm of a matrix and $I_k$ is the identity matrix of size $k$.
 According to Schmidt-Mirsky-Eckard-Young theorem \cite{Antoulas:2005:ALD:1088857}, the solution to the optimization (\ref{eq:MoOr:6}) is equivalent to the truncated singular value decomposition (SVD) of the snapshot matrix $S$ given by
@@
 Suppose that the right hand side function in (\ref{eq:MoOr:1}) is of the form $\mathbf f(t,\mathbf x , \omega) = L\mathbf x + \mathbf g(t,\mathbf x ,\omega)$, where $L\in \mathbb R^{n\times n}$ reflects the linear part, and $\mathbf g$ is a nonlinear function.
-Now suppose a $k$-dimensional reduced basis $V$ is provided, the reduced system takes the form
+Now suppose a $k$-dimensional reduced basis $V$ is provided.
+The reduced system takes the form
 \begin{equation} \label{eq:MoOr:8}
@@
 where $e_{p_i}$ is the $p_i$-th column of the identity matrix $I_n$.
-Hence, multiplying $P$ on $\mathbf g$ selects components $p_1,\dots,p_m$ of $\mathbf g$.
+Hence, multiplying $P$ with $\mathbf g$ selects components $p_1,\dots,p_m$ of $\mathbf g$.
 If we assume that $P^TU$ is non-singular, the coefficient vector $\mathbf c$ can be uniquely determined from
@@
 \end{equation}
-with $\mathbf x$ the configuration vector. In such cases the interpolating index matrix $P$ and the nonlinear function $\mathbf g$ commute. I.e., we have
+with $\mathbf x$ the configuration vector. In such cases the interpolating index matrix $P$ and the nonlinear function $\mathbf g$ commute, i.e.,
 \begin{equation} \label{eq:MoOr:16}

### 2016-11-15 greedy_hamiltonian_new_template/chapters/4.sympmor.tex
fist commit

--- 2016-11-15 greedy_hamiltonian/chapters/4.sympmor.tex
@@
 \end{equation}
-Multiplying both sides with the symplectic inverse of $A$ and using the chain rule we have,
+Multiplying both sides with the symplectic inverse of $A$ and using the chain rule we have
 \begin{equation} \label{eq:SyMo:3}
@@
 Note that the original and the reduced systems are derived from different Hamiltonians.
-In the next proposition we show that the loss in the Hamiltonian is constant in time.
+In the next proposition we show that the error in the Hamiltonian is constant in time.
 
@@
 \end{equation}
-and define $\Phi=[u_i,\dots,u_k]$, where $u_i$ is the $i$-th left singular vector of $S_{\text{combined}}$.
+and define $\Phi=[u_1,\dots,u_k]$, where $u_i$ is the $i$-th left singular vector of $S_{\text{combined}}$.
 It is shown in \cite{Peng:2014di} that among all symplectic bases of the form (\ref{eq:SyMo:9}) cotangent lift minimizes the projection error.
@@
 
-\subsubsection{\bf The Greedy Approach to Symplectic Basis Generation} The greedy generation of the reduced basis is an iterative procedure which, in each iteration, adds the two best possible basis vectors to the symplectic basis to enhance overall accuracy.
+\subsubsection{The Greedy Approach to Symplectic Basis Generation} The greedy generation of the reduced basis is an iterative procedure which, in each iteration, adds the two best possible basis vectors to the symplectic basis to enhance overall accuracy.
 In contrast to the cotangent lift and the complex SVD methods, the greedy approach does not require the symplectic basis to have a specific structure.
@@
 
-{\bf Symplectic Gram-Schmidt:} As discussed in Section \ref{chap:Hasy:1}, any finite dimensional linear symplectic vector space has a symplectic basis that satisfy conditions (\ref{eq:Hasy:4}).
-Suppose that $S=\{ e_1 , \dots , e_k \} \bigcup \{ f_1 , \dots , f_k \}$ is a symplectic basis and $a$ and $b$ are two vectors such that $a,b \not \in \text{span}(S)$ and $a \nparallel b$.
+{\bf Symplectic Gram-Schmidt:} As discussed in Section \ref{chap:Hasy:1}, any finite dimensional symplectic linear vector space has a symplectic basis that satisfies conditions (\ref{eq:Hasy:4}).
+Suppose that $S=\{ e_1 , \dots , e_k \} \bigcup \{ f_1 , \dots , f_k \}$ is a symplectic basis and $a$ and $b$ are two vectors such that $a,b \not \in \text{span}(S)$ and $a \not \not \parallel b$.
 We seek to grow the size of $S$ by the vectors $a$ and $b$.
@@
 Similar to the QR decomposition, the symplectic Gram-Schmidt process leads to a matrix decomposition for any full ranked matrix of size $2n\times 2k$.
-More details about this matrix decomposition are given in the Appendix.
+More details about this matrix decomposition are given in Appendix \ref{chap:Append:1}.
 
@@
 
-We then propagate (\ref{eq:MoOr:1}) in time to produce trajectory snapshots $S=\{ \mathbf z(t_i,\omega_{k+1}) | i = 1,\dots,M \}$.
+We then propagate (\ref{eq:Hasy:7}) in time to produce trajectory snapshots $S=\{ \mathbf z(t_i,\omega_{k+1}) | i = 1,\dots,M \}$.
 The next basis vector is the snapshot that maximises the projection error (\ref{eq:SyMo:8})
@@
 \caption{The greedy algorithm for generation of a symplectic basis} \label{alg:SyMo:3}
-{\bf Input:} Tolerated loss in the Hamiltonian $\delta$, parameter set $\Omega_N = \{\omega_1,\dots,\omega_N\}$, initial condition $\mathbf z_0(\omega)$
+{\bf Input:} Tolerated loss in the Hamiltonian $\delta$, parameter set $\Gamma_N = \{\omega_1,\dots,\omega_N\}$, initial condition $\mathbf z_0(\omega)$
 \begin{enumerate}
@@
 The greedy algorithm on the other hand, is a cheap alternative to the nonlinear optimization and it allows flexibility in terms of the structure.
-\item The greedy method reduces the total cost of the offline stage by solving the original system (\ref{eq:Hasy:7}) only $k$ times to obtain a reduced system of size $2k$, unlike POD based methods that require time integration of (\ref{eq:Hasy:7}) for the entire parameter space.
+\item The greedy method reduces the total cost of the offline stage by solving the original system (\ref{eq:Hasy:7}) only $k$ times to obtain a reduced system of size $2k$, unlike POD-based methods that require time integration of (\ref{eq:Hasy:7}) for the entire parameter space.
 \item If enrichment of the symplectic basis is required, the greedy algorithm can add new basis vectors. The POD-based methods, are bound to re-compute the basis for the refined parameter set.
@@
 \begin{definition}
-Let $S$ be a subset of $\mathbb R^{2n}$ and $Y_n$ be a general $n$-dimensional subspace of $\mathbb R^{2n}$. The angle between $S$ and $Y_n$ is given by
+Let $S$ be a subset of $\mathbb R^{m}$ and $Y_n$, $n\leq m$, be a general $n$-dimensional subspace of $\mathbb R^{m}$. The angle between $S$ and $Y_n$ is given by
 \begin{equation} \label{eq:new4}
@@
 \end{equation}
-The Kolmogorov $n$-width of $S$ in $\mathbb R^n$ is given by
+The Kolmogorov $n$-width of $S$ in $\mathbb R^m$ is given by
 \begin{equation} \label{eq:new5}
-	d_{n}(S,\mathbb{R}^n) := \inf_{Y_n} E(S,Y_n) = \inf_{Y_n} \sup_{s\in S} \inf_{y\in Y_n} \|s-y\|_2
+	d_{n}(S,\mathbb{R}^m) := \inf_{Y_n} E(S,Y_n) = \inf_{Y_n} \sup_{s\in S} \inf_{y\in Y_n} \|s-y\|_2
 \end{equation}
@@
 \begin{aligned}
-	|\alpha_i(s)| = \frac{|\Omega(\bar{\xi}_i,s)|}{|\Omega(\xi_i,\bar{\xi}_i)|} &= \frac{\Omega( \bar{\xi}_i, s - P_{2(k-1)}(s))}{|\Omega(\xi_i,\bar{\xi}_i)|} \\
+	|\alpha_i(s)| = \frac{|\Omega(\bar{\xi}_i,s)|}{|\Omega(\xi_i,\bar{\xi}_i)|} &= \frac{|\Omega( \bar{\xi}_i, s - P_{2(k-1)}(s))|}{|\Omega(\xi_i,\bar{\xi}_i)|} \\
 	&\leq \frac{\|\bar{\xi}_i\|_2 \| s - P_{2(k-1)}(s) \|_2}{ \|\xi_i\|_2 \|\bar{\xi}_i\|_2 } \\
@@
 Looking at the definition of Kolmogorov $n$-width we observe that for any $\theta > 1$ we can find a subspace $Y_{2k}$ such that $E(S,Y_{2k}) \leq \theta d_{2k}(S,\mathbb R^n)$.
-Hence can find vectors $v_1,\dots,v_k,u_1,\dots,u_k\in Y_{2k}$ such that
+Hence we can find vectors $v_1,\dots,v_k,u_1,\dots,u_k\in Y_{2k}$ such that
 \begin{equation} \label{eq:new13}
@@
 	Let $S$ be a compact subset of $\mathbb{R}^{2n}$ with exponentially small Kolmogorov $n$-width $d_{n}\leq c\exp(-\alpha k)$ with $\alpha > \log3$.
-Then there exists $\beta>0$ such that the symplectic subspaces $A_{2k}$ generated by the greedy algorithm provides exponential approximation properties such that
+Then there exists $\beta>0$ such that the symplectic subspaces $A_{2k}$ generated by the greedy algorithm provide exponential approximation properties such that
 \begin{equation} \label{eq:new23}
@@
 \begin{equation} \label{eq:new28}
-	\tilde N(\mathbf y) = A^+ \mathbb J_{2n}(A^+)^T H_2(A \mathbf y) = \mathbb J_{2k} H_2(A \mathbf y),
+	\tilde N(\mathbf y) = A^+ \mathbb J_{2n}(A^+)^T \nabla_{\mathbf y} H_2(A \mathbf y) = \mathbb J_{2k} \nabla_{\mathbf y} H_2(A \mathbf y),
 \end{equation}

### 2016-11-15 greedy_hamiltonian_new_template/chapters/5.numres.tex
fist commit

--- 2016-11-15 greedy_hamiltonian/chapters/5.numres.tex
@@
 \end{equation}
-
 The initial condition is given by
@@
 
-Figure \ref{fig:NuRe:3} shows the solution of the linear wave equation for parameter value $c=0.1019$, chosen to be different from training parameters, at $t=0$, $t = 1$ and $t=2$.
+Figure \ref{fig:NuRe:1a} shows the solution of the linear wave equation for parameter value $c=0.1019$, chosen to be different from training parameters, at $t=0$, $t = 1$ and $t=2$.
 While we see instability and divergence from the exact solution for the POD reduced system, the symplectic methods provide a good approximation of the full model.
 
-The decay of the singular values for the POD, cotangent lift and the complex SVD method are presented in Figure \ref{fig:NuRe:4}(a).
+The decay of the singular values for the POD, cotangent lift and the complex SVD method are presented in Figure \ref{fig:NuRe:2a}.
 The fast decay of the singular values suggests that a low dimensional solution manifold indeed exists.
-However since the linear subspace constructed by the POD is not symplectic, we observe the blow up of the Hamiltonian function in Figure \ref{fig:NuRe:4}(b) and the instability of the solution in Figure \ref{fig:NuRe:3}.
-The symplectic methods (using a reduced basis of the same size as POD) preserve the Hamiltonian function as is seen in Figure \ref{fig:NuRe:4}(b).
+However since the linear subspace constructed by the POD is not symplectic, we observe the blow up of the Hamiltonian function in Figure \ref{fig:NuRe:2b} and the instability of the solution in Figure \ref{fig:NuRe:1}.
+The symplectic methods (using a reduced basis of the same size as POD) preserve the Hamiltonian function as is seen in Figure \ref{fig:NuRe:2b}.
 
-Figure \ref{fig:NuRe:4}(c) shows the $L^2$-error between the solution of the full model and the reduced systems constructed by different methods.
+Figure \ref{fig:NuRe:2c} shows the $L^2$-error between the solution of the full model and the reduced systems constructed by different methods.
 We note that the error for the POD reduced system rapidly increases, confirming that projection based reduced system does not yield a stable solution.
@@
 
-\begin{figure*}[t]
-\begin{center}
-\begin{tabular}{cc}
-	\includegraphics[width=0.4\textwidth]{./figs/wave/solution/solution_t0} &
-	\includegraphics[width=0.4\textwidth]{./figs/wave/solution/solution_t1} \\
-	$t=0$ & $t = 1$
-\end{tabular}
-\end{center}
-\end{figure*}
+%\begin{figure*}[t]
+%\begin{center}
+%\begin{tabular}{cc}
+%	\includegraphics[width=0.4\textwidth]{./figs/wave/solution/solution_t0} &
+%	\includegraphics[width=0.4\textwidth]{./figs/wave/solution/solution_t1} \\
+%	$t=0$ & $t = 1$
+%\end{tabular}
+%\end{center}
+%\end{figure*}
 
-\begin{figure}[t]
+%\begin{figure}[t]
+%\centering
+%\begin{tabular}{cc}
+%\includegraphics[width=0.4\textwidth]{./figs/wave/solution/solution_t2} \\
+%	$t=2$
+%\end{tabular}
+%\caption{The solution $q$ at $t=0$, $t=1$ and $t=2$ of the linear wave equation for parameter value $c= 0.1019$ different from training parameters.
+Here the solution of full system together with the solution of the POD, cotangent lift, complex SVD and the greedy reduced system is presented.}\label{fig:NuRe:3}
+%\end{figure}
+
+%\begin{figure*}[t]
+%\begin{center}
+%\begin{tabular}{cc}
+%	\includegraphics[width=0.5\textwidth]{./figs/wave/singular} &
+%	\includegraphics[width=0.5\textwidth]{./figs/wave/hamiltonian} \\
+%	(a) & (b)
+%\end{tabular}
+%\end{center}
+%\end{figure*}
+
+%\begin{figure}[t] 
+%\begin{center}
+%\begin{tabular}{c}
+%	\includegraphics[width=0.5\textwidth]{./figs/wave/error} \\
+%	(c)
+%\end{tabular}
+%\end{center}
+%\caption{(a) The decay of singular values for POD, cotangent lift and complex SVD methods.
+(b) The $L^2$ error between the solution of th full system and the reduced system for different model reduction methods for $t \in [0,30]$.
+(c) Plot of the Hamiltonian function for $t \in [0,30]$.}\label{fig:NuRe:4}
+%\end{figure}
+
+%\begin{figure*}[t]
+%  \centering
+%  \includegraphics[width=0.5\textwidth]{./figs/wave/hamiltonian}
+%  \label{fig:testfig}
+%\end{figure*}
+
+\begin{figure}
+
+\begin{minipage}{.5\linewidth}
 \centering
-\begin{tabular}{cc}
-\includegraphics[width=0.4\textwidth]{./figs/wave/solution/solution_t2} \\
-	$t=2$
-\end{tabular}
+\subfloat[$t=0$]{\label{fig:NuRe:1a}\includegraphics[width=1\textwidth]{./figs/wave/solution/solution_t0}}
+\end{minipage}%
+\begin{minipage}{.5\linewidth}
+\centering
+\subfloat[$t=1$]{\label{fig:NuRe:1b}\includegraphics[width=\textwidth]{./figs/wave/solution/solution_t1}}
+\end{minipage}\par\medskip
+\centering
+\subfloat[$t=2$]{\label{fig:NuRe:1c}\includegraphics[width=0.5\textwidth]{./figs/wave/solution/solution_t2}}
 \caption{The solution $q$ at $t=0$, $t=1$ and $t=2$ of the linear wave equation for parameter value $c= 0.1019$ different from training parameters.
-Here the solution of full system together with the solution of the POD, cotangent lift, complex SVD and the greedy reduced system is presented.}\label{fig:NuRe:3}
+Here the solution of full system together with the solution of the POD, cotangent lift, complex SVD and the greedy reduced system is presented.}
+\label{fig:NuRe:1}
 \end{figure}
 
-\begin{figure*}[t]
-\begin{center}
-\begin{tabular}{cc}
-	\includegraphics[width=0.5\textwidth]{./figs/wave/singular} &
-	\includegraphics[width=0.5\textwidth]{./figs/wave/hamiltonian} \\
-	(a) & (b)
-\end{tabular}
-\end{center}
-\end{figure*}
+\begin{figure}
 
-\begin{figure}[t] 
-\begin{center}
-\begin{tabular}{c}
-	\includegraphics[width=0.5\textwidth]{./figs/wave/error} \\
-	(c)
-\end{tabular}
-\end{center}
+\begin{minipage}{.5\linewidth}
+\centering
+\subfloat[]{\label{fig:NuRe:2a}\includegraphics[width=1\textwidth]{./figs/wave/singular}}
+\end{minipage}%
+\begin{minipage}{.5\linewidth}
+\centering
+\subfloat[]{\label{fig:NuRe:2b}\includegraphics[width=\textwidth]{./figs/wave/hamiltonian}}
+\end{minipage}\par\medskip
+\centering
+\subfloat[]{\label{fig:NuRe:2c}\includegraphics[width=0.5\textwidth]{./figs/wave/error}}
+
 \caption{(a) The decay of singular values for POD, cotangent lift and complex SVD methods.
-(b) The $L^2$ error between the solution of th full system and the reduced system for different model reduction methods for $t \in [0,30]$.
-(c) Plot of the Hamiltonian function for $t \in [0,30]$.}\label{fig:NuRe:4}
+(b) Plot of the Hamiltonian function for $t \in [0,30]$.
+(c) The $L^2$ error between the solution of the full system and the reduced system for different model reduction methods for $t \in [0,30]$.}
+\label{fig:NuRe:2}
 \end{figure}
@@
 
-We discretize the torus into $N$ equidistant points and take define $\Delta x = L/N$, $x_i = i\Delta x$, $q_i=q(t,x,\epsilon)$ and $p_i = p(t,x_i,\omega)$ for $i = 1 ,\dots,N$.
-A central finite differences scheme is used to discretize (\ref{eq:NuRe:11}) as
+We discretize the torus into $N$ equidistant points and take $\Delta x = L/N$, $x_i = i\Delta x$, $q_i=q(t,x_i,\epsilon)$ and $p_i = p(t,x_i,\omega)$ for $i = 1 ,\dots,N$.
+A central finite differences scheme is used to discretize (\ref{eq:NuRe:12}) as
 \begin{equation}  \label{eq:NuRe:14}
@@
 We compared the reduced system obtained by the greedy algorithm with the cotangent lift, complex SVD, DEIM, symplectic DEIM and also POD.
-For the-SVD based methods, we discretize the parameter space $[0.9,1.1]$ into $M=500$ equidistant grid points to recover the discrete parameter space $\Gamma_M = \{\epsilon_1,\dots,\epsilon_M \}$, and gather trajectory snapshots for each $\epsilon_i$ for $i = 1,\dots,M$ in the snapshots matrix $S$.
+For the SVD-based methods, we discretize the parameter space $[0.9,1.1]$ into $M=500$ equidistant grid points to recover the discrete parameter space $\Gamma_M = \{\epsilon_1,\dots,\epsilon_M \}$, and gather trajectory snapshots for each $\epsilon_i$ for $i = 1,\dots,M$ in the snapshots matrix $S$.
 All reduced system are taken to have identical sizes ($k=90$ for symplectic methods and $k=180$ for the POD method).
@@
 
-\begin{figure*}[t]
-\begin{center}
-\begin{tabular}{cc}
-	\includegraphics[width=0.5\textwidth]{./figs/schrodinger/solution/solution_t0} &
-	\includegraphics[width=0.5\textwidth]{./figs/schrodinger/solution/solution_t10} \\
-	$t = 0$ & $t = 10$
-\end{tabular}
-\end{center}
-\end{figure*}
+%\begin{figure*}[t]
+%\begin{center}
+%\begin{tabular}{cc}
+%	\includegraphics[width=0.5\textwidth]{./figs/schrodinger/solution/solution_t0} &
+%	\includegraphics[width=0.5\textwidth]{./figs/schrodinger/solution/solution_t10} \\
+%	$t = 0$ & $t = 10$
+%\end{tabular}
+%\end{center}
+%\end{figure*}
+%
+%\begin{figure}[t] 
+%\begin{center}
+%\begin{tabular}{c}
+%	\includegraphics[width=0.5\textwidth]{./figs/schrodinger/solution/solution_t20}\\
+%	$$t = 20$$
+%\end{tabular}
+%\end{center}
+%\caption{The solution $|u(t,x)| = \sqrt{q^2 + p^2}$ at $t=0$, $t=10$ and $t=20$ of the Nonlinear Schr\"odinger equation for parameter value $\epsilon = 1.0932$.
+Here the solution of full system together with the solution of the POD, cotangent lift, complex SVD and the greedy reduced system is presented.}\label{fig:NuRe:1}
+%\end{figure}
+%
+%\begin{figure*}[t]
+%\begin{center}
+%\begin{tabular}{cc}
+%	\includegraphics[width=0.5\textwidth]{./figs/schrodinger/singular} &
+%	\includegraphics[width=0.5\textwidth]{./figs/schrodinger/hamiltonian} \\
+%	(a) & (b)
+%\end{tabular}
+%\end{center}
+%\end{figure*}
+%
+%\begin{figure}[t] 
+%\begin{center}
+%\begin{tabular}{c}
+%	\includegraphics[width=0.5\textwidth]{./figs/schrodinger/error} \\
+%	(c)
+%\end{tabular}
+%\end{center}
+%\caption{(a) The decay of singular values for POD, cotangent lift and complex SVD methods.
+(b) The $L^2$ error between the solution of th full system and the reduced system for different model reduction methods for $t \in [0,30]$.
+(c) Plot of the Hamiltonian function for $t \in [0,30]$.}\label{fig:NuRe:2}
+%\end{figure}
 
-\begin{figure}[t] 
-\begin{center}
-\begin{tabular}{c}
-	\includegraphics[width=0.5\textwidth]{./figs/schrodinger/solution/solution_t20}\\
-	$$t = 20$$
-\end{tabular}
-\end{center}
+\begin{figure}
+
+\begin{minipage}{.5\linewidth}
+\centering
+\subfloat[$t=0$]{\label{fig:NuRe:3a}\includegraphics[width=1\textwidth]{./figs/schrodinger/solution/solution_t0}}
+\end{minipage}%
+\begin{minipage}{.5\linewidth}
+\centering
+\subfloat[$t=1$]{\label{fig:NuRe:3b}\includegraphics[width=\textwidth]{./figs/schrodinger/solution/solution_t10}}
+\end{minipage}\par\medskip
+\centering
+\subfloat[$t=2$]{\label{fig:NuRe:3c}\includegraphics[width=0.5\textwidth]{./figs/schrodinger/solution/solution_t20}}
 \caption{The solution $|u(t,x)| = \sqrt{q^2 + p^2}$ at $t=0$, $t=10$ and $t=20$ of the Nonlinear Schr\"odinger equation for parameter value $\epsilon = 1.0932$.
-Here the solution of full system together with the solution of the POD, cotangent lift, complex SVD and the greedy reduced system is presented.}\label{fig:NuRe:1}
+Here the solution of full system together with the solution of the POD, cotangent lift, complex SVD and the greedy reduced system is presented.}
+\label{fig:NuRe:3}
 \end{figure}
 
-\begin{figure*}[t]
-\begin{center}
-\begin{tabular}{cc}
-	\includegraphics[width=0.5\textwidth]{./figs/schrodinger/singular} &
-	\includegraphics[width=0.5\textwidth]{./figs/schrodinger/hamiltonian} \\
-	(a) & (b)
-\end{tabular}
-\end{center}
-\end{figure*}
+\begin{figure}
 
-\begin{figure}[t] 
-\begin{center}
-\begin{tabular}{c}
-	\includegraphics[width=0.5\textwidth]{./figs/schrodinger/error} \\
-	(c)
-\end{tabular}
-\end{center}
+\begin{minipage}{.5\linewidth}
+\centering
+\subfloat[]{\label{fig:NuRe:4a}\includegraphics[width=1\textwidth]{./figs/schrodinger/singular}}
+\end{minipage}%
+\begin{minipage}{.5\linewidth}
+\centering
+\subfloat[]{\label{fig:NuRe:4b}\includegraphics[width=\textwidth]{./figs/schrodinger/hamiltonian}}
+\end{minipage}\par\medskip
+\centering
+\subfloat[]{\label{fig:NuRe:4c}\includegraphics[width=0.5\textwidth]{./figs/schrodinger/error}}
+
 \caption{(a) The decay of singular values for POD, cotangent lift and complex SVD methods.
-(b) The $L^2$ error between the solution of th full system and the reduced system for different model reduction methods for $t \in [0,30]$.
-(c) Plot of the Hamiltonian function for $t \in [0,30]$.}\label{fig:NuRe:2}
+(b) The $L^2$ error between the solution of the full system and the reduced system for different model reduction methods for $t \in [0,30]$.
+(c) Plot of the Hamiltonian function for $t \in [0,30]$.}
+\label{fig:NuRe:4}
 \end{figure}
 
-Figure \ref{fig:NuRe:1} shows the solution of the Schr\"odinger equation for parameter value $\epsilon = 1.0932$ at $t=0$, $t=10$ and $t=20$.
+Figure \ref{fig:NuRe:3} shows the solution of the Schr\"odinger equation for parameter value $\epsilon = 1.0932$ at $t=0$, $t=10$ and $t=20$.
 We first compare the reduced system obtained by the greedy algorithm with POD, cotangent lift and the complex SVD method.
 The size of the reduced systems are taken identical for all methods ($k=180$ for POD and $k=90$ for the rest).
-Although the decay of the singular values in Figure \ref{fig:NuRe:2} suggests that the accuracy of the POD reduced system should be comparable to that of the other methods, we observe instabilities in the solution at $t=10$.
+Although the decay of the singular values in Figure \ref{fig:NuRe:4a} suggests that the accuracy of the POD reduced system should be comparable to that of the other methods, we observe instabilities in the solution at $t=10$.
 The greedy, the cotangent lift and the complex SVD method, on the other hand, generate a stable reduced system that accurately approximates the solution of the full model.
 
-The value of the Hamiltonian function for the reduced systems discussed above as a function of time is illustrated in Figure \ref{fig:NuRe:2}(b).
+The value of the Hamiltonian function for the reduced systems discussed above as a function of time is illustrated in Figure \ref{fig:NuRe:4b}.
 We see that the symplectic methods preserve the Hamiltonian function.
@@
 
-Figure \ref{fig:NuRe:2}(c) illustrates the $L^2$-error between the solution of the full model with the reduced systems generated by different methods.
+Figure \ref{fig:NuRe:4c} illustrates the $L^2$-error between the solution of the full model with the reduced systems generated by different methods.
 We first observe that symplectic methods yield a lower computational error compared to non-symplectic methods.

### 2016-11-15 greedy_hamiltonian_new_template/chapters/6.conclusion.tex
fist commit

\section{Conclusion} \label{chap:Con:1}

In this paper, we present a greedy approach for the construction of a reduced system that preserves the geometric structure of Hamiltonian systems. An iteration of the greedy method comprises searching the parameter space using the error in the Hamiltonian, to find the best basis vectors that increase the overall accuracy of the reduced basis. We argued that for a compact subset with exponentially small Kolmogorov $n$-width we recover exponentially fast convergence of the greedy algorithm. For fast approximation of nonlinear terms, the basis obtained by the greedy was combined with the symplectic DEIM to construct a Hamiltonian reduced system.

%Since the reduced system obtained by the greedy method is Hamiltonian, it preserves stability, and was illustrated in the numerical results.

The numerical results demonstrated that the greedy method can save substantial computational cost in the offline stage as compared to alternative SVD-based techniques. Also since the reduced system obtained by the greedy method is Hamiltonian, the greedy method yields a stable reduced system. Symplectic DEIM also effectively reduced computational cost of approximating nonlinear terms while preserving stability and symplectic structure. Hence, the greedy method is an efficient model reduction technique that provides an accurate and stable reduced system for large-scale parametric Hamiltonian systems.

 % it preserves stability, and was illustrated in the numerical results.
%We showed that the Hamiltonian is preserved in the 
%As the accuracy of the reduced system obtained by the greedy is comparable with the SVD-based methods, the greedy method is a promising dimension reduction technique for parametric Hamiltonian systems.

### 2016-11-15 greedy_hamiltonian_new_template/paper.tex
fist commit

--- 2016-11-15 greedy_hamiltonian/paper.tex
@@
-%%
-%% This is file `docultex.tex', 
-%% Documentation for siam supplemental file macros for use with LaTeX 2e
-%% 
-%% December 19, 2013
-%%
-%% Supplemental Files Supported Version 1.0.0
-%% 
-%% You are not allowed to change this file. 
-%% 
-%% You are allowed to distribute this file under the condition that 
-%% it is distributed together with all of the files in the siam macro 
-%% distribution. These are:
-%%
-%%  siamltex1213.cls (main LaTeX macro for SIAM)
-%%  siam10.clo   (size option for 10pt papers)
-%%  subeqn.clo   (allows equation numbners with lettered subelements)
-%%  siam.bst     (bibliographic style file for BibTeX)
-%%  docultex.tex (this file)
-%%
-%% If you receive only some of these files from someone, complain! 
-%% 
-%% You are NOT ALLOWED to distribute this file alone. You are NOT 
-%% ALLOWED to take money for the distribution or use of either this 
-%% file or a changed version, except for a nominal charge for copying 
-%% etc. 
-%% \CharacterTable
-%%  {Upper-case    \A\B\C\D\E\F\G\H\I\J\K\L\M\N\O\P\Q\R\S\T\U\V\W\X\Y\Z
-%%   Lower-case    \a\b\c\d\e\f\g\h\i\j\k\l\m\n\o\p\q\r\s\t\u\v\w\x\y\z
-%%   Digits        \0\1\2\3\4\5\6\7\8\9
-%%   Exclamation   \!     Double quote  \"     Hash (number) \#
-%%   Dollar        \$     Percent       \%     Ampersand     \&
-%%   Acute accent  \'     Left paren    \(     Right paren   \)
-%%   Asterisk      \*     Plus          \+     Comma         \,
-%%   Minus         \-     Point         \.     Solidus       \/
-%%   Colon         \:     Semicolon     \;     Less than     \<
-%%   Equals        \=     Greater than  \>     Question mark \?
-%%   Commercial at \@     Left bracket  \[     Backslash     \\
-%%   Right bracket \]     Circumflex    \^     Underscore    \_
-%%   Grave accent  \`     Left brace    \{     Vertical bar  \|
-%%   Right brace   \}     Tilde         \~}
+% SIAM Article Template
+\documentclass[review]{siamart}
+\usepackage{mwe}
+\usepackage{subfig}
 
-\documentclass[final,leqno,onefignum,onetabnum]{siamltex1213}
+% Information that is shared between the article and the supplement
+% (title and author information, macros, packages, etc.) goes into
+% ex_shared.tex. If there is no supplement, this file can be included
+% directly.
 
-%-------------------------------------------------------------------------
-%
-%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
-%
-%-------------------------------------------------------------------------
+\input{ex_shared}
 
-\usepackage{./pkg/packages}
+% Optional PDF information
+\ifpdf
+\hypersetup{
+  pdftitle={\TheTitle},
+  pdfauthor={\TheAuthors}
+}
+\fi
 
-\title{Structure Preserving Model Reduction of Parametric Hamiltonian Systems}
+% The next statement enables references to information in the
+% supplement. See the xr-hyperref package for details.
 
-\author{
-	Babak Maboudi Afkham\thanks{Department of Mathematics, Chair of Computational Mathematics and Simulation Schience (MCSS), \'Ecole Polytechnique F\'ed\'erale de Lausanne, Switzerland (babak.maboudi@epfl.ch)}
-	\and 
-	Jan S.
-Hesthaven\thanks{Department of Mathematics, Chair of Computational Mathematics and Simulation Schience (MCSS), \'Ecole Polytechnique F\'ed\'erale de Lausanne, Switzerland (jan.hesthaven@epfl.ch)}}
+\externaldocument{ex_supplement}
+
+% FundRef data to be entered by SIAM
+%<funding-group>
+%<award-group>
+%<funding-source>
+%<named-content content-type="funder-name"> 
+%</named-content> 
+%<named-content content-type="funder-identifier"> 
+%</named-content>
+%</funding-source>
+%<award-id> </award-id>
+%</award-group>
+%</funding-group>
 
 \begin{document}
+
 \maketitle
-\slugger{mms}{xxxx}{xx}{x}{x--x}%slugger should be set to mms, siap, sicomp, sicon, sidma, sima, simax, sinum, siopt, sisc, or sirev
 
+% REQUIRED
 \begin{abstract}
@@
 The efficiency, accuracy, and stability of this model reduction technique is illustrated through simulations of the parametric wave equation and the parametric Schr\"odinger equation.
-
-%Through the greedy selection of basis vectors, we used the Hamiltonian function as an error estimator to search for the
-
-%to add basis vectors, two at a time, and increase the overall accuracy of the reduced basis.
-It therefor requires a single numerical solution to be computed at each iteration which can save computational costs in stage offline stage.
-Under
-
-%We also illustrated that convergence rate of the greedy approach is proportional to the best possible convergence rate of any possible iterative update of the reduced basis.
-Moreover, for nonlinear Hamiltonian systems, we combined the reduced basis obtained from the greedy method with discrete empirical interpolation method and demonstrated the
-
 \end{abstract}
 
-\begin{keywords}Symplectic model reduction, Hamiltonian system, Greedy basis generation, Symplectic Discrete Empirical Interpolation (SDEIM) \end{keywords}
+% REQUIRED
+\begin{keywords}
+  Symplectic model reduction, Hamiltonian system, Greedy basis generation, Symplectic Discrete Empirical Interpolation (SDEIM)
+\end{keywords}
 
-\begin{AMS}\end{AMS}
+% REQUIRED
+\begin{AMS}
 
-\pagestyle{myheadings}
-\thispagestyle{plain}
-\markboth{Babak Maboudi Afkham and Jan S. Hesthaven}{Symplectic Model reduction of Parametric Hamiltonian Systems}
+\end{AMS}
 
@@
 
+\appendix
 \input{./chapters/appendix.tex}
 
+\bibliographystyle{siamplain}
 \bibliography{ref}
-\bibliographystyle{plain}
-
 \end{document}
-%% end of file `docultex.tex'

### 2016-11-15 greedy_hamiltonian_revised/chapters/1.introduction.tex
fist commit

--- 2016-11-15 greedy_hamiltonian_new_template/chapters/1.introduction.tex
@@
 Reduced basis methods have emerged as a powerful approach for the reduction of the intrinsic complexity of such models \cite{Ito:1998up,Ito:1998ch,Ito:2001ev,Peterson:1989ki}.
-These methods contain two stages: the offline and the online.
+These methods contain two stages: {\edit the offline stage and the online stage}.
 In the offline stage, one explors the parameter space to construct a low-dimensional basis that accurately represents the parametrized solution to the partial differential equation.
@@
 proper orthogonal decomposition (POD) \cite{Kunisch:2002er,Atwell:2001ja,Ravindran:2002hn}, require the exploration of the entire parameter space.
-This leads to a very expensive and often impractical offline stages when dealing with multi-dimensional paramtere domains.
+This leads to a very expensive and often impractical offline {\edit stage} when dealing with multi-dimensional paramtere domains.
 On the other hand, sampling techniques, usually of a greedy nature, search through the parameter space selectively often guided by an error estimate to certify the accuracy of the basis.
@@
 
-Beside computational complexity, another aspect of reduced order modeling is the preservation of structures and, in particular, stability of the original model.
+{\edit Besides} computational complexity, another aspect of reduced order modeling is the preservation of structures and, in particular, {\edit the} stability of the original model.
 In general, reduced order models do not guarantee that such properties are preserved \cite{Anonymous:pMn0O0Q4}.
@@
 In this way the geometric structure of the original system is inherited by the reduced system.
-For Hamiltonian systems Peng et al.
+{\edit Model reduction for port-Hamiltonian systems can be found in the works of Beattie et al.
+\cite{Chaturantabut:2016he}, Polyuga et al.
+\cite{Polyuga:2010gj} and the references therein.
+These works construct a reduced port-Hamiltonian system using Krylov or POD methods that carry the passivity and stability of the original system.} For Hamiltonian systems Peng et al.
 \cite{Peng:2014di}, using a symplectic transformation, constructs a reduced Hamiltonian, as an approximation to the Hamiltonian of the original system.
 As a result the reduced system preserves the symplectic structure.
-Although these methods preserve geometric structure, they use a POD-like approach for constructing the reduced basis, and are not well suitable for problems with a high-dimensional parameter domain.
+Although these methods preserve geometric structure, they use a POD-like approach for constructing the reduced basis, and are not well {\edit suited} for problems with a high-dimensional parameter domain.
 
@@
 On the other hand, since time integration of the original system is only required once per iteration, the proposed method saves substantial computational cost during the offline stage when compared to alternative POD-like approaches.
-Moreover, we demonstrate that assumptions, natural for the set of all solutions of the original Hamiltonian system under variation of parameters, lead to exponentially fast convergence of the greedy algorithm.
-For nonlinear Hamiltonian systems, we show how the basis can be combined with the discrete empirical interpolation method (DEIM) to allow a fast evaluation of nonlinear terms while maintaining the symplectic structure.
+{\edit It is well known that structured matrices, e.g.
+symplectic matrices, are not generally well-conditioned \cite{Karow:2006cf}.
+The greedy update of the symplectic basis presented here, yields a orthosymplectic basis, and therefore a norm bounded basis.} Moreover, we demonstrate that assumptions, natural for the set of all solutions of the original Hamiltonian system under variation of parameters, lead to exponentially fast convergence of the greedy algorithm.
+For nonlinear Hamiltonian systems, we show how the basis can be combined with the discrete empirical interpolation method (DEIM) {\edit \cite{Chaturantabut:2010cz,Barrault:2004kz}} to allow a fast evaluation of nonlinear terms while maintaining the symplectic structure.
 

### 2016-11-15 greedy_hamiltonian_revised/chapters/2.mor.tex
fist commit

--- 2016-11-15 greedy_hamiltonian_new_template/chapters/2.mor.tex
@@
 
-We define the solution manifold as the set of all solutions to (\ref{eq:MoOr:1}) under variation of parameter $\omega$
+We define the solution manifold as the set of all solutions to (\ref{eq:MoOr:1}) under variation of {\edit the parameters in $\Gamma$}
 \begin{equation} \label{eq:MoOr:2}
@@
 \end{equation}
-Note that the exact solution and solution manifold is not always available; we assume that we have a numerical integrator in hand that can approximate the solution to (\ref{eq:MoOr:1}) for any realization of $\omega$ with arbitrary accuracy.
+Note that the exact solution and solution manifold is not always available; we assume that we have a numerical integrator in hand that can approximate the solution to (\ref{eq:MoOr:1}) for any realization of $\omega$ {\edit with a given accuracy}.
 By abuse of notation we refer to $\mathbf x$ and $\mathcal M$ as the exact solution and the exact solution manifold, respectively, rather than the discrete solution and discrete solution manifold.
@@
 Let $\mathbf x (t_i,\omega_j)$ with $i=1,\dots,m$ and $j=1,\dots,n$ be a finite number of samples, referred to as \emph{snapshots}, from the solution manifold (\ref{eq:MoOr:2}).
-If we Suppose that a reduced basis $V$ is provided, the projection operator from $\mathbb R^n$ onto the reduced space can be constructed as $VV^T$.
+If we {\edit suppose} that a reduced basis $V$ is provided, the projection operator from $\mathbb R^n$ onto the reduced space can be constructed as $VV^T$.
 The proper orthogonal decomposition (POD) requires the total error of projecting all the snapshots onto the reduced space to be minimum.
@@
 Here $S$ is the snapshot matrix, containing snapshots $\mathbf x(t_i,\omega_j)$ in its columns, $\|\cdot \|_F$ is the Frobenius norm of a matrix and $I_k$ is the identity matrix of size $k$.
-According to Schmidt-Mirsky-Eckard-Young theorem \cite{Antoulas:2005:ALD:1088857}, the solution to the optimization (\ref{eq:MoOr:6}) is equivalent to the truncated singular value decomposition (SVD) of the snapshot matrix $S$ given by
+According to {\edit Schmidt-Mirsky-Eckart-Young theorem \cite{Markovsky:2011:LRA:2103589}}, the solution to the optimization (\ref{eq:MoOr:6}) is equivalent to the truncated singular value decomposition (SVD) of the snapshot matrix $S$ given by
 \begin{equation} \label{eq:MoOr:7}
@@
 \end{equation}
-Here $\sigma_i, u_i$ and $v_i$ are the singular values, left singular vectors and right singular vectors of $S$, respectively.
+Here $\sigma_i, u_i$ and $v_i$ are the singular values, left singular vectors and right singular vectors of $S$, respectively {\edit \cite{Markovsky:2011:LRA:2103589} }.
 
 \subsection{Discrete Empirical Interpolation Method (DEIM)} \label{chap:MoOr.DEIM:1} \nocite{Chaturantabut:2010cz}
-Let us now discuss the inefficiency of evaluating nonlinearities when using projection based reduced model.
+In this section we {\edit discuss the efficiency} of evaluating nonlinearities when using projection based reduced model.
 Suppose that the right hand side function in (\ref{eq:MoOr:1}) is of the form $\mathbf f(t,\mathbf x , \omega) = L\mathbf x + \mathbf g(t,\mathbf x ,\omega)$, where $L\in \mathbb R^{n\times n}$ reflects the linear part, and $\mathbf g$ is a nonlinear function.
@@
 
-To overcome this bottleneck we take an approach similar to what we did above.
+{\edit To overcome this bottleneck we take an approach similar to section \ref{chap:MoOr.PrOr:1} \cite{Chaturantabut:2010cz,Barrault:2004kz}}.
 Assume that the manifold $\mathcal M_{\mathbf g} = \{ \mathbf g(t,\mathbf x , \omega)| t\in \mathbb R, \mathbf x \in \mathbb R , \omega \in \Gamma\}$ is of a low dimension and that $\mathbf g$ can be approximated by a linear subspace of dimension $m\ll n$, spanned by the basis $\{ u_1 , \dots , u_m \}$, i.e.
@@
 
-To obtain the projection basis $U$, the POD can be applied to the ensemble of the discrete evaluation of $\mathbf g$.
-There is no additional cost for computing the nonlinear snapshots since they are generated when computing the trajectory snapshot matrix $S$.
+{\edit To obtain the projection basis $U$, the POD can be applied to the \emph{nonlinear snapshots}.
+The nonlinear snapshots are the ensemble of samples of the nonlinear term $\mathbf g(t_i,\mathbf x, \omega_j)$ with $i=1,\dots,m$ and $j=1,\dots,n$}.
+There is no additional cost for computing the nonlinear snapshots, since they are generated when computing the trajectory snapshot matrix $S$.
 The interpolating indices $p_1,\dots,p_m$ can be constructed as follows.
@@
 
-If we use an implicit time integration scheme, a similar inefficiency occurs for the computation of Newton iteration.
-In addition to the nonlinear term, the Jacobian of $\mathbf g(\mathbf x)$ with respect to the reduced state vector must be computed at each iteration.
-The Jacobian takes the form
+{\edit Numerical solution of (\ref{eq:MoOr:8}) sometimes involve computation of the Jacobian of the nonlinear function $\mathbf g(t,\mathbf x, \omega)$ with respect to the reduced state variable $\mathbf y$}
 \begin{equation} \label{eq:MoOr:9}
-	\mathbf J_{\mathbf y}(\mathbf g) = (WV)^{-1} \mathbf J_{\mathbf x}(\mathbf g) V.
+	\edit \mathbf J_{\mathbf x}(\mathbf g) = (WV)^{-1} \mathbf J_{\mathbf y}(\mathbf g) V,
 \end{equation}
-The complexity of computing (\ref{eq:MoOr:9}) is $\mathcal{O}(\alpha(n) +2n^2k+2nk^2+2nk)$ comprising several matrix-vector multiplications and an evaluation of the Jacobian which depends on the size of the original system.
+{\edit where $\mathbf J_\alpha(\mathbf g)$ is the Jacobian matrix of $\mathbf g$ with respect to the variable $\alpha$.} The complexity of computing (\ref{eq:MoOr:9}) is $\mathcal{O}(\alpha(n) +2n^2k+2nk^2+2nk)$ comprising several matrix-vector multiplications and an evaluation of the Jacobian which depends on the size of the original system.
 Approximating the Jacobian in (\ref{eq:MoOr:9}) is usually both problem and discretization dependent.
@@
 \begin{equation}
-	\mathbf J_{\mathbf y}(\mathbf g) = \underbrace{ (WV)^{-1} U(P^TU)^{-1} }_{k\times m} \underbrace{ \mathbf J_{\mathbf x}(\mathbf g(P^T V \mathbf y) ) }_{m\times m} \underbrace{P^T V}_{k\times k}.
+	\edit \mathbf J_{\mathbf x}(\mathbf g) = \underbrace{ (WV)^{-1} U(P^TU)^{-1} }_{k\times m} \underbrace{ \mathbf J_{\mathbf y}(\mathbf g(P^T V \mathbf y) ) }_{m\times m} \underbrace{P^T V}_{m\times k}.
 \end{equation}
@@
 Hence the overall complexity of computing the Jacobian is now independent of $n$.
+{\edit We refer the reader to \cite{Barrault:2004kz,Chaturantabut:2010cz} for more detail.}

### 2016-11-15 greedy_hamiltonian_revised/chapters/3.ham.tex
fist commit

\section{Hamiltonian Systems and Symplectic Geometry} \label{chap:Hasy:1}
Let $\edit \mathcal M$ be a manifold and $\edit \Omega:\mathcal M \times \mathcal M \to \mathbb R$ be a closed, nondegenerate 2-form on $\edit \mathcal M$. The pair $\edit (\mathcal M,\Omega)$ is called a \emph{symplectic manifold} {\edit \cite{Marsden:1999ck}}.

{\edit Let $(\mathcal M,\Omega)$ be a symplectic manifold and suppose that $H:\mathcal M \to \mathbb R$ is a smooth scalar function. The differential of $H$, denoted by $\mathbf dH$, defines a 1-form on $\mathcal M$. The nondegeneracy of $\Omega$ implies that there is a unique vector field $X_H$, \emph{Hamiltonian vector field }\cite{da2003introduction,Marsden:1999ck}, on $\mathcal M$ such that}
\begin{equation} \label{eq:Hasy:1}
	\edit i_{X_H} \Omega = \mathbf dH, 
\end{equation}
{\edit where $i_{X_H} \Omega$ is the interior product of $X_H$ with $\Omega$, i.e. requiring}
\begin{equation}
	\edit \Omega(X_H,Y) = \mathbf dH(Y),
\end{equation}
{\edit for any vector field $Y$ on $\mathcal M$.} Note that when $\edit \mathcal M$ belongs to a Euclidean space then $\mathbf d H = \nabla_z H$. The equations of evolution are then defined by
\begin{equation} \label{eq:Hasy:2}
	\dot z = X_H(z)
\end{equation}
and known as the \emph{Hamilton's equation} \cite{Marsden:1999ck}. A fundamental feature of Hamiltonian systems is the conservation of the Hamiltonian along integral curves on $P$. To emphasise the importance of this property we state

\begin{theorem} \label{theorem:Hasy:1}
Suppose that $X_H$ is a Hamiltonian vector field with the flow $\phi_t$ on a symplectic manifold $\mathcal M$. Then $H\circ \phi_t = H$ {\edit \cite{Marsden:1999ck}}.
\end{theorem}

\begin{proof}
We show that $H$ is constant along integral curves
\begin{equation} \label{eq:Hasy:3}
\begin{aligned}
	\frac{d}{dt}(H\circ \phi_t)(z) &= \mathbf d H(\phi_t(z)) \cdot( \frac{d}{dt} \phi_t(z) ) \\
	&= \mathbf d H (\phi_t(z))\cdot X_H(\phi_t(z)) \\
	&= \Omega_z( X_H(\phi_t(z)), X_H(\phi_t(z)) ) = 0,
\end{aligned}
\end{equation}
where we used the chain rule and bilinearity of $\Omega$ in the argument.
\end{proof}

For the case where the symplectic manifold is also a linear vector space, the pair $({\edit \mathcal M},\Omega)$ is also referred to as a \emph{symplectic vector space}. We state the following theorems regarding symplectic vector spaces without proof. We refer the reader to \cite{de2006symplectic,Marsden:1999ck,de2006symplectic,Silva01lectureson} for detailed proofs.

{\edit
\begin{theorem} \label{theorem:Hasy:1.1}
If $V$ is a vector space then $\Omega$ is a constant form, that is $\Omega_z$ is independent of $z\in V$ \cite{Marsden:1999ck}. 
\end{theorem}
\begin{theorem} \label{theorem:Hasy:1.2}
If $(V,\Omega)$ is a finite-dimensional symplectic manifold then $V$ is even dimensional \cite{Marsden:1999ck}.
\end{theorem}
\begin{theorem} \label{theorem:Hasy:1.3}
(The Symplectic Gram-Schmidt) If $(V,\Omega)$ is a $2n$-dimensional symplectic vector space, then there is a basis $e_1,\dots e_n,f_1, \dots , f_n$ of $V$ such that
\begin{equation} \label{eq:Hasy:4}
\begin{aligned}
	& \Omega(e_i,e_j) = 0 = \Omega(f_i,f_j), \quad & i\neq j,\\
	& \Omega(e_i,f_j) = \delta_{ij}, & i\leq i,j \leq n.
\end{aligned}
\end{equation}
where $\delta$ is the Kronecker's delta function. Moreover if $V = \mathbb{R}^{2n}$ we can choose the basis vectors $\{e_i,f_i\}_{i=1}^n$ such that
\begin{equation} \label{eq:Hasy:5}
	\Omega(v_1,v_2) = v_1^T \mathbb J_{2n} v_2, \qquad v_1,v_2\in \mathbb R^n,
\end{equation}
with $\mathbb J_{2n}$ being the symplectic matrix, defined as
\begin{equation} \label{eq:Hasy:6}
	\mathbb{J}_{2n} = 
	\begin{pmatrix}
		0_n & I_n \\
		-I_n & 0_n
	\end{pmatrix}.
\end{equation}
Here $I_n$ and $0_n$ is the identity matrix and the zero square matrix of size $n$, respectively \cite{de2006symplectic}.
\end{theorem}
\begin{theorem} \label{theorem:Hasy:1.4}
The classical inner product $\langle \cdot,\cdot \rangle:\mathbb R^{2n}\times \mathbb R^{2n}\to \mathbb R$ can be written in terms of the 2-form as \cite{Marsden:1999ck}
\begin{equation}
	\langle v,u \rangle = \Omega(\mathbb J_{2n}v,u),\quad \forall u,v \in \mathbb R^{2n}.
\end{equation}
\end{theorem}
}

{\edit 
\begin{definition}
Suppose $(V,\Omega)$ is a finite dimensional symplectic vector space and $E\subset V$ is a subspace. Then the symplectic complement of $E$ inside $P$ is defined as
\[
	E^{\perp} := \{ v\in V |\ \Omega(v,e) = 0,\ \forall e\in E \}
\]
\end{definition}
Note that $E \cap E^{\perp}$ is not empty in general \cite{de2006symplectic}. 
\begin{definition}
Suppose $(V,\Omega)$ is a finite dimensional symplectic vector space. A subspace $E\subset V$ is called a Lagrangian subspace inside $V$ if $E = E^\perp$.
\end{definition}
\begin{theorem} \label{theorem:Hasy:1.5}
Suppose $(V,\Omega)$ is a finite dimensional symplectic vector space. If $E\subset V$ is a Lagrangian subspace then $dim(E)=\frac 1 2dim(V)$, where $dim$ denotes the dimension of the subspace \cite{abraham1978foundations}.
\end{theorem}
\begin{definition}
A basis of $(V,\Omega)$ is called orthosymplectic if it is both a symplectic basis and an orthogonal basis with respect to the classical scalar product.
\end{definition}
\begin{theorem} \label{theorem:Hasy:1.6}
Let $E\subset V$ be a Lagrangian vector space. Then there is an orthosymplectic basis for $\mathbb R^{2n}$ \cite{da2003introduction}.
\end{theorem}
\begin{proof}
Starting from a Lagrangain subspace in $E \subset V$ an orthosymplectic basis can be easily constructed.  Suppose that $\{ e'_1,\dots, e'_n \}$ is a basis for $E$, using the classical Gram-Schmidt orthogonalization process we can construct an orthonormal basis $\{ e_1,\dots,e_n \}$. Define the new set of vectors $f_1 = -\mathbb Je_1$, $f_2 = -\mathbb Je_2$, $\dots$, $f_n= -\mathbb Je_n$. We have
\begin{equation}
	\langle f_i, f_j \rangle = e_i^T \mathbb J_{2n}^T \mathbb J_{2n} e_i = \delta_{ij}, \quad \langle f_i, e_j \rangle = -e_i \mathbb J^T_{2n} e_j = 0, \quad i,j=1,\dots,n,
\end{equation}
where the second identity is due to the fact that the basis $\{ e_1,\dots,e_n \}$ forms a Lagrangian subspace. This shows that the set $\{ e_1,\dots,e_n \}\cup \{ f_1,\dots,f_n \}$ forms an orthonormal basis. Also, it can be easily verified that this is a symplectic basis. Thus $\{ e_1,\dots,e_n \}\cup \{ f_1,\dots,f_n \}$ is an orthosymplectic basis.
\end{proof}
\begin{theorem} \label{theorem:Hasy:1.7}
On a finite-dimensional symplectic vector space the relationship (\ref{eq:Hasy:1}) becomes \cite{Marsden:1999ck}
\begin{equation} \label{eq:Hasy:7}
\left\{
\begin{aligned}
	&\dot {\mathbf z} = \mathbb{J}_{2n} \nabla_{\mathbf z} H(\mathbf z), \\
	& \mathbf z(0) = \mathbf z_0.
\end{aligned}
\right.
\end{equation}
or, by introducing the canonical coordinates $\mathbf z = (\mathbf q^T, \mathbf p^T)^T$,
\begin{equation} \label{eq:Hasy:8}
\left\{
\begin{aligned}
	&\dot {\mathbf q} = \nabla_{\mathbf p} H(\mathbf q,\mathbf p),\\
	&\dot {\mathbf p} = - \nabla_{\mathbf q} H(\mathbf q,\mathbf p).
\end{aligned}
\right.
\end{equation}
\end{theorem}
}

We now introduce \emph{symplectic transformations}, i.e. mappings between symplectic manifolds which preserve the two-form $\Omega$. Accurate numerical treatment of Hamiltonian systems often requires preservation of the symmetry expressed in Theorem \ref{theorem:Hasy:1}. Symplectic transformations can be used to construct such symmetry preserving numerical methods.

{\edit
\begin{definition}
Let $(V,\Omega)$ and $(W,\Pi)$ be two linear symplectic vector spaces of dimensions $2n$ and $2k$ respectively. A linear mapping $\phi:V \to W$ is called \emph{symplectic} or \emph{canonical} if
\begin{equation} \label{eq:Hasy:9}
	\Omega = \phi^* \Pi
\end{equation}
where $\phi^* \Pi$ is the pullback of $\Pi$ by $\phi$, i.e. for all $\mathbf{z}_1, \mathbf{z}_2\in V$
\begin{equation}
	\Omega(\mathbf{z}_1,\mathbf{z}_2) = \Pi(\phi(\mathbf{z}_1),\phi(\mathbf{z}_2)).
\end{equation}
\end{definition}

Note if we represent the transformation $\phi$ as a matrix $A\in \mathbb R^{2n\times 2k}$ the condition (\ref{eq:Hasy:9}) is equivalent to \cite{Marsden:1999ck}}

\begin{equation} \label{eq:Hasy:11}
	A^T \mathbb{J}_{2n}A = \mathbb{J}_{2k}.
\end{equation}
A matrix of size $2n\times 2k$ satisfying (\ref{eq:Hasy:11}) is called a \emph{symplectic matrix}.

\begin{definition}
	The \emph{symplectic inverse} of a matrix $A\in \mathbb{R}^{2n\times 2k}$ is denoted by $A^+$ and defined by {\edit \cite{Peng:2014di}}
\begin{equation}\label{eq:Hasy:12}
	A^+ := \mathbb{J}_{2k}^T A^T \mathbb{J}_{2n}.
\end{equation}
\end{definition}
We point out the properties of the symplectic inverse and refer the reader to \cite{Peng:2014di} for detailed proof.
\begin{lemma} \label{lemma:Hasy:1}
Let $A\in \mathbb{R}^{2n\times 2k}$ be a symplectic matrix and $A^+$ its symplectic inverse as defined in (\ref{eq:Hasy:12}). Then ${(A^+)}^T$ is a symplectic matrix and $A^+A = I_{2k}$.
\end{lemma}

{\edit Straight forward calculation verifies} that $AA^+$ is idempotent, i.e. a symplectic projection onto the column span of $A$.

{\edit It natural to expect a numerical integrator that solves (\ref{eq:Hasy:7}) to also satisfy the conservation law in Theorem \ref{theorem:Hasy:1}. Common numerical integrators e.g., Runge-Kutta methods, do not generally preserve the Hamiltonian which results in a wrong qualitative behavior in the solution \cite{Hairer:1250576}. Symplectic integrators are a class of numerical integrators for Hamiltonian systems that preserve the symplectic structure yielding stability in long-time integration.} The Str\"omer-Verlet time stepping scheme is an example of symplectic integrators and is given by
\begin{equation} \label{eq:Hasy:13}
\begin{aligned}
	q_{n+1/2} &= q_n + \frac{\Delta t}{2} \nabla_pH(q_{n+1/2},p_n), \\
	p_{n+1} &= p_n - \frac{\Delta t}{2} \left( \nabla_qH(q_{n+1/2},p_n) + \nabla_qH(q_{n+1/2},p_{n+1}) \right),\\
	q_{n+1} &= q_{n+1/2} + \frac{\Delta t}{2} \nabla_pH(q_{n+1/2},p_{n+1}),
\end{aligned}
\end{equation}
and
\begin{equation} \label{eq:Hasy:14}
\begin{aligned}
	p_{n+1/2} &= p_n - \frac{\Delta t}{2} \nabla_qH(q_{n},p_{n+1/2}), \\
	q_{n+1} &= q_n + \frac{\Delta t}{2} \left( \nabla_pH(q_{n},p_{n+1/2}) + \nabla_pH(q_{n+1},p_{n+1/2}) \right),\\
	p_{n+1} &= p_{n+1/2} - \frac{\Delta t}{2} \nabla_qH(q_{n+1},p_{n+1/2}).
\end{aligned}
\end{equation}
For a general Hamiltonian system, Str\"omer-Verlet scheme is implicit. However, for separable Hamiltonians, i.e. $H(q,p) = K(p) + U(q)$, this {\edit scheme becomes} explicit. We refer the reader to \cite{Hairer:1250576} for more information about the construction and applications of symplectic and geometric numerical integrators.

### 2016-11-15 greedy_hamiltonian_revised/chapters/4.sympmor.tex
fist commit

--- 2016-11-15 greedy_hamiltonian_new_template/chapters/4.sympmor.tex
@@
 By definition $\mathbf z(t) = \phi_t(\mathbf z_0)$ and $\mathbf y(t) = \psi_t(\mathbf y_0)$.
-Using Theorem \ref{theorem:Hasy:1} we have
+Using the definition of the reduced Hamiltonian and Theorem \ref{theorem:Hasy:1} we have
 \begin{equation} \label{eq:SyMo:6}
@@
 \end{equation}
-Also $H(\phi_t(\mathbf z_0)) = H(\mathbf z_0)$. The error in the Hamiltonian can be written in terms of $\mathbf z_0$ and the symplectic basis $A$ as
+The error in the Hamiltonian can then be written in terms of $\mathbf z_0$ and the symplectic basis $A$ as
 \begin{equation} \label{eq:SyMo:7}
@@
 
-The following theorems guarantees stability of the original and the reduced Hamiltonian systems. We refer the reader to \cite{Peng:2014di} for the proofs.
+{\edit 
+The following theorems provide a strong indication of the stability of the reduced system.
 
+\begin{definition} \label{definition:SyMo:1}
+Consider a dynamical system of the form $\dot{\mathbf z} = \mathbf f(\mathbf z)$ and suppose that $\mathbf z_e$ is an equilibrium point for the system so that $\mathbf f(\mathbf z_e) = 0$.
+$\mathbf z_e$ is called nonlinear stable or Lyapunov stable if for any $\epsilon > 0$, we can find $\delta > 0$ such that for any trajectory $\phi_t$, if $\| \phi_0 - \mathbf z_e \|_2 \leq \delta$, then for all $t\geq0$, we have $\| \phi_t - \mathbf z_e \|_2 < \epsilon$, where $\| \cdot \|_2$ is the Euclidean norm.
+\end{definition}	
+The following proposition, also known as Dirichlet's theorem \cite{bhatia2002stability}, states the sufficient condition for an equilibrium point to be Lyapunov stable.
+We refer the reader to \cite{bhatia2002stability} for the proof.
+\begin{proposition} \label{proposition:SyMo:1}
+An equilibrium point $\mathbf z_e$ is Lyapunov stable if there exist a scalar function $W : \mathbb R^{n} \to  \mathbb R$ such that $\nabla W(\mathbf z_e) = 0$, and $\nabla^2 W(\mathbf z_e)$ is positive definite and that for any trajectory $\phi_t$ defined in the neighborhood of $\mathbf z_e$, we have $\frac{d}{dt} W(\phi_t) \leq 0$.
+Here $\nabla^2W$ is the Hessian matrix of $W$.
+\end{proposition}
+The scalar function $W$ is referred to as the \emph{Lyapunov function}.
+In the context of the Hamiltonian systems, a suitable candidate for the Lyapunov function is the Hamiltonian function $H$.
+The following theorem shows that when $H$ (or $-H$) is a Lyapunov function, then the equilibrium points of the original and the reduced system are Liyapunov stable.
 \begin{theorem} \label{theorem:SyMo:1}
-Consider the initial value problem (\ref{eq:Hasy:7}) with its symplectic reduced system (\ref{eq:SyMo:4}).
-If we can find a bounded neighborhood $N$ of $z_0$ in $P$ such that $H(z) < H(z_0)$ or $H(z) > H(z_0)$, for all $z\in \partial N$, then the original system and the reduced system are bounded for all $t\in \mathbb R$.
+Consider a Hamiltonian system of the form (\ref{eq:Hasy:7}) together with the reduced system (\ref{eq:SyMo:4}).
+Suppose $\mathbf z_e$ is an equilibrium point for (\ref{eq:Hasy:7}) and that $\mathbf y_e = A^+\mathbf z_e$.
+If $H$ (or $-H$) is a Lyapunov function satisfying proposition \ref{proposition:SyMo:1}, then $\mathbf z_e$ and $\mathbf y_e$ are Lyapunov stable points for (\ref{eq:Hasy:7}) and (\ref{eq:SyMo:4}) respectively.
 \end{theorem}
-
-\begin{theorem} \label{theorem:SyMo:2}
-If the original system (\ref{eq:Hasy:7}) is a linear Hamiltonian system, i.e.
-$H(x) = \frac 1 2 x^T L x$ with $L$ a $2n\times 2n$ real symmetric matrix, then both the original system and the reduced system (\ref{eq:SyMo:4}) are stable.
-\end{theorem}
+\begin{proof}
+	It is the direct consequence of proposition \ref{proposition:SyMo:1} that $\mathbf z_e$ is a local minimum or maximum of (\ref{eq:Hasy:7}) and also a Lyapunov stable point.
+It can be easily checked that if $\mathbf z_e$ is a local minimum of $H$ then $\mathbf y_e$ is a local minimum for $\tilde H$ and an equilibrium point for (\ref{eq:SyMo:4}).
+Also from the chain rule we have
+\[
+	\nabla^2_{\mathbf y} \tilde H = A^T \nabla^2_{\mathbf z} H A.
+\]
+So for any $\xi\in \mathbb R^{2k}$
+\[
+	\xi^T \nabla^2_{\mathbf y} \tilde H \xi = (A\xi)^T \nabla^2_{\mathbf z} H (A\xi) \geq 0,
+\]
+Where the last inequality is due to the positive definiteness of $H$.
+Therefore $\tilde H$ is also positive definite.
+By proposition \ref{proposition:SyMo:1} we conclude that $\mathbf y_e$ is a Lyapunov stable point.
+\end{proof}
+}
 
@@
 Suppose that a symplectic basis $A$ of size $2n\times2k$ and its symplectic inverse $A^+$ is provided.
-The proper symplectic decomposition requires that the error of the symplectic projection onto the symplectic subspace to be minimum.
+{\edit The Proper Symplectic Decomposition} requires that the error of the symplectic projection onto the symplectic subspace to be minimum.
 Analogous to the minimization (\ref{eq:MoOr:6}), the PSD symplectic basis of size $2k$ is the solution of the optimization problem
@@
 \end{equation}
-Note $A$ is not necessarily an orthogonal matrix.
 Compared to POD, in (\ref{eq:SyMo:8}) the orthogonal projection is replaced with a symplectic projection $AA^+$.
 At first, the minimization looks similar to the one obtained by POD.
-However, the lack of an SVD-like decomposition for symplectic subspaces, makes the minimization (\ref{eq:SyMo:8}) a harder problem than (\ref{eq:MoOr:6}).
+{\edit However, it is well known that symplectic bases are not generally orthogonal, and therefore not norm bounded, this means numerical errors can become dominant in symplectic projection \cite{Karow:2006cf}.
+This makes the minimization (\ref{eq:SyMo:8}) a harder problem than (\ref{eq:MoOr:6}).}
 
@@
 A simplified version of the optimization (\ref{eq:SyMo:8}) can be found in \cite{Peng:2014di}, but there is no guarantee that the method provides a near optimal basis.
-In the following we outline non-direct methods for finding solutions to (\ref{eq:SyMo:8}), proposed by \cite{Peng:2014di}, while assuming a specific structure for $A$.
-Finally, we introduce a greedy approach for symplectic basis generation.
+
+{\edit Finding eigen-spaces of Hamiltonian and symplectic matrices is studied in the context of optimal control problems \cite{Benner:2000ww,Benner:1997ef,Watkins:2004kz,BunseGerstner:1986dg} and model reduction of Riccati equations \cite{Benner:1997ef}, where also an SVD-like decomposition for Hamiltonian and symplectic matrices is proposed by \cite{Xu:2003kx}.
+However, computation of a possibly large snapshot matrix and then using the mentioned methods to compute its eigen-spaces, is usually computationally demanding.
+Also, these methods generally do not guarantee the construction of a well-conditioned symplectic basis.
+
+The greedy approach presented in section \ref{Chap:Symo.PrSy:3}, is an iterative method for construction of a symplectic basis.
+It avoids evaluation of the full snapshot matrix so it saves substantial computational cost in the offline stage of the symplectic model reduction.
+Also, by construction, it yields a orthosymplectic basis and therefore a well-conditioned basis.
+
+In section \ref{chap:SyMo.PrSy:2} we outline non-direct methods for finding solutions to (\ref{eq:SyMo:8}), proposed by \cite{Peng:2014di}, while assuming a specific structure for $A$.
+Finally, in section \ref{Chap:Symo.PrSy:3} we introduce a greedy approach for symplectic basis generation.}
 
@@
 
-%Algorithm \ref{alg:SyMo:1} summarizes the cotengent lift method for finding a symplectic basis.
-%
-%\begin{algorithm} 
-%\caption{Real SVD method for symplectic basis generation} \label{alg:SyMo:1}
-%{\bf Input:} Snapshots $\{ q_i , p_i \}_{i=1}^N$ of the solution manifold and $2k$ the size of the symplectic basis.
-%\begin{enumerate}
-%\item Construct the combined snapshot matrix $S_{\text{combined}}$ in (\ref{eq:SyMo:10}).
-%\item Compute the SVD decomposition of $S_{\text{combined}}=U\Sigma V^T$.
-%\item Define $\Phi = [u_1,\dots,u_k]$ with $u_i$ the $i$-th left singular vector.
-%\item Construct $A$ according to (\ref{eq:SyMo:9}).
-%\end{enumerate}
-%\vspace{0.5cm}
-%{\bf Output:} Symplectic matrix $A$ of the size $2n\times 2k$.
-%\end{algorithm}
-%
-
-\paragraph{\bf Complex SVD} We now consider another form of symplectic basis $A$. Suppose that $A$ takes the form
+\paragraph{\bf Complex SVD} We now consider another form of symplectic basis $A$. Suppose that $A$ takes the form \cite{Peng:2014di}
 \begin{equation} \label{eq:SyMo:11}
@@
 
-%Algorithm \ref{alg:SyMo:2} summarizes the complex SVD method for finding a symplectic basis.
-%
-%\begin{algorithm} 
-%\caption{Complex SVD method for symplectic basis generation} \label{alg:SyMo:2}
-%{\bf Input:} Snapshots $\{ q_i , p_i \}_{i=1}^N$ of the solution manifold and $2k$ the size of the symplectic basis.
-%\begin{enumerate}
-%\item Construct the complex snapshot matrix $S_{\text{complex}}$ in (\ref{eq:SyMo:13}).
-%\item Compute the complex SVD decomposition $S_{\text{complex}} = U \Sigma V^H$.
-%\item Define $\Phi$ and $\Psi$ as in (\ref{eq:SyMo:14}).
-%\item Construct $A$ according to (\ref{eq:SyMo:11}).
-%\end{enumerate}
-%\vspace{0.5cm}
-%{\bf Output:} Symplectic matrix $A$ of the size $2n\times 2k$.
-%\end{algorithm}
-
-\subsubsection{The Greedy Approach to Symplectic Basis Generation} The greedy generation of the reduced basis is an iterative procedure which, in each iteration, adds the two best possible basis vectors to the symplectic basis to enhance overall accuracy.
+\subsubsection{The Greedy Approach to Symplectic Basis Generation} \label{Chap:Symo.PrSy:3} greedy generation of the reduced basis is an iterative procedure which, in each iteration, adds the two best possible basis vectors to the symplectic basis to enhance overall accuracy.
 In contrast to the cotangent lift and the complex SVD methods, the greedy approach does not require the symplectic basis to have a specific structure.
@@
 
-{\bf Symplectic Gram-Schmidt:} As discussed in Section \ref{chap:Hasy:1}, any finite dimensional symplectic linear vector space has a symplectic basis that satisfies conditions (\ref{eq:Hasy:4}).
-Suppose that $S=\{ e_1 , \dots , e_k \} \bigcup \{ f_1 , \dots , f_k \}$ is a symplectic basis and $a$ and $b$ are two vectors such that $a,b \not \in \text{span}(S)$ and $a \not \not \parallel b$.
-We seek to grow the size of $S$ by the vectors $a$ and $b$.
-First we modify $a$ and $b$ to ensure they are symplectic (symplectic orthogonal) to $S$, i.e., we add a linear combination of elements of $S$ to $a$ and $b$, so the resulting vectors are symplectic to $S$.
-This can be achieved since we assumed $a,b\not \in \text{span}(S)$.
-So for $a$ we seek $\alpha_1,\dots,\alpha_k,\beta_1,\dots,\beta_k \in \mathbb R$ such that
-\begin{equation}
-	\Omega\left(a + \sum_{i=1}^k \alpha_i e_i + \sum_{i=1}^k \beta_i f_i , \sum_{i=1}^k \bar{\alpha}_i e_i + \sum_{i=1}^k \bar{\beta}_i f_i \right) = 0,
+{\edit The orthonormalization step is an essential step in most greedy approaches for basis generation in the context of model reduction \cite{Anonymous:2016wl,Quarteroni:2016wi}.
+However common orthonormalization processes, e.g.
+the QR method based on the classical Gram-Schmidt method, destroy the symplectic structure of the original system \cite{BunseGerstner:1986dg}.
+Here we use a variation of the QR method known as the the SR \cite{Salam2014} method which is based on the symplectic Gram-Schmidt method and yields a symplectic basis.
+}
+
+{\edit
+As discussed in Section \ref{chap:Hasy:1}, any finite dimensional symplectic linear vector space has a symplectic basis that satisfies conditions (\ref{eq:Hasy:4}).
+Further theorem \ref{theorem:Hasy:1.6} provides an iterative process for constructing an orthosymplectic basis \cite{Matsuo:2014wl,Salam2014}.
+To briefly describe the SR method, suppose that an orthosymplectic basis
+\begin{equation} \label{eq:SyMo:14.1}
+	A_{2k}=\{ e_1 , \dots , e_k \} \cup \{ -\mathbb J_{2n} e_1 , \dots , -\mathbb J_{2n} e_k \},
+\end{equation}
+and a vector $z\not \in \text{span}(A_{2k})$ is provided.
+We would like to symplectically orthogonalize ($\mathbb J_{2n}$-orthogonalize) $z$ with respect to $A_{2k}$.
+We seek $\alpha_1,\dots,\alpha_k,\beta_1,\dots,\beta_k \in \mathbb R$ such that
+\begin{equation} \label{eq:SyMo:14.2}
+	\Omega\left(z + \sum_{i=1}^k \alpha_i e_i - \sum_{i=1}^k \beta_i \mathbb J_{2n}e_i , \sum_{i=1}^k \bar{\alpha}_i e_i - \sum_{i=1}^k \bar{\beta}_i \mathbb J_{2n}e_i \right) = 0,
 \end{equation}
 for all possible $\bar{\alpha}_1,\dots,\bar{\alpha}_k,\bar{\beta}_1,\dots,\bar{\beta}_k \in \mathbb R$. It is easily seen that the unique solution is as 
-\begin{equation}
-	\alpha_i = - \Omega(a,f_i), \quad \beta_i = \Omega(a,e_i),
+\begin{equation} \label{eq:SyMo:14.3}
+	\alpha_i = - \Omega(z,-\mathbb J_{2n}e_i), \quad \beta_i = \Omega(z,e_i),
 \end{equation}
 for $i=1,\dots,k$. Now we define the modified vectors as
-\begin{equation}
-\begin{aligned}
-	& \tilde a = a - \sum_{i=1}^k \Omega(a,f_i) e_i + \sum_{i=1}^k \Omega(a,e_i) f_i, \\
-	& \tilde b = b - \sum_{i=1}^k \Omega(b,f_i) e_i + \sum_{i=1}^k \Omega(b,e_i) f_i.
-\end{aligned}
+\begin{equation} \label{eq:SyMo:14.4}
+	\tilde z = z - \sum_{i=1}^k \Omega(z,-\mathbb J_{2n}e_i) e_i - \sum_{i=1}^k \Omega(z,e_i) \mathbb J_{2n}e_i.
 \end{equation}
-Finally we construct the new basis vectors as
-\begin{equation}
-\begin{aligned}
-	& e_{k+1} = \text{sign}(\Omega(\tilde a , \tilde b)) \cdot \frac{\tilde a}{ \sqrt{ |\Omega(\tilde a , \tilde b)| } }, \quad f_{k+1} = \frac{\tilde b}{ \sqrt{ |\Omega(\tilde a , \tilde b)| } }.
-\end{aligned}
-\end{equation}
-One easily checks that the new basis vectors $e_1,\dots,e_{k+1},f_1,\dots,f_{k+1}$ satisfy all the conditions (\ref{eq:Hasy:4}).
-Similar to the QR decomposition, the symplectic Gram-Schmidt process leads to a matrix decomposition for any full ranked matrix of size $2n\times 2k$.
-More details about this matrix decomposition are given in Appendix \ref{chap:Append:1}.
+If we introduce $e_{k+1} = \tilde z / \| \tilde z \|_2$, it is easily checked that $e_{k+1}$ is also orthogonal to $A_{2k}$ with respect to the classical inner product.
+Therefore the set span$(\{e_1,\dots,e_{k+1}\})$ forms a Lagrangian subspace and according to theorem \ref{theorem:Hasy:1.6} the basis $A_{2k+2}= A_{2k}\cup \{ e_{k+1} , -\mathbb J_{2n} e_{k+1} \}$ forms an orthosymplectic basis.
 
-In the case of finite symplectic vector fields, there is an alternative way to update the symplectic basis $S$.
-To illustrate, we first note that if $e_i$ or $f_i$ is a basis vector in $S$, we have $\mathbb J_{2n}e_i = -f_i$ and $\mathbb J_{2n}f_i = e_i$.
-Since
-\begin{equation}
-	e_i^T\mathbb J_{2n} f_i = \Omega(e_i,f_i) = 1 \Longleftrightarrow \mathbb J_{2n} e_i = - f_i \text{ and } \mathbb J_{2n} f_i = e_i.
-\end{equation}
-Thus if we consider a vector $v\not \in \text{span}(S)$ then $\mathbb J_{2n}^T v $ also does not belong to span$(S)$.
-Since, if it did, we could write the vector as a linear combination of basis vectors of $S$ as
-\begin{equation}
-	\mathbb J_{2n}^T v = \sum_{i=1}^k \alpha_i e_i + \sum_{i=1}^k \beta_i f_i.
-\end{equation}
-Multiplying both sides by $\mathbb J_{2n}$ yields
-\begin{equation}
-	v = \sum_{i=1}^m \alpha_i \mathbb J_{2n} e_i + \sum_{i=1}^k \beta_i \mathbb J_{2n} f_i,
-\end{equation}
-But all vectors on the right hand side belong to span$(S)$, hence reaching a contradiction.
-For updating $S$ using $v$, we first symplectify (symplectic orthogonalize) $v$ with respect to $S$ to recover
-\begin{equation}
-	\tilde v = v - \sum_{i=1}^k \Omega(v,f_i)e_i + \sum_{i=1}^k \Omega (v,e_i)f_i.
-\end{equation}
-We can write the update for the symplectic basis as
-\begin{equation}
-	e_{k+1} = \frac{\tilde v}{\|\tilde v\|},\quad f_{k+1} = \mathbb J_{2n}^T \tilde v.
-\end{equation}
-It is easily checked that the new set of basis vectors $\{e_1,\dots,e_{k+1},f_1,\dots,f_{k+1} \}$ forms a symplectic basis.
-This update, not only provides a symplectic basis but also an orthonormal basis.
+Note that the $SR$ method can be replaced with a backward stable routines such as the isotropic Arnoldi or the isotropic Lanczos methods \cite{Mehrmann:2000dv}.
+}
 
@@
 
-Suppose that a $2k$ dimensional symplectic basis $A_{2k} = \{e_1,\dots ,e_k,f_1,\dots ,f_k\}$ is generated at the $k$-th step of the greedy method and we seek to enrich it by two additional vectors.
+Suppose that a $2k$ dimensional {\edit orthosymplectic basis (\ref{eq:SyMo:14.1})} is generated at the $k$-th step of the greedy method and we seek to enrich it by two additional vectors.
 Using the error in the Hamiltonian (\ref{eq:SyMo:7}) we search the parameter space to find the value that maximises the error in the Hamiltonian
-\begin{equation}
+\begin{equation} \label{eq:SyMo:14.5}
 	\omega_{k+1} := \underset{\omega\in \Gamma}{\text{argmax }}\Delta H(\omega).
@@
 
-We then propagate (\ref{eq:Hasy:7}) in time to produce trajectory snapshots $S=\{ \mathbf z(t_i,\omega_{k+1}) | i = 1,\dots,M \}$.
+We then propagate (\ref{eq:Hasy:7}) in time to produce trajectory snapshots 
+\begin{equation}
+	S=\{ \mathbf z(t_i,\omega_{k+1}) | i = 1,\dots,M \}.
+\end{equation} 
 The next basis vector is the snapshot that maximises the projection error (\ref{eq:SyMo:8})
-\begin{equation}
-	z := \underset{s\in S}{\text{argmax }} \| s - AA^+s \|.
+{\edit 
+\begin{equation} \label{eq:SyMo:14.6}
+	z := \underset{s\in S}{\text{argmax }} \| s - A_{2k}{A_{2k}}^+s \|.
 \end{equation}
+}
 Finally, we update the basis as
-\begin{equation}
-	e_{k+1} = \tilde z, \quad f_{k+1} = \mathbb J_{2n}^T \tilde z,
+{\edit
+\begin{equation} \label{eq:SyMo:14.7}
+	e_{k+1} = \tilde z, \quad A_{2k+1} = A_{2k}\cup \{ e_{k+1} , -\mathbb J_{2n}e_{k+1} \},
 \end{equation}
+}
 where $\tilde z$ is the vector obtained from applying the symplectic Gram-Schmidt process to $z$.
@@
 
+{\edit
 \begin{algorithm} 
@@
 \item $e_1 \leftarrow \mathbf z_0(\omega^*)$
-\item $f_1 \leftarrow \mathbb J^T_{2n} \mathbf z_0(\omega^*)$
-\item $A \leftarrow [e_1,f_1]$
+\item $A \leftarrow [e_1,-\mathbb J^T_{2n}e_1]$
 \item $k \leftarrow 1$
@@
 \item \hspace{0.5cm} $e_{k+1} \leftarrow \mathbf z^*/ \| z^*\|$
-\item \hspace{0.5cm}$f_{k+1} \leftarrow \mathbb J^T_{2n} \mathbf z^*$
-\item \hspace{0.5cm} $A \leftarrow [e_1,\dots ,e_{k+1} , f_1,\dots,f_{k+1}]$
+\item \hspace{0.5cm} $A \leftarrow [e_1,\dots ,e_{k+1} , -\mathbb J^T_{2n}e_1,\dots,-\mathbb J^T_{2n}e_{k+1}]$
 \item \hspace{0.5cm} $k \leftarrow k+1$
@@
 \end{algorithm}
+}
 
@@
 \begin{itemize}
-\item Lack of a symplectic linear algebra toolbox for solving the minimization in (\ref{eq:SyMo:8}) requires a possibly very expensive nonlinear optimizations to recover near optimal symplectic bases.
+%\item Lack of a symplectic linear algebra toolbox for solving the minimization in (\ref{eq:SyMo:8}) requires a possibly very expensive nonlinear optimizations to recover near optimal symplectic bases.
 Alternative approaches for finding solutions to (\ref{eq:SyMo:8}) requires assumptions on the structure of the symplectic basis.
@@
 \begin{theorem}
-	Let $S$ be a compact subset of $\mathbb{R}^{2n}$ with exponentially small Kolmogorov $n$-width $d_{n}\leq c\exp(-\alpha k)$ with $\alpha > \log3$.
+	Let $S$ be a compact subset of $\mathbb{R}^{2n}$ with exponentially small Kolmogorov $n$-width $\edit d_{k}\leq c\exp(-\alpha k)$ with $\alpha > \log3$.
 Then there exists $\beta>0$ such that the symplectic subspaces $A_{2k}$ generated by the greedy algorithm provide exponential approximation properties such that
@@
 \end{theorem}
+
+{\edit
+\begin{theorem}
+	The basis $A_{2k}$ generated by the greedy algorithm is orthosymplectic.
+\end{theorem}
+\begin{proof}
+	Direct result of theorem \ref{theorem:Hasy:1.6}.
+\end{proof}
+}
 
@@
 Note that the above update only enriches the span of $A$, containing the basis we started with.
-We summarized the symplectic DEIM algorithm in Algorithm \ref{alg:SyMo:4}.
+We summarised the symplectic DEIM algorithm in Algorithm \ref{alg:SyMo:4}.
 
@@
 To overcome this, when index $\mathfrak p_i$ with $\mathfrak p_i\leq n$ or $\mathfrak p_i>n$ is chosen in Algorithm \ref{alg:MoOr:1} we also include $\mathfrak p_i + n$ or $\mathfrak p_i-n$, respectively.
-This guarantees that $\mathbf g$ and $P$ commute.
+{\edit Simple calculations verifies that $\mathbf g$ and $P$ commute.}

### 2016-11-15 greedy_hamiltonian_revised/chapters/5.numres.tex
fist commit

--- 2016-11-15 greedy_hamiltonian_new_template/chapters/5.numres.tex
@@
 
-\subsection{Parametric Linear Wave equation} Consider the stochastic linear wave equation
+\subsection{Parametric Linear Wave equation} Consider the {\edit parametric} linear wave equation
 \begin{equation} \label{eq:NuRe:1}
@@
 
-For numerical time integration we used the Str\"omer-Verlet (\ref{eq:Hasy:13}) scheme, which becomes an explicit scheme since the Hamiltonian is separable for the linear wave-equation.
+For numerical time integration we {\edit use} the Str\"omer-Verlet (\ref{eq:Hasy:13}) scheme, which becomes an explicit scheme since the Hamiltonian is separable for the linear wave-equation.
 The full model uses the following parameter set \vspace{0.5cm}
@@
 We use the Verlet-Str\"omer scheme for symplectic methods and a second order Runge-Kutta method for the POD.
-Finally we use transformation (\ref{eq:SyMo:1}) to transfer the solution of the reduced systems into the high-dimensional space for illustration purposes.
+{\edit The choice of different time integration routines is due to the fact that POD destroys the canonical form of the original equations and a symplectic integrator cannot be applied.
+One can alternatively use separate reduced subspaces for the potential and the momentum spaces, which however is not a standard model approach and needs further analysis.} Finally we use transformation (\ref{eq:SyMo:1}) to transfer the solution of the reduced systems into the high-dimensional space for illustration purposes.
 
@@
 Figure \ref{fig:NuRe:2c} shows the $L^2$-error between the solution of the full model and the reduced systems constructed by different methods.
-We note that the error for the POD reduced system rapidly increases, confirming that projection based reduced system does not yield a stable solution.
+We note that the error for the POD reduced system rapidly increases, confirming that {\edit the} projection based reduced system does not yield a stable solution.
 Furthermore, the symplectic methods provide a better approximation since the geometric structure of the original system is preserved.
-Although the greedy method is almost twice faster than SVD-based method in the offline stage, its accuracy is comparable.
-For complex systems were the solution of the full system is expensive and for parameter domains with higher dimensions, we expect the greedy method to be superior in terms of computational cost.
+Although the greedy method is almost {\edit two times faster than the SVD-based methods} in the offline stage, its accuracy is comparable.
+{\edit For complex systems were the solution of the full system is expensive and for high dimensional parameter domains, POD-based methods become exceedingly expensive \cite{Anonymous:2016wl,Quarteroni:2016wi}.
+However the greedy method requires substantially fewer (proportional to the size of the reduced basis) evaluation of the time integration of the original system.
+Therefore we expect the greedy method to be superior in terms of computational cost.}
 
@@
 \centering
-\subfloat[]{\label{fig:NuRe:2a}\includegraphics[width=1\textwidth]{./figs/wave/singular}}
+\subfloat[]{\label{fig:NuRe:2c}\includegraphics[width=\textwidth]{./figs/wave/error}}
 \end{minipage}%
@@
 \centering
-\subfloat[]{\label{fig:NuRe:2c}\includegraphics[width=0.5\textwidth]{./figs/wave/error}}
+%\subfloat[]{\label{fig:NuRe:2c}\includegraphics[width=0.5\textwidth]{./figs/wave/error}}
 
@@
 
-Regarding computation of the nonlinear terms of reduced systems, We compare the DEIM with the symplectic DEIM.
+Regarding computation of the nonlinear terms of reduced systems, {\edit we} compare the DEIM with the symplectic DEIM.
 For generation of the DEIM reduced basis we apply Algorithm \ref{alg:MoOr:1} to the set of nonlinear snapshots.
@@
 \caption{The solution $|u(t,x)| = \sqrt{q^2 + p^2}$ at $t=0$, $t=10$ and $t=20$ of the Nonlinear Schr\"odinger equation for parameter value $\epsilon = 1.0932$.
-Here the solution of full system together with the solution of the POD, cotangent lift, complex SVD and the greedy reduced system is presented.}
+Here the solution of {\edit the} full system together with the solution of the POD, cotangent lift, complex SVD and the greedy reduced system is presented.}
 \label{fig:NuRe:3}
@@
 \centering
-\subfloat[]{\label{fig:NuRe:4a}\includegraphics[width=1\textwidth]{./figs/schrodinger/singular}}
+\subfloat[]{\label{fig:NuRe:4c}\includegraphics[width=\textwidth]{./figs/schrodinger/error}}
 \end{minipage}%
@@
 \centering
-\subfloat[]{\label{fig:NuRe:4c}\includegraphics[width=0.5\textwidth]{./figs/schrodinger/error}}
+%\subfloat[]{\label{fig:NuRe:4c}\includegraphics[width=0.5\textwidth]{./figs/schrodinger/error}}
 
@@
 
-The value of the Hamiltonian function for the reduced systems discussed above as a function of time is illustrated in Figure \ref{fig:NuRe:4b}.
-We see that the symplectic methods preserve the Hamiltonian function.
-We like to emphasise that the combination of the reduced basis by the greedy and the DEIM does not preserve the symplectic structure as suggested in this figure.
+{\edit In Figure \ref{fig:NuRe:4b} we observe that the symplectic methods preserve the Hamiltonian function, unlike the POD and the DIEM methods.
+We would like to emphasise that using the reduced basis by the greedy and together with the DEIM (purple line) does not preserve the symplectic structure as suggested in this figure.}
 
@@
 Secondly, we observe that although the reduced systems from the cotangent lift and the complex SVD are of the same size, their accuracy is different by an order of magnitude.
-We notice that the greedy algorithm is slightly less accurate than the cotangent lift method while its offline computational cost is less than 20\%.
+We notice that the greedy algorithm is slightly less accurate than the cotangent lift method while its offline computational cost is less than 20\% {\edit compared to the cotangent lift}.
 Lastly we notice that the combination of the greedy reduced basis and DEIM yields large errors in the solution while the solution using the symplectic DEIM is very accurate.

### 2016-11-15 greedy_hamiltonian_revised/chapters/6.conclusion.tex
fist commit

--- 2016-11-15 greedy_hamiltonian_new_template/chapters/6.conclusion.tex
@@
 
-In this paper, we present a greedy approach for the construction of a reduced system that preserves the geometric structure of Hamiltonian systems.
+In this paper, we {\edit presented} a greedy approach for the construction of a reduced system that preserves the geometric structure of Hamiltonian systems.
 An iteration of the greedy method comprises searching the parameter space using the error in the Hamiltonian, to find the best basis vectors that increase the overall accuracy of the reduced basis.

### 2016-11-15 greedy_hamiltonian_revised/paper.tex
fist commit

--- 2016-11-15 greedy_hamiltonian_new_template/paper.tex
@@
 
+\newcommand{\edit}{\color{red}}
+
 \begin{document}
@@
 \begin{abstract}
-While reduced-order models (ROMs) are popular for approximately solving large systems of differential equations, the stability of reduced models over long-time integration remains an open question.
+While reduced-order models (ROMs) {\edit has been} popular for approximately solving large systems of differential equations, the stability of reduced models over long-time integration remains an open question.
 We present a greedy approach for ROM generation of parametric Hamiltonian systems which captures the symplectic structure of Hamiltonian systems to ensure stability of the reduced model.
-Through the greedy selection of basis vectors, two new vectors are added at each iteration to the set of basis vectors to increase the overall accuracy of the reduce basis.
-We used the error in the Hamiltonian function due to model reduction, as an error indicator to search the parameter space and find the next best basis vectors.
+Through the greedy selection of basis vectors, two new vectors are added at each iteration to the set of basis vectors to increase the overall accuracy of the {\edit reduced} basis.
+We {\edit use} the error in the Hamiltonian function due to model reduction, as an error indicator to search the parameter space and find the next best basis vectors.
 We show that the greedy algorithm converges with exponential rate, under natural assumptions on the set of all solutions of the Hamiltonian system under variation of the parameters.
@@
 
-\appendix
-\input{./chapters/appendix.tex}
+%\appendix
+%\input{./chapters/appendix.tex}
 

### 2016-11-17 greedy_hamiltonian_revised/chapters/5.numres.tex
paper finished

@@
 In this section, we illustrate the performance of the greedy generation of a symplectic basis.
-First, the parametric linear wave equation is considered to compare SVD based methods with the greedy method.
+The parametric linear wave equation is considered to compare SVD based methods with the greedy method.
 The nonlinear model order reduction using the combination of DIEM and the symplectic basis is then illustrated by considering the parametric nonlinear Schr\"odinger equation.
+{\edit Finally we discuss the numerical convergence of the greedy method introduced in Algorithm \ref{alg:SyMo:3}.}
 
-\subsection{Parametric Linear Wave equation} Consider the {\edit parametric} linear wave equation
+\subsection{Parametric Linear Wave equation} \label{chap:NuRe:1.1} Consider the {\edit parametric} linear wave equation
 \begin{equation} \label{eq:NuRe:1}
@@
 
-The decay of the singular values for the POD, cotangent lift and the complex SVD method are presented in Figure \ref{fig:NuRe:2a}.
-The fast decay of the singular values suggests that a low dimensional solution manifold indeed exists.
+{\edit The decay of the singular values for the POD ca be found in Figure \ref{fig:NuRe:5a}.
+The decay} of the singular values suggests that a low dimensional solution manifold indeed exists.
 However since the linear subspace constructed by the POD is not symplectic, we observe the blow up of the Hamiltonian function in Figure \ref{fig:NuRe:2b} and the instability of the solution in Figure \ref{fig:NuRe:1}.
@@
 
-\caption{(a) The decay of singular values for POD, cotangent lift and complex SVD methods.
-(b) Plot of the Hamiltonian function for $t \in [0,30]$.
-(c) The $L^2$ error between the solution of the full system and the reduced system for different model reduction methods for $t \in [0,30]$.}
+\caption{{\edit (a) The $L^2$ error between the solution of the full system and the reduced system for different model reduction methods for $t \in [0,30]$.} (b) Plot of the Hamiltonian function for $t \in [0,30]$.
+}
 \label{fig:NuRe:2}
@@
 
-\subsection{Nonlinear Schr\"odinger equation} Let us consider the one-dimensional parametric Schr\"odinger equation
+\subsection{Nonlinear Schr\"odinger equation} \label{chap:NuRe:1.2} Let us consider the one-dimensional parametric Schr\"odinger equation
 \begin{equation} \label{eq:NuRe:10}
@@
 
-\caption{(a) The decay of singular values for POD, cotangent lift and complex SVD methods.
-(b) The $L^2$ error between the solution of the full system and the reduced system for different model reduction methods for $t \in [0,30]$.
-(c) Plot of the Hamiltonian function for $t \in [0,30]$.}
+\caption{{\edit (a) Plot of the Hamiltonian function for $t \in [0,30]$.} (b) The $L^2$ error between the solution of the full system and the reduced system for different model reduction methods for $t \in [0,30]$.
+}
 \label{fig:NuRe:4}
@@
 The size of the reduced systems are taken identical for all methods ($k=180$ for POD and $k=90$ for the rest).
-Although the decay of the singular values in Figure \ref{fig:NuRe:4a} suggests that the accuracy of the POD reduced system should be comparable to that of the other methods, we observe instabilities in the solution at $t=10$.
+Although the decay of the singular values in Figure \ref{fig:NuRe:5b} suggests that the accuracy of the POD reduced system should be comparable to that of the other methods, we observe instabilities in the solution at $t=10$.
 The greedy, the cotangent lift and the complex SVD method, on the other hand, generate a stable reduced system that accurately approximates the solution of the full model.
@@
 We note that the symplectic DEIM is even more accurate than the greedy itself since it has been enriched by the nonlinear snapshots.
+
+\subsection{Numerical Convergence} {\edit In this section we discuss the numerical convergence of the symplectic greedy method introduced in Section \ref{chap:SyMo:1}.
+The exponential convergence properties of the conventional greedy \cite{Quarteroni:2016wi} is presented in \cite{Buffa:2012iz,Binev:2011fj}.
+Theorem \ref{theorem:SyMo:2} suggests that the symplectic greedy method has similar properties.
+To illustrate this we compared the convergence of the conventional greedy with the convergence of the symplectic greedy method through the numerical simulations in sections \ref{chap:NuRe:1.1} and \ref{chap:NuRe:1.2}.
+
+The decay of the singular values of the snapshot matrix for the parametric wave equation and the nonlinear Schr\"odinger equation can be found in Figure \ref{fig:NuRe:5}.
+The rate of the decay of the singular values is a strong indicator for the rate of the decay of the Kolmogorov $n$-width of the solution manifold.
+We expect that the conventional greedy method and the symplectic greedy method provide a similar rate in the decay of the error.
+
+Figure \ref{fig:NuRe:5} shows the maximum $L_2$ error between the original system and the reduced system at each iteration of different greedy methods.
+In this figure we find the conventional greedy with orthogonal projection error as a basis selection criterion (orange), the symplectic greedy method with symplectic projection error as a basis selection criterion (green) and the symplectic greedy method with energy loss $\Delta H$ as a basis selection criterion (red).
+For simplicity, we may refer to these routines as the greedy with the orthogonal projection, the greedy with the symplectic projection and the greedy with the energy loss, respectively.
+
+It is observed that the decay rate of the error for greedy with the orthogonal projection and the greedy with the symplectic projection is similar to the decay of the singular values.
+This matches our expectation from Theorem \ref{theorem:SyMo:2}.
+We also notice that the greedy method with the energy loss provides excellent error indication as a basis selection criterion.
+}
+
+\begin{figure}
+
+\begin{minipage}{.5\linewidth}
+\centering
+\subfloat[]{\label{fig:NuRe:5a}\includegraphics[width=\textwidth]{./figs/wave/convergence}}
+\end{minipage}%
+\begin{minipage}{.5\linewidth}
+\centering
+\subfloat[]{\label{fig:NuRe:5b}\includegraphics[width=\textwidth]{./figs/schrodinger/convergence}}
+\end{minipage}\par\medskip
+\centering
+%\subfloat[]{\label{fig:NuRe:4c}\includegraphics[width=0.5\textwidth]{./figs/schrodinger/error}}
+
+\caption{ (a) Convergence of the greedy method for the wave equation. (b) Convergence of the greedy method for the nonlinear Schr\"odinger equation equation. }
+\label{fig:NuRe:5}
+\end{figure}

### 2016-11-17 response/response.tex
all done, response to the referees also added

\documentclass[a4paper]{article}
\usepackage{url}
\usepackage[margin=1in]{geometry}
\newcommand{\breview}{\begin{quotation}\begin{bf}\noindent}
\newcommand{\ereview}{\end{bf}\end{quotation}}
\newcommand{\reviewbullet}[1]{\breview \begin{itemize}\item #1 \end{itemize}\ereview}
\begin{document}

Dear Editor, \\[1cm]

Attached you will find the revised version of manuscript \#M106699, titled ''Structure Preserving Model Reduction of Parametric Hamiltonian Systems''.

Yours Sincerely,\\[1.0cm]
Babak Maboudi Afkham and
Jan S. Hesthaven
\\[1cm]

\section*{Reviewer 1}

\breview
1. There is a global lack of rigour in the mathematical presentation: It starts with the very first two equations. If (1) is clear, the definition of M in the equation (2) is obscure: what is the role of the time? Is there a finite time horizon? Then in equation (3) what are the v\_i? are they made of snapshots as explained later? But then what is the role of the numerical discretisation ?
\ereview

We revised major parts of the background and the introduction of the manuscript and corrected the mathematical notations for better comprehension. The authors believe that the introduction of the solution manifold, equation (2), and the reduced basis, equation (3), is essential in the understanding of model reduction techniques. The role of time was missing from the definition which is now added. However, the generation of the reduced basis, v\_i, is carefully and in much detail is discussed in chapters 3 and 4.

\breview
2. Similarly, in section 4, there are some notations used that are never defined (line 376 what do the assumptions on a and b mean?) All this lack of rigorous explanation (which contrasts with the heavy Hamiltonian formalism used in Section 3) makes the paper difficult to read and understand. 
\ereview

The explanation of the symplectic Gram-Schmidt routine (the part pointed by the reviewer) is revised for better understanding.

\breview
3. I was expecting much more convincing numerical experiments. The first stochastic example is not really understandable.

The example concerning the nonlinear Schroedinger equation is also relatively easy. If I understand correctly, there are 500 points in the parameter space and 90 in the reduced basis. Is there really a gain between computing all the trajectories and finding the reduced basis?
\ereview

We agree with the reviewer that the numerical examples presented are relatively simple. However, alternative methods are either unstable or too expensive for even these simple test cases. Therefore, a more complex numerical experiment does not contribute to a better understanding of the method.

\breview
4. I cannot find out what is plotted on Figure 1. The expectation of the solution? One realization? Then the authors talk about a gain of computational cost (which should be the goal of this new algorithm) but nothing is really shown.

What about a comparison between the different computational cost of the algorithm? 
\ereview

The purpose of Figure 1 is to illustrate how the proposed method can preserve stability with the eyeball norm, and so using just a single realization. To eliminate any confusion, we have added Figure 5 that presents the maximum error in time and parameter domain. Also the qualitative computation cost is discussed in Section 5.

\breview
A comparison with Monte Carlo algorithm is also missing. Moreover, t = 2 is not a long time for a 1D linear wave equation and the gain of symplectic algorithms is unclear. 
\ereview

The purpose of model reduction is to construct approximated solution with substantial lower cost. It is not clear to the authors, possibly due to a miss understanding, how Monte Carlo method enters this scope.

\breview
What about an illustration of the exponential convergence result given in Section 4? 
\ereview

We agree with the reviewer that a better illustration of the exponential convergence is needed. We added section 5.3 to numerically discuss this matter.

\section*{Reviewer 2}

\breview
1. The topic of projection methods for Hamiltonian systems is a widely studied topic in the context of optimal control problems where almost all the concepts discussed in this paper have already been studied, symplectic QR, symplectic SVD, symplectic Krylov and Lanc- zos techniques, model reduction for Hamiltonian systems (Riccati equations) etc. All these techniques are reinvented in this paper without reference to the previous work. This is unac- ceptable and in some sense even unethical, since any google search would have immediately revealed these papers by e.g. Benner, Breiten, Bunse-Gerstner, Freund, Kressner, Mehrmann, Poloni, Saak, Schoeder, Simoncini, Watkins etc. Many of these are also dealing with model reduction and particularly adress the preservation of Hamiltonian and symplectic structure in projection methods.
\ereview

Most of the topics and references introduced above deal with eigenvalue problems of Hamiltonian, skew-Hamiltonian and symplectic matrices. The main focus of this manuscript however, is the model reduction of Hamiltonian systems specifically the greedy approach to structure preserving model reduction. This, although strongly related, is not reflected in the references suggested by the reviewer. Nonetheless we agree that including these topics in the paper is essential as they comprise a major part of the development of the field in optimal control. Therefore, most of these references are included in the revision of the paper.

\breview
2. There are also some mathematical questions that need to be addressed. It is well-known that symplectic bases are not norm bounded, so special precautions have to be taken to guarantee that they do not become unbounded. This can be done by the isotropic Arnoldi/Gram Schmidt and needs to be done here. Unfortunately this procedure requires the existence of Lagrange subspaces which is not the case here.

I do not see the reflection of this unboundedness in the convergence proof, I guess it is reflected in the constant, but this should be analyzed. 
\ereview
The symplectic greedy method suggested in the paper, produces an orthosymplectic (symplectic and orthonormal) basis by construction. The orthonormality of the basis guarantees a numerically stable basis. Although this was briefly discussed in the paper, there was clearly a need for a detailed discussion. To this end, it is shown in the revision of the paper, as pointed by the reviewer, that the greedy method constructs a Lagrangian subspace in each iteration and then expands it to an orthosymplectic basis. We emphasised on this in the revision of the paper in form of Theorem 21.

\breview
Another mathematical problem is the stability of the projected system, this is to my opinion not guaranteed and needs to be proved, e.g. it needs to be guaranteed that there does not exists a zero Lyapunov exponent associated with crossing trajectories and with mixed sign/Krein characteristic which puts the Hamiltonian system on the boundary of the stable region and small perturbations will make it unstable. Here structure preservation helps but does not solve the problem.
\ereview

It was briefly discussed in the paper that stable critical points of a Hamiltonian system remains stable in the reduced system. This matter is discussed in more detail in the revision of the paper. This together with the numerical experiments in the paper, although do not guarantee general stability, but are strong indication of the stability of the reduced system.

\end{document}

### 2016-11-17 greedy_hamiltonian_revised/chapters/3.ham.tex
all done, response to the referees also added

@@
 \section{Hamiltonian Systems and Symplectic Geometry} \label{chap:Hasy:1}
-Let $\edit \mathcal M$ be a manifold and $\edit \Omega:\mathcal M \times \mathcal M \to \mathbb R$ be a closed, nondegenerate 2-form on $\edit \mathcal M$.
+Let $\edit \mathcal M$ be a manifold and $\edit \Omega:\mathcal M \times \mathcal M \to \mathbb R$ be a closed, nondegenerate {\edit and skew-symmetric} 2-form on $\edit \mathcal M$.
 The pair $\edit (\mathcal M,\Omega)$ is called a \emph{symplectic manifold} {\edit \cite{Marsden:1999ck}}.
@@
 \end{equation}
-where $\delta$ is the Kronecker's delta function. Moreover if $V = \mathbb{R}^{2n}$ we can choose the basis vectors $\{e_i,f_i\}_{i=1}^n$ such that
+where $\delta$ is the Kronecker's delta function. Moreover if $V = \mathbb{R}^{2n}$ we can choose basis vectors $\{e_i,f_i\}_{i=1}^n$ such that
 \begin{equation} \label{eq:Hasy:5}
@@
 \begin{theorem} \label{theorem:Hasy:1.6}
-Let $E\subset V$ be a Lagrangian vector space. Then there is an orthosymplectic basis for $\mathbb R^{2n}$ \cite{da2003introduction}.
+Suppose $(V,\Omega)$ is a $2n$ dimensional symplectic vector space and $E\subset V$ is a Lagrangian subspace. Then there is an orthosymplectic basis for $V$ \cite{da2003introduction}.
 \end{theorem}
@@
 Starting from a Lagrangain subspace in $E \subset V$ an orthosymplectic basis can be easily constructed.
+By Theorem \ref{theorem:Hasy:1.5} $E$ is $n$ dimensional.
 Suppose that $\{ e'_1,\dots, e'_n \}$ is a basis for $E$, using the classical Gram-Schmidt orthogonalization process we can construct an orthonormal basis $\{ e_1,\dots,e_n \}$.
-Define the new set of vectors $f_1 = -\mathbb Je_1$, $f_2 = -\mathbb Je_2$, $\dots$, $f_n= -\mathbb Je_n$.
+Define the new set of vectors $f_1 = \mathbb J_{2n}^Te_1$, $f_2 =\mathbb J_{2n}^T e_2$, $\dots$, $f_n= \mathbb J_{2n}^Te_n$.
 We have
 \begin{equation}
-	\langle f_i, f_j \rangle = e_i^T \mathbb J_{2n}^T \mathbb J_{2n} e_i = \delta_{ij}, \quad \langle f_i, e_j \rangle = -e_i \mathbb J^T_{2n} e_j = 0, \quad i,j=1,\dots,n,
+	\langle f_i, f_j \rangle = e_i^T \mathbb J_{2n} {\mathbb J_{2n}}^T e_i = \delta_{ij}, \quad \langle f_i, e_j \rangle = e_i^T \mathbb J_{2n} e_j = 0, \quad i,j=1,\dots,n,
 \end{equation}
-where the second identity is due to the fact that the basis $\{ e_1,\dots,e_n \}$ forms a Lagrangian subspace.
+where we used the fact that $\mathbb J_{2n} {\mathbb J_{2n}}^T = I_{2n}$ in the first identity and the second identity is due to the fact that the basis $\{ e_1,\dots,e_n \}$ forms a Lagrangian subspace.
 This shows that the set $\{ e_1,\dots,e_n \}\cup \{ f_1,\dots,f_n \}$ forms an orthonormal basis.
@@
 
-{\edit It natural to expect a numerical integrator that solves (\ref{eq:Hasy:7}) to also satisfy the conservation law in Theorem \ref{theorem:Hasy:1}.
+{\edit It is natural to expect a numerical integrator that solves (\ref{eq:Hasy:7}) to also satisfy the conservation law in Theorem \ref{theorem:Hasy:1}.
 Common numerical integrators e.g., Runge-Kutta methods, do not generally preserve the Hamiltonian which results in a wrong qualitative behavior in the solution \cite{Hairer:1250576}.

### 2016-11-17 greedy_hamiltonian_revised/chapters/4.sympmor.tex
paper finished
all done, response to the referees also added

@@
 In the context of the Hamiltonian systems, a suitable candidate for the Lyapunov function is the Hamiltonian function $H$.
-The following theorem shows that when $H$ (or $-H$) is a Lyapunov function, then the equilibrium points of the original and the reduced system are Liyapunov stable.
+The following theorem shows that when $H$ (or $-H$) is a Lyapunov function, then the equilibrium points of the original and the reduced system are Liyapunov stable \cite{abraham1978foundations}.
 \begin{theorem} \label{theorem:SyMo:1}
@@
 Suppose $\mathbf z_e$ is an equilibrium point for (\ref{eq:Hasy:7}) and that $\mathbf y_e = A^+\mathbf z_e$.
-If $H$ (or $-H$) is a Lyapunov function satisfying proposition \ref{proposition:SyMo:1}, then $\mathbf z_e$ and $\mathbf y_e$ are Lyapunov stable points for (\ref{eq:Hasy:7}) and (\ref{eq:SyMo:4}) respectively.
+If $H$ (or $-H$) is a Lyapunov function satisfying proposition \ref{proposition:SyMo:1}, then $\mathbf z_e$ and $\mathbf y_e$ are Lyapunov stable equilibrium points for (\ref{eq:Hasy:7}) and (\ref{eq:SyMo:4}) respectively.
 \end{theorem}
@@
 As discussed in Section \ref{chap:Hasy:1}, any finite dimensional symplectic linear vector space has a symplectic basis that satisfies conditions (\ref{eq:Hasy:4}).
-Further theorem \ref{theorem:Hasy:1.6} provides an iterative process for constructing an orthosymplectic basis \cite{Matsuo:2014wl,Salam2014}.
+Further, theorem \ref{theorem:Hasy:1.6} provides an iterative process for constructing an orthosymplectic basis \cite{Matsuo:2014wl,Salam2014}.
 To briefly describe the SR method, suppose that an orthosymplectic basis
 \begin{equation} \label{eq:SyMo:14.1}
-	A_{2k}=\{ e_1 , \dots , e_k \} \cup \{ -\mathbb J_{2n} e_1 , \dots , -\mathbb J_{2n} e_k \},
+	A_{2k}=\{ e_1 , \dots , e_k \} \cup \{ \mathbb J_{2n}^T e_1 , \dots , \mathbb J_{2n}^T e_k \},
 \end{equation}
@@
 \begin{equation} \label{eq:SyMo:14.2}
-	\Omega\left(z + \sum_{i=1}^k \alpha_i e_i - \sum_{i=1}^k \beta_i \mathbb J_{2n}e_i , \sum_{i=1}^k \bar{\alpha}_i e_i - \sum_{i=1}^k \bar{\beta}_i \mathbb J_{2n}e_i \right) = 0,
+	\Omega\left(z + \sum_{i=1}^k \alpha_i e_i + \sum_{i=1}^k \beta_i \mathbb J_{2n}^Te_i , \sum_{i=1}^k \bar{\alpha}_i e_i + \sum_{i=1}^k \bar{\beta}_i \mathbb J_{2n}^Te_i \right) = 0,
 \end{equation}
@@
 \begin{equation} \label{eq:SyMo:14.3}
-	\alpha_i = - \Omega(z,-\mathbb J_{2n}e_i), \quad \beta_i = \Omega(z,e_i),
+	\alpha_i = - \Omega(z,\mathbb J_{2n}^Te_i), \quad \beta_i = \Omega(z,e_i),
 \end{equation}
@@
 \begin{equation} \label{eq:SyMo:14.4}
-	\tilde z = z - \sum_{i=1}^k \Omega(z,-\mathbb J_{2n}e_i) e_i - \sum_{i=1}^k \Omega(z,e_i) \mathbb J_{2n}e_i.
+	\tilde z = z - \sum_{i=1}^k \Omega(z,\mathbb J_{2n}^Te_i) e_i + \sum_{i=1}^k \Omega(z,e_i) \mathbb J_{2n}^Te_i.
 \end{equation}
 If we introduce $e_{k+1} = \tilde z / \| \tilde z \|_2$, it is easily checked that $e_{k+1}$ is also orthogonal to $A_{2k}$ with respect to the classical inner product.
-Therefore the set span$(\{e_1,\dots,e_{k+1}\})$ forms a Lagrangian subspace and according to theorem \ref{theorem:Hasy:1.6} the basis $A_{2k+2}= A_{2k}\cup \{ e_{k+1} , -\mathbb J_{2n} e_{k+1} \}$ forms an orthosymplectic basis.
+Therefore the set span$(\{e_1,\dots,e_{k+1}\})$ forms a Lagrangian subspace and according to theorem \ref{theorem:Hasy:1.6} the basis $A_{2k+2}= A_{2k}\cup \{ e_{k+1} , \mathbb J_{2n}^T e_{k+1} \}$ forms an orthosymplectic basis.
 
@@
 \begin{equation} \label{eq:SyMo:14.7}
-	e_{k+1} = \tilde z, \quad A_{2k+1} = A_{2k}\cup \{ e_{k+1} , -\mathbb J_{2n}e_{k+1} \},
+	e_{k+1} = \tilde z, \quad A_{2k+1} = A_{2k}\cup \{ e_{k+1} , \mathbb J_{2n}^Te_{k+1} \},
 \end{equation}
@@
 
-{\edit
 \begin{algorithm} 
@@
 \item \hspace{0.5cm} $e_{k+1} \leftarrow \mathbf z^*/ \| z^*\|$
-\item \hspace{0.5cm} $A \leftarrow [e_1,\dots ,e_{k+1} , -\mathbb J^T_{2n}e_1,\dots,-\mathbb J^T_{2n}e_{k+1}]$
+\item \hspace{0.5cm} $A \leftarrow [e_1,\dots ,e_{k+1} , \mathbb J^T_{2n}e_1,\dots,\mathbb J^T_{2n}e_{k+1}]$
 \item \hspace{0.5cm} $k \leftarrow k+1$
@@
 \end{algorithm}
-}
 
@@
 This establishes the following theorem.
-\begin{theorem}
+\begin{theorem} \label{theorem:SyMo:2}
 	Let $S$ be a compact subset of $\mathbb{R}^{2n}$ with exponentially small Kolmogorov $n$-width $\edit d_{k}\leq c\exp(-\alpha k)$ with $\alpha > \log3$.
@@
 {\edit
-\begin{theorem}
+\begin{theorem} \label{theorem:SyMo:3}
 	The basis $A_{2k}$ generated by the greedy algorithm is orthosymplectic.

### 2016-11-17 greedy_hamiltonian_revised/chapters/5.numres.tex
fixed minor error
all done, response to the referees also added

@@
 \centering
-\subfloat[$t=1$]{\label{fig:NuRe:3b}\includegraphics[width=\textwidth]{./figs/schrodinger/solution/solution_t10}}
+\subfloat[$t=10$]{\label{fig:NuRe:3b}\includegraphics[width=\textwidth]{./figs/schrodinger/solution/solution_t10}}
 \end{minipage}\par\medskip
 \centering
-\subfloat[$t=2$]{\label{fig:NuRe:3c}\includegraphics[width=0.5\textwidth]{./figs/schrodinger/solution/solution_t20}}
+\subfloat[$t=20$]{\label{fig:NuRe:3c}\includegraphics[width=0.5\textwidth]{./figs/schrodinger/solution/solution_t20}}
 \caption{The solution $|u(t,x)| = \sqrt{q^2 + p^2}$ at $t=0$, $t=10$ and $t=20$ of the Nonlinear Schr\"odinger equation for parameter value $\epsilon = 1.0932$.
@@
 
-Figure \ref{fig:NuRe:4c} illustrates the $L^2$-error between the solution of the full model with the reduced systems generated by different methods.
+Figure \ref{fig:NuRe:4c} illustrates the $L_2$-error between the solution of the full model with the reduced systems generated by different methods.
 We first observe that symplectic methods yield a lower computational error compared to non-symplectic methods.
@@
 The decay of the singular values of the snapshot matrix for the parametric wave equation and the nonlinear Schr\"odinger equation can be found in Figure \ref{fig:NuRe:5}.
-The rate of the decay of the singular values is a strong indicator for the rate of the decay of the Kolmogorov $n$-width of the solution manifold.
+The decay rate of the singular values is a strong indicator for the decay rate of the Kolmogorov $n$-width of the solution manifold.
 We expect that the conventional greedy method and the symplectic greedy method provide a similar rate in the decay of the error.
@@
 This matches our expectation from Theorem \ref{theorem:SyMo:2}.
-We also notice that the greedy method with the energy loss provides excellent error indication as a basis selection criterion.
+We also notice that the greedy method with the energy loss provides an excellent error indication as a basis selection criterion.
 }

### 2016-12-09 greedy_hamiltonian_revised/chapters/1.introduction.tex
Jan's comments included

--- 2016-11-15 greedy_hamiltonian_new_template/chapters/1.introduction.tex
@@
 \section{Introduction}
-Parameterized partial differential equations are very often used as tools to model many problems in engineering and the applied sciences.
-While the need for more accuracy has led to the development of exceedingly complex models, the computational and storage limitations often make direct approaches impossible.
+{\edit Parameterized partial differential equations often arise as a model in many problems in engineering and the applied sciences}.
+While the need for more accuracy has led to the development of exceedingly complex models, the {\edit limitations in computational cost and storage} often make direct approaches {\edit impractical}.
 Hence, we must seek alternative methods that allow us to approximate the desired output under variation of the input parameters while keeping the computational costs to a minimum.
@@
 Reduced basis methods have emerged as a powerful approach for the reduction of the intrinsic complexity of such models \cite{Ito:1998up,Ito:1998ch,Ito:2001ev,Peterson:1989ki}.
-These methods contain two stages: the offline and the online.
-In the offline stage, one explors the parameter space to construct a low-dimensional basis that accurately represents the parametrized solution to the partial differential equation.
-In this stage, evaluation of the solution of the original model for multiple parameter values is required.
-The online stage comprises a Galerkin projection onto the span of the reduced basis, which allows exploration of the parameter space at significantly reduced complexity \cite{Antoulas:2005:ALD:1088857,Anonymous:2016wl}.
+These methods contain two stages: {\edit the offline stage and the online stage}.
+In the offline stage, one explores the parameter space to construct a low-dimensional basis that accurately represents the parametrized solution to the partial differential equation.
+In this stage, the evaluation of the solution of the original model for multiple parameter values is required.
+The online stage comprises a Galerkin projection onto the span of the reduced basis, which allows exploration of the parameter space at a significantly reduced complexity \cite{Antoulas:2005:ALD:1088857,Anonymous:2016wl}.
 
-Convectional reduced basis techniques e.g.
-proper orthogonal decomposition (POD) \cite{Kunisch:2002er,Atwell:2001ja,Ravindran:2002hn}, require the exploration of the entire parameter space.
-This leads to a very expensive and often impractical offline stages when dealing with multi-dimensional paramtere domains.
-On the other hand, sampling techniques, usually of a greedy nature, search through the parameter space selectively often guided by an error estimate to certify the accuracy of the basis.
-This approach, accompanied with an efficient sampling procedure, trades considerable amount of computation with overall accuracy of the reduced-basis \cite{Cuong:2005gd,Rozza:2005ie,Anonymous:2016wl}.
+Convectional reduced basis techniques, such as proper orthogonal decomposition (POD) \cite{Kunisch:2002er,Atwell:2001ja,Ravindran:2002hn}, require the exploration of the entire parameter space.
+This leads to a very expensive and often impractical offline {\edit stage} when dealing with multi-dimensional parameter domains.
+On the other hand, sampling techniques, usually of a greedy nature, search through the parameter space selectively, guided by an error estimate to certify the accuracy of the basis.
+This approach, accompanied with an efficient sampling procedure, {\edit balances the cost} of computation with the overall accuracy of the reduced-basis \cite{Cuong:2005gd,Rozza:2005ie,Anonymous:2016wl}.
 
-Beside computational complexity, another aspect of reduced order modeling is the preservation of structures and, in particular, stability of the original model.
+{\edit Besides} computational complexity, another aspect of reduced order modeling is the preservation of structure and, in particular, {\edit the} stability of the original model.
 In general, reduced order models do not guarantee that such properties are preserved \cite{Anonymous:pMn0O0Q4}.
 
-In the context of Hamiltonian and Lagrangian systems, recent work suggests modifications of POD to preserve geometric structure.
+In the context of Hamiltonian and Lagrangian systems, recent work suggests modifications of POD to preserve {\edit some} geometric structure.
 Lall et al.
 \cite{Lall:2003iy} and Carlberg et al.
-\cite{Carlberg:2014ky} suggests that the reduced-order system should be identified by a Lagrangian function on a low-ordered configuration space.
-In this way the geometric structure of the original system is inherited by the reduced system.
-For Hamiltonian systems Peng et al.
+\cite{Carlberg:2014ky} suggests that the reduced-order system should be identified by a Lagrangian function on a low-dimensional configuration space.
+In this way, the geometric structure of the original system is inherited by the reduced system.
+{\edit Model reduction for port-Hamiltonian systems can be found in the works of Beattie et al.
+\cite{Chaturantabut:2016he}, Polyuga et al.
+\cite{Polyuga:2010gj} and references therein.
+These works construct a reduced port-Hamiltonian system using Krylov or POD methods that inherit the passivity and stability of the original system.} For Hamiltonian systems, Peng et al.
 \cite{Peng:2014di}, using a symplectic transformation, constructs a reduced Hamiltonian, as an approximation to the Hamiltonian of the original system.
-As a result the reduced system preserves the symplectic structure.
-Although these methods preserve geometric structure, they use a POD-like approach for constructing the reduced basis, and are not well suitable for problems with a high-dimensional parameter domain.
+As a result, the reduced system preserves the symplectic structure.
+Although these methods preserve {\edit the} geometric structure, they use a POD-like approach for constructing the reduced basis and are not well {\edit suited} for problems with a high-dimensional parameter domain.
 
-In this paper, we present a greedy approach for the construction of a reduced system that preserves the geometric structure of Hamiltonian systems, as an iterative routine that successively enriches the reduced basis at each iteration.
-On one hand this technique results in a reduced Hamiltonian system that mimics symplectic properties of the original system and preserves the Hamiltonian structure and stability over the course of time integration.
+In this paper, we present a greedy approach for the construction of a reduced system that preserves the geometric structure of Hamiltonian systems.
+This technique results in a reduced Hamiltonian system that mimics the symplectic properties of the original system and preserves the Hamiltonian structure and stability over the course of time.
 On the other hand, since time integration of the original system is only required once per iteration, the proposed method saves substantial computational cost during the offline stage when compared to alternative POD-like approaches.
-Moreover, we demonstrate that assumptions, natural for the set of all solutions of the original Hamiltonian system under variation of parameters, lead to exponentially fast convergence of the greedy algorithm.
-For nonlinear Hamiltonian systems, we show how the basis can be combined with the discrete empirical interpolation method (DEIM) to allow a fast evaluation of nonlinear terms while maintaining the symplectic structure.
+{\edit It is well known that structured matrices, e.g.
+symplectic matrices, generally are not well-conditioned \cite{Karow:2006cf}.
+The greedy update of the symplectic basis presented here, yields a orthosymplectic basis and, therefore, a norm bounded basis.} Moreover, we demonstrate that assumptions, natural for the set of all solutions of the original Hamiltonian system under variation of parameters, lead to exponentially fast convergence of the greedy algorithm.
+For nonlinear Hamiltonian systems, we show how the basis can be combined with the discrete empirical interpolation method (DEIM) {\edit \cite{Chaturantabut:2010cz,Barrault:2004kz}} to enable a fast evaluation of nonlinear terms while maintaining the symplectic structure.
 
@@
 Accuracy, stability, and efficiency of the greedy method compared to other SVD-based methods are discussed in Section \ref{chap:NuRe:1}.
-Finally we offer the conclusion in Section \ref{chap:Con:1}.
+{\edit Finally we offer some conclusive remarks in Section \ref{chap:Con:1}}.

### 2016-12-09 greedy_hamiltonian_revised/chapters/2.mor.tex
all done, response to the referees also added
Jan's comments included

@@
 \section{Model Order Reduction} \label{chap:MoOr:1}
-Consider a parameterized, finite dimensional dynamical system that can be described by a set of first order ordinary differential equations on the form
+Consider a parameterized, finite dimensional dynamical system described by a set of first order ordinary differential equations
 \begin{equation} \label{eq:MoOr:1}
@@
 \end{equation}
-Here $\mathbf x \in \mathbb R^n$ is the state vector, $\omega \in \Gamma$ is a vector containing all the parameters of the system that belongs to a compact set $\Gamma$ ($\subset \mathbb R^d$) and $\mathbf f : \mathbb R \times \mathbb R^n \times \Gamma \to \mathbb R^n$ is a general function of the state variables and parameters.
+Here $\mathbf x \in \mathbb R^n$ is the state vector, $\omega \in \Gamma$ is a vector containing all the parameters of the system {\edit belonging} to a compact set $\Gamma$ ($\subset \mathbb R^d$) and $\mathbf f : \mathbb R \times \mathbb R^n \times \Gamma \to \mathbb R^n$ is a general {\edit vector valued} function of the state variables and parameters.
 
@@
 \begin{equation} \label{eq:MoOr:2}
-	\mathcal M = \{ \mathbf x(\omega) | \omega \in \Gamma \} \subset \mathbb R^n.
+	\edit \mathcal M = \{ \mathbf x(t,\omega) | \omega \in \Gamma,\ t\geq 0 \} \subset \mathbb R^n.
 \end{equation}
-Note that the exact solution and solution manifold is not always available; we assume that we have a numerical integrator in hand that can approximate the solution to (\ref{eq:MoOr:1}) for any realization of $\omega$ {\edit with a given accuracy}.
-By abuse of notation we refer to $\mathbf x$ and $\mathcal M$ as the exact solution and the exact solution manifold, respectively, rather than the discrete solution and discrete solution manifold.
+Note that the exact solution and solution manifold is {\edit often not} available; we assume that we have a numerical integrator that can approximate the solution to (\ref{eq:MoOr:1}) for any realization of $\omega$ {\edit with a given accuracy}.
+By abuse of notation, we refer to $\mathbf x$ and $\mathcal M$ as the exact solution and the exact solution manifold, respectively, rather than the discrete solution and discrete solution manifold.
 
-Model order reduction is based on the assumption that $\mathcal M$ is of low dimension \cite{Anonymous:2016wl,Antoulas:2005:ALD:1088857}, and that the span of appropriately chosen basis vectors $\{v_i\}_{i=1}^k$ covers most of the solution manifold with a small error.
+Model order reduction is based on the assumption that $\mathcal M$ is of low dimension \cite{Anonymous:2016wl,Antoulas:2005:ALD:1088857} and that the span of appropriately chosen basis vectors $\{v_i\}_{i=1}^k$ covers most of the solution manifold {\edit to within} a small error.
 The set $\{v_i\}_{i=1}^k$ is denoted as the \emph{reduced basis} and its span as the \emph{reduced space}.
@@
 \end{equation}
-where $V$ is a matrix containing the reduced basis vectors as its columns and $\mathbf y$ is the coordinates of the approximation in this basis.
+where $V$ is a matrix containing the reduced basis vectors as its columns and $\mathbf y$ {\edit contains} the coordinates of the approximation in this basis.
 By substituting (\ref{eq:MoOr:3}) into (\ref{eq:MoOr:1}) we obtain the overdetermined system
@@
 \end{equation}
-Here we added the residual $\mathbf r$ to emphasise that (\ref{eq:MoOr:4}) is an approximation of (\ref{eq:MoOr:1}).
-Taking the Petrov-Galerkin projection \cite{Antoulas:2005:ALD:1088857} we construct a basis $W$ of size $n-k$ that is orthogonal to the residual $\mathbf r$ and assume $W^T V$ is invertible.
+Here we added the residual $\mathbf r$ to emphasize that (\ref{eq:MoOr:4}) is an approximation of (\ref{eq:MoOr:1}).
+Taking the Petrov-Galerkin projection \cite{Antoulas:2005:ALD:1088857} we construct a basis $W$ of size $n-k$ that is orthogonal to the residual $\mathbf r$ and {\edit requires that} $W^T V$ is invertible.
 This yields
@@
 Equation (\ref{eq:MoOr:5}) consists of $k$ equations and is called the reduced system.
-Solving the reduced system instead of the original system can reduce the computational costs since $k$ is significantly smaller than $n$.
+Solving the reduced system instead of the original system can reduce the computational costs {\edit provided} $k$ is significantly smaller than $n$.
 For nonlinear systems, the evaluation of $\mathbf f$ may still have computational complexity that depends on $n$.
@@
 Let $\mathbf x (t_i,\omega_j)$ with $i=1,\dots,m$ and $j=1,\dots,n$ be a finite number of samples, referred to as \emph{snapshots}, from the solution manifold (\ref{eq:MoOr:2}).
-If we {\edit suppose} that a reduced basis $V$ is provided, the projection operator from $\mathbb R^n$ onto the reduced space can be constructed as $VV^T$.
-The proper orthogonal decomposition (POD) requires the total error of projecting all the snapshots onto the reduced space to be minimum.
-The POD basis of size $k$ is thus the solution of the optimization problem
+If we {\edit assume} that a reduced basis $V$ is provided, the projection operator from $\mathbb R^n$ onto the reduced space can be constructed as $VV^T$.
+The proper orthogonal decomposition (POD) requires the total error of projecting all the snapshots onto the reduced space to be {\edit minimized}.
+The POD basis of size $k$ is thus the solution to the optimization problem
 \begin{equation} \label{eq:MoOr:6}
@@
 \end{equation}
-Here $S$ is the snapshot matrix, containing snapshots $\mathbf x(t_i,\omega_j)$ in its columns, $\|\cdot \|_F$ is the Frobenius norm of a matrix and $I_k$ is the identity matrix of size $k$.
-According to {\edit Schmidt-Mirsky-Eckart-Young theorem \cite{Markovsky:2011:LRA:2103589}}, the solution to the optimization (\ref{eq:MoOr:6}) is equivalent to the truncated singular value decomposition (SVD) of the snapshot matrix $S$ given by
+Here $S$ is the snapshot matrix, containing snapshots $\mathbf x(t_i,\omega_j)$ in its columns, $\|\cdot \|_F$ is the Frobenius norm and $I_k$ is the identity matrix of size $k$.
+According to {\edit Schmidt-Mirsky-Eckart-Young theorem \cite{Markovsky:2011:LRA:2103589}}, the solution to (\ref{eq:MoOr:6}) is equivalent to the truncated singular value decomposition (SVD) of the snapshot matrix $S$ given by
 \begin{equation} \label{eq:MoOr:7}
@@
 \end{equation}
-Here $\sigma_i, u_i$ and $v_i$ are the singular values, left singular vectors and right singular vectors of $S$, respectively {\edit \cite{Markovsky:2011:LRA:2103589} }.
+Here $\sigma_i, u_i$ and $v_i$ are the singular values, the left singular vectors, and the right singular vectors of $S$, respectively {\edit \cite{Markovsky:2011:LRA:2103589} }.
 
 \subsection{Discrete Empirical Interpolation Method (DEIM)} \label{chap:MoOr.DEIM:1} \nocite{Chaturantabut:2010cz}
-In this section we {\edit discuss the efficiency} of evaluating nonlinearities when using projection based reduced model.
-Suppose that the right hand side function in (\ref{eq:MoOr:1}) is of the form $\mathbf f(t,\mathbf x , \omega) = L\mathbf x + \mathbf g(t,\mathbf x ,\omega)$, where $L\in \mathbb R^{n\times n}$ reflects the linear part, and $\mathbf g$ is a nonlinear function.
-Now suppose a $k$-dimensional reduced basis $V$ is provided.
+In this section we {\edit discuss the efficiency} of evaluating nonlinearities {\edit in the context of} projection based reduced models.
+Suppose that the right hand side in (\ref{eq:MoOr:1}) is of the form $\mathbf f(t,\mathbf x , \omega) = L\mathbf x + \mathbf g(t,\mathbf x ,\omega)$, where $L\in \mathbb R^{n\times n}$ reflects the linear part, and $\mathbf g$ is a nonlinear function.
+Now {\edit assume that} a $k$-dimensional reduced basis $V$ is provided.
 The reduced system takes the form
@@
 Here, $\tilde L$ is a $k\times k$ matrix which can be computed before time integration of the reduced system.
-However, evaluation of $\tilde N (\mathbf y)$ has a complexity that depends on $n$, the size of the original system.
-Suppose that evaluation of $\mathbf g$ with $n$ components has complexity $\alpha(n)$, for some function $\alpha$.
-Then the complexity of evaluating $\tilde N(\mathbf y)$ is $\mathcal{O}(\alpha(n) + 4nk)$ which consists of 2 matrix-vector operations and the evaluation of the nonlinear function.
-This means that estimation of the nonlinear terms can be as expensive as solving the original system.
+However, {\edit the} evaluation of $\tilde N (\mathbf y)$ has a complexity that depends on $n$, the size of the original system.
+Suppose that the evaluation of $\mathbf g$ with $n$ components has the complexity $\alpha(n)$, for some function $\alpha$.
+Then the complexity of evaluating $\tilde N(\mathbf y)$ is $\mathcal{O}(\alpha(n) + 4nk)$ which consists of 2 matrix-vector operations and the {\edit evaluation of the nonlinear function, i.e.
+the evaluation of the nonlinear terms can be as expensive as solving the original system.}
 
-{\edit To overcome this bottleneck we take an approach similar to section \ref{chap:MoOr.PrOr:1} \cite{Chaturantabut:2010cz,Barrault:2004kz}}.
+{\edit To overcome this bottleneck we take an approach similar to that of Section \ref{chap:MoOr.PrOr:1} \cite{Chaturantabut:2010cz,Barrault:2004kz}}.
 Assume that the manifold $\mathcal M_{\mathbf g} = \{ \mathbf g(t,\mathbf x , \omega)| t\in \mathbb R, \mathbf x \in \mathbb R , \omega \in \Gamma\}$ is of a low dimension and that $\mathbf g$ can be approximated by a linear subspace of dimension $m\ll n$, spanned by the basis $\{ u_1 , \dots , u_m \}$, i.e.
@@
 \end{equation}
-Here $U$ contains basis vectors $u_i$ and $\mathbf c$ is the corresponding vector of coefficients. Now suppose $p_1,\dots,p_m$ are $m$ indices from $\{1,\dots,n\}$ and define an $n\times m$ matrix
+Here $U$ contains basis vectors $u_i$ and $\mathbf c$ is the vector of coefficients. Now suppose $p_1,\dots,p_m$ are $m$ indices from $\{1,\dots,n\}$ and define an $n\times m$ matrix
 \begin{equation} \label{eq:MoOr:11}
@@
 where $e_{p_i}$ is the $p_i$-th column of the identity matrix $I_n$.
-Hence, multiplying $P$ with $\mathbf g$ selects components $p_1,\dots,p_m$ of $\mathbf g$.
+Multiplying $P$ with $\mathbf g$ selects components $p_1,\dots,p_m$ of $\mathbf g$.
 If we assume that $P^TU$ is non-singular, the coefficient vector $\mathbf c$ can be uniquely determined from
@@
 \end{equation}
-which is referred to as the \emph{Discrete Empirical Interpolation} (DEIM) approximation \cite{Chaturantabut:2010cz}. Applying DEIM on the reduced system (\ref{eq:MoOr:5}) yields
+which is referred to as the \emph{Discrete Empirical Interpolation} (DEIM) approximation \cite{Chaturantabut:2010cz}. Applying DEIM to the reduced system (\ref{eq:MoOr:5}) yields
 \begin{equation} \label{eq:MoOr:14}
@@
 
-{\edit To obtain the projection basis $U$, the POD can be applied to the \emph{nonlinear snapshots}.
-The nonlinear snapshots are the ensemble of samples of the nonlinear term $\mathbf g(t_i,\mathbf x, \omega_j)$ with $i=1,\dots,m$ and $j=1,\dots,n$}.
-There is no additional cost for computing the nonlinear snapshots, since they are generated when computing the trajectory snapshot matrix $S$.
+{\edit To obtain the projection basis $U$, the POD can be applied to the ensemble of samples of the nonlinear term $\mathbf g(t_i,\mathbf x, \omega_j)$ with $i=1,\dots,m$ and $j=1,\dots,n$}.
+There is no additional cost {\edit associated with} computing the nonlinear snapshots, since they are generated when computing the trajectory snapshot matrix $S$.
 The interpolating indices $p_1,\dots,p_m$ can be constructed as follows.
@@
 It is shown in \cite{Chaturantabut:2010cz} that if the residual vector is a nonzero vector in each iteration then $P^TU$ is non-singular and (\ref{eq:MoOr:13}) is well defined.
-We summarize the algorithm of generating interpolating indices in Algorithm \ref{alg:MoOr:1}.
 
@@
 
-{\edit Numerical solution of (\ref{eq:MoOr:8}) sometimes involve computation of the Jacobian of the nonlinear function $\mathbf g(t,\mathbf x, \omega)$ with respect to the reduced state variable $\mathbf y$}
+{\edit The numerical solution of (\ref{eq:MoOr:8}) may involve the computation of the Jacobian of the nonlinear function $\mathbf g(t,\mathbf x, \omega)$ with respect to the reduced state variable $\mathbf y$}
 \begin{equation} \label{eq:MoOr:9}
-	\edit \mathbf J_{\mathbf x}(\mathbf g) = (WV)^{-1} \mathbf J_{\mathbf y}(\mathbf g) V,
+	\edit \mathbf J_{\mathbf y}(\mathbf g) = (WV)^{-1} \mathbf J_{\mathbf x}(\mathbf g) V,
 \end{equation}
-{\edit where $\mathbf J_\alpha(\mathbf g)$ is the Jacobian matrix of $\mathbf g$ with respect to the variable $\alpha$.} The complexity of computing (\ref{eq:MoOr:9}) is $\mathcal{O}(\alpha(n) +2n^2k+2nk^2+2nk)$ comprising several matrix-vector multiplications and an evaluation of the Jacobian which depends on the size of the original system.
+{\edit where $\mathbf J_\alpha(\mathbf g)$ is the Jacobian matrix of $\mathbf g$ with respect to the variable $\alpha$.} The complexity of (\ref{eq:MoOr:9}) is $\mathcal{O}(\alpha(n) +2n^2k+2nk^2+2nk)$, comprising several matrix-vector multiplications and an evaluation of the Jacobian which depends on the size of the original system.
 Approximating the Jacobian in (\ref{eq:MoOr:9}) is usually both problem and discretization dependent.
-Very often the nonlinear function $\mathbf g$ is evaluated component-wise i.e.
+Often the nonlinear function $\mathbf g$ is evaluated component-wise i.e.
 \begin{equation} \label{eq:MoOr:15}
@@
 \end{equation}
-with $\mathbf x$ the configuration vector. In such cases the interpolating index matrix $P$ and the nonlinear function $\mathbf g$ commute, i.e.,
+In such cases the interpolating index matrix $P$ and the nonlinear function $\mathbf g$ commute, i.e.,
 \begin{equation} \label{eq:MoOr:16}
@@
 \begin{equation}
-	\edit \mathbf J_{\mathbf x}(\mathbf g) = \underbrace{ (WV)^{-1} U(P^TU)^{-1} }_{k\times m} \underbrace{ \mathbf J_{\mathbf y}(\mathbf g(P^T V \mathbf y) ) }_{m\times m} \underbrace{P^T V}_{m\times k}.
+	\edit \mathbf J_{\mathbf y}(\mathbf g) = \underbrace{ (WV)^{-1} U(P^TU)^{-1} }_{k\times m} \underbrace{ \mathbf J_{\mathbf x}(\mathbf g(P^T V \mathbf y) ) }_{m\times m} \underbrace{P^T V}_{m\times k}.
 \end{equation}
-The matrix $(WV)^{-1} U(P^TU)^{-1}$ can be computed offline and the Jacobian is evaluated only at $m\times m$ components.
+The matrix $(WV)^{-1} U(P^TU)^{-1}$ can be computed offline and the Jacobian is evaluated only for $m\times m$ components.
 Hence the overall complexity of computing the Jacobian is now independent of $n$.

### 2016-12-09 greedy_hamiltonian_revised/chapters/3.ham.tex
Jan's comments included

@@
 \end{equation}
-{\edit where $i_{X_H} \Omega$ is the interior product of $X_H$ with $\Omega$, i.e. requiring}
+{\edit where $i_{X_H} \Omega$ is the interior product of $X_H$ with $\Omega$, i.e., that requiring}
 \begin{equation}
@@
 \end{equation}
-and known as the \emph{Hamilton's equation} \cite{Marsden:1999ck}.
+and known as \emph{Hamilton's equation} \cite{Marsden:1999ck}.
 A fundamental feature of Hamiltonian systems is the conservation of the Hamiltonian along integral curves on $P$.
-To emphasise the importance of this property we state
+To emphasize the importance of this property we recall {\edit \cite{Marsden:1999ck}}
 
 \begin{theorem} \label{theorem:Hasy:1}
-Suppose that $X_H$ is a Hamiltonian vector field with the flow $\phi_t$ on a symplectic manifold $\mathcal M$. Then $H\circ \phi_t = H$ {\edit \cite{Marsden:1999ck}}.
+Suppose that $X_H$ is a Hamiltonian vector field with the flow $\phi_t$ on a symplectic manifold $\mathcal M$. Then $H\circ \phi_t = H$.
 \end{theorem}
@@
 \begin{proof}
-We show that $H$ is constant along integral curves
+$H$ is constant along integral curves since
 \begin{equation} \label{eq:Hasy:3}
@@
 \end{equation}
-where we used the chain rule and bilinearity of $\Omega$ in the argument.
+{\edit by using} the chain rule and bilinearity of $\Omega$ in the argument.
 \end{proof}
@@
 For the case where the symplectic manifold is also a linear vector space, the pair $({\edit \mathcal M},\Omega)$ is also referred to as a \emph{symplectic vector space}.
-We state the following theorems regarding symplectic vector spaces without proof.
-We refer the reader to \cite{de2006symplectic,Marsden:1999ck,de2006symplectic,Silva01lectureson} for detailed proofs.
+We {\edit need} the following theorems regarding symplectic vector spaces and refer the reader to \cite{de2006symplectic,Marsden:1999ck,de2006symplectic,Silva01lectureson} for detailed proofs.
 
 {\edit
-\begin{theorem} \label{theorem:Hasy:1.1}
-If $V$ is a vector space then $\Omega$ is a constant form, that is $\Omega_z$ is independent of $z\in V$ \cite{Marsden:1999ck}. 
+\begin{theorem} \label{theorem:Hasy:1.1} \cite{Marsden:1999ck}
+If $V$ is a vector space then $\Omega$ is a constant form, that is $\Omega_z$ is independent of $z\in V$. 
 \end{theorem}
-\begin{theorem} \label{theorem:Hasy:1.2}
-If $(V,\Omega)$ is a finite-dimensional symplectic manifold then $V$ is even dimensional \cite{Marsden:1999ck}.
+\begin{theorem} \label{theorem:Hasy:1.2} \cite{Marsden:1999ck}
+If $(V,\Omega)$ is a finite-dimensional symplectic manifold then $V$ is even dimensional.
 \end{theorem}
-\begin{theorem} \label{theorem:Hasy:1.3}
+\begin{theorem} \label{theorem:Hasy:1.3} \cite{de2006symplectic}
 (The Symplectic Gram-Schmidt) If $(V,\Omega)$ is a $2n$-dimensional symplectic vector space, then there is a basis $e_1,\dots e_n,f_1, \dots , f_n$ of $V$ such that
@@
 \end{equation}
-Here $I_n$ and $0_n$ is the identity matrix and the zero square matrix of size $n$, respectively \cite{de2006symplectic}.
+Here $I_n$ and $0_n$ is the identity matrix and the zero square matrix of size $n$, respectively.
 \end{theorem}
-\begin{theorem} \label{theorem:Hasy:1.4}
-The classical inner product $\langle \cdot,\cdot \rangle:\mathbb R^{2n}\times \mathbb R^{2n}\to \mathbb R$ can be written in terms of the 2-form as \cite{Marsden:1999ck}
+\begin{theorem} \label{theorem:Hasy:1.4} \cite{Marsden:1999ck}
+The classical inner product $\langle \cdot,\cdot \rangle:\mathbb R^{2n}\times \mathbb R^{2n}\to \mathbb R$ can be written in terms of the 2-form as
 \begin{equation}
@@
 {\edit 
-\begin{definition}
+\begin{definition}  \cite{de2006symplectic}
 Suppose $(V,\Omega)$ is a finite dimensional symplectic vector space and $E\subset V$ is a subspace. Then the symplectic complement of $E$ inside $P$ is defined as
@@
 \end{definition}
-Note that $E \cap E^{\perp}$ is not empty in general \cite{de2006symplectic}. 
-\begin{definition}
+Note that $E \cap E^{\perp}$ is not empty in general. 
+\begin{definition} \cite{de2006symplectic}
 Suppose $(V,\Omega)$ is a finite dimensional symplectic vector space. A subspace $E\subset V$ is called a Lagrangian subspace inside $V$ if $E = E^\perp$.
 \end{definition}
-\begin{theorem} \label{theorem:Hasy:1.5}
-Suppose $(V,\Omega)$ is a finite dimensional symplectic vector space.
-If $E\subset V$ is a Lagrangian subspace then $dim(E)=\frac 1 2dim(V)$, where $dim$ denotes the dimension of the subspace \cite{abraham1978foundations}.
+\begin{theorem} \label{theorem:Hasy:1.5} \cite{abraham1978foundations}
+Suppose $(V,\Omega)$ is a finite dimensional symplectic vector space. If $E\subset V$ is a Lagrangian subspace then $dim(E)=\frac 1 2dim(V)$. Here $dim$ denotes the dimension of the subspace.
 \end{theorem}
@@
 \end{definition}
-\begin{theorem} \label{theorem:Hasy:1.6}
-Suppose $(V,\Omega)$ is a $2n$ dimensional symplectic vector space and $E\subset V$ is a Lagrangian subspace. Then there is an orthosymplectic basis for $V$ \cite{da2003introduction}.
+\begin{theorem} \label{theorem:Hasy:1.6}  \cite{da2003introduction}
+Suppose $(V,\Omega)$ is a $2n$ dimensional symplectic vector space and $E\subset V$ is a Lagrangian subspace. Then there is an orthosymplectic basis for $V$.
 \end{theorem}
@@
 Suppose that $\{ e'_1,\dots, e'_n \}$ is a basis for $E$, using the classical Gram-Schmidt orthogonalization process we can construct an orthonormal basis $\{ e_1,\dots,e_n \}$.
-Define the new set of vectors $f_1 = \mathbb J_{2n}^Te_1$, $f_2 =\mathbb J_{2n}^T e_2$, $\dots$, $f_n= \mathbb J_{2n}^Te_n$.
+Define a new set of vectors $f_1 = \mathbb J_{2n}^Te_1$, $f_2 =\mathbb J_{2n}^T e_2$, $\dots$, $f_n= \mathbb J_{2n}^Te_n$.
 We have
@@
 \end{proof}
-\begin{theorem} \label{theorem:Hasy:1.7}
-On a finite-dimensional symplectic vector space the relationship (\ref{eq:Hasy:1}) becomes \cite{Marsden:1999ck}
+\begin{theorem} \label{theorem:Hasy:1.7} \cite{Marsden:1999ck}
+On a finite-dimensional symplectic vector space the relationship (\ref{eq:Hasy:1}) becomes 
 \begin{equation} \label{eq:Hasy:7}
@@
 
-We now introduce \emph{symplectic transformations}, i.e.
-mappings between symplectic manifolds which preserve the two-form $\Omega$.
-Accurate numerical treatment of Hamiltonian systems often requires preservation of the symmetry expressed in Theorem \ref{theorem:Hasy:1}.
+{\edit Let} us now introduce \emph{symplectic transformations}, i.e., mappings between symplectic manifolds which preserve the two-form $\Omega$.
+The accurate numerical treatment of Hamiltonian systems often requires preservation of the symmetry expressed in Theorem \ref{theorem:Hasy:1}.
 Symplectic transformations can be used to construct such symmetry preserving numerical methods.
@@
 \begin{definition}
-Let $(V,\Omega)$ and $(W,\Pi)$ be two linear symplectic vector spaces of dimensions $2n$ and $2k$ respectively. A linear mapping $\phi:V \to W$ is called \emph{symplectic} or \emph{canonical} if
+Let $(V,\Omega)$ and $(W,\Pi)$ be two linear symplectic vector spaces of dimensions $2n$ and $2k$, respectively. A linear mapping $\phi:V \to W$ is called \emph{symplectic} or \emph{canonical} if
 \begin{equation} \label{eq:Hasy:9}
@@
 
-Note if we represent the transformation $\phi$ as a matrix $A\in \mathbb R^{2n\times 2k}$ the condition (\ref{eq:Hasy:9}) is equivalent to \cite{Marsden:1999ck}}
+Note that if we represent the transformation $\phi$ as a matrix $A\in \mathbb R^{2n\times 2k}$ condition (\ref{eq:Hasy:9}) is equivalent to \cite{Marsden:1999ck}}
 
@@
 
-{\edit Straight forward calculation verifies} that $AA^+$ is idempotent, i.e. a symplectic projection onto the column span of $A$.
+{\edit A straight-forward calculation verifies} that $AA^+$ is idempotent, i.e., a symplectic projection onto the column span of $A$.
 
 {\edit It is natural to expect a numerical integrator that solves (\ref{eq:Hasy:7}) to also satisfy the conservation law in Theorem \ref{theorem:Hasy:1}.
-Common numerical integrators e.g., Runge-Kutta methods, do not generally preserve the Hamiltonian which results in a wrong qualitative behavior in the solution \cite{Hairer:1250576}.
-Symplectic integrators are a class of numerical integrators for Hamiltonian systems that preserve the symplectic structure yielding stability in long-time integration.} The Str\"omer-Verlet time stepping scheme is an example of symplectic integrators and is given by
+Common numerical integrators e.g., Runge-Kutta methods, do not generally preserve the Hamiltonian which results in a qualitative wrong behavior of the solution \cite{Hairer:1250576}.
+Symplectic integrators are a class of numerical integrators for Hamiltonian systems that preserve the symplectic structure and ensure stability in long-time integration.} The Str\"omer-Verlet time stepping scheme is an example of symplectic integrators and is given by
 \begin{equation} \label{eq:Hasy:13}
@@
 \end{equation}
-For a general Hamiltonian system, Str\"omer-Verlet scheme is implicit.
+For a general Hamiltonian system, the Str\"omer-Verlet scheme is implicit.
 However, for separable Hamiltonians, i.e.

### 2016-12-09 greedy_hamiltonian_revised/chapters/4.sympmor.tex
Jan's comments included

@@
 \section{Symplectic Model Reduction} \label{chap:SyMo:1}
-We now propose how to modify reduced order modeling to ensure that they preserve the symplectic structure of Hamiltonian system.
+We now {\edit discuss} how to modify reduced order modeling to ensure that {\edit the resulting scheme preserves} the symplectic structure of the Hamiltonian system.
 
-Consider a Hamiltonian system (\ref{eq:Hasy:7}) on a $2n$-dimensional symplectic vector space $(P,\Omega)$.
-Suppose that the solution manifold $\mathcal M_H$ is well approximated, by a low dimensional symplectic subspace $(Q,\Omega)$ of dimension $2k$ $(k\ll n)$.
-We can construct a symplectic basis $A$ for $Q$ and approximate the solution to (\ref{eq:Hasy:7}) as
+Consider a Hamiltonian system (\ref{eq:Hasy:7}) on a $2n$-dimensional symplectic vector space $\edit (V,\Omega)$.
+Suppose that the solution manifold $\mathcal M_H$ is well approximated by a low dimensional symplectic subspace $\edit (W,\Omega)$ of dimension $2k$ $(k\ll n)$.
+We can {\edit then} construct a symplectic basis $A$ for $\edit W$ and approximate the solution to (\ref{eq:Hasy:7}) as
 \begin{equation} \label{eq:SyMo:1}
@@
 \end{equation}
-Since $A$ is a symplectic basis, Lemma \ref{lemma:Hasy:1} ensures that $(A^+)^T$ is a symplectic matrix i.e.
-$A^+ \mathbb J_{2n} (A^+)^T = \mathbb{J}_{2k}$.
+Since $A$ is a symplectic basis, Lemma \ref{lemma:Hasy:1} ensures that $(A^+)^T$ is a symplectic matrix i.e., $A^+ \mathbb J_{2n} (A^+)^T = \mathbb{J}_{2k}$.
 By defining the reduced Hamiltonian $\tilde H:\mathbb R^{2k} \to \mathbb R$ as $\tilde H (y) = H(Ay)$ we obtain the reduced system
@@
 \end{equation}
-The system obtained from the Petrov-Galerkin projection in (\ref{eq:MoOr:5}) is not a Hamiltonian system and therefore does not guarantee conservation of the symplectic structure.
-On the other hand we observe that the reduced system in (\ref{eq:SyMo:4}) is of the form (\ref{eq:Hasy:7}) and, hence, is a Hamiltonian system, i.e.
+The system obtained from the Petrov-Galerkin projection in (\ref{eq:MoOr:5}) is not a Hamiltonian system and does not guarantee conservation of the symplectic structure.
+On the other hand, we observe that the reduced system in (\ref{eq:SyMo:4}) is of the form (\ref{eq:Hasy:7}) and, hence, is a Hamiltonian system, i.e.
 the symplectic structure will be conserved along integral curves of (\ref{eq:SyMo:4}).
-Note that the original and the reduced systems are derived from different Hamiltonians.
+Note that the original and the reduced systems are {\edit endowed with} different Hamiltonians.
 In the next proposition we show that the error in the Hamiltonian is constant in time.
@@
 
-\begin{definition} \label{definition:SyMo:1}
+\begin{definition} \label{definition:SyMo:1} \cite{bhatia2002stability}
 Consider a dynamical system of the form $\dot{\mathbf z} = \mathbf f(\mathbf z)$ and suppose that $\mathbf z_e$ is an equilibrium point for the system so that $\mathbf f(\mathbf z_e) = 0$.
-$\mathbf z_e$ is called nonlinear stable or Lyapunov stable if for any $\epsilon > 0$, we can find $\delta > 0$ such that for any trajectory $\phi_t$, if $\| \phi_0 - \mathbf z_e \|_2 \leq \delta$, then for all $t\geq0$, we have $\| \phi_t - \mathbf z_e \|_2 < \epsilon$, where $\| \cdot \|_2$ is the Euclidean norm.
+$\mathbf z_e$ is called nonlinearly stable or Lyapunov stable if, for any $\epsilon > 0$, we can find $\delta > 0$ such that for any trajectory $\phi_t$, if $\| \phi_0 - \mathbf z_e \|_2 \leq \delta$, then for all $0 \leq t < \infty$, we have $\| \phi_t - \mathbf z_e \|_2 < \epsilon$, where $\| \cdot \|_2$ is the Euclidean norm.
 \end{definition}	
@@
 We refer the reader to \cite{bhatia2002stability} for the proof.
-\begin{proposition} \label{proposition:SyMo:1}
-An equilibrium point $\mathbf z_e$ is Lyapunov stable if there exist a scalar function $W : \mathbb R^{n} \to  \mathbb R$ such that $\nabla W(\mathbf z_e) = 0$, and $\nabla^2 W(\mathbf z_e)$ is positive definite and that for any trajectory $\phi_t$ defined in the neighborhood of $\mathbf z_e$, we have $\frac{d}{dt} W(\phi_t) \leq 0$.
+\begin{proposition} \label{proposition:SyMo:1} \cite{bhatia2002stability}
+An equilibrium point $\mathbf z_e$ is Lyapunov stable if there exists a scalar function $W : \mathbb R^{n} \to  \mathbb R$ such that $\nabla W(\mathbf z_e) = 0$, $\nabla^2 W(\mathbf z_e)$ is positive definite, and that for any trajectory $\phi_t$ defined in the neighborhood of $\mathbf z_e$, we have $\frac{d}{dt} W(\phi_t) \leq 0$.
 Here $\nabla^2W$ is the Hessian matrix of $W$.
@@
 In the context of the Hamiltonian systems, a suitable candidate for the Lyapunov function is the Hamiltonian function $H$.
-The following theorem shows that when $H$ (or $-H$) is a Lyapunov function, then the equilibrium points of the original and the reduced system are Liyapunov stable \cite{abraham1978foundations}.
+The following theorem shows that when $H$ (or $-H$) is a Lyapunov function, then the equilibrium points of the original and the reduced system are Lyapunov stable \cite{abraham1978foundations}.
 \begin{theorem} \label{theorem:SyMo:1}
@@
 Suppose $\mathbf z_e$ is an equilibrium point for (\ref{eq:Hasy:7}) and that $\mathbf y_e = A^+\mathbf z_e$.
-If $H$ (or $-H$) is a Lyapunov function satisfying proposition \ref{proposition:SyMo:1}, then $\mathbf z_e$ and $\mathbf y_e$ are Lyapunov stable equilibrium points for (\ref{eq:Hasy:7}) and (\ref{eq:SyMo:4}) respectively.
+If $H$ (or $-H$) is a Lyapunov function satisfying Proposition \ref{proposition:SyMo:1}, then $\mathbf z_e$ and $\mathbf y_e$ are Lyapunov stable equilibrium points for (\ref{eq:Hasy:7}) and (\ref{eq:SyMo:4}), respectively.
 \end{theorem}
 \begin{proof}
-	It is the direct consequence of proposition \ref{proposition:SyMo:1} that $\mathbf z_e$ is a local minimum or maximum of (\ref{eq:Hasy:7}) and also a Lyapunov stable point.
+	It is a direct consequence of Proposition \ref{proposition:SyMo:1} that $\mathbf z_e$ is a local minimum or maximum of (\ref{eq:Hasy:7}) and also a Lyapunov stable point.
 It can be easily checked that if $\mathbf z_e$ is a local minimum of $H$ then $\mathbf y_e$ is a local minimum for $\tilde H$ and an equilibrium point for (\ref{eq:SyMo:4}).
@@
 \[
-	\xi^T \nabla^2_{\mathbf y} \tilde H \xi = (A\xi)^T \nabla^2_{\mathbf z} H (A\xi) \geq 0,
+	\xi^T \nabla^2_{\mathbf y} \tilde H \xi = (A\xi)^T \nabla^2_{\mathbf z} H (A\xi) \geq 0.
 \]
-Where the last inequality is due to the positive definiteness of $H$.
+Here the last inequality is due to the positive definiteness of $H$.
 Therefore $\tilde H$ is also positive definite.
-By proposition \ref{proposition:SyMo:1} we conclude that $\mathbf y_e$ is a Lyapunov stable point.
+By Proposition \ref{proposition:SyMo:1} we conclude that $\mathbf y_e$ is a Lyapunov stable point.
 \end{proof}
@@
 Suppose that a symplectic basis $A$ of size $2n\times2k$ and its symplectic inverse $A^+$ is provided.
-{\edit The Proper Symplectic Decomposition} requires that the error of the symplectic projection onto the symplectic subspace to be minimum.
-Analogous to the minimization (\ref{eq:MoOr:6}), the PSD symplectic basis of size $2k$ is the solution of the optimization problem
+{\edit The Proper Symplectic Decomposition} requires that the error of the symplectic projection onto the symplectic subspace {\edit be minimized}.
+Hence, the PSD symplectic basis of size $2k$ is the solution to the optimization problem
 
@@
 At first, the minimization looks similar to the one obtained by POD.
-{\edit However, it is well known that symplectic bases are not generally orthogonal, and therefore not norm bounded, this means numerical errors can become dominant in symplectic projection \cite{Karow:2006cf}.
-This makes the minimization (\ref{eq:SyMo:8}) a harder problem than (\ref{eq:MoOr:6}).}
+{\edit However, it is well known that symplectic bases are not generally orthogonal, and therefore not norm bounded.
+This means that numerical errors may become dominant in the symplectic projection \cite{Karow:2006cf} which makes the minimization (\ref{eq:SyMo:8}) a harder problem than (\ref{eq:MoOr:6}).}
 
-As the optimization problem (\ref{eq:SyMo:8}) is nonlinear, direct solution usually expensive.
+As the optimization problem (\ref{eq:SyMo:8}) is nonlinear, the direct solution is usually expensive.
 A simplified version of the optimization (\ref{eq:SyMo:8}) can be found in \cite{Peng:2014di}, but there is no guarantee that the method provides a near optimal basis.
 
-{\edit Finding eigen-spaces of Hamiltonian and symplectic matrices is studied in the context of optimal control problems \cite{Benner:2000ww,Benner:1997ef,Watkins:2004kz,BunseGerstner:1986dg} and model reduction of Riccati equations \cite{Benner:1997ef}, where also an SVD-like decomposition for Hamiltonian and symplectic matrices is proposed by \cite{Xu:2003kx}.
-However, computation of a possibly large snapshot matrix and then using the mentioned methods to compute its eigen-spaces, is usually computationally demanding.
+{\edit Finding eigen-spaces of Hamiltonian and symplectic matrices is studied in the context of optimal control problems \cite{Benner:2000ww,Benner:1997ef,Watkins:2004kz,BunseGerstner:1986dg} and model reduction of Riccati equations \cite{Benner:1997ef}, where also an SVD-like decomposition for Hamiltonian and symplectic matrices has been proposed \cite{Xu:2003kx}.
+However, the computation of a large snapshot matrix and use of the mentioned methods to compute its eigen-spaces, is usually computationally demanding.
 Also, these methods generally do not guarantee the construction of a well-conditioned symplectic basis.
 
-The greedy approach presented in section \ref{Chap:Symo.PrSy:3}, is an iterative method for construction of a symplectic basis.
-It avoids evaluation of the full snapshot matrix so it saves substantial computational cost in the offline stage of the symplectic model reduction.
+The greedy approach presented in Section \ref{Chap:Symo.PrSy:3} is an iterative method for construction of a symplectic basis.
+It avoids the evaluation of the full snapshot matrix, hence substantially reduces the computational cost in the offline stage of the symplectic model reduction.
 Also, by construction, it yields a orthosymplectic basis and therefore a well-conditioned basis.
 
-In section \ref{chap:SyMo.PrSy:2} we outline non-direct methods for finding solutions to (\ref{eq:SyMo:8}), proposed by \cite{Peng:2014di}, while assuming a specific structure for $A$.
-Finally, in section \ref{Chap:Symo.PrSy:3} we introduce a greedy approach for symplectic basis generation.}
+In Section \ref{chap:SyMo.PrSy:2} we briefly outline non-direct methods for finding solutions to (\ref{eq:SyMo:8}), proposed by \cite{Peng:2014di}, and assuming a specific structure for $A$.
+In Section \ref{Chap:Symo.PrSy:3} we introduce a greedy approach for the symplectic basis generation.}
 
@@
 
-\paragraph{\bf Complex SVD} We now consider another form of symplectic basis $A$. Suppose that $A$ takes the form \cite{Peng:2014di}
+\paragraph{\bf Complex SVD} Suppose instead that $A$ takes the form \cite{Peng:2014di}
 \begin{equation} \label{eq:SyMo:11}
@@
 \end{equation}
-such that $\Phi$ and $\Psi$ are real matrices of the size $n\times k$ satisfying conditions
+while $\Phi$ and $\Psi$ are real matrices of size $n\times k$ satisfying conditions
 \begin{equation} \label{eq:SyMo:12}
@@
 \end{equation}
-The SVD decomposition of $S_{\text{complex}}$ can be written as $S_{\text{complex}} = U \Sigma V^H$ with the superscript $H$ indicating the conjugate transpose of a complex valued matrix.
-Each left singular vector of $S_{\text{complex}}$ takes the form $u_m = r_m + i s_m$.
-We define
+Each left singular vector of $S_{\text{complex}}$ now takes the form $u_m = r_m + i s_m$. We define
 \begin{equation} \label{eq:SyMo:14}
@@
 \end{equation}
-Then we can easily check that conditions (\ref{eq:SyMo:12}) is satisfied since $U$ is unitary.
+One can easily check that (\ref{eq:SyMo:12}) is satisfied {\edit since the matrix of singular vectors is unitary}.
 It is shown in \cite{Peng:2014di} that among all symplectic bases of the form (\ref{eq:SyMo:11}) the complex SVD minimizes the projection error.
 
-\subsubsection{The Greedy Approach to Symplectic Basis Generation} \label{Chap:Symo.PrSy:3} greedy generation of the reduced basis is an iterative procedure which, in each iteration, adds the two best possible basis vectors to the symplectic basis to enhance overall accuracy.
+\subsubsection{The Greedy Approach to Symplectic Basis Generation} \label{Chap:Symo.PrSy:3} Greedy generation of the reduced basis is an iterative procedure which, in each iteration, adds the two best possible basis vectors to the symplectic basis to enhance overall accuracy.
 In contrast to the cotangent lift and the complex SVD methods, the greedy approach does not require the symplectic basis to have a specific structure.
-This results in more compact bases and/or more accurate reduced systems.
-In addition, for parametric problems, the greedy approach only requires one numerical solution to be computed per iteration which saves substantial computational cost in the offline stage.
+This typically results in a more compact basis and/or more accurate reduced systems.
+For parametric problems, the greedy approach only requires one numerical solution to be computed per iteration hence saving substantial computational cost in the offline stage.
 
@@
 However common orthonormalization processes, e.g.
-the QR method based on the classical Gram-Schmidt method, destroy the symplectic structure of the original system \cite{BunseGerstner:1986dg}.
-Here we use a variation of the QR method known as the the SR \cite{Salam2014} method which is based on the symplectic Gram-Schmidt method and yields a symplectic basis.
+the QR method, destroy the symplectic structure of the original system \cite{BunseGerstner:1986dg}.
+Here we use a variation of the QR method known as the SR \cite{Salam2014} method which is based on the symplectic Gram-Schmidt method and yields a symplectic basis.
 }
@@
 As discussed in Section \ref{chap:Hasy:1}, any finite dimensional symplectic linear vector space has a symplectic basis that satisfies conditions (\ref{eq:Hasy:4}).
-Further, theorem \ref{theorem:Hasy:1.6} provides an iterative process for constructing an orthosymplectic basis \cite{Matsuo:2014wl,Salam2014}.
+Further, Theorem \ref{theorem:Hasy:1.6} provides an iterative process for constructing an orthosymplectic basis \cite{Matsuo:2014wl,Salam2014}.
 To briefly describe the SR method, suppose that an orthosymplectic basis
@@
 and a vector $z\not \in \text{span}(A_{2k})$ is provided.
-We would like to symplectically orthogonalize ($\mathbb J_{2n}$-orthogonalize) $z$ with respect to $A_{2k}$.
-We seek $\alpha_1,\dots,\alpha_k,\beta_1,\dots,\beta_k \in \mathbb R$ such that
+We aim to symplectically orthogonalize ($\mathbb J_{2n}$-orthogonalize) $z$ with respect to $A_{2k}$ and seek $\alpha_1,\dots,\alpha_k,\beta_1,\dots,\beta_k \in \mathbb R$ such that
 \begin{equation} \label{eq:SyMo:14.2}
@@
 \end{equation}
-for all possible $\bar{\alpha}_1,\dots,\bar{\alpha}_k,\bar{\beta}_1,\dots,\bar{\beta}_k \in \mathbb R$. It is easily seen that the unique solution is as 
+for all possible $\bar{\alpha}_1,\dots,\bar{\alpha}_k,\bar{\beta}_1,\dots,\bar{\beta}_k \in \mathbb R$. It is easily seen that the unique solution is 
 \begin{equation} \label{eq:SyMo:14.3}
@@
 \end{equation}
-for $i=1,\dots,k$. Now we define the modified vectors as
+for $i=1,\dots,k$. Now define the modified vectors as
 \begin{equation} \label{eq:SyMo:14.4}
@@
 If we introduce $e_{k+1} = \tilde z / \| \tilde z \|_2$, it is easily checked that $e_{k+1}$ is also orthogonal to $A_{2k}$ with respect to the classical inner product.
-Therefore the set span$(\{e_1,\dots,e_{k+1}\})$ forms a Lagrangian subspace and according to theorem \ref{theorem:Hasy:1.6} the basis $A_{2k+2}= A_{2k}\cup \{ e_{k+1} , \mathbb J_{2n}^T e_{k+1} \}$ forms an orthosymplectic basis.
+Therefore span$\{e_1,\dots,e_{k+1}\}$ forms a Lagrangian subspace and according to Theorem \ref{theorem:Hasy:1.6} the basis $A_{2k+2}= A_{2k}\cup \{ e_{k+1} , \mathbb J_{2n}^T e_{k+1} \}$ forms an orthosymplectic basis.
 
-Note that the $SR$ method can be replaced with a backward stable routines such as the isotropic Arnoldi or the isotropic Lanczos methods \cite{Mehrmann:2000dv}.
+Note that the $SR$ method can be replaced with backward stable routines such as the isotropic Arnoldi or the isotropic Lanczos methods \cite{Mehrmann:2000dv}.
 }
@@
 In the framework of symplectic model reduction, one possible candidate is the error in the Hamiltonian (\ref{eq:SyMo:5}).
-Correctly approximating symplectic systems relies on preservation of the Hamiltonian, hence the error in the Hamiltonian appears to be a natural choice.
-Moreover, since the error in the Hamiltonian depends on the initial condition and the reduced symplectic basis, evaluation of error does not require time integration of the full system.
+Correctly approximating symplectic systems relies on preservation of the Hamiltonian, hence the error in the Hamiltonian {\edit arises as a} a natural choice.
+Moreover, since the error in the Hamiltonian depends on the initial condition and the reduced symplectic basis, evaluation of the error does not require the time integration of the full system.
 
-Suppose that a $2k$ dimensional {\edit orthosymplectic basis (\ref{eq:SyMo:14.1})} is generated at the $k$-th step of the greedy method and we seek to enrich it by two additional vectors.
-Using the error in the Hamiltonian (\ref{eq:SyMo:7}) we search the parameter space to find the value that maximises the error in the Hamiltonian
+Suppose that a $2k$-dimensional {\edit orthosymplectic basis (\ref{eq:SyMo:14.1})} is generated at the $k$-th step of the greedy method and we seek to enrich it by two additional vectors.
+Using the error in the Hamiltonian (\ref{eq:SyMo:7}) we search the parameter space to identify the value that maximizes the error in the Hamiltonian
 \begin{equation} \label{eq:SyMo:14.5}
@@
 }
-where $\tilde z$ is the vector obtained from applying the symplectic Gram-Schmidt process to $z$.
+where $\tilde z$ is the vector obtained {\edit after} applying the symplectic Gram-Schmidt process to $z$.
 
 Since the maximization over the entire parameter space $\Gamma$ is impossible, we discretize the parameter set into a grid with $N$ points: $\Gamma_N = \{ \omega_1,\dots,\omega_N\}$.
-However, since the selection of parameters only require the evaluation of the error in the Hamiltonian and not time integration of the original system, then $\Gamma_N$ can be chosen very rich.
-Other sampling techniques can be found in \cite{Quarteroni:2016wi}.
+However, since the selection of parameters only require the evaluation of the error in the Hamiltonian and not time integration of the original system, then $\Gamma_N$ can be chosen {\edit to be} very rich.
 
-We summarized the greedy algorithm for generation of a symplectic basis in Algorithm \ref{alg:SyMo:3}.
+We summarize the greedy algorithm for the generation of a symplectic basis in Algorithm \ref{alg:SyMo:3}.
 
@@
 \item \hspace{0.5cm} Apply symplectic Gram-Schmidt on $\mathbf z^*$
-\item \hspace{0.5cm} $e_{k+1} \leftarrow \mathbf z^*/ \| z^*\|$
+\item \hspace{0.5cm} $e_{k+1} \leftarrow \mathbf z^*/ \| \mathbf  z^*\|$
 \item \hspace{0.5cm} $A \leftarrow [e_1,\dots ,e_{k+1} , \mathbb J^T_{2n}e_1,\dots,\mathbb J^T_{2n}e_{k+1}]$
@@
 
-The convergence and convergence rate of the greedy algorithm with respect to Kolmogorov $n$-width is investigated in Section \ref{chap:SyMo.PrSy:3}.
+%The convergence and convergence rate of the greedy algorithm with respect to Kolmogorov $n$-width is investigated in Section \ref{chap:SyMo.PrSy:3}.
 We finish this section by stating the advantage of using the greedy algorithm over the POD-based methods:
-\begin{itemize}
+%\begin{itemize}
 %\item Lack of a symplectic linear algebra toolbox for solving the minimization in (\ref{eq:SyMo:8}) requires a possibly very expensive nonlinear optimizations to recover near optimal symplectic bases.
@@
 The greedy algorithm on the other hand, is a cheap alternative to the nonlinear optimization and it allows flexibility in terms of the structure.
-\item The greedy method reduces the total cost of the offline stage by solving the original system (\ref{eq:Hasy:7}) only $k$ times to obtain a reduced system of size $2k$, unlike POD-based methods that require time integration of (\ref{eq:Hasy:7}) for the entire parameter space.
-\item If enrichment of the symplectic basis is required, the greedy algorithm can add new basis vectors. The POD-based methods, are bound to re-compute the basis for the refined parameter set.
-\item POD-based methods, e.g., the cotangent lift and the complex SVD, require performing SVD decompositions on possibly very large matrices. The greedy algorithm avoids this.
-\end{itemize}
+%\item The greedy method reduces the total cost of the offline stage by solving the original system (\ref{eq:Hasy:7}) only $k$ times to obtain a reduced system of size $2k$, unlike POD-based methods that require time integration of (\ref{eq:Hasy:7}) for the entire parameter space.
+%\item If enrichment of the symplectic basis is required, the greedy algorithm can add new basis vectors. The POD-based methods, are bound to re-compute the basis for the refined parameter set.
+%\item POD-based methods, e.g., the cotangent lift and the complex SVD, require performing SVD decompositions on possibly very large matrices. The greedy algorithm avoids this.
+%\end{itemize}
 
@@
 
-To show convergence of the greedy method we define a slightly different version based on the projection error.
-The error in the Hamiltonian is then introduced as a cheap surrogate to the projection error to accelerate parameter selection.
+To show convergence of the greedy method we {\edit consider} a slightly different version based on the projection error.
+The error in the Hamiltonian is then introduced as a cheap surrogate to the projection error to accelerate the parameter selection.
 
-Suppose that we are given with a compact subset $S$ of $\mathbb R^{2n}$.
-Our intention is to find a set of vectors $A=\{e_1,\dots,e_k,f_1,\dots,f_k\}$, such that $A$ forms a symplectic basis and any $s\in S$ is well approximated by elements of the subspace span$(A)$.
+Suppose that we are given a compact subset $S$ of $\mathbb R^{2n}$.
+Our intention is to find a set of vectors $A=\{e_1,\dots,e_k,f_1,\dots,f_k\}$ such that $A$ forms {\edit an orthosymplectic} basis and any $s\in S$ is well approximated by elements of the subspace span$(A)$.
 The modified greedy method for generating basis vectors $e_i$ and $f_i$ is as follows.
-In the initial step we pick $e_1$ such that $\Omega(e_1,-\mathbb{J}_{2n}^T e_1) = \|e_1\|_2 = \max_{s\in S} \|s\|_2$, where $\|\cdot\|_2$ is the Euclidean norm.
-Then define $f_1 = -\mathbb{J}_{2n}^T e_1$.
-It is easy to check that the span of $A_2 = \{e_1,f_1\}$ is symplectic, so $A_2$ is the first subspace that approximates elements of $S$.
+In the initial step we pick $e_1$ such that $\edit \|e_1\|_2 = \max_{s\in S} \|s\|_2$.
+Then define $f_1 = \mathbb{J}_{2n}^T e_1$.
+It is easy to check that the span of $A_2 = \{e_1,f_1\}$ is {\edit orthosymplectic}, so $A_2$ is the first subspace that approximates elements of $S$.
 In the $k$-th step of the greedy method, suppose we have a basis $A_{2k} = \{ e_1,\dots, e_k , f_1,\dots ,f_k \}$.
@@
 \begin{equation} \label{eq:new3}
-\begin{aligned}
-	& e_{k+1} := \underset{s\in S}{\text{argmax }}\sigma_{2k}(s), \\
-	& f_{k+1} := \mathbb{J}^T e_{k+1}.
-\end{aligned}
+	e_{k+1} := \underset{s\in S}{\text{argmax }}\sigma_{2k}(s), \quad f_{k+1} := \mathbb{J}^T e_{k+1}.
 \end{equation}
-We emphasise that the sequence of basis vectors generated by the greedy is generally not unique.
+We emphasisze that the sequence of basis vectors generated by the greedy is generally not unique.
 
-To estimate the quality of the reduced subspace, it is natural to compare it with the best possible $2k$-dimensional subspace in the sense of minimum projection (not necessary symplectic) error.
-For this we need to introduce the Kolmogorov $n$-width \cite{Kolmogoroff:1936fj,Pinkus:1985vy}.
+To estimate the quality of the reduced subspace, it is natural to compare it with the best possible $2k$-dimensional subspace in the sense of the minimum projection (not necessary symplectic) error.
+For this we introduce the Kolmogorov $n$-width \cite{Kolmogoroff:1936fj,Pinkus:1985vy}.
 
@@
 For a given subspace $Y_n$, the angle between $S$ and $Y_n$ measures the worst possible projection error of elements in $S$ onto $Y_n$.
-Hence the Kolmogorov $n$-width quantifies how well $S$ can be approximated by any $n$-dimensional subspace.
+Hence the Kolmogorov $n$-width quantifies how well $S$ can be approximated by {\edit an} $n$-dimensional subspace.
 
@@
 \end{equation}
-Now the projection of a vector $s\in S$ onto span$(A_{2k})$ can be written using the symplectic basis as
+The projection of a vector $s\in S$ onto span$(A_{2k})$ can be written using the symplectic basis as
 \begin{equation} \label{eq:new7}
@@
 \begin{aligned}
-	|\alpha_i(s)| = \frac{|\Omega(\bar{\xi}_i,s)|}{|\Omega(\xi_i,\bar{\xi}_i)|} &= \frac{|\Omega( \bar{\xi}_i, s - P_{2(k-1)}(s))|}{|\Omega(\xi_i,\bar{\xi}_i)|} \\
-	&\leq \frac{\|\bar{\xi}_i\|_2 \| s - P_{2(k-1)}(s) \|_2}{ \|\xi_i\|_2 \|\bar{\xi}_i\|_2 } \\
+	|\alpha_i(s)| = \frac{|\Omega(\bar{\xi}_i,s)|}{|\Omega(\xi_i,\bar{\xi}_i)|} = \frac{|\Omega( \bar{\xi}_i, s - P_{2(k-1)}(s))|}{|\Omega(\xi_i,\bar{\xi}_i)|}  &\leq \frac{\|\bar{\xi}_i\|_2 \| s - P_{2(k-1)}(s) \|_2}{ \|\xi_i\|_2 \|\bar{\xi}_i\|_2 } \\
 	&= \frac{\| s - P_{2(k-1)}(s) \|_2}{\| e_i - P_{2(k-1)}(e_i) \|_2} \leq 1.
@@
 \end{equation}
-Here, we use the fact that $|\Omega(\xi_i,\bar{\xi}_i)| = |\xi_i^T \mathbb{J}_{2n}\bar{\xi}_i| = \| \xi_i \|^2 = \|\bar{\xi}_i\|^2$ and the last inequality follows from the greedy algorithm which maximizes $e_i$.
+Here, we use the fact that $\edit |\Omega(\xi_i,\bar{\xi}_i)| = \| \xi_i \|^2_2 = \|\bar{\xi}_i\|^2_2$ with the last inequality following from the greedy algorithm which maximizes $e_i$.
 Similarly we deduce that $|\bar{\alpha}_i(s)|\leq 1$.
@@
 	&\mu_i^j = \sum_{l=i}^{j-1}\left( - \alpha_l(f_j) \mu_i^l  + \bar{\alpha}_l(f_j) \gamma_i^l \right), \quad\gamma_i^j = \sum_{l=i}^{j-1}\left( - \alpha_l(f_j) \gamma_i^l  + \bar{\alpha}_l(f_j) \mu_i^l \right), \\
-	&\lambda^j_i = - \gamma ^j_i, \quad \eta^j_i = \mu^j_i.
+	&\lambda^j_i = - \gamma ^j_i, \quad \eta^j_i = \mu^j_i,
 \end{aligned}
@@
 for $j = 1,\dots,k+1$.
-Note that since all $u_i$ and $v_i$ belong to $Y_{2k}$ so does their linear combination including all $\zeta_j$ and $\bar{\zeta}_j$.
+Note that since $u_i$ and $v_i$ belong to $Y_{2k}$ so does their linear combination including all $\zeta_j$ and $\bar{\zeta}_j$.
 We can use the inequality (\ref{eq:new12}) to write
@@
 \end{equation}
-Moreover since $Y_{2k}$ is of dimension $2k$ we can find $\kappa_i$, $i=1,\dots,2(k+1)$ such that
+Moreover since $Y_{2k}$ is of dimension $2k$ we find $\kappa_i$, $i=1,\dots,2(k+1)$ such that
 \begin{equation} \label{eq:new15}
-	\sum_{i=1}^{2(k+1)} \kappa_i^2 = 1,
-\end{equation}
-and
-\begin{equation} \label{eq:new16} 
-	\sum_{i=1}^{k+1} \kappa_i \zeta_i + \sum_{i=1}^{k+1} \kappa_{i+k+1} \bar{\zeta}_i = 0.
+	\sum_{i=1}^{2(k+1)} \kappa_i^2 = 1, \quad\sum_{i=1}^{k+1} \kappa_i \zeta_i + \sum_{i=1}^{k+1} \kappa_{i+k+1} \bar{\zeta}_i = 0.
 \end{equation}
@@
 \end{equation}
-Define $c = \kappa_j^{-1} \sum_{i=1,i\neq j}^{k+1} \kappa_i \xi_i + \kappa_j^{-1}\sum_{i=1}^{k+1} \kappa_{i+k+1} \bar{\xi}_i$.
-And note that $\mathbb{J}_{2n}^T c$ is symplectic to $\xi_j$.
-From this we deduce
+Define $c = \kappa_j^{-1} \sum_{i=1,i\neq j}^{k+1} \kappa_i \xi_i + \kappa_j^{-1}\sum_{i=1}^{k+1} \kappa_{i+k+1} \bar{\xi}_i$ and note that $\mathbb{J}_{2n}^T c$ is symplectic to $\xi_j$. We recover
 \begin{equation} \label{eq:new19}
 \begin{aligned}
-	\| \xi_j \|_2 &\leq \| \xi_j \|_2 + \| c \|_2 \\
-		&= \Omega(\xi_j,\mathbb{J}_{2n}^T \xi_j) + \Omega(c,\mathbb{J}_{2n}^T c) \\
+	\| \xi_j \|_2 &\leq \| \xi_j \|_2 + \| c \|_2 = \Omega(\xi_j,\mathbb{J}_{2n}^T \xi_j) + \Omega(c,\mathbb{J}_{2n}^T c) \\
 	       &= \Omega(\xi_j,\mathbb{J}_{2n}^T \xi_j) + \Omega(c,\mathbb{J}_{2n}^T c) + \Omega(\xi_j,\mathbb{J}_{2n}^T c) + \Omega(c,\mathbb{J}_{2n}^T \xi_j) \\
-       	&= \Omega(\xi_j + c, \mathbb{J}^T_{2n} (\xi_j + c)) \\
-	&= \| \xi_j + c \|_2	
+       	&= \Omega(\xi_j + c, \mathbb{J}^T_{2n} (\xi_j + c)) = \| \xi_j + c \|_2	
 \end{aligned}
 \end{equation}
-Combining this with inequality (\ref{eq:new18}) yields
+Combining this with (\ref{eq:new18}) yields
 \begin{equation} \label{eq:new20}
@@
 
-{\edit
-\begin{theorem} \label{theorem:SyMo:3}
-	The basis $A_{2k}$ generated by the greedy algorithm is orthosymplectic.
-\end{theorem}
-\begin{proof}
-	Direct result of theorem \ref{theorem:Hasy:1.6}.
-\end{proof}
-}
+%{\edit
+%\begin{theorem} \label{theorem:SyMo:3}
+%	The basis $A_{2k}$ generated by the greedy algorithm is orthosymplectic.
+%\end{theorem}
+%\begin{proof}
+%	Direct result of theorem \ref{theorem:Hasy:1.6}.
+%\end{proof}
+%}
 
@@
 \end{equation}
-One may argue that if DEIM's projection basis $V$ is rich enough then the approximation of the nonlinear term does not affect the symplectic structure significantly since $\mathbf g$ is already a symplectic function.
-Nonetheless, for a general choice of $V$ the system (\ref{eq:new25}) is not guaranteed to be a Hamiltonian system.
+For a general choice of $V$ the system (\ref{eq:new25}) is not guaranteed to be a Hamiltonian system, impacting long time accuracy and stability.
 However, we can guarantee that (\ref{eq:new25}) is a Hamiltonian system by choosing $V=(A^+)^T$.
-To understand why, we note that the system (\ref{eq:new25}) is a Hamiltonian system if and only if $\tilde N(\mathbf y) = \mathbb J_{2k} \nabla_{\mathbf y} \mathbf g(\mathbf y)$.
+To {\edit see this}, we note that the system (\ref{eq:new25}) is a Hamiltonian system if and only if $\tilde N(\mathbf y) = \mathbb J_{2k} \nabla_{\mathbf y} \mathbf g(\mathbf y)$.
 Also we have
@@
 \end{equation}
-using that $(A^+)^T$ is a symplectic matrix. Hence, $V = (A^+)^T$ is a sufficient condition for (\ref{eq:new25}) to be a Hamiltonian system.
+since $(A^+)^T$ is a symplectic matrix. Hence, $V = (A^+)^T$ is a sufficient condition for (\ref{eq:new25}) to {\edit be Hamiltonian}.
 
@@
 Let $S_{\mathbf g} = \{ \mathbf g (\mathbf x(t_i,\omega_j)) \}$ with $i = 1,\dots,M$ and $ j = 1 ,\dots,N$ be the nonlinear snapshots that were gathered in the greedy algorithm.
-We then form $(A^+)^T = \{ e'_1,\dots, e'_k,f'_1,\dots,f'_k\}$ and use a greedy approach for adding new basis vectors to $(A^+)^T$.
-At the $i$-th iteration of the symplectic DEIM, we use $(A^+)^T$ to approximate elements in $S_{\mathbf g}$ and choose the vector that maximizes error as the next basis vector.
+We then form $(A^+)^T = \{ e'_1,\dots, e'_k,f'_1,\dots,f'_k\}$ and use a greedy approach to add new basis vectors to $(A^+)^T$.
+At the $i$-th iteration of the symplectic DEIM, we use $(A^+)^T$ to approximate elements in $S_{\mathbf g}$ and choose the vector that maximizes the error as the next basis vector
 \begin{equation}
@@
 \end{equation}
-Finally when $(A^+)^T$ approximates elements $S_{\mathbf g}$ with the desired accuracy, we transpose and symplecticaly inverse $(A^+)^T$ to obtain $A$, yielding a Hamiltonian reduced system.
-Note that the above update only enriches the span of $A$, containing the basis we started with.
-We summarised the symplectic DEIM algorithm in Algorithm \ref{alg:SyMo:4}.
+Finally when $(A^+)^T$ approximates elements $S_{\mathbf g}$ with the desired accuracy, we transpose and symplectically invert $(A^+)^T$ to obtain $A$.
+We summarize the symplectic DEIM algorithm in Algorithm \ref{alg:SyMo:4}.
 
@@
 When using an implicit time integration scheme we face inefficiencies when evaluating the Jacobian of nonlinear terms, as discussed in Section \ref{chap:MoOr.DEIM:1}.
-We recall that the key to fast approximation of the Jacobian is that the interpolating index matrix $P$ obtained in the DEIM approximation, commutes with the nonlinear function.
+We recall that the key to fast approximation of the Jacobian is that the interpolating index matrix $P$, obtained in the DEIM approximation, commutes with the nonlinear function.
 Nonlinear terms in Hamiltonian systems often take the from

### 2016-12-09 greedy_hamiltonian_revised/chapters/5.numres.tex
Jan's comments included

@@
 \end{equation}
-Each $\omega_l$ is uniformly distributed in $[0,1]$ and $c\in \mathbb{R}$ is a constant number.
+{\edit Here $\omega_l \in [0,1]$ for $l=1,\dots,4$} and $c\in \mathbb{R}$ is a constant number.
 By rewriting (\ref{eq:NuRe:1}) in canonical form, using the change of variable $q = u$ and $\partial q/ \partial t= p$, we obtain the symplectic form
@@
 \end{equation}
-with the associated Hamiltonian given by
+with the associated Hamiltonian
 \begin{equation} \label{eq:NuRe:4}
@@
 \end{equation}
-We discretized the torus into $N$ equidistant points and define $\Delta x = L/N$, $x_i = i\Delta x$, $q_i=q(t,x_i,\omega)$ and $p_i=p(t,x_i,\omega)$ for $i = 1, \dots, N$.
+We discretize the torus into $N$ equidistant points and define $\Delta x = L/N$, $x_i = i\Delta x$, $q_i=q(t,x_i,\omega)$ and $p_i=p(t,x_i,\omega)$ for $i = 1, \dots, N$.
 Furthermore, we discretize (\ref{eq:NuRe:3}) using a standard central finite differences scheme to obtain
@@
 \end{equation}
-with $D_{xx}$ the central finite differences matrix operator. The discrete Hamiltonian can finally be written as
+where $D_{xx}$ is the central finite differences matrix operator. The discrete Hamiltonian can finally be written as
 \begin{equation} \label{eq:NuRe:7}
@@
 
-For numerical time integration we {\edit use} the Str\"omer-Verlet (\ref{eq:Hasy:13}) scheme, which becomes an explicit scheme since the Hamiltonian is separable for the linear wave-equation.
+For numerical time integration we {\edit use} the Str\"omer-Verlet (\ref{eq:Hasy:13}) scheme, {\edit which is explicit since} the Hamiltonian is separable for the linear wave-equation.
 The full model uses the following parameter set \vspace{0.5cm}
@@
 Time discretization size & $\Delta t = 0.01$ \\
+Wave speed & $c^2 = 0.1$ \\
 \hline
@@
 \vspace{0.5cm}
-
 We compare the reduced system obtained by the greedy algorithm with the methods based on SVD.
@@
 For the SVD based methods and POD, snapshots are gathered in the snapshot matrices $S$, $S_{\text{combined}}$ and $S_{\text{complex}}$, respectively, and the SVD is performed to construct the reduced basis.
-The greedy method is applied according to Algorithm \ref{alg:SyMo:3}; as input, the tolerance for the error in the Hamiltonian is set to $\delta = 5 \times 10^{-3}$.
+The greedy method is applied following Algorithm \ref{alg:SyMo:3}; as input, the tolerance for the error in the Hamiltonian is set to $\delta = 5 \times 10^{-3}$.
 All reduced systems are taken to have an identical size ($k=80$ for POD and $k=40$ for the symplectic methods).
-We use the Verlet-Str\"omer scheme for symplectic methods and a second order Runge-Kutta method for the POD.
-{\edit The choice of different time integration routines is due to the fact that POD destroys the canonical form of the original equations and a symplectic integrator cannot be applied.
-One can alternatively use separate reduced subspaces for the potential and the momentum spaces, which however is not a standard model approach and needs further analysis.} Finally we use transformation (\ref{eq:SyMo:1}) to transfer the solution of the reduced systems into the high-dimensional space for illustration purposes.
+We use the {\edit Str\"omer-Verlet} scheme for symplectic methods and a second order Runge-Kutta method for the POD.
+{\edit The choice of different time integration routines is due to the fact that the POD destroys the canonical form of the original equations and a symplectic integrator cannot be applied.
+One can alternatively use separate reduced subspaces for the potential and the momentum spaces, which however is not a standard model reduction approach and requires further analysis.} Finally we use transformation (\ref{eq:SyMo:1}) to transfer the solution of the reduced systems into the high-dimensional space for illustration purposes.
 
@@
 
-Figure \ref{fig:NuRe:1a} shows the solution of the linear wave equation for parameter value $c=0.1019$, chosen to be different from training parameters, at $t=0$, $t = 1$ and $t=2$.
+\begin{figure}
+
+\begin{minipage}{.5\linewidth}
+\centering
+\subfloat[$t=0$]{\label{fig:NuRe:1a}\includegraphics[width=1\textwidth]{./figs/wave/solution/solution_t0}}
+\end{minipage}%
+\begin{minipage}{.5\linewidth}
+\centering
+\subfloat[$t=1$]{\label{fig:NuRe:1b}\includegraphics[width=\textwidth]{./figs/wave/solution/solution_t1}}
+\end{minipage}\par\medskip
+\centering
+\subfloat[$t=2$]{\label{fig:NuRe:1c}\includegraphics[width=0.5\textwidth]{./figs/wave/solution/solution_t2}}
+\caption{The solution $q$ at $t=0$, $t=1$ and $t=2$ of the linear wave equation for parameter value $c= 0.1019$ different from training parameters.
+Here, the solution of the full system together with the solution of the POD, cotangent lift, complex SVD and the greedy reduced system is shown.}
+\label{fig:NuRe:1}
+\end{figure}
+
+\begin{figure}
+
+\begin{minipage}{.5\linewidth}
+\centering
+\subfloat[]{\label{fig:NuRe:2c}\includegraphics[width=\textwidth]{./figs/wave/error}}
+\end{minipage}%
+\begin{minipage}{.5\linewidth}
+\centering
+\subfloat[]{\label{fig:NuRe:2b}\includegraphics[width=\textwidth]{./figs/wave/hamiltonian}}
+\end{minipage}\par\medskip
+\centering
+%\subfloat[]{\label{fig:NuRe:2c}\includegraphics[width=0.5\textwidth]{./figs/wave/error}}
+
+\caption{{\edit (a) The $L^2$-error between the solution of the full system and the reduced system for different model reduction methods for $t \in [0,30]$.} (b) Plot of the Hamiltonian function for $t \in [0,30]$.
+}
+\label{fig:NuRe:2}
+\end{figure}
+
+Figure \ref{fig:NuRe:1a} shows the solution of the linear wave equation for parameter values {\edit $(\omega_1,\omega_2,\omega_3,\omega_4) = (0.8456,0.1320,0.9328,0.5809)$ or $\kappa(\omega) = 0.1019$}, chosen to be different from training parameters, at $t=0$, $t = 1$ and $t=2$.
 While we see instability and divergence from the exact solution for the POD reduced system, the symplectic methods provide a good approximation of the full model.
 
-{\edit The decay of the singular values for the POD ca be found in Figure \ref{fig:NuRe:5a}.
+{\edit The decay of the singular values for the POD are shown in Figure \ref{fig:NuRe:5a}.
 The decay} of the singular values suggests that a low dimensional solution manifold indeed exists.
-However since the linear subspace constructed by the POD is not symplectic, we observe the blow up of the Hamiltonian function in Figure \ref{fig:NuRe:2b} and the instability of the solution in Figure \ref{fig:NuRe:1}.
-The symplectic methods (using a reduced basis of the same size as POD) preserve the Hamiltonian function as is seen in Figure \ref{fig:NuRe:2b}.
+However, since the linear subspace, constructed by the POD, is not symplectic, we observe blow up of the Hamiltonian function in Figure \ref{fig:NuRe:2b} and the instability of the solution in Figure \ref{fig:NuRe:1}.
+The symplectic methods (using a reduced basis of the same size as POD) preserve the Hamiltonian function as shown in Figure \ref{fig:NuRe:2b}.
 
@@
 Furthermore, the symplectic methods provide a better approximation since the geometric structure of the original system is preserved.
-Although the greedy method is almost {\edit two times faster than the SVD-based methods} in the offline stage, its accuracy is comparable.
-{\edit For complex systems were the solution of the full system is expensive and for high dimensional parameter domains, POD-based methods become exceedingly expensive \cite{Anonymous:2016wl,Quarteroni:2016wi}.
-However the greedy method requires substantially fewer (proportional to the size of the reduced basis) evaluation of the time integration of the original system.
-Therefore we expect the greedy method to be superior in terms of computational cost.}
+Although the greedy method is almost {\edit twice faster than the SVD-based methods} in the offline stage, its accuracy is comparable.
+{\edit The cotangent lift method provides a more accurate solution, on the other hand the cotangent lift basis (\ref{eq:SyMo:9}) takes a less general form and usually computationally more demanding than the greedy method.}
+
+{\edit For complex systems were the solution of the full system is expensive and for high dimensional parameter domains, POD-based methods become impractical \cite{Anonymous:2016wl,Quarteroni:2016wi}.
+However, the greedy method requires substantially fewer (proportional to the size of the reduced basis) evaluation of the time integration of the original system.}
 
@@
 
-\begin{figure}
-
-\begin{minipage}{.5\linewidth}
-\centering
-\subfloat[$t=0$]{\label{fig:NuRe:1a}\includegraphics[width=1\textwidth]{./figs/wave/solution/solution_t0}}
-\end{minipage}%
-\begin{minipage}{.5\linewidth}
-\centering
-\subfloat[$t=1$]{\label{fig:NuRe:1b}\includegraphics[width=\textwidth]{./figs/wave/solution/solution_t1}}
-\end{minipage}\par\medskip
-\centering
-\subfloat[$t=2$]{\label{fig:NuRe:1c}\includegraphics[width=0.5\textwidth]{./figs/wave/solution/solution_t2}}
-\caption{The solution $q$ at $t=0$, $t=1$ and $t=2$ of the linear wave equation for parameter value $c= 0.1019$ different from training parameters.
-Here the solution of full system together with the solution of the POD, cotangent lift, complex SVD and the greedy reduced system is presented.}
-\label{fig:NuRe:1}
-\end{figure}
-
-\begin{figure}
-
-\begin{minipage}{.5\linewidth}
-\centering
-\subfloat[]{\label{fig:NuRe:2c}\includegraphics[width=\textwidth]{./figs/wave/error}}
-\end{minipage}%
-\begin{minipage}{.5\linewidth}
-\centering
-\subfloat[]{\label{fig:NuRe:2b}\includegraphics[width=\textwidth]{./figs/wave/hamiltonian}}
-\end{minipage}\par\medskip
-\centering
-%\subfloat[]{\label{fig:NuRe:2c}\includegraphics[width=0.5\textwidth]{./figs/wave/error}}
-
-\caption{{\edit (a) The $L^2$ error between the solution of the full system and the reduced system for different model reduction methods for $t \in [0,30]$.} (b) Plot of the Hamiltonian function for $t \in [0,30]$.
-}
-\label{fig:NuRe:2}
-\end{figure}
-
 \subsection{Nonlinear Schr\"odinger equation} \label{chap:NuRe:1.2} Let us consider the one-dimensional parametric Schr\"odinger equation
@@
 where $u$ is a complex valued wave function, $i$ is the imaginary unit, $|\cdot|$ is the modulus operator and $\epsilon$ is a parameter that belongs to the interval $\Gamma = [0.9,1.1]$.
-We consider periodic boundary conditions, i.e., $x$ belongs to a one dimensional torus of length $L$.
+We consider periodic boundary conditions, i.e., $x$ belongs to a one-dimensional torus of length $L$.
 We consider the initial condition
 \begin{equation} \label{eq:NuRe:11}
-	u_0(x) = \frac{\sqrt 2}{\cosh(x - x_0)} \exp(i\frac{c(x-x0)}{2}),
+	u_0(x) = \frac{\sqrt 2}{\cosh(x - x_0)} \exp(i\frac{c(x-x_0)}{2}),
 \end{equation}
@@
 \end{equation}
-
 We discretize the torus into $N$ equidistant points and take $\Delta x = L/N$, $x_i = i\Delta x$, $q_i=q(t,x_i,\epsilon)$ and $p_i = p(t,x_i,\omega)$ for $i = 1 ,\dots,N$.
@@
 		0_N & D_{xx}
-	\end{pmatrix}
+	\end{pmatrix}.
 \end{equation}
-with $D_{xx}$ being the central finite differences matrix operator for periodic boundary conditions. $\mathbf g$ is a vector valued nonlinear function defined as
+Here $\mathbf g$ is a vector valued nonlinear function defined as
 \begin{equation}  \label{eq:NuRe:16}
@@
 \begin{equation}  \label{eq:NuRe:17}
-	H_{\Delta x}(\mathbf z) = {\Delta x}\sum_{i=1}^{N} \left( \frac{q_i q_{i-1} - q_i^2}{\Delta x ^2} + \frac{p_i p_{i-1} - p_i^2}{\Delta x ^2} + \frac \epsilon 4 (p_i^2 + q_i^2)^2  \right).
+	H_{\Delta x}(\mathbf z) = {\Delta x}\sum_{i=1}^{N} \left( \frac{q_i q_{i-1} - q_i^2}{\Delta x ^2} + \frac{p_i p_{i-1} - p_i^2}{\Delta x ^2} + \frac \epsilon 4 (p_i^2 + q_i^2)^2  \right),
 \end{equation}
-We use a Str\"omer-Verlet (\ref{eq:Hasy:13}) scheme for time integration.
+and use a Str\"omer-Verlet (\ref{eq:Hasy:13}) scheme for time integration.
 Since the Hamiltonian function (\ref{eq:NuRe:17}) is non-separable, this scheme becomes implicit so in each time iteration, a system of nonlinear equations is solved using Newton's iteration.
@@
 \vspace{0.5cm}
-
 Regarding computation of the nonlinear terms of reduced systems, {\edit we} compare the DEIM with the symplectic DEIM.
@@
 
-We compared the reduced system obtained by the greedy algorithm with the cotangent lift, complex SVD, DEIM, symplectic DEIM and also POD.
-For the SVD-based methods, we discretize the parameter space $[0.9,1.1]$ into $M=500$ equidistant grid points to recover the discrete parameter space $\Gamma_M = \{\epsilon_1,\dots,\epsilon_M \}$, and gather trajectory snapshots for each $\epsilon_i$ for $i = 1,\dots,M$ in the snapshots matrix $S$.
-All reduced system are taken to have identical sizes ($k=90$ for symplectic methods and $k=180$ for the POD method).
-Following algorithm \ref{alg:SyMo:3} we construct the reduced system using the same discrete parameter space $\Gamma_M$.
+We compare the reduced system obtained using the greedy algorithm with the cotangent lift, complex SVD, DEIM, symplectic DEIM and also POD.
+For the SVD-based methods, we discretize the parameter space $[0.9,1.1]$ into $M=500$ equidistant grid points {\edit across} the discrete parameter space $\Gamma_M = \{\epsilon_1,\dots,\epsilon_M \}$, and gather trajectory snapshots for each $\epsilon_i$ for $i = 1,\dots,M$ in the snapshots matrix $S$.
+All reduced systems are taken to have identical sizes ($k=90$ for the symplectic methods and $k=180$ for the POD method).
+Following Algorithm \ref{alg:SyMo:3} we construct the reduced system using the same discrete parameter space $\Gamma_M$.
 The tolerance for the error in the Hamiltonian is set to $\delta = 10^{-3}$.
 Moreover, for DEIM and symplectic DEIM, we construct bases of size $k'=80$.
-Note that in the reduced system, generated in the symplectic DEIM, will be of size $k+k'=170$.
+Note that the reduced system, generated in the symplectic DEIM, will be of size $k+k'=170$.
 
 The cost of the offline stage is reduced to 20\% when using the greedy method for constructing a symplectic basis of size $k=90$, as compared to the SVD-based methods.
-The online stage, i.e., time integration for a new parameter in $\Gamma$, is generally more than 3 times faster than the original system.
-We point out that the efficiency of reduced systems are implementation and platform dependent.
+The online stage, i.e., time integration for a new parameter in $\Gamma$, is generally more than 3 times faster than {\edit for} the original system.
+We point out that the efficiency of reduced systems are implementation and platform dependent {\edit and we expect further reduction as the size of the problem increases.}
 
@@
 \caption{The solution $|u(t,x)| = \sqrt{q^2 + p^2}$ at $t=0$, $t=10$ and $t=20$ of the Nonlinear Schr\"odinger equation for parameter value $\epsilon = 1.0932$.
-Here the solution of {\edit the} full system together with the solution of the POD, cotangent lift, complex SVD and the greedy reduced system is presented.}
+Here the solution of {\edit the} full system, together with the solution of the POD, cotangent lift, complex SVD and the greedy reduced system, is shown.}
 \label{fig:NuRe:3}
@@
 Figure \ref{fig:NuRe:3} shows the solution of the Schr\"odinger equation for parameter value $\epsilon = 1.0932$ at $t=0$, $t=10$ and $t=20$.
-We first compare the reduced system obtained by the greedy algorithm with POD, cotangent lift and the complex SVD method.
+We first compare the reduced system obtained by the greedy algorithm with the POD, the cotangent lift, and the complex SVD method.
 The size of the reduced systems are taken identical for all methods ($k=180$ for POD and $k=90$ for the rest).
@@
 
-{\edit In Figure \ref{fig:NuRe:4b} we observe that the symplectic methods preserve the Hamiltonian function, unlike the POD and the DIEM methods.
-We would like to emphasise that using the reduced basis by the greedy and together with the DEIM (purple line) does not preserve the symplectic structure as suggested in this figure.}
+{\edit In Figure \ref{fig:NuRe:4b} we observe that the symplectic methods preserve the Hamiltonian function, unlike the POD and the DEIM methods.
+We emphasise that using the reduced basis, obtained by the greedy, together with the DEIM (purple line) does not preserve the symplectic structure as suggested in this figure.}
 
-Figure \ref{fig:NuRe:4c} illustrates the $L_2$-error between the solution of the full model with the reduced systems generated by different methods.
-We first observe that symplectic methods yield a lower computational error compared to non-symplectic methods.
+Figure \ref{fig:NuRe:4c} illustrates the $L^2$-error between the solution of the full model with the reduced systems, generated by different methods.
+We first observe that symplectic methods yield a lower computational error {\edit when} compared to non-symplectic methods.
 Secondly, we observe that although the reduced systems from the cotangent lift and the complex SVD are of the same size, their accuracy is different by an order of magnitude.
-We notice that the greedy algorithm is slightly less accurate than the cotangent lift method while its offline computational cost is less than 20\% {\edit compared to the cotangent lift}.
+We notice that the greedy algorithm is slightly less accurate than the cotangent lift method while its offline computational cost is reduced to 20\% {\edit when compared to the cotangent lift}.
 Lastly we notice that the combination of the greedy reduced basis and DEIM yields large errors in the solution while the solution using the symplectic DEIM is very accurate.
@@
 Theorem \ref{theorem:SyMo:2} suggests that the symplectic greedy method has similar properties.
-To illustrate this we compared the convergence of the conventional greedy with the convergence of the symplectic greedy method through the numerical simulations in sections \ref{chap:NuRe:1.1} and \ref{chap:NuRe:1.2}.
+To illustrate this we compare the convergence of the conventional greedy with the convergence of the symplectic greedy method through the numerical simulations in Sections \ref{chap:NuRe:1.1} and \ref{chap:NuRe:1.2}.
 
-The decay of the singular values of the snapshot matrix for the parametric wave equation and the nonlinear Schr\"odinger equation can be found in Figure \ref{fig:NuRe:5}.
+The decay of the singular values of the snapshot matrix for the parametric wave equation and the nonlinear Schr\"odinger equation are given in Figure \ref{fig:NuRe:5}.
 The decay rate of the singular values is a strong indicator for the decay rate of the Kolmogorov $n$-width of the solution manifold.
@@
 
-Figure \ref{fig:NuRe:5} shows the maximum $L_2$ error between the original system and the reduced system at each iteration of different greedy methods.
-In this figure we find the conventional greedy with orthogonal projection error as a basis selection criterion (orange), the symplectic greedy method with symplectic projection error as a basis selection criterion (green) and the symplectic greedy method with energy loss $\Delta H$ as a basis selection criterion (red).
-For simplicity, we may refer to these routines as the greedy with the orthogonal projection, the greedy with the symplectic projection and the greedy with the energy loss, respectively.
+Figure \ref{fig:NuRe:5} shows the maximum $L^2$ error between the original system and the reduced system at each iteration of different greedy methods.
+In this figure we find the conventional greedy with orthogonal projection error as a basis selection criterion (orange), the symplectic greedy method with a symplectic projection error as a basis selection criterion (green), and the symplectic greedy method with energy loss $\Delta H$ as a basis selection criterion (red).
 
@@
 This matches our expectation from Theorem \ref{theorem:SyMo:2}.
-We also notice that the greedy method with the energy loss provides an excellent error indication as a basis selection criterion.
+We also notice that the greedy method with the loss in Hamiltonian provides an excellent error indication as a basis selection criterion.
 }

### 2016-12-09 greedy_hamiltonian_revised/paper.tex
paper finished
Jan's comments included

@@
 \begin{abstract}
-While reduced-order models (ROMs) {\edit has been} popular for approximately solving large systems of differential equations, the stability of reduced models over long-time integration remains an open question.
-We present a greedy approach for ROM generation of parametric Hamiltonian systems which captures the symplectic structure of Hamiltonian systems to ensure stability of the reduced model.
-Through the greedy selection of basis vectors, two new vectors are added at each iteration to the set of basis vectors to increase the overall accuracy of the {\edit reduced} basis.
-We {\edit use} the error in the Hamiltonian function due to model reduction, as an error indicator to search the parameter space and find the next best basis vectors.
-We show that the greedy algorithm converges with exponential rate, under natural assumptions on the set of all solutions of the Hamiltonian system under variation of the parameters.
+While reduced-order models (ROMs) {\edit have been} popular for {\edit efficiently} solving large systems of differential equations, the stability of reduced models over long-time integration {\edit is of present challenges}.
+We present a greedy approach for ROM generation of parametric Hamiltonian systems {\edit that} captures the symplectic structure of Hamiltonian systems to ensure stability of the reduced model.
+Through the greedy selection of basis vectors, two new vectors are added at each iteration to the set of basis vectors to increase the accuracy of the {\edit reduced} basis.
+We {\edit use} the error in the Hamiltonian due to model reduction as an error indicator to search the parameter space and identify the next best basis vectors.
+Under natural assumptions on the set of all solutions of the Hamiltonian system under variation of the parameters, we show that the greedy algorithm converges with exponential rate.
 Moreover, we demonstrate that combining the greedy basis with the discrete empirical interpolation method also preserves the symplectic structure.
-This enables the reduction of computational cost for nonlinear Hamiltonian systems.
+This enables the reduction of the computational cost for nonlinear Hamiltonian systems.
 The efficiency, accuracy, and stability of this model reduction technique is illustrated through simulations of the parametric wave equation and the parametric Schr\"odinger equation.
@@
 
+\section*{Acknowledgments}
+{\edit We would like to thank the referees for providing us with very useful comments which served to improve the paper.}
+
 %\appendix

### 2016-12-09 response/response.tex
responed was revised

@@
 
-We revised major parts of the background and the introduction of the manuscript and corrected the mathematical notations for better comprehension.
+We revised major parts of the background and the introduction of the manuscript and corrected the mathematical notations.
 The authors believe that the introduction of the solution manifold, equation (2), and the reduced basis, equation (3), is essential in the understanding of model reduction techniques.
-The role of time was missing from the definition which is now added.
-However, the generation of the reduced basis, v\_i, is carefully and in much detail is discussed in chapters 3 and 4.
+The role of time was missing from the definition and is now added.
+However, the generation of the reduced basis, v\_i, is discussed in detail in chapters 3 and 4.
 
@@
 
-The explanation of the symplectic Gram-Schmidt routine (the part pointed by the reviewer) is revised for better understanding.
+The explanation of the symplectic Gram-Schmidt routine (the part pointed to by the reviewer) is revised for better understanding.
 
@@
 We agree with the reviewer that the numerical examples presented are relatively simple.
-However, alternative methods are either unstable or too expensive for even these simple test cases.
+However, alternative methods are either unstable or too expensive for even such simple test cases.
 Therefore, a more complex numerical experiment does not contribute to a better understanding of the method.
@@
 
-What about a comparison between the different computational cost of the algorithm? 
+What about a comparison between the different computational cost of the algorithm? \dots the gain of symplectic algorithms is unclear. 
 \ereview
@@
 To eliminate any confusion, we have added Figure 5 that presents the maximum error in time and parameter domain.
-Also the qualitative computation cost is discussed in Section 5.
+Also the gain in computation cost is discussed in Section 5.
 
 \breview
-A comparison with Monte Carlo algorithm is also missing. Moreover, t = 2 is not a long time for a 1D linear wave equation and the gain of symplectic algorithms is unclear. 
+5. A comparison with Monte Carlo algorithm is also missing.  
 \ereview
 
-The purpose of model reduction is to construct approximated solution with substantial lower cost.
-It is not clear to the authors, possibly due to a miss understanding, how Monte Carlo method enters this scope.
+The purpose of model reduction is to construct an approximated solution with a substantially lower cost for parametric PDEs.
+The goal here is to cheaply evaluate an accurate solution for specific parameter values.
+However, the Monte Carlo method finds the mean/average solution and does not enter this scope.
+Bad notations in Chapter 5 might have caused a misunderstanding that the numerical examples are stochastic PDEs (which then a comparison with the Monte Carlo method would be meaningful).
+To remove any confusion, we revised the notations in this chapter.
 
 \breview
-What about an illustration of the exponential convergence result given in Section 4? 
+6. What about an illustration of the exponential convergence result given in Section 4? 
 \ereview
 
-We agree with the reviewer that a better illustration of the exponential convergence is needed. We added section 5.3 to numerically discuss this matter.
+We agree with the reviewer that a better illustration of the exponential convergence is needed. We added section 5.3 to discuss this matter.
 
@@
 1.
-The topic of projection methods for Hamiltonian systems is a widely studied topic in the context of optimal control problems where almost all the concepts discussed in this paper have already been studied, symplectic QR, symplectic SVD, symplectic Krylov and Lanc- zos techniques, model reduction for Hamiltonian systems (Riccati equations) etc.
+The topic of projection methods for Hamiltonian systems is a widely studied topic in the context of optimal control problems where almost all the concepts discussed in this paper have already been studied, symplectic QR, symplectic SVD, symplectic Krylov and Lanczos techniques, model reduction for Hamiltonian systems (Riccati equations) etc.
 All these techniques are reinvented in this paper without reference to the previous work.
-This is unac- ceptable and in some sense even unethical, since any google search would have immediately revealed these papers by e.g.
+This is unacceptable and in some sense even unethical, since any google search would have immediately revealed these papers by e.g.
 Benner, Breiten, Bunse-Gerstner, Freund, Kressner, Mehrmann, Poloni, Saak, Schoeder, Simoncini, Watkins etc.
@@
 Most of the topics and references introduced above deal with eigenvalue problems of Hamiltonian, skew-Hamiltonian and symplectic matrices.
-The main focus of this manuscript however, is the model reduction of Hamiltonian systems specifically the greedy approach to structure preserving model reduction.
-This, although strongly related, is not reflected in the references suggested by the reviewer.
-Nonetheless we agree that including these topics in the paper is essential as they comprise a major part of the development of the field in optimal control.
+The main focus of this manuscript, however, is the model reduction of Hamiltonian systems and the development of a greedy approach to structure preserving model reduction.
+This, although related, is not reflected in the references suggested by the reviewer.
+Nonetheless we agree that including these topics in the paper is valuable as they comprise a major part of the development of the field in optimal control.
 Therefore, most of these references are included in the revision of the paper.
@@
 The orthonormality of the basis guarantees a numerically stable basis.
-Although this was briefly discussed in the paper, there was clearly a need for a detailed discussion.
+This was only briefly discussed in the paper and there was clearly a need for a detailed discussion.
 To this end, it is shown in the revision of the paper, as pointed by the reviewer, that the greedy method constructs a Lagrangian subspace in each iteration and then expands it to an orthosymplectic basis.
-We emphasised on this in the revision of the paper in form of Theorem 21.
+We emphasize this in the revision of the paper.
 
@@
 This matter is discussed in more detail in the revision of the paper.
-This together with the numerical experiments in the paper, although do not guarantee general stability, but are strong indication of the stability of the reduced system.
+This, together with the numerical experiments in the paper, do not guarantee general stability, but are strong indication of the stability of the reduced system.
 

### 2016-12-09 response/response.tex
minor errors fixed

@@
 
-Attached you will find the revised version of manuscript \#M106699, titled ''Structure Preserving Model Reduction of Parametric Hamiltonian Systems''.
+Attached you will find the revised version of manuscript \#M106699, titled ``Structure Preserving Model Reduction of Parametric Hamiltonian Systems".
 
@@
 We revised major parts of the background and the introduction of the manuscript and corrected the mathematical notations.
-The authors believe that the introduction of the solution manifold, equation (2), and the reduced basis, equation (3), is essential in the understanding of model reduction techniques.
+The authors believe that the introduction of the solution manifold, equation (2), and the reduced basis, equation (3), is essential to the understanding of model reduction techniques.
 The role of time was missing from the definition and is now added.
@@
 To eliminate any confusion, we have added Figure 5 that presents the maximum error in time and parameter domain.
-Also the gain in computation cost is discussed in Section 5.
+Also, the gain in computation cost is discussed in Section 5.
 
@@
 
-We agree with the reviewer that a better illustration of the exponential convergence is needed. We added section 5.3 to discuss this matter.
+We agree with the reviewer that a better illustration of the exponential convergence is needed. We added Section 5.3 to discuss this matter.
 
@@
 This, although related, is not reflected in the references suggested by the reviewer.
-Nonetheless we agree that including these topics in the paper is valuable as they comprise a major part of the development of the field in optimal control.
+Nonetheless,e we agree that including these topics in the paper is valuable as they comprise a major part of the development of the field in optimal control.
 Therefore, most of these references are included in the revision of the paper.

### 2016-12-09 response/response.tex
Minor Updates

@@
 \newcommand{\reviewbullet}[1]{\breview \begin{itemize}\item #1 \end{itemize}\ereview}
+
+\usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
+
 \begin{document}
@@
 
-We revised major parts of the background and the introduction of the manuscript and corrected the mathematical notations.
-The authors believe that the introduction of the solution manifold, equation (2), and the reduced basis, equation (3), is essential to the understanding of model reduction techniques.
-The role of time was missing from the definition and is now added.
+We revised major parts of the background and the introduction \todo{parts of the introduction and the background} of the manuscript and corrected the mathematical notations.
+The authors \todo{We believe that} believe that the introduction of the solution manifold, equation (2), and the reduced basis, equation (3), is\todo{are} essential to the understanding of \todo{the model reduction} model reduction techniques.
+
+\todo[inline]{We believe that the introduction of the solution manifold, reduced basis and equations (2) and (3) are essential}
+
+The role of time was missing from the definition and is now added\todo{The role of time is added to ...}.
 However, the generation of the reduced basis, v\_i, is discussed in detail in chapters 3 and 4.

### 2016-12-09 response/response.tex
final edits - done

--- 2016-12-09 response/response.tex
@@
 We revised major parts of the background and the introduction of the manuscript and corrected the mathematical notations.
-The authors believe that the introduction of the solution manifold, equation (2), and the reduced basis, equation (3), is essential to the understanding of model reduction techniques.
-The role of time was missing from the definition and is now added.
-However, the generation of the reduced basis, v\_i, is discussed in detail in chapters 3 and 4.
+We believe that the introduction of the solution manifold, equation (2), and the reduced basis, equation (3), is essential to the understanding of model reduction techniques.
+
+The role of time was missing from the definition and is now added. However, the generation of the reduced basis, v\_i, is discussed in detail in chapters 3 and 4.
 
@@
 
-The purpose of Figure 1 is to illustrate how the proposed method can preserve stability with the eyeball norm, and so using just a single realization.
+The purpose of Figure 1 is to illustrate how the proposed method can preserve stability with the eyeball norm, and so the plot shows the solution for a single parameter value.
 To eliminate any confusion, we have added Figure 5 that presents the maximum error in time and parameter domain.
@@
 This, although related, is not reflected in the references suggested by the reviewer.
-Nonetheless,e we agree that including these topics in the paper is valuable as they comprise a major part of the development of the field in optimal control.
+Nonetheless, we agree that including these topics in the paper is valuable as they comprise a major part of the development of the field in optimal control.
 Therefore, most of these references are included in the revision of the paper.

### 2016-12-09 greedy_hamiltonian_revised/chapters/1.introduction.tex
final edits - done

@@
 
-In the context of Hamiltonian and Lagrangian systems, recent work suggests modifications of POD to preserve {\edit some} geometric structure.
+In the context of Hamiltonian and Lagrangian systems, recent work suggests modifications of POD to preserve {\edit some} geometric structures.
 Lall et al.
@@
 In this paper, we present a greedy approach for the construction of a reduced system that preserves the geometric structure of Hamiltonian systems.
-This technique results in a reduced Hamiltonian system that mimics the symplectic properties of the original system and preserves the Hamiltonian structure and stability over the course of time.
+This technique results in a reduced Hamiltonian system that mimics the symplectic properties of the original system and preserves the Hamiltonian structure and its stability over the course of time.
 On the other hand, since time integration of the original system is only required once per iteration, the proposed method saves substantial computational cost during the offline stage when compared to alternative POD-like approaches.

### 2016-12-09 greedy_hamiltonian_revised/chapters/3.ham.tex
final edits - done

@@
 and known as \emph{Hamilton's equation} \cite{Marsden:1999ck}.
-A fundamental feature of Hamiltonian systems is the conservation of the Hamiltonian along integral curves on $P$.
+A fundamental feature of Hamiltonian systems is the conservation of the Hamiltonian along integral curves on $\edit \mathcal M$.
 To emphasize the importance of this property we recall {\edit \cite{Marsden:1999ck}}
@@
 \begin{theorem} \label{theorem:Hasy:1.1} \cite{Marsden:1999ck}
-If $V$ is a vector space then $\Omega$ is a constant form, that is $\Omega_z$ is independent of $z\in V$. 
+If $(V,\Omega)$ is a symplectic vector space then $\Omega$ is a constant form, that is $\Omega_z$ is independent of $z\in V$. 
 \end{theorem}
@@
 \begin{definition}  \cite{de2006symplectic}
-Suppose $(V,\Omega)$ is a finite dimensional symplectic vector space and $E\subset V$ is a subspace. Then the symplectic complement of $E$ inside $P$ is defined as
+Suppose $(V,\Omega)$ is a finite dimensional symplectic vector space and $E\subset V$ is a subspace. Then the symplectic complement of $E$ inside $V$ is defined as
 \[
@@
 
-{\edit Let} us now introduce \emph{symplectic transformations}, i.e., mappings between symplectic manifolds which preserve the two-form $\Omega$.
+{\edit Let} us now introduce \emph{symplectic transformations}, i.e., mappings between symplectic manifolds which preserve the 2-form $\Omega$.
 The accurate numerical treatment of Hamiltonian systems often requires preservation of the symmetry expressed in Theorem \ref{theorem:Hasy:1}.

### 2016-12-09 greedy_hamiltonian_revised/chapters/5.numres.tex
final edits - done

@@
 \end{equation}
-where $D_{xx}$ is the central finite differences matrix operator. The discrete Hamiltonian can finally be written as
+with $D_{xx}$ the central finite differences matrix operator. The discrete Hamiltonian can finally be written as
 \begin{equation} \label{eq:NuRe:7}
@@
 
-We compare the reduced system obtained using the greedy algorithm with the cotangent lift, complex SVD, DEIM, symplectic DEIM and also POD.
+We compare the reduced system obtained using the greedy algorithm with the cotangent lift, the complex SVD, DEIM, the symplectic DEIM and also the POD.
 For the SVD-based methods, we discretize the parameter space $[0.9,1.1]$ into $M=500$ equidistant grid points {\edit across} the discrete parameter space $\Gamma_M = \{\epsilon_1,\dots,\epsilon_M \}$, and gather trajectory snapshots for each $\epsilon_i$ for $i = 1,\dots,M$ in the snapshots matrix $S$.

### 2016-12-09 greedy_hamiltonian_revised/paper.tex
final edits - done

@@
 
-\newcommand{\edit}{\color{red}}
+\newcommand{\edit}{\color{black}}
 
@@
 We present a greedy approach for ROM generation of parametric Hamiltonian systems {\edit that} captures the symplectic structure of Hamiltonian systems to ensure stability of the reduced model.
-Through the greedy selection of basis vectors, two new vectors are added at each iteration to the set of basis vectors to increase the accuracy of the {\edit reduced} basis.
+Through the greedy selection of basis vectors, {\edit two new vectors are added at each iteration to the linear vector space} to increase the accuracy of the {\edit reduced} basis.
 We {\edit use} the error in the Hamiltonian due to model reduction as an error indicator to search the parameter space and identify the next best basis vectors.

### 2017-02-09 dissipative_hamiltonian/chapters/2.Hamil.tex
dissipative hamiltonians added

\section{Dissipative Hamiltonian Systems} \label{sec:2}
In this section we first discuss Hamiltonian systems. Add more later

\subsection{Hamiltonian Systems} Suppose that $(\mathbb{R}^{2n},\Omega)$ is a symplectic linear vector space, where $\mathbb{R}^{2n}$ is the configuration space and $\Omega:\mathbb{R}^{2n}\times \mathbb{R}^{2n} \to \mathbb R$ is a closed, skew-symmetric and non-degenerate 2-form on $\mathbb{R}^{2n}$. Given a smooth Hamiltonian function $H:\mathbb{R}^{2n}\to \mathbb R$, \emph{Hamilton}'s equations of evolution are given as
\begin{equation} \label{eq:2.1}
	\dot {\mathbf x} = \mathbb J_{2n} \nabla_{\mathbf x} H,
\end{equation}
where $\mathbf x\in\mathbb R^{2n}$ is the configuration coordinates and $\mathbb J$ is a $2n\times 2n$ matrix such that $\Omega(\mathbf x,\mathbf y) = \mathbf x^T \mathbb J_{2n} \mathbf y$ \cite{Marsden:2010:IMS:1965128}. One by using the \emph{Symplectic Gram-Schmidt} \cite{de2006symplectic} can construct a coordinate system in which $\mathbb J_{2n}$ takes the form
\begin{equation} \label{eq:2.2}
	\mathbb{J}_{2n} = 
	\begin{pmatrix}
		0_n & I_n \\
		-I_n & 0_n
	\end{pmatrix},
\end{equation}
where $I_n$ and $0_n$ are the identity matrix and the zero matrix of size $n$ respectively. A main feature of Hamiltonian systems is the conservation of the Hamiltonian along the integral curves.
\begin{theorem} \label{theorem:2.1}
\cite{Marsden:2010:IMS:1965128} Consider the flow $\phi_t:\mathbb R\times \mathbb R^{2n} \to \mathbb R^{2n}$ of the Hamiltonian system (\ref{eq:2.1}). Then $H\circ \phi_t = H$.
\end{theorem}

In many physical problems $H$ represents the system energy and is bounded below. Here, we mostly assume that $H$ takes the form $H(\mathbf x) = \frac 1 2 \mathbf x^T K^T K \mathbf x$, where $K$ is a full rank $2n\times 2n$ matrix. This assumption leads to a linear system of evolution (\ref{eq:2.1}). Note that the Hamiltonian extension in section \ref{} and the model reduction for dissipative Hamiltonian systems in section \ref{} can be naturally extended to Hamiltonians of the form $H(\mathbf x) = \frac 1 2 \mathbf x^T K^T K \mathbf x + g(\mathbf x)$, where $g:\mathbb R^{2n} \to \mathbb R$ is arbitrary function of $\mathbf x$.

It is natural to expect a numerical integrator that solves (\ref{eq:2.1}) to also satisfy the conservation law \ref{theorem:2.1}. Common numerical integrators, e.g. the Runge-Kutta methos, do not generally preserve the Hamiltonian. Symplectic numerical integrators are a class of numerical integrators for Hamiltonian systems that preserves the symplectic structure and ensure stability in long-time integration. The Str\:omer-Verlet time stepping scheme is an example of such numerical integrators and is given by

\begin{equation} \label{eq:2.2}
\begin{aligned}
	p_{n+1/2} &= p_n - \frac{\Delta t}{2} \nabla_qH(q_{n},p_{n+1/2}), \\
	q_{n+1} &= q_n + \frac{\Delta t}{2} \left( \nabla_pH(q_{n},p_{n+1/2}) + \nabla_pH(q_{n+1},p_{n+1/2}) \right),\\
	p_{n+1} &= p_{n+1/2} - \frac{\Delta t}{2} \nabla_qH(q_{n+1},p_{n+1/2}).
\end{aligned}
\end{equation}
More information on construction and application of symplectic integrators can be found in \cite{Hairer:1250576}.

\subsection{Dissipative Hamiltonian Systems and Hamiltonian Extensions}

### 2017-02-09 dissipative_hamiltonian/paper.tex
dissipative hamiltonians added

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}

%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%
\begin{document}

\title{Structure-Preserving Model-Reduction of Dissipative Hamiltonian Systems%\thanks{Grants or other notes
%about the article that should go on the front page should be
%placed here. General acknowledgments should be placed at the end of the article.}
}
%\subtitle{Do you have a subtitle?\\ If so, write it here}

%\titlerunning{Short form of title}        % if too long for running head

\author{Babak Maboudi Afkham         \and
        Jan S. Hesthaven %etc.
}

%\authorrunning{Short form of author list} % if too long for running head

\institute{Babak Maboudi Afkham \at
              EPFL SB MATH MCSS, MA C1 645 (B\^atiment MA), Station 8, CH-1015 Lausanne, Switzerland \\
              Tel.: +41-21-6935905 \\
              \email{babak.maboudi@epfl.ch}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
           \and
           Jan S. Hesthaven \at
           EPFL SB MATH MCSS, MA C2 652 (B\^atiment MA), Station 8, CH-1015 Lausanne, Switzerland
}

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

\begin{abstract}
Insert your abstract here. Include keywords, PACS and mathematical
subject classification numbers as needed.
\keywords{First keyword \and Second keyword \and More}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\input{./chapters/1.intro.tex}
\input{./chapters/2.Hamil.tex}

%\begin{acknowledgements}
%If you'd like to thank anyone, place your comments here
%and remove the percent signs.
%\end{acknowledgements}

% BibTeX users please use one of
%\bibliographystyle{spbasic}      % basic style, author-year citations
%\bibliographystyle{spmpsci}      % mathematics and physical sciences
%\bibliographystyle{spphys}       % APS-like style for physics
%\bibliography{}   % name your BibTeX data base

% Non-BibTeX users please use
%\begin{thebibliography}{}
%%
%% and use \bibitem to create references. Consult the Instructions
%% for authors for reference list style.
%%
%\bibitem{RefJ}
%% Format for Journal Reference
%Author, Article title, Journal, Volume, page numbers (year)
%% Format for books
%\bibitem{RefB}
%Author, Book title, page numbers. Publisher, place (year)
%% etc
%\end{thebibliography}

\bibliographystyle{siamplain}
\bibliography{ref}

\end{document}
% end of file template.tex

### 2017-02-21 dissipative_hamiltonian/chapters/2.Hamil.tex
added some material fo dissipative hamiltonian systems

@@
 
-\subsection{Hamiltonian Systems} Suppose that $(\mathbb{R}^{2n},\Omega)$ is a symplectic linear vector space, where $\mathbb{R}^{2n}$ is the configuration space and $\Omega:\mathbb{R}^{2n}\times \mathbb{R}^{2n} \to \mathbb R$ is a closed, skew-symmetric and non-degenerate 2-form on $\mathbb{R}^{2n}$.
+\subsection{Hamiltonian Systems} Suppose that $(\mathbb{R}^{2n},\Omega)$ is a symplectic linear vector space, where $\mathbb{R}^{2n}$ is a configuration space and $\Omega:\mathbb{R}^{2n}\times \mathbb{R}^{2n} \to \mathbb R$ is a closed, skew-symmetric and non-degenerate 2-form on $\mathbb{R}^{2n}$.
 Given a smooth Hamiltonian function $H:\mathbb{R}^{2n}\to \mathbb R$, \emph{Hamilton}'s equations of evolution are given as
 \begin{equation} \label{eq:2.1}
-	\dot {\mathbf x} = \mathbb J_{2n} \nabla_{\mathbf x} H,
+	\begin{aligned}
+	\dot {\mathbf x}(t) &= \mathbb J_{2n} \nabla_{\mathbf x} H, \\
+	\mathbf x(0) &= \mathbf x_0
+	\end{aligned}
 \end{equation}
-where $\mathbf x\in\mathbb R^{2n}$ is the configuration coordinates and $\mathbb J$ is a $2n\times 2n$ matrix such that $\Omega(\mathbf x,\mathbf y) = \mathbf x^T \mathbb J_{2n} \mathbf y$ \cite{Marsden:2010:IMS:1965128}.
+where $\mathbf x\in\mathbb R^{2n}$ is the configuration coordinates and $\mathbb J_{2n}$ is a $2n\times 2n$ matrix such that $\Omega(\mathbf x,\mathbf y) = \mathbf x^T \mathbb J_{2n} \mathbf y$ \cite{Marsden:2010:IMS:1965128}.
 One by using the \emph{Symplectic Gram-Schmidt} \cite{de2006symplectic} can construct a coordinate system in which $\mathbb J_{2n}$ takes the form
@@
 In many physical problems $H$ represents the system energy and is bounded below.
-Here, we mostly assume that $H$ takes the form $H(\mathbf x) = \frac 1 2 \mathbf x^T K^T K \mathbf x$, where $K$ is a full rank $2n\times 2n$ matrix.
+Here, we mostly assume that $H$ is a quadratic Hamiltonian, i.e., it takes the form $H(\mathbf x) = \frac 1 2 \mathbf x^T K^T K \mathbf x$, where $K$ is a full rank $2n\times 2n$ matrix.
 This assumption leads to a linear system of evolution (\ref{eq:2.1}).
-Note that the Hamiltonian extension in section \ref{} and the model reduction for dissipative Hamiltonian systems in section \ref{} can be naturally extended to Hamiltonians of the form $H(\mathbf x) = \frac 1 2 \mathbf x^T K^T K \mathbf x + g(\mathbf x)$, where $g:\mathbb R^{2n} \to \mathbb R$ is arbitrary function of $\mathbf x$.
+Note that the Hamiltonian extension in section \ref{} and the model reduction for dissipative Hamiltonian systems in section \ref{} can be naturally extended to Hamiltonians of the form $H(\mathbf x) = \frac 1 2 \mathbf x^T K^T K \mathbf x + g(\mathbf x)$, where $g:\mathbb R^{2n} \to \mathbb R$ is an arbitrary function of $\mathbf x$.
 
@@
 
-\begin{equation} \label{eq:2.2}
+\begin{equation} \label{eq:2.3}
 \begin{aligned}
@@
 \subsection{Dissipative Hamiltonian Systems and Hamiltonian Extensions}
+
+Many systems in engineering and physic appear as a perturbation of a Hamiltonian system, where the perturbation is regarded as dissipation.
+In these systems, the energy tends to decrease over time evolution, and the conservation law in Theorem \ref{theorem:2.1} does not hold anymore.
+Therefore, it is common to take the conservation of energy as a fundamental principle and consider the dissipative system coupled with a heat bath that absorbs the dissipated energy of the original system.
+
+To account dissipation for a quadratic Hamiltonian $H(\mathbf x) = \frac 1 2 \mathbf x^T K^T K \mathbf x$, we rewrite (\ref{eq:2.1}) as a time dispersive and dissipative (TDD) \cite{Figotin:2006jy} system
+\begin{equation} \label{eq:2.4}
+	\begin{aligned}
+		& \dot {\mathbf x} = \mathbb J_{2n} K^T f(t), \\
+		& f(t) + \int_0^t \chi(t-s) \cdot f(s)\ ds = K\mathbf x,
+	\end{aligned}
+\end{equation}
+with the initial condition $\mathbf x(0) = \mathbf x_0$.
+Here $\chi:\mathbb R^+\to \mathbb R^{2n\times 2n}$ is a bounded and symmetric matrix valued function with respect to the Frobenius norm and is called the \emph{general susceptibility}.
+Note that integral term in (\ref{eq:2.3}) corresponds to the dissipation, whereas if $\chi(s) = 0$ then (\ref{eq:2.3}) is equivalent to (\ref{eq:2.1}).
+Note that under suitable assumptions on $K$, both the systems (\ref{eq:2.1}) and (\ref{eq:2.3}) are well-posed \cite{Figotin:2006jy}.
+
+\begin{example}
+Consider the dynamics of the damped harmonic oscillator
+\begin{equation} \label{eq:2.5}
+	\ddot q + r \dot q + k q = 0
+\end{equation}
+where $k$ is the Hooke's constant and $r$ is the spring's damping factor.
+Note that without a damping term, (\ref{eq:2.5}) is a Hamiltonian system.
+The TDD formulation for the damped harmonic oscillator takes the form
+\begin{equation}
+	\begin{aligned}
+		& \dot q(t) = f(t), \\
+		& \dot p(t) = - k q(t), \\
+		& f(t) + \int_0^t r f(s) \ ds = p(t).
+	\end{aligned}
+\end{equation}
+\end{example}
+Here the susceptibility is the constant function $r$.

### 2017-02-22 dissipative_hamiltonian/chapters/2.Hamil.tex
Hamiltonian extension in progress

@@
 	\begin{aligned}
-	\dot {\mathbf x}(t) &= \mathbb J_{2n} \nabla_{\mathbf x} H, \\
-	\mathbf x(0) &= \mathbf x_0
+	\dot {z}(t) &= \mathbb J_{2n} \nabla_{z} H, \\
+	z(0) &= z_0,
 	\end{aligned}
 \end{equation}
-where $\mathbf x\in\mathbb R^{2n}$ is the configuration coordinates and $\mathbb J_{2n}$ is a $2n\times 2n$ matrix such that $\Omega(\mathbf x,\mathbf y) = \mathbf x^T \mathbb J_{2n} \mathbf y$ \cite{Marsden:2010:IMS:1965128}.
+where $z \in\mathbb R^{2n}$ is the configuration coordinates and $\mathbb J_{2n}$ is a $2n\times 2n$ matrix such that $\Omega(x,y) = x^T \mathbb J_{2n} y$ \cite{Marsden:2010:IMS:1965128}.
 One by using the \emph{Symplectic Gram-Schmidt} \cite{de2006symplectic} can construct a coordinate system in which $\mathbb J_{2n}$ takes the form
@@
 In many physical problems $H$ represents the system energy and is bounded below.
-Here, we mostly assume that $H$ is a quadratic Hamiltonian, i.e., it takes the form $H(\mathbf x) = \frac 1 2 \mathbf x^T K^T K \mathbf x$, where $K$ is a full rank $2n\times 2n$ matrix.
+Here, we mostly assume that $H$ is a quadratic Hamiltonian, i.e., it takes the form $H(z) = \frac 1 2 z^T K^T K z$, where $K$ is a full rank $2n\times 2n$ matrix.
 This assumption leads to a linear system of evolution (\ref{eq:2.1}).
-Note that the Hamiltonian extension in section \ref{} and the model reduction for dissipative Hamiltonian systems in section \ref{} can be naturally extended to Hamiltonians of the form $H(\mathbf x) = \frac 1 2 \mathbf x^T K^T K \mathbf x + g(\mathbf x)$, where $g:\mathbb R^{2n} \to \mathbb R$ is an arbitrary function of $\mathbf x$.
+Note that the Hamiltonian extension in section \ref{} and the model reduction for dissipative Hamiltonian systems in section \ref{} can be naturally extended to Hamiltonians of the form $H(z) = \frac 1 2 z^T K^T K z + g(z)$, where $g:\mathbb R^{2n} \to \mathbb R$ is an arbitrary function of $z$.
 
@@
 	q_{n+1} &= q_n + \frac{\Delta t}{2} \left( \nabla_pH(q_{n},p_{n+1/2}) + \nabla_pH(q_{n+1},p_{n+1/2}) \right),\\
-	p_{n+1} &= p_{n+1/2} - \frac{\Delta t}{2} \nabla_qH(q_{n+1},p_{n+1/2}).
+	p_{n+1} &= p_{n+1/2} - \frac{\Delta t}{2} \nabla_qH(q_{n+1},p_{n+1/2}),
 \end{aligned}
 \end{equation}
-More information on construction and application of symplectic integrators can be found in \cite{Hairer:1250576}.
+where $p$ and $q$ are the canonical coordinates. More information on construction and application of symplectic integrators can be found in \cite{Hairer:1250576}.
 
@@
 
-To account dissipation for a quadratic Hamiltonian $H(\mathbf x) = \frac 1 2 \mathbf x^T K^T K \mathbf x$, we rewrite (\ref{eq:2.1}) as a time dispersive and dissipative (TDD) \cite{Figotin:2006jy} system
+To account dissipation for a quadratic Hamiltonian $H(z) = \frac 1 2 z^T K^T K  z$, we rewrite (\ref{eq:2.1}) as a time dispersive and dissipative (TDD) \cite{Figotin:2006jy} system 
 \begin{equation} \label{eq:2.4}
 	\begin{aligned}
-		& \dot {\mathbf x} = \mathbb J_{2n} K^T f(t), \\
-		& f(t) + \int_0^t \chi(t-s) \cdot f(s)\ ds = K\mathbf x,
+		& \dot {z} = \mathbb J_{2n} K^T f(t), \\
+		& z(0) = z_0,
 	\end{aligned}
 \end{equation}
-with the initial condition $\mathbf x(0) = \mathbf x_0$.
+where $f$ is the solution to the integral equation
+\begin{equation} \label{eq:2.4.1}
+	f(t) + \int_0^t \chi(t-s) \cdot f(s)\ ds = K z.
+\end{equation}
 Here $\chi:\mathbb R^+\to \mathbb R^{2n\times 2n}$ is a bounded and symmetric matrix valued function with respect to the Frobenius norm and is called the \emph{general susceptibility}.
-Note that integral term in (\ref{eq:2.3}) corresponds to the dissipation, whereas if $\chi(s) = 0$ then (\ref{eq:2.3}) is equivalent to (\ref{eq:2.1}).
-Note that under suitable assumptions on $K$, both the systems (\ref{eq:2.1}) and (\ref{eq:2.3}) are well-posed \cite{Figotin:2006jy}.
+Note that integral term in (\ref{eq:2.4.1}) corresponds to the dissipation, whereas if $\chi(s) = 0$ then (\ref{eq:2.3}) is equivalent to (\ref{eq:2.1}).
+Note that under suitable assumptions on $K$, both the systems (\ref{eq:2.1}) and (\ref{eq:2.4}) are well-posed \cite{Figotin:2006jy}.
 
@@
 The TDD formulation for the damped harmonic oscillator takes the form
-\begin{equation}
-	\begin{aligned}
-		& \dot q(t) = f(t), \\
-		& \dot p(t) = - k q(t), \\
-		& f(t) + \int_0^t r f(s) \ ds = p(t).
-	\end{aligned}
+\begin{equation} \label{eq:2.6}
+	\dot q(t) = f(t), \quad \dot p(t) = - k q(t), \quad f(t) + \int_0^t r f(s) \ ds = p(t).
 \end{equation}
+Here $(q,p)$ are the canonical coordinates and the susceptibility is the constant function $r$.
 \end{example}
-Here the susceptibility is the constant function $r$.
+
+It is shown in \cite{Figotin:2006jy,Figotin:2005} that under natural assumptions on a linear susceptibility $\chi(t)$ (see below), one can couple a TDD system of the form (\ref{eq:2.4}) with a canonical heat bath where the dissipated energy is captured in the heat bath in a canonical sense.
+In other words, one can construct a Hilbert space $\mathcal H$ and an isometric injection $I:\mathbb R^{2n} \to \mathbb R^{2n}\times \mathcal H$ where the solution $z$ to (\ref{eq:2.4}) is the projection of $x$ onto $\mathbb R^{2n}$, where $x$ is the solution to
+\begin{equation} \label{eq:2.7}
+	\frac{d}{dt} x = J \frac{\delta h}{\delta x}.
+\end{equation}
+Here $J$ is the symplastic operator and $h:\mathbb R^{2n}\times \mathcal H \to \mathbb R$ is an extended quadratic Hamiltonian function defined on $\mathbb R^{2n}\times \mathcal H$.
+
+\begin{theorem}
+Suppose that $K$ is full ranked and $\chi(t)$ is symmetric. There exists a quadratic extension of the form (\ref{eq:2.8}) to (\ref{eq:2.4}), if
+\begin{equation} \label{eq:2.8}
+	\text{Im}(\xi\hat{\chi}(\xi)) \geq 0, \quad \forall \xi = \omega + i\eta, \ \eta \geq 0,
+\end{equation}
+where $\hat{\chi}$ is the Fourier-Laplace transform of $\chi$
+\begin{equation} \label{eq:2.9}
+	\hat{\chi}(\xi) = \int_0^\infty e^{i\xi t} \chi(t)\ dt.
+\end{equation}
+\end{theorem}
+
+\begin{proof}
+Here we prove the theorem for the case where $\chi$ is a constant symmetric matrix, where condition (\ref{eq:2.8}) corresponds to $\chi$ being positive semi-definite.
+We refer the reader to \cite{Figotin:2006jy} for the proof of the general case.
+Consider the Hamiltonian system
+\begin{subequations}
+\begin{align}
+		\label{eq:2.10.a} & \dot{z}(t) = \mathbb J_{2n} K^T f(t), \\
+		\label{eq:2.10.b} & \partial_t \theta(t,x) = \partial_x^2 \phi(t,x) + \sqrt 2 \delta_0(x) \cdot \sqrt{\chi}  f(t), \\
+		\label{eq:2.10.c} & \partial_t \phi(t,x) = \theta(t,x),
+\end{align}
+\end{subequations}
+together with the initial condition
+\begin{equation} \label{eq:2.10.1}
+	z(0) = z_0,\quad \theta(0,\cdot) = 0, \quad \phi(0,\cdot) = 0.
+\end{equation}
+Here $\theta$ and $\phi$ are vector valued functions with $2n$ entries, $\delta_0(s)$ is the Dirac's delta function, $\sqrt{ \chi}$ is the matrix square root of $\chi$ and $f$ is the solution to the equation
+\begin{equation} \label{eq:2.11}
+	f(t) + \sqrt{2} \cdot \sqrt{ \chi } \phi(t,0) = Kz(t).
+\end{equation}
+To show that the Hamiltonian system (\ref{eq:2.10.a}) to (\ref{eq:2.10.c}) is an extension to (\ref{eq:2.4}) in the sense discussed before, it is enough to show that the solution $f$ to the equation (\ref{eq:2.11}) also satisfies (\ref{eq:2.4.1}).
+Equations (\ref{eq:2.10.b}) and (\ref{eq:2.10.c}) is a forced wave, with the analytical solution
+\begin{equation} \label{eq:2.12}
+	\phi(t,x) = \frac {\sqrt 2} 2 \int_0^{t-|x|} \sqrt{\chi} f(s)\ ds,\quad \theta(t,x) = \frac{\sqrt 2}{2} \cdot \sqrt{\chi} f(t - |x|).
+\end{equation}
+\end{proof}
+We can recover (\ref{eq:2.4.1}) by substituting (\ref{eq:2.12}) into (\ref{eq:2.11}). {!!! mention Hamiltonian of the extended Hamiltonian system !!!}

### 2017-02-28 dissipative_hamiltonian/chapters/2.Hamil.tex
model reduction of dissipative Hamiltonians started

@@
 This assumption leads to a linear system of evolution (\ref{eq:2.1}).
-Note that the Hamiltonian extension in section \ref{} and the model reduction for dissipative Hamiltonian systems in section \ref{} can be naturally extended to Hamiltonians of the form $H(z) = \frac 1 2 z^T K^T K z + g(z)$, where $g:\mathbb R^{2n} \to \mathbb R$ is an arbitrary function of $z$.
+Note that the Hamiltonian extension in section ??
+and the model reduction for dissipative Hamiltonian systems in section ??
+can be naturally extended to Hamiltonians of the form $H(z) = \frac 1 2 z^T K^T K z + g(z)$, where $g:\mathbb R^{2n} \to \mathbb R$ is an arbitrary function of $z$.
 
-It is natural to expect a numerical integrator that solves (\ref{eq:2.1}) to also satisfy the conservation law \ref{theorem:2.1}.
+Under general coordinate changes, equations of evolution do not necessarily take the form of (\ref{eq:2.1}).
+Let $(\mathbb R^{2n},\Omega)$ and $(\mathbb R^{2k},\Lambda)$ be two symplectic linear vector spaces.
+A linear transformation $\alpha :\mathbb R^{2n} \to \mathbb R^{2k}$ is called a \emph{symplectic transformation} \cite{Marsden:2010:IMS:1965128} if
+\begin{equation}
+	\Omega(x,y) = \Lambda(\alpha(x),\alpha(y)), \quad \text{for all } x,y\in \mathbb R^{2n}.
+\end{equation}
+Alternatively in matrix notation, $A\in \mathbb R^{2n\times 2k}$ is called a \emph{symplectic matrix} if
+\begin{equation}
+	A^T \mathbb{J}_{2n} A = \mathbb{J}_{2k},
+\end{equation}
+where the superscript $T$ represents transpose operator.
+As the symplectic 2-form is preserved under symplectic transformations, the form of the equations of evolution remains invariant \cite{Marsden:2010:IMS:1965128}.
+The \emph{symplectic inverse} of $A$, is a form of pseudo-inverse given by
+\begin{equation}
+	A^+ = \mathbb{\mathbb J}_{2k}^T A^T \mathbb J_{2n}.
+\end{equation}
+It is shown in \cite{Peng:2014di} that $A^+A = I_{2k}$ and that $(A^+)^T$ is a symplectic matrix.
+Further, one can easily check that $AA^+$ is idempotent and so is a projection operator onto the column span of $A$.
+
+It is natural to expect a numerical integrator that solves (\ref{eq:2.1}) to also satisfy the conservation law in theorem \ref{theorem:2.1}.
 Common numerical integrators, e.g.
@@
 Symplectic numerical integrators are a class of numerical integrators for Hamiltonian systems that preserves the symplectic structure and ensure stability in long-time integration.
-The Str\:omer-Verlet time stepping scheme is an example of such numerical integrators and is given by
+The Str\"omer-Verlet time stepping scheme is an example of such numerical integrators and is given by
 
@@
 \end{equation}
-Here $\chi:\mathbb R^+\to \mathbb R^{2n\times 2n}$ is a bounded and symmetric matrix valued function with respect to the Frobenius norm and is called the \emph{general susceptibility}.
+Here $\chi:\mathbb R^+\to \mathbb R^{2n\times 2n}$ is a bounded matrix valued function with respect to the Frobenius norm and is called the \emph{general susceptibility}.
 Note that integral term in (\ref{eq:2.4.1}) corresponds to the dissipation, whereas if $\chi(s) = 0$ then (\ref{eq:2.3}) is equivalent to (\ref{eq:2.1}).
@@
 
-\begin{example}
+\begin{example} \label{example:2.1}
 Consider the dynamics of the damped harmonic oscillator
@@
 \end{proof}
-We can recover (\ref{eq:2.4.1}) by substituting (\ref{eq:2.12}) into (\ref{eq:2.11}). {!!! mention Hamiltonian of the extended Hamiltonian system !!!}
+We can recover (\ref{eq:2.4.1}) by substituting (\ref{eq:2.12}) into (\ref{eq:2.11}). The extended Hamiltonian for the system (\ref{eq:2.10.a}) to (\ref{eq:2.10.c}) takes the quadratic from
+\begin{equation}
+	\tilde h(z,\phi,\theta) = \frac 1 2 \left( \| Kz - \phi(t,0) \|_2^2 + \| \theta(t) \|^2_{\mathcal H } + \| \partial_x\phi(t)\|^2_{\mathcal H }\right)
+\end{equation}
+where $\| \cdot \|_2$ is the Euclidean norm on $\mathbb R^{2n}$ and $\| \cdot \|_{\mathcal H}$ is the induced norm from the inner product on $\mathcal H$.
+
+Equations (\ref{eq:2.10.a}) and (\ref{eq:2.10.c}) are called the \emph{hidden strings}.
+The dissipation of energy in the original system (\ref{eq:2.4}) is carried away, as vibrations, along the added strings which makes the extended system conservative.
+Hamiltonian extension of the damped harmonic oscillator in Example \ref{example:2.1} is exactly the Lamb model \cite{lamb:1900}.
+The Lamb model is a harmonic oscillator coupled with a vibrating string, and the tension in the string causes linear dissipation in the dynamics of the harmonic oscillator.

### 2017-03-06 dissipative_hamiltonian/chapters/2.Hamil.tex
MOR of dissipative Hamiltonian systems is updated

@@
 It is shown in \cite{Figotin:2006jy,Figotin:2005} that under natural assumptions on a linear susceptibility $\chi(t)$ (see below), one can couple a TDD system of the form (\ref{eq:2.4}) with a canonical heat bath where the dissipated energy is captured in the heat bath in a canonical sense.
-In other words, one can construct a Hilbert space $\mathcal H$ and an isometric injection $I:\mathbb R^{2n} \to \mathbb R^{2n}\times \mathcal H$ where the solution $z$ to (\ref{eq:2.4}) is the projection of $x$ onto $\mathbb R^{2n}$, where $x$ is the solution to
+In other words, one can construct a Hilbert space $\mathcal H$ and an isometric injection $I:\mathbb R^{2n} \to \mathbb R^{2n}\times \mathcal H^{2n}$ where the solution $z$ to (\ref{eq:2.4}) is the projection of $x$ onto $\mathbb R^{2n}$, where $x$ is the solution to
 \begin{equation} \label{eq:2.7}
-	\frac{d}{dt} x = J \frac{\delta h}{\delta x}.
+	\frac{d}{dt} x = J_{2n} \frac{\delta h}{\delta x}.
 \end{equation}
-Here $J$ is the symplastic operator and $h:\mathbb R^{2n}\times \mathcal H \to \mathbb R$ is an extended quadratic Hamiltonian function defined on $\mathbb R^{2n}\times \mathcal H$.
+Here $J_2n$ is the symplastic operator and $h:\mathbb R^{2n}\times \mathcal H^{2n} \to \mathbb R$ is an extended quadratic Hamiltonian function defined on $\mathbb R^{2n}\times \mathcal H^{2n}$.
 
@@
 \end{equation}
-Here $\theta$ and $\phi$ are vector valued functions with $2n$ entries, $\delta_0(s)$ is the Dirac's delta function, $\sqrt{ \chi}$ is the matrix square root of $\chi$ and $f$ is the solution to the equation
+Here $\theta$ and $\phi$ are vector valued functions in $\mathcal H^{2n}$, $\delta_0(s)$ is the Dirac's delta function, $\sqrt{ \chi}$ is the matrix square root of $\chi$ and $f$ is the solution to the equation
 \begin{equation} \label{eq:2.11}
@@
 \begin{equation}
-	\tilde h(z,\phi,\theta) = \frac 1 2 \left( \| Kz - \phi(t,0) \|_2^2 + \| \theta(t) \|^2_{\mathcal H } + \| \partial_x\phi(t)\|^2_{\mathcal H }\right)
+	\tilde h(z,\phi,\theta) = \frac 1 2 \left( \| Kz - \phi(t,0) \|_2^2 + \| \theta(t) \|^2_{\mathcal H^{2n} } + \| \partial_x\phi(t)\|^2_{\mathcal H^{2n} }\right)
 \end{equation}
-where $\| \cdot \|_2$ is the Euclidean norm on $\mathbb R^{2n}$ and $\| \cdot \|_{\mathcal H}$ is the induced norm from the inner product on $\mathcal H$.
+where $\| \cdot \|_2$ is the Euclidean norm on $\mathbb R^{2n}$ and $\| \cdot \|_{\mathcal H^{2n}}$ is the induced norm from the inner product on $\mathcal H^{2n}$.
 

### 2017-03-06 dissipative_hamiltonian/chapters/3.model.tex
dissipative Hamiltonians finally added

\section{Model Order Reduction}
In this section we first explain the main results of \cite{Peng:2014di} regarding model reduction of Hamiltonian systems. In section ?? we introduce the model reduction of dissipative Hamiltonian systems.

\subsection{Symplectic Model Order Reduction}
Consider a Hamiltonian system of the form (\ref{eq:2.1}) together with a quadratic Hamiltonian of the form $H(z) = \frac 1 2 z^T K^T K z$. In this paper we focus on reducing the complexity of numerical evaluation of (\ref{eq:2.1}) with respect to time $t$. Nevertheless, one can naturally extend the results of this paper to a Hamiltonian system that depends on a set of physical or geometrical parameters belonging to a compact subset of a Euclidean space.

The main idea behind model order reduction is that the manifold, $\mathcal M = \{ z(t) : t \in [0,T] \}$ can be approximated by a low dimensional symplectic linear subspace \cite{hesthaven2015certified,quarteroni2015reduced}. A basis for such a subspace is called a \emph{reduced basis}, and its span is referred to as the \emph{reduced space} \cite{hesthaven2015certified}.

Suppose that a reduced symplectic basis $A \in \mathbb R^{2n\times 2k}$ is provided with $k \ll n$. The approximated solution to (\ref{eq:2.1}) in this basis is given by
\begin{equation} \label{eq:3.1}
	z =Ay,
\end{equation}
where $y\in \mathbb R^{2k}$ is the coordinates of the approximation with respect to the basis $A$. Substituting (\ref{eq:3.1}) into (\ref{eq:2.1}) yields
\begin{equation} \label{eq:3.2}
	A \dot y = \mathbb{J}_{2n} \nabla_{z} H(A y).
\end{equation}
Multiplying both sides with the symplectic inverse of $A$ and using the chain rule we obtain
\begin{equation} \label{eq:3.3}
	\dot y = A^+ \mathbb{J}_{2n} (A^+)^T \nabla_{y} H(A y).
\end{equation}
As $(A^+)^T$ is a symplectic matrix it yields that $A^+ \mathbb{J}_{2n} (A^+)^T = \mathbb{J}_{2k}$. By defining the reduced Hamiltonian $\tilde H : \mathbb{R}^{2k} \to \mathbb R$, defined as $\tilde H (y) = H(Ay)$, we obtain the reduced system
\begin{equation} \label{eq:3.4}
	\begin{aligned}
	\dot {y}(t) &= \mathbb J_{2k} \nabla_{y} \tilde H, \\
	y(0) &= A^+ z_0,
	\end{aligned}
\end{equation}
Equation (\ref{eq:3.4}) is called the \emph{symplectic Galerkin projection} \cite{Peng:2014di} of the Hamiltonian system (\ref{eq:2.1}). Conventional model reduction routines, e.g. Galerkin or Petrov-Galerkin methods \cite{hesthaven2015certified,quarteroni2015reduced}, do not yield a Hamiltonian reduced system. Hence, the reduced system does not necessarily preserve the conservation law in theorem \ref{theorem:2.1}. This results in a qualitatively wrong and often unstable solution for the reduced system {\color{red} put reference here}. On the other hand the reduced system obtained by the symplectic Galerkin projection, is a Hamiltonian system, therefore system energy is preserved over time integration \cite{Peng:2014di}.

\begin{definition}
\cite{bhatia2002stability} {\color{red} discuss stability}.
\end{definition}

In the next section we introduced the Proper Symplastic Decomposition to construct a symplectic reduced basis.

\subsection{Proper Symplectic Decomposition (PSD)} Consider the set $S = \{z(t_i)| t_i\in[0,T],\ i=1,\dots,N\}$, of samples of the solution to (\ref{eq:2.1}), referred to as the \emph{snapshots}. The PSD, requires a symplectic basis $A$ of size $2k$ to have the minimum symplectic projection error onto the space space of $S$ . In other words, a PSD symplectic basis is the solution to the optimization problem
\begin{equation} \label{eq:3.4}
	\begin{aligned}
	& \underset{A \in \mathbb R^{2n\times 2k}}{\text{minimize}, }
	& & \sum_{s\in S} \| s - AA^+s\|_2^2, \\
	& \text{subject to}
	& & A^T \mathbb{J}_{2n}A = \mathbb{J}_{2k}.
	\end{aligned}
\end{equation}

The Proper Orthogonal Decomposition (POD) \cite{hesthaven2015certified,quarteroni2015reduced} method, is based on a similar idea where the projection operator is an orthogonal projection and the constrain requires $A$ to be an ortho-normal basis. For the POD, the Schmidt-Mirsky-Eckart-Young theorem \cite{Markovsky:2011:LRA:2103589} provides a computationally efficient solution to the minimization problem which requires the SVD decomposition of the snapshot matrix $S$. As the direct solution to (\ref{eq:3.4}) is usually expensive, we must seek alternative methods to obtain near optimal solutions.

A heuristic approach to find a solution to (\ref{eq:3.4}) is proposed by \cite{Peng:2014di}, suggesting that the symplectic basis $A$ should take the form
\begin{equation} \label{eq:3.5}
	A = 
	\begin{pmatrix}
		\Phi & 0 \\
		0 & \Phi
	\end{pmatrix},
\end{equation}
where $\Phi \in \mathbb{R}^{n\times k}$ is an orthonormal matrix. By constructing the \emph{combined snapshot matrix}
\begin{equation} \label{eq:3.6}
	S_{\text{combined}} = [q_1,\dots,q_N,p_1,\dots,p_N], \quad (q_i^T,p_i^T)^T = z(t_i),
\end{equation}
we define $\Phi=[u_1,\dots,u_k]$, where $u_i$ is the $i$-th left singular vector of $S_{\text{combined}}$. The symplectic basis obtained this way is referred to as the \emph{Cotangent Lift}. It is shown in \cite{Peng:2014di} that among all symplectic bases of the form (\ref{eq:3.5}) the cotangent lift minimizes the projection error.

\subsection{Model Reduction of Dissipative Hamiltonian Systems}

The fundamental dynamics of a dissipative Hamiltonian systems are based on a Hamiltonian formulation. Thus, over the course of model reduction, we expect that the evolution of energy is not correctly translated to the reduce system. This is especially expected in a near-zero dissipation regime.

On the other hand, as the conservation law in theorem \ref{theorem:2.1} does not hold for a general dissipative Hamiltonian system, the symplectic model reduction discussed in section \ref{} cannot be applied directly. Therefore, we need to seek alternative methods to transfer the correct evolution of energy into the reduced system.

Consider an extended Hamiltonian system of the form (\ref{eq:2.10.a}) with a quadratic Hamiltonian, $H(z) = z^TK^TKz$. Since $K^TK$ is symmetric and positive definite, it has a unique Cholesky factorization $K^TK = L^T L$ where $L$ is upper triangular \cite{strang09}. So we can write 
\begin{equation} \label{eq:3.7}
H(z) = z^T L^T L z.
\end{equation}
Further suppose that the solution $z(t)$ to the original TDD equation lies on a low-dimensional symplectic subspace such that $z = Ay$, where $A$ is an ortho-symplectic matrix of the size $2n\times 2k$ and $y$ is the expansion coefficients of $z$ in the basis of $A$. The Hamiltonian $H$ then takes the form
\begin{equation} \label{eq:3.8}
	H(z) = H(Ay) = y^T A^T L^T L A y.
\end{equation}
Since $A^T L^T L A$ is a symmetric and positive definite matrix, it has a unique Cholesky factorization $\tilde L^T \tilde L$ where $\tilde L = A^T L A$ is a upper triangular matrix of size $2k \times 2k$. Writing (\ref{eq:2.10.a}) in terms of the reduced coordinates $y$ and the Hamiltonian (\ref{eq:3.7}) reads
\begin{equation} \label{eq:3.9}
		A\dot{y}(t) = \mathbb J_{2n} L^T f(t),
\end{equation}
together with the complementary equation
\begin{equation} \label{eq:3.10}
	f(t) + \sqrt 2 \cdot \sqrt{\chi} \phi(t,0) = LAy.
\end{equation}
Multiplying (\ref{eq:3.9}) with $A^+$ and (\ref{eq:3.10}) with $A^T$ yields
\begin{align} \label{eq:3.11}
	& \dot y(t) = \mathbb J_{2k} A^T L^T f(t), \\
	& A^T f(t) + \sqrt{2} A^T \sqrt{\chi} \phi(t,0) = A^T L A y
\end{align}
Here we used the fact that $A^+\mathbb J_{2n} = \mathbb{J}_{2k} A^T$. If we define $f = A \tilde f$, $\phi = A \tilde \phi$, $\theta = A\tilde \theta$ and the \emph{reduced susceptibility} as $\tilde \chi = A^T \chi A$ we recover the reduced Hamiltonian system
\begin{subequations}
\begin{align}
		\label{eq:3.12.a} & \dot{y}(t) = \mathbb J_{2k} {\tilde L}^T \tilde f(t), \\
		\label{eq:3.12.b} & \partial_t \tilde \theta(t,x) = \partial_x^2 \tilde \phi(t,x) + \sqrt 2 \delta_0(x) \cdot \sqrt{\tilde \chi}  \tilde f(t), \\
		\label{eq:3.12.c} & \partial_t \tilde \phi(t,x) = \tilde \theta(t,x),
\end{align}
\end{subequations}
together with the auxiliary equation
\begin{equation} \label{eq:3.13}
	\tilde f(t) + \sqrt{2} \sqrt{\tilde \chi} \tilde \phi(t,0) = \tilde L y.
\end{equation}
Equations (\ref{eq:3.12.a}) to (\ref{eq:3.12.c}) is a Hamiltonian system on the symplectic linear vector space $\mathbb R^{2k} \times \mathcal H^{2k}$ and contributes to the \emph{reduced TDD system}
\begin{equation}
	\dot {y} = \mathbb J_{2k} \tilde L^T \tilde f(t), \quad \tilde f(t) + \int_0^t \tilde \chi\cdot \tilde f(s)\ ds = \tilde L y.
\end{equation}
Therefore, the system energy will be conserved along integral curves of (\ref{eq:3.12.a}) to (\ref{eq:3.12.c}).

We would like to point out that the transformation that transforms (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) to (\ref{eq:3.12.a})-(\ref{eq:3.12.c}) is given by
\begin{equation}
	\mathcal A = \begin{pmatrix}
		A& 0 \\
		0& A
	\end{pmatrix} : \mathbb R^{2n} \times \mathcal H^{2n} \to \mathbb R^{2k} \times \mathcal H^{2k}.
\end{equation}
This is a symplectic transformation, since $\mathcal A^T J_{2n} \mathcal A = J_{2k}$. Further, the hidden strings $(\tilde \phi,\tilde \theta)$

### 2017-03-07 dissipative_hamiltonian/chapters/2.Hamil.tex
Numerical result added

@@
 \end{equation}
-where the superscript $T$ represents transpose operator.
+where the superscript $T$ represents the transpose operator.
 As the symplectic 2-form is preserved under symplectic transformations, the form of the equations of evolution remains invariant \cite{Marsden:2010:IMS:1965128}.
@@
 Further, one can easily check that $AA^+$ is idempotent and so is a projection operator onto the column span of $A$.
+
+It is known that symplectic matrices are usually ill-conditioned \cite{Karow:2006cf}.
+Under some conditions on a symplectic space \cite{da2003introduction}, one can construct a symplectic basis which is also ortho-normal, and thus norm bounded.
+A basis which is both symplectic and orthogonal is referred to as an \emph{ortho-symplectic basis}.
+We refer the reader to \cite{da2003introduction} for conditions on existence and construction of an ortho-symplectic basis.
 
@@
 To show that the Hamiltonian system (\ref{eq:2.10.a}) to (\ref{eq:2.10.c}) is an extension to (\ref{eq:2.4}) in the sense discussed before, it is enough to show that the solution $f$ to the equation (\ref{eq:2.11}) also satisfies (\ref{eq:2.4.1}).
-Equations (\ref{eq:2.10.b}) and (\ref{eq:2.10.c}) is a forced wave, with the analytical solution
+Equations (\ref{eq:2.10.b}) and (\ref{eq:2.10.c}) are equations to a vibrating string, and can be solved analytically
 \begin{equation} \label{eq:2.12}
@@
 \end{proof}
-We can recover (\ref{eq:2.4.1}) by substituting (\ref{eq:2.12}) into (\ref{eq:2.11}). The extended Hamiltonian for the system (\ref{eq:2.10.a}) to (\ref{eq:2.10.c}) takes the quadratic from
-\begin{equation}
-	\tilde h(z,\phi,\theta) = \frac 1 2 \left( \| Kz - \phi(t,0) \|_2^2 + \| \theta(t) \|^2_{\mathcal H^{2n} } + \| \partial_x\phi(t)\|^2_{\mathcal H^{2n} }\right)
+We can recover (\ref{eq:2.4.1}) by substituting (\ref{eq:2.12}) into (\ref{eq:2.11}).
+The extended Hamiltonian $H_\text{ex}$ for the system (\ref{eq:2.10.a}) to (\ref{eq:2.10.c}) takes the quadratic from
+\begin{equation} \label{eq:2.13}
+	H_\text{ex}(z,\phi,\theta) = \frac 1 2 \left( \| Kz - \phi(t,0) \|_2^2 + \| \theta(t) \|^2_{\mathcal H^{2n} } + \| \partial_x\phi(t)\|^2_{\mathcal H^{2n} }\right)
 \end{equation}
@@
 The Lamb model is a harmonic oscillator coupled with a vibrating string, and the tension in the string causes linear dissipation in the dynamics of the harmonic oscillator.
+
+Note that time integration of (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) involves the integration of $f$ in (\ref{eq:2.12}).
+In general $f(t)$ must be stored and may cause serious storage limitation in long-time integration.
+However, we are interested solely in finding $z(t)$ which depends on $f(t)$, and $\phi(t,0)$ that is the integral of the history of $f(t)$.
+So by carefully choosing a quadrature rule that uses the same quadrature nodes as the time integrator we can avoid storing the history of $f$.
+For example for the trapezoidal rule, we recover the recursive relation
+\begin{equation} \label{eq:2.14}
+	\int_{0}^{t_n} f(s) \ ds \approx \frac{\Delta t}{2} f(t_n) + \frac{\Delta t}{2} f(t_{n-1}) + \int_{0}^{t_{n-1}} f(s) \ ds,
+\end{equation}
+where $\Delta t$ is the time discretization step.
+The recursive relation in (\ref{eq:2.12}) suggests that only storing the value of the integral term together with the state of $f$ in the previous time step is required to evaluate the integral for the new time step.
+For other interpolation based quadrature rules, we can construct similar recursive rules of the form
+\begin{equation}
+	\int_{0}^{t_n} f(s) \ ds \approx \sum_{i=0}^{k} \omega_i f(t_{n-i})  + \int_{0}^{t_{n-k}} f(s) \ ds
+\end{equation}
+for some quadrature weights $\omega_i$, $i=1,\dots,k$ with $k\ll n$. Thus, time integration of (\ref{eq:2.10.a})-(\ref{eq:2.10.c}), requires storing only $k$ evaluations of $f$.

### 2017-03-07 dissipative_hamiltonian/chapters/3.model.tex
Numerical result added

@@
 It is shown in \cite{Peng:2014di} that among all symplectic bases of the form (\ref{eq:3.5}) the cotangent lift minimizes the projection error.
+Note that the basis $A$ constructed using the cotangent lift is an ortho-symplectic basis.
 
@@
 
-We would like to point out that the transformation that transforms (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) to (\ref{eq:3.12.a})-(\ref{eq:3.12.c}) is given by
+We point out that the transformation that transforms (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) to (\ref{eq:3.12.a})-(\ref{eq:3.12.c}) is given by
 \begin{equation}
@@
 \end{equation}
-This is a symplectic transformation, since $\mathcal A^T J_{2n} \mathcal A = J_{2k}$. Further, the hidden strings $(\tilde \phi,\tilde \theta)$
+This is a symplectic transformation, since $\mathcal A^T J_{2n} \mathcal A = J_{2k}$.
+Further, the dissipation of energy in the reduced system only depends on the reduced susceptibility.
+Thus, the choice of $A$ should be independent of the hidden strings $(\tilde \phi, \tilde \theta)$.
+In other words, if the reduced space is chosen to be a symplectic subspace, then the actions of model reduction and Hamiltonian extension commute.
+We summarize the algorithm for model reduction of dissipative Hamiltonian systems in Algorithm \ref{alg:3.1}.
+
+\begin{algorithm}
+\caption{Model Reduction For Dissipative Hamiltonian Systems} \label{alg:3.1}
+\begin{enumerate}
+	\item Construct the Hamiltonian extension (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) to the original TDD system (\ref{eq:2.4}).
+	\item Collect the snapshots $z(t_i)$, $i=1,\dots,N$ through time integration of the extended Hamiltonian.
+	\item Construct an ortho-symplectic basis $A$.
+	\item Define $\tilde L = A^T L A$, $\tilde \chi = A^T \chi A$ and construct the reduced dissipative Hamiltonian system (\ref{eq:3.12.a})-(\ref{eq:3.12.c})
+\end{enumerate}
+\end{algorithm}
+
+Note that Algorithm \ref{alg:3.1} does not depend on the choice of the method to construct an ortho-symplectic basis $A$.
+Thus, an SVD-based method, such as the cotangent lift, or a greedy-natured method can be applied.

### 2017-03-07 dissipative_hamiltonian/chapters/4.numer.tex
Numerical results started

\section{Numerical Results}
In this section we illustrate the performance of the discussed metthod through numerical solution of the dissipative wave equation and a port-Hamiltonian model for a dissipative circuit.

\subsection{Dissipative wave equation}

Consider the dissipative linear wave equation
\begin{equation} \label{eq:4.1}
	\begin{aligned}
		q_{t}(t,x) &= p(t,x), \\
		p_{t}(t,x) &= c^2 q_{xx}(t,x) - r(x)  p(t,x) , \\
		q(0,x) &= q_0(x), \\
		p(0,x) &= 0.
	\end{aligned}
\end{equation}
where $x$ belongs to a one-dimensional torus of length $L$ and $r:[0,1]\to[0,1]$ is a positive semi-definite real valued function.

We discretize the torus into $N_{\Delta x}$ equidistant points and define $\Delta x = L/N_{\Delta x}$, $x_i = i\Delta x$, $q_i=q(t,x_i)$ and $p_i=p(t,x_i)$ for $i = 1, \dots, N_{\Delta x}$. The discretization of $r$ corresponds to a diagonal and semi-positive definite matrix $r_\Delta$. Further, we discretize (\ref{eq:4.1}) using a standard centeral finite differences schemes to obtain
\begin{equation} \label{eq:4.2}
	\dot z = \mathbb J_{2n} K^T K z - R z,
\end{equation}
where $z = (q_1,\dots,q_{N_{\Delta x}},p_1,\dots,p_{N_{\Delta x}})$ and $K$ and $R$ are given as
\begin{equation} \label{eq:4.3}
	K^T K =
	\begin{pmatrix}
		I & 0 \\
		0 & c^2D_x^TD_x
	\end{pmatrix} , \quad
	R =
	\begin{pmatrix}
		0 & 0 \\
		0 & r_\Delta
	\end{pmatrix},
\end{equation}
with $D_x^TD_x = D_{xx}$ the central finite differences matrix operator. Writing (\ref{eq:4.2}) in a TDD formulation gives
\begin{equation} \label{eq:4.4}
\begin{aligned}
	& \dot z = \mathbb J_{2n} K^T f(t) \\
	& f(t) + R \int_0^t f(t) = K z.
\end{aligned}
\end{equation}
Here, since $R$ is not time dependent, it commutes with the integration operator. The Hamiltonian extension of (\ref{eq:4.4}), then takes the form (\ref{eq:2.10.a})-(\ref{eq:2.10.c}).

The initial condition used is given by
\begin{equation} \label{eq:4.5}
	q_i(0) = h( 10\times|x_i - \frac{1}{2}| ), \quad p_i = 0, \quad i=1,\dots,N
\end{equation}
where $h(s)$ is the cubic spline function
\begin{equation} \label{eq:4.6}
h(s) = 
\left\{
\begin{aligned}
& 1 - \frac{3}{2}s^2 + \frac{3}{4}s^3, \quad & 0\leq s \leq 1, \\
& \frac{1}{4}(2-s)^3, & 1< s \leq 2, \\
& 0, & s > 2.
\end{aligned}
\right.
\end{equation}

For the numerical time integration of the extended Hamiltonian system, the Str\"omer-Verlet time stepping scheme (\ref{eq:2.3}) is used. Further in each time step, the system of linear equations (\ref{eq:2.4.1}) is solved to recover $z$. System parameters are summarized below.
\vspace{0.5cm}
\begin{center}
\begin{tabular}{|l|l|}
\hline
Domain length & $L = 1$ \\
No. grid points & $N = 500$ \\
Space discretization size & $\Delta x = 0.002$ \\
Time discretization size & $\Delta t = 0.01$ \\
Wave speed & $c^2 = 0.1$ \\
\hline
\end{tabular}
\end{center}
\vspace{0.5cm}

The first numerical experiment, corresponds to an inhomogeneous dissipative media. Here, $r_{\Delta} = \text{diag}(r_1,\dots,r_{N_{\Delta x}})$, with $r_i = 0.1 + 0.9(i/N_{\Delta x})$, and diag is the MATLAB notation for constructing a diagonal matrix.

We construct the reduced dissipative Hamiltonian system according to Algorithm \ref{alg:3.1}. The cotangent lift is used to construct an ortho-symplectic reduced basis $A$. The performance of the reduced dissipative Hamiltonian (RDH) system method is then compared to the POD, and the method proposed in \cite{peng2016geometric} referred to as the PSD.

### 2017-03-09 dissipative_hamiltonian/chapters/4.numer.tex
The first section of the results is done also the figures are added

@@
 
-We construct the reduced dissipative Hamiltonian system according to Algorithm \ref{alg:3.1}.
-The cotangent lift is used to construct an ortho-symplectic reduced basis $A$.
-The performance of the reduced dissipative Hamiltonian (RDH) system method is then compared to the POD, and the method proposed in \cite{peng2016geometric} referred to as the PSD.
+Figure \ref{fig:4.1}.a shows the solution of the original dissipative wave equation (\ref{eq:4.1}) at $t \in \{0,2.5,5,7.5\}$.
+For a nonzero $r_\Delta$ the solution will converge to $(q(t=\infty,x),p(t=\infty,x)) = (\rho,0)$ where $\rho$ is the center of mass of $q_0$.
+
+To construct the reduced basis for different methods, we gather the snapshots of (\ref{eq:4.1}).
+We construct RDH reduced system according to the Algorithm \ref{alg:3.1}.
+The cotangent lift is used to construct an ortho-symplectic reduced basis $A$ (\ref{eq:3.5}).
+The performance of the RDH method is then compared to the POD, and the method proposed in \cite{peng2016geometric} referred to as the PSD.
+
+Figure \ref{fig:4.1}.b Shows the decay of the singular values of the snapshot matrix for the POD, PSD and the RDH methods.
+Note that the snapshots for the PSD and the RDH are different since they have different canonical representations.
+The fast decay of the eigenvalues in all methods is a strong indicator for the existence of a low dimensional reduced system.
+The reduced bases are then constructed using 20, 40 and 60 number of modes.
+
+The $L^2$ error between the full systems and the RDH, the PSD and the POD method are presented in Figure \ref{fig:4.1}.c.
+We notice that symplectic methods provide a more accurate solution with respect to the POD method.
+In fact, the POD method does not yield a stable reduced system.
+Further it is seen that enriching the PSD reduced basis does not yield a significant enhancement in the accuracy of the reduced system.
+This is because in the PSD method, a non-conservative system is being numerically integrated with a symplectic integrator.
+This results in an incorrect evolution of energy and eventually, in a qualitatively wrong numerical solution.
+
+On the other hand, We notice that the RDH method with 40 modes provides a significantly more accurate solution compared to the PSD method with 60 modes.
+The RDH method provides a conservative reduced system where the dissipated energy is trapped in the hidden strings.
+The conservation of energy is then guaranteed using a symplectic integrator.
+Due to this reason, we observe remarkable increase in the accuracy by enriching the RDH reduced basis.
+
+Figure \ref{fig:4.1}.d shows the preservation of the energy in different methods.
+The conservation law expressed in theorem \ref{theorem:2.1} is destroyed through the POD model reduction.
+That is why we observe the blow-up of the system energy.
+The symplectic methods however, preserves the energy significantly better.
+Due to the reason discussed above, enriching the PSD basis does not significantly affect the preservation of energy.
+On the contrary, The RDH not only provides a noteworthy accuracy in preserving the energy, but also provides significant improvement under enrichment.
+
+in Figure \ref{fig:4.1}.e we show the transfer of the energy from the TDD system to the hidden strings, for the full system and the RDH reduced system.
+We notice that the RDH method preserves the total energy of the extended Hamiltonian system.
+Further, the transfer of the energy to the hidden strings in the full model is correctly translated in the reduced system.
+
+\begin{figure}[t]
+\begin{tabular}{cc}
+\includegraphics[width=0.5\textwidth]{./figs/wave/solution} & 
+\includegraphics[width=0.5\textwidth]{./figs/wave/singular} \\
+(a) & (b) \\
+\includegraphics[width=0.5\textwidth]{./figs/wave/error} & 
+\includegraphics[width=0.5\textwidth]{./figs/wave/energy} \\
+(c) & (d) \\
+\multicolumn{2} {c} {\includegraphics[width=0.7\textwidth]{./figs/wave/energy_conserved}} \\
+\multicolumn{2} {c} {(e)} 
+\end{tabular}
+\caption{(a) The solution to the original dissipative wave equation (\ref{eq:4.1}), (b) The decay of the singular values for the POD, the PSD and the RDH methods, (c) The $L^2$ error for the different methods, (d) Energy preservation for different methods, (e) Energy preservation of the Hamiltonian extension for the original and the reduced system.
+``FM'' and ``RM'' stand for the full model and the reduced model respectively.} \label{fig:4.1}
+\end{figure}
+
+The second numerical experiment,

### 2017-03-14 dissipative_hamiltonian/chapters/4.numer.tex
started the section for port hamiltonian

@@
 
-in Figure \ref{fig:4.1}.e we show the transfer of the energy from the TDD system to the hidden strings, for the full system and the RDH reduced system.
+In Figure \ref{fig:4.1}.e we show the transfer of the energy from the TDD system to the hidden strings, for the full system and the RDH reduced system.
 We notice that the RDH method preserves the total energy of the extended Hamiltonian system.
@@
 
-The second numerical experiment,
+The second numerical experiment is the dissipative wave equation (\ref{eq:4.1}) in a near-zero dissipation regime.
+The purpose of this numerical experiment is to show that the RDH yields a stable solution in the elimination of the dissipation.
+Also to demonstrate that the PSD converges to the RDH as $\| \chi \|_{\infty}\to 0$.
+
+The numerical setting is taken to be identical to the previous numerical experiment, but with the difference that $r_i = 0.001$, for $i=1,\dots,N_{\Delta x}$.
+
+\subsection{Port-Hamiltonian Systems}
+Port-Hamiltonian systems are popular in network modeling and electrical engineering.
+In the framework of port-Hamiltonian modelling, Energy conservation and Hamiltonian structure is the fundamental principle of the dynamics of the system.
+Ports in the system network then allows the exchange of energy with the environment in the form of input forces, capacitors and dissipations \cite{vanderSchaft:2014:PST:2693645.2693646}.
+Port-Hamiltonian systems can be viewed as a forced and dissipative Hamiltonian system.
+\begin{figure}[t]
+\begin{center}
+	\includegraphics[width=0.7\textwidth]{./figs/porthamil/circuit}
+\end{center}
+\caption{$n$-dimensional ladder network} \label{fig:4.2}
+\end{figure}
+
+Consider the $n$-dimensional linear ladder network in Figure \ref{fig:4.2}.
+Here $C_i$, $L_i$ and $R_i$, $i=1,\dots,n$, are the capacitance, inductance and resistance of the corresponding capacitors, inductors and resistors respectively and $R_{n+1}$ is the load capacitor.
+The port-Hamiltonian model of the linear ladder network takes the form
+\begin{equation} \label{eq:4.7}
+		\dot x = (J_{2n} - R)Q^TQx + u.
+\end{equation}
+Here $x = (c_1,\phi_1,\dots,c_n,\phi_n)^T$, $Q = I_{2n\times 2n}$, $R = \text{diag}(0,R_1,\dots,0,R_n+R_{n+1})$, $u=(1,0,\dots,0)^T$ is the input vector and $J_{2n}$ is a skew-symmetric $2n\times 2n$ matrix with -1 and 1 on the first diagonal above and the first diagonal below the main diagonal respectively.
+
+Since $J_{2n}$ is full ranked, one can always find a coordinate transformation $ x= T \tilde x$ such that $T^{-1} J_{2n} T^{-T} = \mathbb J_{2n}$.
+The dissipative Hamiltonian formulation for (\ref{eq:4.7}) takes the form
+\begin{equation}
+	\dot {\tilde x} = \mathbb J_{2n} \tilde Q^T\tilde Q \tilde x - \tilde Rx + \tilde u,
+\end{equation}
+where $\tilde Q = QT$, $\tilde R = T^{-1}RT^{-T}Q^TQ$ and $\tilde u = T^{-1} u$.
+Note that in this case, since $T$ is orthogonal and $R$ is diagonal, then $\tilde R$ becomes symmetric and semi-positive definite.

### 2017-03-15 dissipative_hamiltonian/chapters/2.Hamil.tex
progress in porthamiltonian section

@@
 \begin{equation} \label{eq:2.7}
-	\frac{d}{dt} x = J_{2n} \frac{\delta h}{\delta x}.
+	\frac{d}{dt} x = J_{2n} \frac{\delta H_{\text{ex}}}{\delta x}.
 \end{equation}
-Here $J_2n$ is the symplastic operator and $h:\mathbb R^{2n}\times \mathcal H^{2n} \to \mathbb R$ is an extended quadratic Hamiltonian function defined on $\mathbb R^{2n}\times \mathcal H^{2n}$.
+Here $J_2n$ is the symplastic operator and $H_{\text{ex}}:\mathbb R^{2n}\times \mathcal H^{2n} \to \mathbb R$ is an extended quadratic Hamiltonian function defined on $\mathbb R^{2n}\times \mathcal H^{2n}$.
 

### 2017-03-15 dissipative_hamiltonian/chapters/4.numer.tex
progress in porthamiltonian section

@@
 \end{equation}
-Here $x = (c_1,\phi_1,\dots,c_n,\phi_n)^T$, $Q = I_{2n\times 2n}$, $R = \text{diag}(0,R_1,\dots,0,R_n+R_{n+1})$, $u=(1,0,\dots,0)^T$ is the input vector and $J_{2n}$ is a skew-symmetric $2n\times 2n$ matrix with -1 and 1 on the first diagonal above and the first diagonal below the main diagonal respectively.
+Here $x = (c_1,\phi_1,\dots,c_n,\phi_n)^T$ where $c_i$ and $\phi_i$, for $i=1,\dots,n$, are the charge and the flux of $C_i$ and $L_i$ respectively. $Q$ and $R$ are given as
+\begin{equation}
+	Q = \text{diag}(C_1^{-1},L_1^{-1},\dots,C_n^{-n},L_n^{-n}), \quad R = \text{diag}(0,R_1,\dots,0,R_n+R_{n+1}),
+\end{equation}
+$u=(1,0,\dots,0)^T$ is a constant input current and $J_{2n}$ is a skew-symmetric $2n\times 2n$ matrix with -1 and 1 on the superdiagonal and subdiagonal respectively.
+
+The energy associated to a port-Hamiltonian system of the form (\ref{eq:4.7}) at time $t$, is given as $H(x(t)) = \frac 1 2 x^T Q^T Q x$.
+Since $J_{2n}$ is skew symmetric it implies that $\frac d {dt} H(x) = u^T Q^T Q x - x^T Q^T Q R Q^T Q x \leq u^T Q^T Q x$ which is referred to as the \emph{passivity} of the system (\ref{eq:4.7}) \cite{vanderSchaft:1996es,Willems:1972ek}.
 
@@
 The dissipative Hamiltonian formulation for (\ref{eq:4.7}) takes the form
-\begin{equation}
+\begin{equation} \label{eq:4.8}
 	\dot {\tilde x} = \mathbb J_{2n} \tilde Q^T\tilde Q \tilde x - \tilde Rx + \tilde u,
@@
 where $\tilde Q = QT$, $\tilde R = T^{-1}RT^{-T}Q^TQ$ and $\tilde u = T^{-1} u$.
-Note that in this case, since $T$ is orthogonal and $R$ is diagonal, then $\tilde R$ becomes symmetric and semi-positive definite.
+Note that in this case, since $T$ is orthogonal and $R$ is diagonal, then $\tilde R$ is symmetric and semi-positive definite.
+The TDD formulation of (\ref{eq:4.8}) takes the form
+\begin{equation} \label{eq:4.9}
+	\dot{\tilde x} = \mathbb{J}_{2n} \tilde Q^T f(t) + \tilde u, \quad f(t) + \tilde R \int_0^t f(t) = \tilde Q \tilde x.
+\end{equation}
+The extended Hamiltonian formulation (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) with a quadratic Hamiltonian $H_{\text{ex}}$ can be then carried out according to section \ref{??}.
+We note that due to the input $\tilde u$, the Hamiltonian $H_{\text{ex}}$ is time dependent.
+As the matter of fact $\frac{d}{dt} H_{\text{ex}} = \tilde u^T\tilde Q^T \tilde Q \tilde x$.
+If we define $\overset{\circ}{H}_{\text{ex}} : \mathbb R^{2n}\times \mathcal H^{2n}\times \mathbb R^{2}\to \mathbb R$ as
+\begin{equation} \label{eq:4.10}
+	\overset{\circ}{H}_{\text{ex}}(\tilde x,\phi,\theta,t,e) = H_{\text{ex}}(\tilde x,\phi,\theta,t) + e, \quad \dot e = - \partial_t H_{\text{ex}},
+\end{equation}
+it is easily checked that $\frac d {dt} \overset{\circ}{H}_{\text{ex}} =0$ \cite{Hairer:1250576}.
+However for the time integration of the Hamiltonian system related to $\overset{\circ}{H}_{\text{ex}}$, since the evolution of $\tilde x$, $\phi$ and $\theta$ does not explicitly depend on $e$, we can apply a symplectic integrator directly on (\ref{eq:4.10}).
+Thus, the passivity of (\ref{eq:4.7}) will be preserved through a symplectic time integration of (\ref{eq:4.9}).
+
+Using an ortho-symplectic reduced basis $A$ The reduced dissipative Hamiltonian method can be applied to (\ref{eq:4.9}) to construct a reduced system of the form (\ref{eq:3.12.a})-(\ref{eq:3.12.c}) together with the extended Hamiltonian $\tilde H_{\text{ex}}$.
+We immediately notice that $\frac{d}{dt} \tilde H_{\text{ex}} = (A^+ \tilde u)^T A^T \tilde Q^T \tilde Q A y$, meaning that the reduced system is passive.
+Further the dissipative Hamiltonian structure of the reduced system indicates that the reduced system also carries a port-Hamiltonian structure.
+
+We consider a 100-dimensional ($n=50$) port-Hamiltonian system for the ladder network discussed above.
+We take $C_i=1$, $L_i = 1$, $R_i=0.2$ for $i=1,\dots,50$, and $R_{51} = 0.4$.
+The cotangent lift is used to construct a reduced basis and the RDH method is used to construct a reduced system according to Algorithm (\ref{alg:3.1}).
+
+The solution of the RDH method is compared to the main results of \cite{Polyuga:2010gj}, where a passivity-preserving model reduction is developed using the moment matching method at infinity.
+
+\begin{figure}[t]
+\begin{tabular}{cc}
+\includegraphics[width=0.5\textwidth]{./figs/porthamil/error_capac} & 
+\includegraphics[width=0.5\textwidth]{./figs/porthamil/error_flux} \\
+(a) capacitors & (b) inductors
+\end{tabular}
+\caption{Error between the full model and the reduced model obtained by the reduced dissipating Hamiltonian method ``RDH'' and the moment matching method ``MM''.
+(a) The average error of charge in capacitors in time.
+(b) the average error of the flux in inductors in time.} \label{fig:4.3}
+\end{figure}

### 2017-03-16 dissipative_hamiltonian/chapters/4.numer.tex
result section completed

@@
 (c) & (d) \\
-\multicolumn{2} {c} {\includegraphics[width=0.7\textwidth]{./figs/wave/energy_conserved}} \\
-\multicolumn{2} {c} {(e)} 
+\includegraphics[width=0.55\textwidth]{./figs/wave/energy_conserved} &
+\includegraphics[width=0.45\textwidth]{./figs/wave/error_homo} \\
+(e) & (f) 
 \end{tabular}
 \caption{(a) The solution to the original dissipative wave equation (\ref{eq:4.1}), (b) The decay of the singular values for the POD, the PSD and the RDH methods, (c) The $L^2$ error for the different methods, (d) Energy preservation for different methods, (e) Energy preservation of the Hamiltonian extension for the original and the reduced system.
-``FM'' and ``RM'' stand for the full model and the reduced model respectively.} \label{fig:4.1}
+``FM'' and ``RM'' stand for the full model and the reduced model respectively.
+(f) The $L^2$ error between the solution to the reduced system and the full in a near-zero dissipation regime.} \label{fig:4.1}
 \end{figure}
@@
 The second numerical experiment is the dissipative wave equation (\ref{eq:4.1}) in a near-zero dissipation regime.
-The purpose of this numerical experiment is to show that the RDH yields a stable solution in the elimination of the dissipation.
-Also to demonstrate that the PSD converges to the RDH as $\| \chi \|_{\infty}\to 0$.
-
-The numerical setting is taken to be identical to the previous numerical experiment, but with the difference that $r_i = 0.001$, for $i=1,\dots,N_{\Delta x}$.
+The numerical setting is taken to be identical to the previous numerical experiment, but with the difference that $r_i = 10^{-5}$, for $i=1,\dots,N_{\Delta x}$.
+Figure \ref{fig:4.1}.f shows the $L^2$ error between the solution to the reduced system and the full system, for the POD, the PSD and the RDH methods.
+We notice that the POD does not yield a reduced system as the symplectic structure is lost via model reduction.
+Further we notice that error for the PSD and the RDH coincide since the two methods become identical as $\| \chi \|_{\infty}\to 0$.
 
@@
 
-The solution of the RDH method is compared to the main results of \cite{Polyuga:2010gj}, where a passivity-preserving model reduction is developed using the moment matching method at infinity.
+The solution of the RDH method is compared to the main results of \cite{Polyuga:2010gj}, where a passivity-preserving model reduction is developed using a moment matching method at infinity.
+The charge in $C_1$ is chosen to be the single out put for the moment matching method.
 
@@
 \includegraphics[width=0.5\textwidth]{./figs/porthamil/error_flux} \\
-(a) capacitors & (b) inductors
+(a) capacitors & (b) inductors \\
+\multicolumn{2} {c} {\includegraphics[width=0.5\textwidth]{./figs/porthamil/error1}} \\
+\multicolumn{2} {c} {(c) charge in $C_1$} 
 \end{tabular}
@@
 (a) The average error of charge in capacitors in time.
-(b) the average error of the flux in inductors in time.} \label{fig:4.3}
+(b) the average error of the flux in inductors in time.
+(c) Error in $C_1$.} \label{fig:4.3}
 \end{figure}
+
+Reduced bases of size $2k = 10$, $2k = 20$ and $2k = 30$ are constructed with the RDH and the moment matching method.
+Figure \ref{fig:4.3}.c shows the error in the charge in $C_1$ for the two methods.
+We observe that although the moment matching method is bounded, over long-time integration, the RDH method provides a significantly more accurate solution.
+This is because in the moment matching method, passivity of the reduced system implies that the energy of the system will be bounded by the input energy.
+However, there is no guarantee that the dissipation of energy in the reduced system mimics the one of the original system.
+On the other hand, the RDH method allows correct dissipation of energy through the hidden strings.
+Also the symplectic time integration in the RDH method guarantees that the total energy will be preserved.
+
+Over short-time integration, we notice that the moment matching method with 10 modes provides a more accurate solution than the RDH with 10 modes.
+Further, the moment matching method with 20 and 30 modes provide a comparable accuracy to the RDH method with 20 and 30 modes.
+But, the RDH method maintains the high accuracy over the long-time integration, while the moment matching method loses up to 3 orders of magnitude in the accuracy independent of the number of modes.
+
+Figure \ref{fig:4.3}.a and Figure \ref{fig:4.3}.b show the average error in the charge and flux of the capacitors and inductors respectively, for these methods.
+The RDH method provides a significantly better accuracy compared to the moment matching method.
+This is because the charge of $C_1$ is specified as the output of interest in the moment matching method so it is expected that that method provides low accuracy for computing other outputs.
+On the other hand, the RDH method not only provides high accuracy in computing the charge for $C_1$ but also it has high accuracy for computing other outputs of interest.

### 2017-03-18 dissipative_hamiltonian/chapters/1.intro.tex
introduction added

\section{Introduction} \label{sec:1}

The need for more accuracy has led to ever more complex models in engineering and physics. As a consequence, direct numerical methods for solving partial differential equations, a powerful modeling tool, have become inefficient due to limitations in computational power and resources.

Over the past decade, reduced basis methods have emerged as a popular approach to reduce the complexity of large systems of partial differential equations \cite{Ito:1998up,Ito:1998ch,Ito:2001ev}. These methods construct a low dimensional subspace, the reduced space, that approximately represent the solution to the system of differential equations. The projection of the original system onto the reduced space then allows the exploration of the solution with a significantly reduced computational complexity \cite{hesthaven2015certified,quarteroni2015reduced}.

Hamiltonian systems are an important class of systems that appear in engineering and physics. In these systems, preserving system energy is essential in a correct numerical solutions. Therefore, developing model reduction techniques that preserves the symplectic symmetry is essential. The classical reduced basis methods, do not generally preserve the conservation laws and symmetries, e.g. the system energy, of the full system \cite{Amsallem:2014ef,prajna2003pod}. This often results in an unstable or qualitatively wrong solution.

It is demonstrated by \cite{Lall:2003iy,Carlberg:2014ky,Peng:2014di}, that if the basis to the reduced space is not chosen carefully, the symplastic symmetry of Hamiltonian and Lagrangian systems will be destroyed over the course of model reduction. To resolve this issue Peng et al. \cite{Peng:2014di} and Lall et al. \cite{Lall:2003iy} construct a reduced ordered configuration space that inherits symmetries of the full configuration space. Then, by using a proper time integrator scheme one can preserve symmetries.

Most models in engineering, appear as a dissipative perturbation of a Hamiltonian system. In these systems, conservation of energy is taken as a fundamental principle of the system dynamics, while deriving and dissipative forces, e.g. friction, can change the energy level of the system \cite{vanderSchaft:2014:PST:2693645.2693646}. As the system energy is no longer preserved for such systems, the methods mentioned above can no longer be applied directly \cite{peng2016geometric}.

For dissipative and forced Hamiltonian system, Peng et al. \cite{peng2016geometric} suggest a symplectic model reduction method that preserves Hamiltonian and the dissipative structure of the original system. However, since this method uses a symplectic integrator for a non-conservative system, there is no guarantee that the evolution of the energy is translated correctly to the reduced system.

In the context of network modeling and circuit simulation, a considerable attention is geared toward developing structure preserving, and in particular energy preserving, model reduction techniques. Model reduction of port-Hamiltonian system is given \cite{Polyuga:2010gj,beattie2011structure,chaturantabut2016structure} and the references therein. These methods construct use a Krylov or a Proper Orthogonal Decomposition (POD) method to construct a reduced port-Hamiltonian system that preserves the passivity, and thus the stability of the original system. In general these methods do not guarantee the correct distribution of the energy among the energy consuming and energy storing units. Further, over long time integration, accumulation of local errors might produce an erroneous solution.

In this paper we propose the Reduced Dissipative Hamiltonian Method (RDH), a structure preserving model reduction approach for dissipative Hamiltonian systems. By considering a canonical heat bath, also known as the hidden strings \cite{Figotin:2006jy,Figotin:2005}, the reduced system is extended to a closed and conservative system. Therefore, a symplectic time integrator can guarantee the conservation of the system energy and the dissipated energy. Further, the hidden strings assure that the local errors in the dissipation of energy do not accumulate, resulting in a correct evolution of the system energy.

This paper is organized as follow. Section \ref{sec:2} covers the required topics on Hamiltonian systems, dissipative Hamiltonian system and Hamiltonian extension. In Section \ref{sec:3} we discussion the symplectic model reduction for Hamiltonian systems and introduce the Reduced Dissipative Hamiltonian Method. Accuracy, stability, and efficiency of the RDH method is then discussed in Section \ref{sec:4}, through simulation of the dissipative wave equation and a linear port-Hamiltonian system of an electrical circuit. We offer conclusive remarks in Section \ref{sec:5}.

### 2017-03-18 dissipative_hamiltonian/chapters/3.model.tex
progress in porthamiltonian section
introduction added

@@
-\section{Model Order Reduction}
+\section{Model Order Reduction} \label{sec:3}
 In this section we first explain the main results of \cite{Peng:2014di} regarding model reduction of Hamiltonian systems.
@@
 
-\subsection{Model Reduction of Dissipative Hamiltonian Systems}
+\subsection{The Reduced Dissipative Hamiltonian Method}
 
@@
 \begin{algorithm}
-\caption{Model Reduction For Dissipative Hamiltonian Systems} \label{alg:3.1}
+\caption{The Reduced Dissipative Hamiltonian Method (RDH)} \label{alg:3.1}
 \begin{enumerate}

### 2017-03-18 dissipative_hamiltonian/chapters/4.numer.tex
introduction added

@@
-\section{Numerical Results}
+\section{Numerical Results} \label{sec:4}
 In this section we illustrate the performance of the discussed metthod through numerical solution of the dissipative wave equation and a port-Hamiltonian model for a dissipative circuit.

### 2017-03-18 dissipative_hamiltonian/chapters/5.conc.tex
introduction added

\section{Conclusion} \label{sec:5}

Here comes the conclusion

### 2017-03-18 dissipative_hamiltonian/paper.tex
model reduction of dissipative Hamiltonians started
MOR of dissipative Hamiltonian systems is updated
Numerical result added
The first section of the results is done also the figures are added
result section completed
abstract added

@@
 \usepackage{amssymb}
+\usepackage{color}
+\usepackage{algorithm}
+\usepackage{subfig}
 
@@
 \begin{abstract}
-Insert your abstract here. Include keywords, PACS and mathematical
-subject classification numbers as needed.
+Reduced basis methods are popular for approximately solving large and complex systems of differential equations.
+However, conventional methods do not generally preserve conservation laws and symmetries of the full order model.
+Here, we present The Reduced Dissipative Hamiltonian Method, that preserves the symplectic symmetry of dissipative Hamiltonian systems over the course of model reduction.
+This method constructs a closed reduced Hamiltonian system by coupling the full model with a canonical heat bath.
+This results in a correct dissipation of energy, preservation of total energy and ultimately in the stability of the solution.
+Accuracy and stability of the method is illustrated through the numerical simulation of the dissipative wave equation and a port-Hamiltonian model of an electric circuit.
+
 \keywords{First keyword \and Second keyword \and More}
@@
 \input{./chapters/2.Hamil.tex}
+\input{./chapters/3.Model.tex}
+\input{./chapters/4.Numer.tex}
+\input{./chapters/5.conc.tex}
 

### 2017-03-19 dissipative_hamiltonian/chapters/1.intro.tex
first draft finished

@@
 
-The need for more accuracy has led to ever more complex models in engineering and physics.
-As a consequence, direct numerical methods for solving partial differential equations, a powerful modeling tool, have become inefficient due to limitations in computational power and resources.
+The need for more accuracy has led to ever more complex models and systems of partial differential equations in engineering and physics.
+As a consequence, direct numerical methods for solving PDEs have become inefficient due to limitations in computational power and resources.
 
 Over the past decade, reduced basis methods have emerged as a popular approach to reduce the complexity of large systems of partial differential equations \cite{Ito:1998up,Ito:1998ch,Ito:2001ev}.
-These methods construct a low dimensional subspace, the reduced space, that approximately represent the solution to the system of differential equations.
+These methods construct a low-dimensional subspace, the reduced space, that approximately represent the solution to the system of differential equations.
 The projection of the original system onto the reduced space then allows the exploration of the solution with a significantly reduced computational complexity \cite{hesthaven2015certified,quarteroni2015reduced}.
@@
 Hamiltonian systems are an important class of systems that appear in engineering and physics.
-In these systems, preserving system energy is essential in a correct numerical solutions.
-Therefore, developing model reduction techniques that preserves the symplectic symmetry is essential.
-The classical reduced basis methods, do not generally preserve the conservation laws and symmetries, e.g.
-the system energy, of the full system \cite{Amsallem:2014ef,prajna2003pod}.
-This often results in an unstable or qualitatively wrong solution.
+In these systems, preserving system energy is essential in a correct numerical solution.
+Therefore, developing model reduction techniques that preserve the symplectic symmetry is essential.
+The classical reduced basis methods, do not generally preserve the conservation laws and intrinsic symmetries of Hamiltonian systems \cite{Amsallem:2014ef,prajna2003pod}.
+This often results in an unstable or a qualitatively wrong solution.
 
-It is demonstrated by \cite{Lall:2003iy,Carlberg:2014ky,Peng:2014di}, that if the basis to the reduced space is not chosen carefully, the symplastic symmetry of Hamiltonian and Lagrangian systems will be destroyed over the course of model reduction.
-To resolve this issue Peng et al.
+It is demonstrated by \cite{Lall:2003iy,Carlberg:2014ky,Peng:2014di}, that if the basis for the reduced space is not chosen carefully, the symplastic symmetry of Hamiltonian and Lagrangian systems will be destroyed over the course of model reduction.
+To resolve this issue, Peng et al.
 \cite{Peng:2014di} and Lall et al.
 \cite{Lall:2003iy} construct a reduced ordered configuration space that inherits symmetries of the full configuration space.
-Then, by using a proper time integrator scheme one can preserve symmetries.
+Then, by using a proper time integrator scheme, the symmetries are preserved in the reduced system.
 
@@
 
-In the context of network modeling and circuit simulation, a considerable attention is geared toward developing structure preserving, and in particular energy preserving, model reduction techniques.
-Model reduction of port-Hamiltonian system is given \cite{Polyuga:2010gj,beattie2011structure,chaturantabut2016structure} and the references therein.
-These methods construct use a Krylov or a Proper Orthogonal Decomposition (POD) method to construct a reduced port-Hamiltonian system that preserves the passivity, and thus the stability of the original system.
-In general these methods do not guarantee the correct distribution of the energy among the energy consuming and energy storing units.
+In the context of network modeling and circuit simulation, a considerable work has been done on developing structure preserving, and in particular energy preserving, model reduction techniques.
+Model reduction for port-Hamiltonian systems are given in \cite{Polyuga:2010gj,beattie2011structure,chaturantabut2016structure} and the references therein.
+These methods use a Krylov or a Proper Orthogonal Decomposition (POD) approach to construct a reduced port-Hamiltonian system that preserves the passivity, and thus the stability of the original system.
+In general, these methods do not guarantee the correct distribution of the energy among the energy consuming and energy storing units.
 Further, over long time integration, accumulation of local errors might produce an erroneous solution.
 
-In this paper we propose the Reduced Dissipative Hamiltonian Method (RDH), a structure preserving model reduction approach for dissipative Hamiltonian systems.
+In this paper, we present the Reduced Dissipative Hamiltonian (RDH) method, a structure preserving model reduction approach for dissipative Hamiltonian systems.
 By considering a canonical heat bath, also known as the hidden strings \cite{Figotin:2006jy,Figotin:2005}, the reduced system is extended to a closed and conservative system.
-Therefore, a symplectic time integrator can guarantee the conservation of the system energy and the dissipated energy.
+Therefore, a symplectic time integrator can guarantee the conservation of the system energy and the correct dissipation of energy.
 Further, the hidden strings assure that the local errors in the dissipation of energy do not accumulate, resulting in a correct evolution of the system energy.
@@
 This paper is organized as follow.
-Section \ref{sec:2} covers the required topics on Hamiltonian systems, dissipative Hamiltonian system and Hamiltonian extension.
-In Section \ref{sec:3} we discussion the symplectic model reduction for Hamiltonian systems and introduce the Reduced Dissipative Hamiltonian Method.
-Accuracy, stability, and efficiency of the RDH method is then discussed in Section \ref{sec:4}, through simulation of the dissipative wave equation and a linear port-Hamiltonian system of an electrical circuit.
+Section \ref{sec:2} covers the required topics on Hamiltonian systems, dissipative Hamiltonian systems, and the Hamiltonian extension.
+In Section \ref{sec:3} we discussion the symplectic model reduction for Hamiltonian systems and introduce the Reduced Dissipative Hamiltonian method.
+Accuracy, stability and efficiency of the RDH method is then discussed in Section \ref{sec:4}, through simulation of the dissipative wave equation and a linear port-Hamiltonian system of an electrical circuit.
 We offer conclusive remarks in Section \ref{sec:5}.

### 2017-03-19 dissipative_hamiltonian/chapters/2.Hamil.tex
first draft finished

--- 2017-03-07 dissipative_hamiltonian/chapters/2.Hamil.tex
@@
 \section{Dissipative Hamiltonian Systems} \label{sec:2}
-In this section we first discuss Hamiltonian systems. Add more later
+In this section, we first summarize basic concepts in the geometry of symplectic linear vector space.
+Then, we discuss Hamiltonian systems under dissipative forces and introduce the Hamiltonian extension of dissipative Hamiltonian systems.
 
-\subsection{Hamiltonian Systems} Suppose that $(\mathbb{R}^{2n},\Omega)$ is a symplectic linear vector space, where $\mathbb{R}^{2n}$ is a configuration space and $\Omega:\mathbb{R}^{2n}\times \mathbb{R}^{2n} \to \mathbb R$ is a closed, skew-symmetric and non-degenerate 2-form on $\mathbb{R}^{2n}$.
+\subsection{Hamiltonian Systems} \label{sec:2.1} Suppose that $(\mathbb{R}^{2n},\Omega)$ is a symplectic linear vector space, where $\mathbb{R}^{2n}$ is a configuration space and $\Omega:\mathbb{R}^{2n}\times \mathbb{R}^{2n} \to \mathbb R$ is a closed, skew-symmetric and non-degenerate 2-form on $\mathbb{R}^{2n}$.
 Given a smooth Hamiltonian function $H:\mathbb{R}^{2n}\to \mathbb R$, \emph{Hamilton}'s equations of evolution are given as
@@
 \end{equation}
-where $z \in\mathbb R^{2n}$ is the configuration coordinates and $\mathbb J_{2n}$ is a $2n\times 2n$ matrix such that $\Omega(x,y) = x^T \mathbb J_{2n} y$ \cite{Marsden:2010:IMS:1965128}.
+where $z \in\mathbb R^{2n}$ is the configuration coordinates and $\mathbb J_{2n}$ is a $2n\times 2n$ matrix such that $\Omega(x,y) = x^T \mathbb J_{2n} y$, for all $x,y\in \mathbb R^{2n}$ \cite{Marsden:2010:IMS:1965128}.
 One by using the \emph{Symplectic Gram-Schmidt} \cite{de2006symplectic} can construct a coordinate system in which $\mathbb J_{2n}$ takes the form
@@
 
-In many physical problems $H$ represents the system energy and is bounded below.
-Here, we mostly assume that $H$ is a quadratic Hamiltonian, i.e., it takes the form $H(z) = \frac 1 2 z^T K^T K z$, where $K$ is a full rank $2n\times 2n$ matrix.
+In many physical problems, $H$ represents the system energy and is bounded below.
+Here, to avoid difficulties with well-posedness of Hamiltonian systems,  we mostly assume that $H$ is a quadratic Hamiltonian, i.e., it takes the form $H(z) = \frac 1 2 z^T K^T K z$, where $K$ is a full rank $2n\times 2n$ matrix.
 This assumption leads to a linear system of evolution (\ref{eq:2.1}).
-Note that the Hamiltonian extension in section ??
-and the model reduction for dissipative Hamiltonian systems in section ??
-can be naturally extended to Hamiltonians of the form $H(z) = \frac 1 2 z^T K^T K z + g(z)$, where $g:\mathbb R^{2n} \to \mathbb R$ is an arbitrary function of $z$.
+Note that the Hamiltonian extension in section \ref{sec:2.2} and the Reduced Dissipative Hamiltonian method in section \ref{sec:3.3} can be naturally extended to Hamiltonians of the form $H(z) = \frac 1 2 z^T K^T K z + g(z)$, where $g:\mathbb R^{2n} \to \mathbb R$ is an arbitrary function of $z$.
 
-Under general coordinate changes, equations of evolution do not necessarily take the form of (\ref{eq:2.1}).
+Under general coordinate transformations, the equations of evolution do not necessarily take the form of (\ref{eq:2.1}).
 Let $(\mathbb R^{2n},\Omega)$ and $(\mathbb R^{2k},\Lambda)$ be two symplectic linear vector spaces.
@@
 where the superscript $T$ represents the transpose operator.
-As the symplectic 2-form is preserved under symplectic transformations, the form of the equations of evolution remains invariant \cite{Marsden:2010:IMS:1965128}.
-The \emph{symplectic inverse} of $A$, is a form of pseudo-inverse given by
+As the symplectic 2-form is preserved under a symplectic transformation, the form of the equations of evolution remains invariant through a symplectic coordinate change  \cite{Marsden:2010:IMS:1965128}.
+The \emph{symplectic inverse} of $A$, is a pseudo-inverse given by
 \begin{equation}
@@
 \end{equation}
-It is shown in \cite{Peng:2014di} that $A^+A = I_{2k}$ and that $(A^+)^T$ is a symplectic matrix.
-Further, one can easily check that $AA^+$ is idempotent and so is a projection operator onto the column span of $A$.
+It is shown in \cite{Peng:2014di} that $A^+A = I_{2k}$ and that $(A^+)^T$ are symplectic matrices.
+Further, one can easily check that $AA^+$ is idempotent, meaning it is a projection operator onto the column span of $A$.
 
@@
 Under some conditions on a symplectic space \cite{da2003introduction}, one can construct a symplectic basis which is also ortho-normal, and thus norm bounded.
-A basis which is both symplectic and orthogonal is referred to as an \emph{ortho-symplectic basis}.
+A basis which is both symplectic and orthogonal is called a \emph{ortho-symplectic basis}.
 We refer the reader to \cite{da2003introduction} for conditions on existence and construction of an ortho-symplectic basis.
@@
 \end{equation}
-where $p$ and $q$ are the canonical coordinates. More information on construction and application of symplectic integrators can be found in \cite{Hairer:1250576}.
+where $p$ and $q$ are the canonical coordinates $z = (q^T,p^T)^T$. More information on construction and application of symplectic integrators can be found in \cite{Hairer:1250576}.
 
-\subsection{Dissipative Hamiltonian Systems and Hamiltonian Extensions}
+\subsection{Dissipative Hamiltonian Systems and Hamiltonian Extensions} \label{sec:2.2}
 
 Many systems in engineering and physic appear as a perturbation of a Hamiltonian system, where the perturbation is regarded as dissipation.
-In these systems, the energy tends to decrease over time evolution, and the conservation law in Theorem \ref{theorem:2.1} does not hold anymore.
+In these systems, the energy tends to decrease over time evolution, and thus, the conservation law in Theorem \ref{theorem:2.1} does not hold anymore.
 Therefore, it is common to take the conservation of energy as a fundamental principle and consider the dissipative system coupled with a heat bath that absorbs the dissipated energy of the original system.
@@
 Here $\chi:\mathbb R^+\to \mathbb R^{2n\times 2n}$ is a bounded matrix valued function with respect to the Frobenius norm and is called the \emph{general susceptibility}.
-Note that integral term in (\ref{eq:2.4.1}) corresponds to the dissipation, whereas if $\chi(s) = 0$ then (\ref{eq:2.3}) is equivalent to (\ref{eq:2.1}).
-Note that under suitable assumptions on $K$, both the systems (\ref{eq:2.1}) and (\ref{eq:2.4}) are well-posed \cite{Figotin:2006jy}.
+Note that the integral term in (\ref{eq:2.4.1}) corresponds to the dissipation, whereas if $\chi(s) = 0$ then (\ref{eq:2.4}) is equivalent to (\ref{eq:2.1}).
+Furthermore, under suitable assumptions on $K$, both the systems (\ref{eq:2.1}) and (\ref{eq:2.4}) are well-posed \cite{Figotin:2006jy}.
 
@@
 
-It is shown in \cite{Figotin:2006jy,Figotin:2005} that under natural assumptions on a linear susceptibility $\chi(t)$ (see below), one can couple a TDD system of the form (\ref{eq:2.4}) with a canonical heat bath where the dissipated energy is captured in the heat bath in a canonical sense.
+It is shown in \cite{Figotin:2006jy,Figotin:2005} that under natural assumptions on the linear susceptibility $\chi(t)$ (see below), one can couple a TDD system of the form (\ref{eq:2.4}) with a canonical heat bath where the dissipated energy is captured in the heat bath in a canonical sense.
 In other words, one can construct a Hilbert space $\mathcal H$ and an isometric injection $I:\mathbb R^{2n} \to \mathbb R^{2n}\times \mathcal H^{2n}$ where the solution $z$ to (\ref{eq:2.4}) is the projection of $x$ onto $\mathbb R^{2n}$, where $x$ is the solution to
 \begin{equation} \label{eq:2.7}
-	\frac{d}{dt} x = J_{2n} \frac{\delta h}{\delta x}.
+	\frac{d}{dt} x = \mathcal J_{2n} \frac{\delta H_{\text{ex}}}{\delta x}.
 \end{equation}
-Here $J_2n$ is the symplastic operator and $h:\mathbb R^{2n}\times \mathcal H^{2n} \to \mathbb R$ is an extended quadratic Hamiltonian function defined on $\mathbb R^{2n}\times \mathcal H^{2n}$.
+Here $H_{\text{ex}}:\mathbb R^{2n}\times \mathcal H^{2n} \to \mathbb R$ is an extended quadratic Hamiltonian function and $\mathcal J_{2n}$ is the symplastic operator defined on $\mathbb R^{2n}\times \mathcal H^{2n}$ respectively.
 
 \begin{theorem}
-Suppose that $K$ is full ranked and $\chi(t)$ is symmetric. There exists a quadratic extension of the form (\ref{eq:2.8}) to (\ref{eq:2.4}), if
+Suppose that $K$ is full ranked and $\chi(t)$ is symmetric. There exists a quadratic extension of the form (\ref{eq:2.7}) to (\ref{eq:2.4}), if
 \begin{equation} \label{eq:2.8}
@@
 \end{equation}
-To show that the Hamiltonian system (\ref{eq:2.10.a}) to (\ref{eq:2.10.c}) is an extension to (\ref{eq:2.4}) in the sense discussed before, it is enough to show that the solution $f$ to the equation (\ref{eq:2.11}) also satisfies (\ref{eq:2.4.1}).
-Equations (\ref{eq:2.10.b}) and (\ref{eq:2.10.c}) are equations to a vibrating string, and can be solved analytically
+To show that the Hamiltonian system (\ref{eq:2.10.a}-\ref{eq:2.10.c}) is an extension to (\ref{eq:2.4}) in the sense discussed above, it is enough to show that the solution $f$ to the equation (\ref{eq:2.11}) also satisfies (\ref{eq:2.4.1}).
+Equations (\ref{eq:2.10.b}-\ref{eq:2.10.c}) are equations to a vibrating string, and can be solved analytically
 \begin{equation} \label{eq:2.12}
@@
 \end{proof}
-We can recover (\ref{eq:2.4.1}) by substituting (\ref{eq:2.12}) into (\ref{eq:2.11}).
-The extended Hamiltonian $H_\text{ex}$ for the system (\ref{eq:2.10.a}) to (\ref{eq:2.10.c}) takes the quadratic from
+Then, we can recover (\ref{eq:2.4.1}) by substituting (\ref{eq:2.12}) into (\ref{eq:2.11}).
+The extended Hamiltonian $H_\text{ex}$ for the system (\ref{eq:2.10.a}-\ref{eq:2.10.c}) takes the quadratic from
 \begin{equation} \label{eq:2.13}
@@
 
-Equations (\ref{eq:2.10.a}) and (\ref{eq:2.10.c}) are called the \emph{hidden strings}.
+Equations (\ref{eq:2.10.b}) and (\ref{eq:2.10.c}) are called the \emph{hidden strings}.
 The dissipation of energy in the original system (\ref{eq:2.4}) is carried away, as vibrations, along the added strings which makes the extended system conservative.
-Hamiltonian extension of the damped harmonic oscillator in Example \ref{example:2.1} is exactly the Lamb model \cite{lamb:1900}.
+The Hamiltonian extension of the damped harmonic oscillator in Example \ref{example:2.1} is exactly the Lamb model \cite{lamb:1900}.
 The Lamb model is a harmonic oscillator coupled with a vibrating string, and the tension in the string causes linear dissipation in the dynamics of the harmonic oscillator.
 
-Note that time integration of (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) involves the integration of $f$ in (\ref{eq:2.12}).
-In general $f(t)$ must be stored and may cause serious storage limitation in long-time integration.
-However, we are interested solely in finding $z(t)$ which depends on $f(t)$, and $\phi(t,0)$ that is the integral of the history of $f(t)$.
+Note that the time integration of (\ref{eq:2.10.a}-\ref{eq:2.10.c}) involves the integration of $f$ in (\ref{eq:2.12}).
+In general, the history of $f(t)$ must be stored and may cause serious storage limitation in long-time integration.
+However, we are interested solely in finding $z(t)$ which depends on $f$ at time $t$, and $\phi(t,0)$, that is the integral of the history of $f$.
 So by carefully choosing a quadrature rule that uses the same quadrature nodes as the time integrator we can avoid storing the history of $f$.
@@
 \end{equation}
-for some quadrature weights $\omega_i$, $i=1,\dots,k$ with $k\ll n$. Thus, time integration of (\ref{eq:2.10.a})-(\ref{eq:2.10.c}), requires storing only $k$ evaluations of $f$.
+for some quadrature weights $\omega_i$, $i=1,\dots,k$ with $k\ll n$. Thus, time integration of (\ref{eq:2.10.a}-\ref{eq:2.10.c}), requires storing only $k$ evaluations of $f$.

### 2017-03-19 dissipative_hamiltonian/chapters/3.model.tex
first draft finished

@@
 In this section we first explain the main results of \cite{Peng:2014di} regarding model reduction of Hamiltonian systems.
-In section ??
-we introduce the model reduction of dissipative Hamiltonian systems.
+In section \ref{sec:3.3} we introduce the Reduced Dissipative Hamiltonian method.
 
-\subsection{Symplectic Model Order Reduction}
+\subsection{Symplectic Model Order Reduction} \label{sec:3.1}
 Consider a Hamiltonian system of the form (\ref{eq:2.1}) together with a quadratic Hamiltonian of the form $H(z) = \frac 1 2 z^T K^T K z$.
@@
 As $(A^+)^T$ is a symplectic matrix it yields that $A^+ \mathbb{J}_{2n} (A^+)^T = \mathbb{J}_{2k}$.
-By defining the reduced Hamiltonian $\tilde H : \mathbb{R}^{2k} \to \mathbb R$, defined as $\tilde H (y) = H(Ay)$, we obtain the reduced system
-\begin{equation} \label{eq:3.4}
+By defining the reduced Hamiltonian $\tilde H : \mathbb{R}^{2k} \to \mathbb R$, defined as $\tilde H (y) = H(Ay)$, we recover the reduced system
+\begin{equation} \label{eq:3.3.1}
 	\begin{aligned}
@@
 \end{equation}
-Equation (\ref{eq:3.4}) is called the \emph{symplectic Galerkin projection} \cite{Peng:2014di} of the Hamiltonian system (\ref{eq:2.1}).
+Equation (\ref{eq:3.3.1}) is called the \emph{symplectic Galerkin projection} \cite{Peng:2014di} of the Hamiltonian system (\ref{eq:2.1}).
 Conventional model reduction routines, e.g.
@@
 Hence, the reduced system does not necessarily preserve the conservation law in theorem \ref{theorem:2.1}.
-This results in a qualitatively wrong and often unstable solution for the reduced system {\color{red} put reference here}.
-On the other hand the reduced system obtained by the symplectic Galerkin projection, is a Hamiltonian system, therefore system energy is preserved over time integration \cite{Peng:2014di}.
+This results in a qualitatively wrong and often unstable solution \cite{Peng:2014di}.
+On the other hand the reduced system obtained by the symplectic Galerkin projection, is a Hamiltonian system, therefore the system energy is preserved over time integration \cite{Peng:2014di}.
+The following theorem from \cite{Peng:2014di} guarantees the boundedness of the solution of the reduced system obtained by the symplectic Galerkin projection.
 
-\begin{definition}
-\cite{bhatia2002stability} {\color{red} discuss stability}.
-\end{definition}
+\begin{theorem}
+\cite{bhatia2002stability} Let $S$ be a bounded open subset of $\mathbb R^{2n}$ that contain $z_0$.
+Further assume that $H(z_0)<H(z)$ or $H(z_0)>H(z)$ for all $z\in \partial S$ the boundary of $S$.
+For a given symplectic basis $A$, if $z_0$ is in the range of $A$, then both the original system and the reduced system obtained by the symplectic Galerkin projection are bounded.
+\end{theorem}
+
+\begin{proof}
+The proof is a direct consequence of the conservation law in theorem \ref{theorem:2.1} and the fact that the image of S under the projection $AA^+$ remains open and bounded.
+\end{proof}
 
@@
 
-\subsection{Proper Symplectic Decomposition (PSD)} Consider the set $S = \{z(t_i)| t_i\in[0,T],\ i=1,\dots,N\}$, of samples of the solution to (\ref{eq:2.1}), referred to as the \emph{snapshots}.
-The PSD, requires a symplectic basis $A$ of size $2k$ to have the minimum symplectic projection error onto the space space of $S$ .
+\subsection{Proper Symplectic Decomposition (PSD)} \label{sec:3.2} Consider the set $S = \{z(t_i)| t_i\in[0,T],\ i=1,\dots,N\}$, of samples of the solution to (\ref{eq:2.1}), referred to as the \emph{snapshots}.
+The PSD, requires a symplectic basis $A$ of size $2k$ to have the minimum symplectic projection error over $S$ .
 In other words, a PSD symplectic basis is the solution to the optimization problem
@@
 
-\subsection{The Reduced Dissipative Hamiltonian Method}
+\subsection{The Reduced Dissipative Hamiltonian Method} \label{sec:3.3}
 
-The fundamental dynamics of a dissipative Hamiltonian systems are based on a Hamiltonian formulation.
-Thus, over the course of model reduction, we expect that the evolution of energy is not correctly translated to the reduce system.
-This is especially expected in a near-zero dissipation regime.
+Since the theory of symplectic model reduction in section \ref{sec:3.2} is based on the conservation law in theorem \ref{theorem:2.1}, it can no longer be applied to dissipative Hamiltonian systems.
+We expect that for dissipative systems, the evolution of energy is not correctly translated into the reduced system.
+The Reduced Dissipative Hamiltonian method considers a Hamiltonian extension to a dissipative Hamiltonian system to construct a closed system.
+A symplectic model reduction can then be naturally applied to conserve the total energy.
 
-On the other hand, as the conservation law in theorem \ref{theorem:2.1} does not hold for a general dissipative Hamiltonian system, the symplectic model reduction discussed in section \ref{} cannot be applied directly.
-Therefore, we need to seek alternative methods to transfer the correct evolution of energy into the reduced system.
-
-Consider an extended Hamiltonian system of the form (\ref{eq:2.10.a}) with a quadratic Hamiltonian, $H(z) = z^TK^TKz$.
+Consider a dissipative Hamiltonian system of the form (\ref{eq:2.4}) with a quadratic Hamiltonian, $H(z) = z^TK^TKz$.
 Since $K^TK$ is symmetric and positive definite, it has a unique Cholesky factorization $K^TK = L^T L$ where $L$ is upper triangular \cite{strang09}.
@@
 \end{equation}
-Further suppose that the solution $z(t)$ to the original TDD equation lies on a low-dimensional symplectic subspace such that $z = Ay$, where $A$ is an ortho-symplectic matrix of the size $2n\times 2k$ and $y$ is the expansion coefficients of $z$ in the basis of $A$.
+Further, suppose that the solution $z(t)$ lies on a low-dimensional symplectic subspace such that $z = Ay$, where $A$ is an ortho-symplectic matrix of the size $2n\times 2k$ and $y$ is the expansion coefficients of $z$ in the basis of $A$.
 The Hamiltonian $H$ then takes the form
@@
 \end{equation}
-Since $A^T L^T L A$ is a symmetric and positive definite matrix, it has a unique Cholesky factorization $\tilde L^T \tilde L$ where $\tilde L = A^T L A$ is a upper triangular matrix of size $2k \times 2k$.
-Writing (\ref{eq:2.10.a}) in terms of the reduced coordinates $y$ and the Hamiltonian (\ref{eq:3.7}) reads
+Since $A^T L^T L A$ is a symmetric and positive definite matrix, it has a unique Cholesky factorization $\tilde L^T \tilde L$ where $\tilde L = A^T L A$ is an upper triangular matrix of size $2k \times 2k$.
+Writing (\ref{eq:2.4}) in terms of the reduced coordinates $y$ and the Hamiltonian (\ref{eq:3.7}) reads
 \begin{equation} \label{eq:3.9}
@@
 \end{equation}
-Therefore, the system energy will be conserved along integral curves of (\ref{eq:3.12.a}) to (\ref{eq:3.12.c}).
+Therefore, the system energy will be conserved along integral curves of (\ref{eq:3.12.a}-\ref{eq:3.12.c}).
 
-We point out that the transformation that transforms (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) to (\ref{eq:3.12.a})-(\ref{eq:3.12.c}) is given by
+We point out that the transformation that transforms (\ref{eq:2.10.a}-\ref{eq:2.10.c}) to (\ref{eq:3.12.a}-\ref{eq:3.12.c}) is given by
 \begin{equation}
@@
 Further, the dissipation of energy in the reduced system only depends on the reduced susceptibility.
-Thus, the choice of $A$ should be independent of the hidden strings $(\tilde \phi, \tilde \theta)$.
+Thus, the choice of $A$ should be independent of the hidden strings $(\phi, \theta)$.
 In other words, if the reduced space is chosen to be a symplectic subspace, then the actions of model reduction and Hamiltonian extension commute.
@@
 \begin{enumerate}
-	\item Construct the Hamiltonian extension (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) to the original TDD system (\ref{eq:2.4}).
+	\item Construct the Hamiltonian extension (\ref{eq:2.10.a}-\ref{eq:2.10.c}) to the original TDD system (\ref{eq:2.4}).
 	\item Collect the snapshots $z(t_i)$, $i=1,\dots,N$ through time integration of the extended Hamiltonian.
 	\item Construct an ortho-symplectic basis $A$.
-	\item Define $\tilde L = A^T L A$, $\tilde \chi = A^T \chi A$ and construct the reduced dissipative Hamiltonian system (\ref{eq:3.12.a})-(\ref{eq:3.12.c})
+	\item Define $\tilde L = A^T L A$, $\tilde \chi = A^T \chi A$ and construct the reduced dissipative Hamiltonian system (\ref{eq:3.12.a}-\ref{eq:3.12.c})
 \end{enumerate}

### 2017-03-19 dissipative_hamiltonian/chapters/4.numer.tex
first draft finished

@@
 \begin{equation} \label{eq:4.4}
-\begin{aligned}
-	& \dot z = \mathbb J_{2n} K^T f(t) \\
-	& f(t) + R \int_0^t f(t) = K z.
-\end{aligned}
+	\dot z = \mathbb J_{2n} K^T f(t), \quad f(t) + R \int_0^t f(t) = K z.
 \end{equation}
@@
 \begin{equation} \label{eq:4.5}
-	q_i(0) = h( 10\times|x_i - \frac{1}{2}| ), \quad p_i = 0, \quad i=1,\dots,N
+	q_i(0) = h( 10\times|x_i - \frac{1}{2}| ), \quad p_i = 0, \quad i=1,\dots,N,
 \end{equation}
@@
 For the numerical time integration of the extended Hamiltonian system, the Str\"omer-Verlet time stepping scheme (\ref{eq:2.3}) is used.
-Further in each time step, the system of linear equations (\ref{eq:2.4.1}) is solved to recover $z$.
+Further in each time step, the system of linear equations (\ref{eq:2.11}) is solved to recover $z$.
 System parameters are summarized below.
@@
 Space discretization size & $\Delta x = 0.002$ \\
-Time discretization size & $\Delta t = 0.01$ \\
+Time discretization size & $\Delta t = 0.002$ \\
 Wave speed & $c^2 = 0.1$ \\
@@
 To construct the reduced basis for different methods, we gather the snapshots of (\ref{eq:4.1}).
-We construct RDH reduced system according to the Algorithm \ref{alg:3.1}.
+We construct the RDH reduced system according to the Algorithm \ref{alg:3.1}.
 The cotangent lift is used to construct an ortho-symplectic reduced basis $A$ (\ref{eq:3.5}).
@@
 
-The $L^2$ error between the full systems and the RDH, the PSD and the POD method are presented in Figure \ref{fig:4.1}.c.
+The $L^2$ error between the full systems and the RDH, the PSD and the POD methods are presented in Figure \ref{fig:4.1}.c.
 We notice that symplectic methods provide a more accurate solution with respect to the POD method.
@@
 On the other hand, We notice that the RDH method with 40 modes provides a significantly more accurate solution compared to the PSD method with 60 modes.
-The RDH method provides a conservative reduced system where the dissipated energy is trapped in the hidden strings.
+The RDH method provides a conservative reduced system where the dissipated energy is absorbed by the hidden strings.
 The conservation of energy is then guaranteed using a symplectic integrator.
@@
 ``FM'' and ``RM'' stand for the full model and the reduced model respectively.
-(f) The $L^2$ error between the solution to the reduced system and the full in a near-zero dissipation regime.} \label{fig:4.1}
+(f) The $L^2$ error between the solution to the reduced system and the full system in a near-zero dissipation regime.} \label{fig:4.1}
 \end{figure}
@@
 \end{equation}
-The extended Hamiltonian formulation (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) with a quadratic Hamiltonian $H_{\text{ex}}$ can be then carried out according to section \ref{??}.
+The extended Hamiltonian formulation (\ref{eq:2.10.a}-\ref{eq:2.10.c}) with a quadratic Hamiltonian $H_{\text{ex}}$ can be then carried out according to section \ref{sec:2.2}.
 We note that due to the input $\tilde u$, the Hamiltonian $H_{\text{ex}}$ is time dependent.
@@
 
-Using an ortho-symplectic reduced basis $A$ The reduced dissipative Hamiltonian method can be applied to (\ref{eq:4.9}) to construct a reduced system of the form (\ref{eq:3.12.a})-(\ref{eq:3.12.c}) together with the extended Hamiltonian $\tilde H_{\text{ex}}$.
+Using an ortho-symplectic reduced basis $A$ The Reduced Dissipative Hamiltonian method can be applied to (\ref{eq:4.9}) to construct a reduced system of the form (\ref{eq:3.12.a}-\ref{eq:3.12.c}) together with the extended Hamiltonian $\tilde H_{\text{ex}}$.
 We immediately notice that $\frac{d}{dt} \tilde H_{\text{ex}} = (A^+ \tilde u)^T A^T \tilde Q^T \tilde Q A y$, meaning that the reduced system is passive.
@@
 We take $C_i=1$, $L_i = 1$, $R_i=0.2$ for $i=1,\dots,50$, and $R_{51} = 0.4$.
-The cotangent lift is used to construct a reduced basis and the RDH method is used to construct a reduced system according to Algorithm (\ref{alg:3.1}).
+The cotangent lift is used to construct a reduced basis and the RDH method is used to construct a reduced system according to Algorithm \ref{alg:3.1}.
 
@@
 We observe that although the moment matching method is bounded, over long-time integration, the RDH method provides a significantly more accurate solution.
-This is because in the moment matching method, passivity of the reduced system implies that the energy of the system will be bounded by the input energy.
+This is because in the moment matching method, the passivity of the reduced system implies that the energy of the system will be bounded by the input energy.
 However, there is no guarantee that the dissipation of energy in the reduced system mimics the one of the original system.
@@
 
-Figure \ref{fig:4.3}.a and Figure \ref{fig:4.3}.b show the average error in the charge and flux of the capacitors and inductors respectively, for these methods.
+Figure \ref{fig:4.3}.a and Figure \ref{fig:4.3}.b show the average error in the charge and flux of the capacitors and inductors respectively.
 The RDH method provides a significantly better accuracy compared to the moment matching method.
-This is because the charge of $C_1$ is specified as the output of interest in the moment matching method so it is expected that that method provides low accuracy for computing other outputs.
+This is because the charge of $C_1$ is specified as the output of interest in the moment matching method and so it is expected that that method provides low accuracy for computing other outputs.
 On the other hand, the RDH method not only provides high accuracy in computing the charge for $C_1$ but also it has high accuracy for computing other outputs of interest.

### 2017-03-19 dissipative_hamiltonian/chapters/5.conc.tex
first draft finished

@@
 
-Here comes the conclusion
+In this paper we present the Reduced Dissipative Hamiltonian method.
+The method preserves the symplectic structure of dissipative Hamiltonian systems and guarantees the correct dissipation of energy through time integration.
+The canonical heat bath, coupled with the full system, constructs a closed system of evolution, therefore a symplectic time integrator can be applied to preserve the system energy and stability over-long time integration.
+Moreover, the method allows the usage of a generic ortho-symplectic basis for model reduction.
+
+The numerical simulations illustrate that the RDH method can preserve the system energy with significant more accuracy than other methods.
+Further they show that the hidden strings assure that the dissipation of energy in the reduce system mimics the dissipation of energy in the full system.
+This ensures that the local error do not accumulate over long-time integration.
+
+All in all, the Reduced Dissipative Hamiltonian method is a promising model reduction technique for dissipative Hamiltonian systems and can save significant computational cost while preserving the symplectic structure and the dissipation structure of the full system.

### 2017-03-19 dissipative_hamiltonian/paper.tex
first draft finished

@@
 This results in a correct dissipation of energy, preservation of total energy and ultimately in the stability of the solution.
-Accuracy and stability of the method is illustrated through the numerical simulation of the dissipative wave equation and a port-Hamiltonian model of an electric circuit.
+Accuracy and stability of the method are illustrated through the numerical simulation of the dissipative wave equation and a port-Hamiltonian model of an electric circuit.
 
-\keywords{First keyword \and Second keyword \and More}
+\keywords{Model order reduction \and Symplectic model reduction \and The Reduced Dissipative Hamiltonian method}
 % \PACS{PACS code1 \and PACS code2 \and more}

### 2017-03-25 dissipative_hamiltonian/chapters/1.intro.tex
first edit of the paper

@@
 
-The need for more accuracy has led to ever more complex models and systems of partial differential equations in engineering and physics.
-As a consequence, direct numerical methods for solving PDEs have become inefficient due to limitations in computational power and resources.
+The need for increased accuracy has led to more complex models and the use of large systems of partial differential equations in engineering and science.
+As a consequence, direct numerical methods for solving PDEs have become computationally demanding and, at times impractical.
 
-Over the past decade, reduced basis methods have emerged as a popular approach to reduce the complexity of large systems of partial differential equations \cite{Ito:1998up,Ito:1998ch,Ito:2001ev}.
-These methods construct a low-dimensional subspace, the reduced space, that approximately represent the solution to the system of differential equations.
+During the past decade, reduced basis methods have emerged as a powerful approach to reduce the cost of evaluating large systems of partial differential equations \cite{Ito:1998up,Ito:1998ch,Ito:2001ev}.
+These methods construct a low-dimensional linear subspace, the reduced space, that approximately represents the solution to the system of differential equations.
 The projection of the original system onto the reduced space then allows the exploration of the solution with a significantly reduced computational complexity \cite{hesthaven2015certified,quarteroni2015reduced}.
 
-Hamiltonian systems are an important class of systems that appear in engineering and physics.
-In these systems, preserving system energy is essential in a correct numerical solution.
-Therefore, developing model reduction techniques that preserve the symplectic symmetry is essential.
-The classical reduced basis methods, do not generally preserve the conservation laws and intrinsic symmetries of Hamiltonian systems \cite{Amsallem:2014ef,prajna2003pod}.
+Hamiltonian systems are an important class of systems that appear in engineering and science.
+In these systems, preserving system energy is essential to obtain a correct numerical solution.
+Therefore, the development of model reduction techniques that preserve the symplectic symmetry is crucial.
+However, the classical reduced basis methods do not generally preserve the conservation laws and intrinsic symmetries of Hamiltonian systems \cite{Amsallem:2014ef,prajna2003pod}.
 This often results in an unstable or a qualitatively wrong solution.
 
-It is demonstrated by \cite{Lall:2003iy,Carlberg:2014ky,Peng:2014di}, that if the basis for the reduced space is not chosen carefully, the symplastic symmetry of Hamiltonian and Lagrangian systems will be destroyed over the course of model reduction.
-To resolve this issue, Peng et al.
-\cite{Peng:2014di} and Lall et al.
-\cite{Lall:2003iy} construct a reduced ordered configuration space that inherits symmetries of the full configuration space.
-Then, by using a proper time integrator scheme, the symmetries are preserved in the reduced system.
+It is demonstrated in \cite{Maboudi:2016,Lall:2003iy,Carlberg:2014ky,Peng:2014di}, that if the basis for the reduced space is not chosen carefully, the symplectic symmetry of Hamiltonian and Lagrangian systems will be destroyed by model reduction.
+To resolve this issue, in \cite{Maboudi:2016,Peng:2014di,Lall:2003iy} a reduced order configuration space is constructed that inherits symmetries of the full configuration space.
+By using a proper time integrator scheme, the symmetries are preserved in the reduced system.
+A greedy-type algorithem is developed in \cite{Maboudi:2016} for construction of a basis for such a reduced configuration space.
 
-Most models in engineering, appear as a dissipative perturbation of a Hamiltonian system.
-In these systems, conservation of energy is taken as a fundamental principle of the system dynamics, while deriving and dissipative forces, e.g.
-friction, can change the energy level of the system \cite{vanderSchaft:2014:PST:2693645.2693646}.
-As the system energy is no longer preserved for such systems, the methods mentioned above can no longer be applied directly \cite{peng2016geometric}.
+Most models in engineering appear as a dissipative perturbation of a Hamiltonian system.
+In these systems, conservation of energy is taken as a fundamental principle of the system dynamics, while dissipative forces, e.g.
+friction, can change the energy of the system \cite{vanderSchaft:2014:PST:2693645.2693646}.
+As the energy is no longer preserved for such systems, existing methods can no longer be applied directly \cite{peng2016geometric}.
 
 For dissipative and forced Hamiltonian system, Peng et al.
-\cite{peng2016geometric} suggest a symplectic model reduction method that preserves Hamiltonian and the dissipative structure of the original system.
+\cite{peng2016geometric} suggest a symplectic model reduction method that preserves the Hamiltonian and the dissipative structure of the original system.
 However, since this method uses a symplectic integrator for a non-conservative system, there is no guarantee that the evolution of the energy is translated correctly to the reduced system.
 
-In the context of network modeling and circuit simulation, a considerable work has been done on developing structure preserving, and in particular energy preserving, model reduction techniques.
+In the context of network modeling and circuit simulation, considerable work has been done in the development of structure preserving, and in particular energy preserving, model reduction techniques.
 Model reduction for port-Hamiltonian systems are given in \cite{Polyuga:2010gj,beattie2011structure,chaturantabut2016structure} and the references therein.
-These methods use a Krylov or a Proper Orthogonal Decomposition (POD) approach to construct a reduced port-Hamiltonian system that preserves the passivity, and thus the stability of the original system.
-In general, these methods do not guarantee the correct distribution of the energy among the energy consuming and energy storing units.
-Further, over long time integration, accumulation of local errors might produce an erroneous solution.
+These methods use a Krylov or a Proper Orthogonal Decomposition (POD) approach to construct a reduced port-Hamiltonian system that preserves the passivity, and, thus, the stability of the original system.
+However, these methods do not generally guarantee the correct distribution of the energy among the energy consuming and energy storing units.
+Furthermore, over long time integration, accumulation of local errors might produce an erroneous solution.
 
-In this paper, we present the Reduced Dissipative Hamiltonian (RDH) method, a structure preserving model reduction approach for dissipative Hamiltonian systems.
-By considering a canonical heat bath, also known as the hidden strings \cite{Figotin:2006jy,Figotin:2005}, the reduced system is extended to a closed and conservative system.
-Therefore, a symplectic time integrator can guarantee the conservation of the system energy and the correct dissipation of energy.
-Further, the hidden strings assure that the local errors in the dissipation of energy do not accumulate, resulting in a correct evolution of the system energy.
+In this paper, we present the Reduced Dissipative Hamiltonian (RDH) method as a structure preserving model reduction approach for dissipative Hamiltonian systems.
+A key difference between this method and the other existing methods is that the RDH enables the reduced system to be integrated using a symplectic integrator.
+By considering a canonical heat bath, also known as hidden strings \cite{Figotin:2006jy,Figotin:2005}, the reduced system is extended to a closed and conservative system.
+Therefore, a symplectic time integrator can be used to guarantee conservation of the system energy and the correct dissipation of energy.
+Furthermore, the hidden strings assure that the local errors in the dissipation of energy do not accumulate, resulting in a correct evolution of the system energy.
 
-This paper is organized as follow.
-Section \ref{sec:2} covers the required topics on Hamiltonian systems, dissipative Hamiltonian systems, and the Hamiltonian extension.
+What remains of this paper is organized as follow.
+Section \ref{sec:2} covers the required background on Hamiltonian systems, dissipative Hamiltonian systems, and the Hamiltonian extension.
 In Section \ref{sec:3} we discussion the symplectic model reduction for Hamiltonian systems and introduce the Reduced Dissipative Hamiltonian method.
-Accuracy, stability and efficiency of the RDH method is then discussed in Section \ref{sec:4}, through simulation of the dissipative wave equation and a linear port-Hamiltonian system of an electrical circuit.
+Accuracy, stability and efficiency of the RDH method is discussed in Section \ref{sec:4}, and illustrated through simulation of the dissipative wave equation and a linear port-Hamiltonian system of an electrical circuit.
 We offer conclusive remarks in Section \ref{sec:5}.

### 2017-03-25 dissipative_hamiltonian/chapters/2.Hamil.tex
first edit of the paper

@@
 \section{Dissipative Hamiltonian Systems} \label{sec:2}
-In this section, we first summarize basic concepts in the geometry of symplectic linear vector space.
-Then, we discuss Hamiltonian systems under dissipative forces and introduce the Hamiltonian extension of dissipative Hamiltonian systems.
+We first summarize needed concepts around the geometry of a symplectic linear vector space and then discuss Hamiltonian systems subject to dissipative forces.
+Finally, we introduce the Hamiltonian extension of dissipative Hamiltonian systems.
 
@@
 \end{equation}
-where $z \in\mathbb R^{2n}$ is the configuration coordinates and $\mathbb J_{2n}$ is a $2n\times 2n$ matrix such that $\Omega(x,y) = x^T \mathbb J_{2n} y$, for all $x,y\in \mathbb R^{2n}$ \cite{Marsden:2010:IMS:1965128}.
-One by using the \emph{Symplectic Gram-Schmidt} \cite{de2006symplectic} can construct a coordinate system in which $\mathbb J_{2n}$ takes the form
+where $z \in\mathbb R^{2n}$ is the state coordinates and $\mathbb J_{2n}$ is a $2n\times 2n$ matrix such that $\Omega(x,y) = x^T \mathbb J_{2n} y$, for all $x,y\in \mathbb R^{2n}$ \cite{Marsden:2010:IMS:1965128}.
+By using the \emph{Symplectic Gram-Schmidt} \cite{de2006symplectic} method, one can construct a coordinate system in which $\mathbb J_{2n}$ takes the form
 \begin{equation} \label{eq:2.2}
@@
 \end{equation}
-where $I_n$ and $0_n$ are the identity matrix and the zero matrix of size $n$ respectively. A main feature of Hamiltonian systems is the conservation of the Hamiltonian along the integral curves.
+where $I_n$ and $0_n$ are the identity matrix and the zero matrix of size $n$, respectively. A main feature of Hamiltonian systems is the conservation of the Hamiltonian along the integral curves.
 \begin{theorem} \label{theorem:2.1}
@@
 
-In many physical problems, $H$ represents the system energy and is bounded below.
-Here, to avoid difficulties with well-posedness of Hamiltonian systems,  we mostly assume that $H$ is a quadratic Hamiltonian, i.e., it takes the form $H(z) = \frac 1 2 z^T K^T K z$, where $K$ is a full rank $2n\times 2n$ matrix.
+In many physical problems, $H$ represents the system energy and is bounded from below.
+Here, to avoid difficulties with well-posedness of Hamiltonian systems,  we often assume that $H$ is a quadratic Hamiltonian, i.e., it takes the form $H(z) = \frac 1 2 z^T K^T K z$, where $K$ is a full rank $2n\times 2n$ matrix.
 This assumption leads to a linear system of evolution (\ref{eq:2.1}).
-Note that the Hamiltonian extension in section \ref{sec:2.2} and the Reduced Dissipative Hamiltonian method in section \ref{sec:3.3} can be naturally extended to Hamiltonians of the form $H(z) = \frac 1 2 z^T K^T K z + g(z)$, where $g:\mathbb R^{2n} \to \mathbb R$ is an arbitrary function of $z$.
+We emphasise that the Hamiltonian extension in Section \ref{sec:2.2} and the Reduced Dissipative Hamiltonian method in Section \ref{sec:3.3} can be naturally extended to Hamiltonians of the form $H(z) = \frac 1 2 z^T K^T K z + g(z)$, where $g:\mathbb R^{2n} \to \mathbb R$ is an arbitrary function of $z$.
 
-Under general coordinate transformations, the equations of evolution do not necessarily take the form of (\ref{eq:2.1}).
+Under general coordinate transformations, the equations of evolution may not take the form in (\ref{eq:2.1}).
 Let $(\mathbb R^{2n},\Omega)$ and $(\mathbb R^{2k},\Lambda)$ be two symplectic linear vector spaces.
@@
 \end{equation}
-Alternatively in matrix notation, $A\in \mathbb R^{2n\times 2k}$ is called a \emph{symplectic matrix} if
+In matrix notation, $A\in \mathbb R^{2n\times 2k}$ is called a \emph{symplectic matrix} if
 \begin{equation}
@@
 where the superscript $T$ represents the transpose operator.
-As the symplectic 2-form is preserved under a symplectic transformation, the form of the equations of evolution remains invariant through a symplectic coordinate change  \cite{Marsden:2010:IMS:1965128}.
+As the symplectic 2-form is preserved under a symplectic transformation, the form of the equations of evolution remains invariant through a symplectic coordinate transformation \cite{Marsden:2010:IMS:1965128}.
 The \emph{symplectic inverse} of $A$, is a pseudo-inverse given by
@@
 \end{equation}
-It is shown in \cite{Peng:2014di} that $A^+A = I_{2k}$ and that $(A^+)^T$ are symplectic matrices.
-Further, one can easily check that $AA^+$ is idempotent, meaning it is a projection operator onto the column span of $A$.
+It is shown in \cite{Peng:2014di} that $A^+A = I_{2k}$ and that $(A^+)^T$ is a symplectic matrix.
+Furthermore, one easily checks that $AA^+$ is idempotent, i.e.
+it is a projection operator onto the column span of $A$.
 
@@
 
-It is natural to expect a numerical integrator that solves (\ref{eq:2.1}) to also satisfy the conservation law in theorem \ref{theorem:2.1}.
-Common numerical integrators, e.g.
-the Runge-Kutta methos, do not generally preserve the Hamiltonian.
-Symplectic numerical integrators are a class of numerical integrators for Hamiltonian systems that preserves the symplectic structure and ensure stability in long-time integration.
-The Str\"omer-Verlet time stepping scheme is an example of such numerical integrators and is given by
-
+It is natural to expect the numerical integrator that solves (\ref{eq:2.1}) also satisfy the conservation law in Theorem \ref{theorem:2.1}.
+However, common numerical integrators, e.g.
+the Runge-Kutta method, do not generally preserve the Hamiltonian.
+Symplectic numerical integrators are a class of numerical integrators for Hamiltonian systems that preserves the symplectic structure and ensure stability during long-time integration.
+The Str\"omer-Verlet time-stepping scheme
 \begin{equation} \label{eq:2.3}
@@
 \end{equation}
-where $p$ and $q$ are the canonical coordinates $z = (q^T,p^T)^T$. More information on construction and application of symplectic integrators can be found in \cite{Hairer:1250576}.
+where $p$ and $q$ are the canonical coordinates $z = (q^T,p^T)^T$, is an example of such numerical integrators.
+More information on the construction and application of symplectic integrators can be found in \cite{Hairer:1250576}.
 
@@
 
-Many systems in engineering and physic appear as a perturbation of a Hamiltonian system, where the perturbation is regarded as dissipation.
-In these systems, the energy tends to decrease over time evolution, and thus, the conservation law in Theorem \ref{theorem:2.1} does not hold anymore.
+Many systems in engineering and science appear as a perturbation of a Hamiltonian system, where the perturbation can be regarded as dissipation.
+In these systems, the energy tends to decrease over time, and thus, the conservation law in Theorem \ref{theorem:2.1} does not hold.
 Therefore, it is common to take the conservation of energy as a fundamental principle and consider the dissipative system coupled with a heat bath that absorbs the dissipated energy of the original system.
 
-To account dissipation for a quadratic Hamiltonian $H(z) = \frac 1 2 z^T K^T K  z$, we rewrite (\ref{eq:2.1}) as a time dispersive and dissipative (TDD) \cite{Figotin:2006jy} system 
+To account for dissipation in a quadratic Hamiltonian $H(z) = \frac 1 2 z^T K^T K  z$, we rewrite (\ref{eq:2.1}) as a time dispersive and dissipative (TDD) \cite{Figotin:2006jy} system 
 \begin{equation} \label{eq:2.4}
@@
 \end{equation}
-where $f$ is the solution to the integral equation
+where $f$ is the solution to the Volterra integral equation \cite{corduneanu1991integral}
 \begin{equation} \label{eq:2.4.1}
@@
 Here $\chi:\mathbb R^+\to \mathbb R^{2n\times 2n}$ is a bounded matrix valued function with respect to the Frobenius norm and is called the \emph{general susceptibility}.
-Note that the integral term in (\ref{eq:2.4.1}) corresponds to the dissipation, whereas if $\chi(s) = 0$ then (\ref{eq:2.4}) is equivalent to (\ref{eq:2.1}).
-Furthermore, under suitable assumptions on $K$, both the systems (\ref{eq:2.1}) and (\ref{eq:2.4}) are well-posed \cite{Figotin:2006jy}.
+Note that the integral term in (\ref{eq:2.4.1}) accounts to the accumulation of the dissipation, whereas $\chi(s) = 0$ implies (\ref{eq:2.4}) is equivalent to (\ref{eq:2.1}).
+Furthermore, under suitable assumptions on $K$, both (\ref{eq:2.1}) and (\ref{eq:2.4}) are well-posed \cite{Figotin:2006jy}.
 
@@
 It is shown in \cite{Figotin:2006jy,Figotin:2005} that under natural assumptions on the linear susceptibility $\chi(t)$ (see below), one can couple a TDD system of the form (\ref{eq:2.4}) with a canonical heat bath where the dissipated energy is captured in the heat bath in a canonical sense.
-In other words, one can construct a Hilbert space $\mathcal H$ and an isometric injection $I:\mathbb R^{2n} \to \mathbb R^{2n}\times \mathcal H^{2n}$ where the solution $z$ to (\ref{eq:2.4}) is the projection of $x$ onto $\mathbb R^{2n}$, where $x$ is the solution to
+In other words, one can construct a Hilbert space $\mathcal H$ and an isometric injection $I:\mathbb R^{2n} \to \mathbb R^{2n}\times \mathcal H^{2n}$ where the solution $z$ to (\ref{eq:2.4}) is the projection of $x$ onto $\mathbb R^{2n}$, and $x$ is the solution to
 \begin{equation} \label{eq:2.7}
-	\frac{d}{dt} x = \mathcal J_{2n} \frac{\delta H_{\text{ex}}}{\delta x}.
+	\dot x = \mathcal J_{2n} \frac{\delta H_{\text{ex}}}{\delta x}.
 \end{equation}
-Here $H_{\text{ex}}:\mathbb R^{2n}\times \mathcal H^{2n} \to \mathbb R$ is an extended quadratic Hamiltonian function and $\mathcal J_{2n}$ is the symplastic operator defined on $\mathbb R^{2n}\times \mathcal H^{2n}$ respectively.
+Here $H_{\text{ex}}:\mathbb R^{2n}\times \mathcal H^{2n} \to \mathbb R$ is an extended quadratic Hamiltonian function and $\mathcal J_{2n}$ is the symplectic operator defined on $\mathbb R^{2n}\times \mathcal H^{2n}$ respectively.
 
 \begin{theorem}
-Suppose that $K$ is full ranked and $\chi(t)$ is symmetric. There exists a quadratic extension of the form (\ref{eq:2.7}) to (\ref{eq:2.4}), if
+Suppose that $K$ is full rank and $\chi(t)$ is symmetric. Then there is a quadratic extension to (\ref{eq:2.4}) of the form (\ref{eq:2.7}), if
 \begin{equation} \label{eq:2.8}
@@
 \end{theorem}
-
-\begin{proof}
-Here we prove the theorem for the case where $\chi$ is a constant symmetric matrix, where condition (\ref{eq:2.8}) corresponds to $\chi$ being positive semi-definite.
+\emph{Proof.} Here we prove the theorem for the case where $\chi$ is a constant symmetric matrix, where condition (\ref{eq:2.8}) corresponds to $\chi$ being positive semi-definite.
 We refer the reader to \cite{Figotin:2006jy} for the proof of the general case.
@@
 		\label{eq:2.10.a} & \dot{z}(t) = \mathbb J_{2n} K^T f(t), \\
-		\label{eq:2.10.b} & \partial_t \theta(t,x) = \partial_x^2 \phi(t,x) + \sqrt 2 \delta_0(x) \cdot \sqrt{\chi}  f(t), \\
-		\label{eq:2.10.c} & \partial_t \phi(t,x) = \theta(t,x),
+		\label{eq:2.10.c} & \partial_t \phi(t,x) = \theta(t,x), \\
+		\label{eq:2.10.b} & \partial_t \theta(t,x) = \partial_x^2 \phi(t,x) + \sqrt 2 \delta_0(x) \cdot \sqrt{\chi}  f(t), 
 \end{align}
@@
 \begin{equation} \label{eq:2.10.1}
-	z(0) = z_0,\quad \theta(0,\cdot) = 0, \quad \phi(0,\cdot) = 0.
+	z(0) = z_0,\quad \phi(0,\cdot) = 0, \quad \theta(0,\cdot) = 0.
 \end{equation}
@@
 \end{equation}
-To show that the Hamiltonian system (\ref{eq:2.10.a}-\ref{eq:2.10.c}) is an extension to (\ref{eq:2.4}) in the sense discussed above, it is enough to show that the solution $f$ to the equation (\ref{eq:2.11}) also satisfies (\ref{eq:2.4.1}).
-Equations (\ref{eq:2.10.b}-\ref{eq:2.10.c}) are equations to a vibrating string, and can be solved analytically
+To show that the Hamiltonian system (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) is an extension to (\ref{eq:2.4}) in the sense discussed above, it is enough to show that the solution $f$ to equation (\ref{eq:2.11}) also satisfies (\ref{eq:2.4.1}).
+Equations (\ref{eq:2.10.c}) and (\ref{eq:2.10.b}) are equations for a vibrating string, and can be solved analytically
 \begin{equation} \label{eq:2.12}
@@
 \end{equation}
-\end{proof}
-Then, we can recover (\ref{eq:2.4.1}) by substituting (\ref{eq:2.12}) into (\ref{eq:2.11}).
-The extended Hamiltonian $H_\text{ex}$ for the system (\ref{eq:2.10.a}-\ref{eq:2.10.c}) takes the quadratic from
+Then, we recover (\ref{eq:2.4.1}) by substituting (\ref{eq:2.12}) into (\ref{eq:2.11}).
+The extended Hamiltonian $H_\text{ex}$ for the system (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) takes the quadratic from
 \begin{equation} \label{eq:2.13}
@@
 
-Equations (\ref{eq:2.10.b}) and (\ref{eq:2.10.c}) are called the \emph{hidden strings}.
-The dissipation of energy in the original system (\ref{eq:2.4}) is carried away, as vibrations, along the added strings which makes the extended system conservative.
-The Hamiltonian extension of the damped harmonic oscillator in Example \ref{example:2.1} is exactly the Lamb model \cite{lamb:1900}.
-The Lamb model is a harmonic oscillator coupled with a vibrating string, and the tension in the string causes linear dissipation in the dynamics of the harmonic oscillator.
+Equations (\ref{eq:2.10.c}) and (\ref{eq:2.10.b}) are called the \emph{hidden strings}.
+The dissipation of energy in the original system (\ref{eq:2.4}) is carried away, as vibrations, along the added strings making the extended system conservative.
+The Hamiltonian extension of the damped harmonic oscillator in Example \ref{example:2.1} is exactly the Lamb model \cite{lamb:1900} which is a harmonic oscillator coupled with a vibrating string, and the tension in the string causes linear dissipation in the dynamics of the harmonic oscillator.
 
-Note that the time integration of (\ref{eq:2.10.a}-\ref{eq:2.10.c}) involves the integration of $f$ in (\ref{eq:2.12}).
-In general, the history of $f(t)$ must be stored and may cause serious storage limitation in long-time integration.
-However, we are interested solely in finding $z(t)$ which depends on $f$ at time $t$, and $\phi(t,0)$, that is the integral of the history of $f$.
+Note that the time integration of (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) involves the integration of $f$ in (\ref{eq:2.12}).
+In general, the history of $f(t)$ must be stored and may cause storage limitation in long-time integration.
+However, we are interested solely in finding $z(t)$ which depends on $f$ at time $t$, and $\phi(t,0)$, i.e.
+the integral of the history of $f$.
 So by carefully choosing a quadrature rule that uses the same quadrature nodes as the time integrator we can avoid storing the history of $f$.
@@
 \end{equation}
-where $\Delta t$ is the time discretization step.
-The recursive relation in (\ref{eq:2.12}) suggests that only storing the value of the integral term together with the state of $f$ in the previous time step is required to evaluate the integral for the new time step.
+where $\Delta t$ is the time step.
+The recursive relation in (\ref{eq:2.12}) suggests that storing the value of the integral term together with the state of $f$ in the previous time step suffices to evaluate the integral for the new time step.
 For other interpolation based quadrature rules, we can construct similar recursive rules of the form
@@
 \end{equation}
-for some quadrature weights $\omega_i$, $i=1,\dots,k$ with $k\ll n$. Thus, time integration of (\ref{eq:2.10.a}-\ref{eq:2.10.c}), requires storing only $k$ evaluations of $f$.
+for some quadrature weights $\omega_i$, $i=1,\dots,k$ with $k\ll n$. Thus, time integration of (\ref{eq:2.10.a})-(\ref{eq:2.10.c}), only requires storage of $k$ evaluations of $f$.

### 2017-03-25 dissipative_hamiltonian/chapters/3.model.tex
first edit of the paper

@@
 \section{Model Order Reduction} \label{sec:3}
-In this section we first explain the main results of \cite{Peng:2014di} regarding model reduction of Hamiltonian systems.
-In section \ref{sec:3.3} we introduce the Reduced Dissipative Hamiltonian method.
+In this section we first explain the main results of \cite{Maboudi:2016,Peng:2014di} regarding model reduction of Hamiltonian systems and, subsequently we introduce the Reduced Dissipative Hamiltonian method.
 
@@
 Consider a Hamiltonian system of the form (\ref{eq:2.1}) together with a quadratic Hamiltonian of the form $H(z) = \frac 1 2 z^T K^T K z$.
-In this paper we focus on reducing the complexity of numerical evaluation of (\ref{eq:2.1}) with respect to time $t$.
-Nevertheless, one can naturally extend the results of this paper to a Hamiltonian system that depends on a set of physical or geometrical parameters belonging to a compact subset of a Euclidean space.
+In this paper we focus on reducing the complexity of the numerical evaluation of (\ref{eq:2.1}) with respect to time $t$.
+Nevertheless, one can extend the results of this paper to a Hamiltonian system that depends on a set of physical or geometrical parameters belonging to a compact subset of a Euclidean space.
 
-The main idea behind model order reduction is that the manifold, $\mathcal M = \{ z(t) : t \in [0,T] \}$ can be approximated by a low dimensional symplectic linear subspace \cite{hesthaven2015certified,quarteroni2015reduced}.
+The main idea behind model order reduction is that the solution manifold, $\mathcal M = \{ z(t) : t \in [0,T] \}$ can be approximated by a low dimensional linear subspace \cite{hesthaven2015certified,quarteroni2015reduced}.
 A basis for such a subspace is called a \emph{reduced basis}, and its span is referred to as the \emph{reduced space} \cite{hesthaven2015certified}.
 
-Suppose that a reduced symplectic basis $A \in \mathbb R^{2n\times 2k}$ is provided with $k \ll n$. The approximated solution to (\ref{eq:2.1}) in this basis is given by
+Suppose that a reduced symplectic basis $A \in \mathbb R^{2n\times 2k}$ is provided with $k \ll n$. The approximate solution to (\ref{eq:2.1}) in this basis is expressed as
 \begin{equation} \label{eq:3.1}
@@
 \end{equation}
-where $y\in \mathbb R^{2k}$ is the coordinates of the approximation with respect to the basis $A$. Substituting (\ref{eq:3.1}) into (\ref{eq:2.1}) yields
+where $y\in \mathbb R^{2k}$ are the coordinates of the approximation with respect to the basis $A$. Substituting (\ref{eq:3.1}) into (\ref{eq:2.1}) yields
 \begin{equation} \label{eq:3.2}
@@
 \end{equation}
-As $(A^+)^T$ is a symplectic matrix it yields that $A^+ \mathbb{J}_{2n} (A^+)^T = \mathbb{J}_{2k}$.
-By defining the reduced Hamiltonian $\tilde H : \mathbb{R}^{2k} \to \mathbb R$, defined as $\tilde H (y) = H(Ay)$, we recover the reduced system
+As $(A^+)^T$ is a symplectic matrix it implies that $A^+ \mathbb{J}_{2n} (A^+)^T = \mathbb{J}_{2k}$.
+By defining the reduced Hamiltonian $\tilde H : \mathbb{R}^{2k} \to \mathbb R$, as $\tilde H (y) = H(Ay)$, we recover the reduced system
 \begin{equation} \label{eq:3.3.1}
@@
 Equation (\ref{eq:3.3.1}) is called the \emph{symplectic Galerkin projection} \cite{Peng:2014di} of the Hamiltonian system (\ref{eq:2.1}).
-Conventional model reduction routines, e.g.
-Galerkin or Petrov-Galerkin methods \cite{hesthaven2015certified,quarteroni2015reduced}, do not yield a Hamiltonian reduced system.
-Hence, the reduced system does not necessarily preserve the conservation law in theorem \ref{theorem:2.1}.
-This results in a qualitatively wrong and often unstable solution \cite{Peng:2014di}.
-On the other hand the reduced system obtained by the symplectic Galerkin projection, is a Hamiltonian system, therefore the system energy is preserved over time integration \cite{Peng:2014di}.
-The following theorem from \cite{Peng:2014di} guarantees the boundedness of the solution of the reduced system obtained by the symplectic Galerkin projection.
+Conventional model reduction techniques, e.g.
+Galerkin or Petrov-Galerkin methods \cite{hesthaven2015certified,quarteroni2015reduced}, do not yield a Hamiltonian reduced system and the reduced system does not necessarily preserve the conservation law in Theorem \ref{theorem:2.1} which results in a qualitatively wrong and often unstable solution \cite{Peng:2014di}.
+On the other hand the reduced system, obtained by the symplectic Galerkin projection, is a Hamiltonian system, and the system energy is therefore preserved over time \cite{Peng:2014di}.
+The following theorem guarantees the boundedness of the solution of the reduced system obtained by the symplectic Galerkin projection for quadratic Hamiltonians.
 
 \begin{theorem}
-\cite{bhatia2002stability} Let $S$ be a bounded open subset of $\mathbb R^{2n}$ that contain $z_0$.
-Further assume that $H(z_0)<H(z)$ or $H(z_0)>H(z)$ for all $z\in \partial S$ the boundary of $S$.
-For a given symplectic basis $A$, if $z_0$ is in the range of $A$, then both the original system and the reduced system obtained by the symplectic Galerkin projection are bounded.
+\cite{Peng:2014di} Let $S$ be a bounded open subset of $\mathbb R^{2n}$ that contain $z_0$.
+Furthermore, assume that $H(z_0)<H(z)$ or $H(z_0)>H(z)$ for all $z\in \partial S$, the boundary of $S$.
+For a given symplectic basis $A$, provided $z_0$ is in the range of $A$, then both the original system and the reduced system obtained by the symplectic Galerkin projection are bounded.
 \end{theorem}
 
-\begin{proof}
-The proof is a direct consequence of the conservation law in theorem \ref{theorem:2.1} and the fact that the image of S under the projection $AA^+$ remains open and bounded.
-\end{proof}
+In the next section we introduced the greedy method to construct a symplectic reduced basis \cite{Maboudi:2016}.
 
-In the next section we introduced the Proper Symplastic Decomposition to construct a symplectic reduced basis.
+\subsection{The Greedy Approach to Symplectic Basis Generation} \label{sec:3.2}
+Greedy generation of the reduced basis is an iterative procedure which, in each iteration, adds the two best possible basis vectors to the symplectic basis to enhance overall accuracy.
+Suppose that at the $k$-th generic step of the greedy basis selection, an ortho-symplectic basis $A_{2k} = \{ e_1,\dots,e_k,f_1,\dots,f_k\}$ of size $2k$ is provided.
+The first step in an iteration of the greedy basis selection comprises finding $t^{k+1}\in[0,T]$ such $z(t^{k+1})$ is worst approximated by the current reduced space.
+In other words
+\begin{equation} \label{eq:3.4}
+	t^{k+1} := \underset{t\in [0,T]}{\text{argmax }} \| z(t) - A_{2k}{A_{2k}}^+z(t) \|.
+\end{equation}
+Suppose that $e_{k+1}$ is the vector obtained by $\mathbb J_{2n}$-orthogonalization (the symplectic Gram-Schmidt process \cite{Salam2014}) of $z(t^{k+1})$ with respect to $A_{2k}$.
+Then the enriched basis $A_{2k+2}$ takes the form
+\begin{equation}
+	A_{2k+2} = \{ e_1,\dots,e_k, e_{k+1},f_1,\dots,f_k,\mathbb J_{2n}^T e_{k+1}\}.
+\end{equation}
+It is easily checked that $A_{2k+2}$ is ortho-symplectic.
+Furthermore, it is shown in \cite{Maboudi:2016} that under natural assumptions on the solution manifold $\mathcal M$, the greedy method converges exponentially fast.
 
-\subsection{Proper Symplectic Decomposition (PSD)} \label{sec:3.2} Consider the set $S = \{z(t_i)| t_i\in[0,T],\ i=1,\dots,N\}$, of samples of the solution to (\ref{eq:2.1}), referred to as the \emph{snapshots}.
+For parametric problems, evaluation of the projection error in (\ref{eq:3.4}) for the entire parameter space is computationally demanding.
+One can use the loss in the Hamiltonian function as a cheap surrogate to the projection error, i.e.
+\begin{equation}
+	\omega^{k+1} := \underset{\omega\in \Omega}{\text{argmax }} | H(z(\omega)) - H(AA^+z(\omega)) |.
+\end{equation}
+Here $\Omega$ is a closed and bounded set of parameters for the original Hamiltonian system.
+Note that since the Hamiltonian function is constant in time, then $\omega^{k+1}$ can be identified prior to time integration, i.e.
+$H(z) = H(z_0)$ and $H(AA^+z) = H(AA^+z_0)$.
+We summarize the greedy method for symplectic basis generation in Algorithm \ref{alg:added1} and refer the reader to \cite{Maboudi:2016} for further details.
+
+\begin{algorithm} 
+\caption{The greedy algorithm for generation of a symplectic basis} \label{alg:added1}
+{\bf Input:} Tolerated projection error $\delta$, initial condition $ z_0$
+\begin{enumerate}
+\item $t^1 \leftarrow t=0$
+\item $e_1 \leftarrow z_0$
+\item $A \leftarrow [e_1,\mathbb J^T_{2n}e_1]$
+\item $k \leftarrow 1$
+\item \textbf{while} $\| z(t) - A_{2k}{A_{2k}}^+z(t) \| > \delta$ for all $t \in [0,T]$
+\item \hspace{0.5cm} $t^{k+1} := \underset{t\in [0,T]}{\text{argmax }} \| z(t) - A_{2k}{A_{2k}}^+z(t) \|$
+\item \hspace{0.5cm} $\mathbb J_{2n}$-orthogonalize $ z(t^{k+1})$ to obtain $e_{k+1}$
+\item \hspace{0.5cm} $A \leftarrow [e_1,\dots ,e_{k+1} , \mathbb J^T_{2n}e_1,\dots,\mathbb J^T_{2n}e_{k+1}]$
+\item \hspace{0.5cm} $k \leftarrow k+1$
+\item \textbf{end while}
+\end{enumerate}
+\vspace{0.5cm}
+{\bf Output:} Symplectic basis $A$.
+\end{algorithm}
+
+%In the next section we introduced the Proper Symplectic Decomposition to construct a symplectic reduced basis.
+%
+%\subsection{Proper Symplectic Decomposition (PSD)} \label{sec:3.2} Consider the set $S = \{z(t_i)| t_i\in[0,T],\ i=1,\dots,N\}$, of samples of the solution to (\ref{eq:2.1}), referred to as the \emph{snapshots}.
 The PSD, requires a symplectic basis $A$ of size $2k$ to have the minimum symplectic projection error over $S$ .
 In other words, a PSD symplectic basis is the solution to the optimization problem
-\begin{equation} \label{eq:3.4}
-	\begin{aligned}
-	& \underset{A \in \mathbb R^{2n\times 2k}}{\text{minimize}, }
-	& & \sum_{s\in S} \| s - AA^+s\|_2^2, \\
-	& \text{subject to}
-	& & A^T \mathbb{J}_{2n}A = \mathbb{J}_{2k}.
-	\end{aligned}
-\end{equation}
-
-The Proper Orthogonal Decomposition (POD) \cite{hesthaven2015certified,quarteroni2015reduced} method, is based on a similar idea where the projection operator is an orthogonal projection and the constrain requires $A$ to be an ortho-normal basis.
+%\begin{equation} \label{eq:3.4}
+%	\begin{aligned}
+%	& \underset{A \in \mathbb R^{2n\times 2k}}{\text{minimize}, }
+%	& & \sum_{s\in S} \| s - AA^+s\|_2^2, \\
+%	& \text{subject to}
+%	& & A^T \mathbb{J}_{2n}A = \mathbb{J}_{2k}.
+%	\end{aligned}
+%\end{equation}
+%
+%The Proper Orthogonal Decomposition (POD) \cite{hesthaven2015certified,quarteroni2015reduced} method, is based on a similar idea where the projection operator is an orthogonal projection and the constrain requires $A$ to be an ortho-normal basis.
 For the POD, the Schmidt-Mirsky-Eckart-Young theorem \cite{Markovsky:2011:LRA:2103589} provides a computationally efficient solution to the minimization problem which requires the SVD decomposition of the snapshot matrix $S$.
 As the direct solution to (\ref{eq:3.4}) is usually expensive, we must seek alternative methods to obtain near optimal solutions.
-
-A heuristic approach to find a solution to (\ref{eq:3.4}) is proposed by \cite{Peng:2014di}, suggesting that the symplectic basis $A$ should take the form
-\begin{equation} \label{eq:3.5}
-	A = 
-	\begin{pmatrix}
-		\Phi & 0 \\
-		0 & \Phi
-	\end{pmatrix},
-\end{equation}
-where $\Phi \in \mathbb{R}^{n\times k}$ is an orthonormal matrix. By constructing the \emph{combined snapshot matrix}
-\begin{equation} \label{eq:3.6}
-	S_{\text{combined}} = [q_1,\dots,q_N,p_1,\dots,p_N], \quad (q_i^T,p_i^T)^T = z(t_i),
-\end{equation}
-we define $\Phi=[u_1,\dots,u_k]$, where $u_i$ is the $i$-th left singular vector of $S_{\text{combined}}$.
+%
+%A heuristic approach to find a solution to (\ref{eq:3.4}) is proposed by \cite{Peng:2014di}, suggesting that the symplectic basis $A$ should take the form
+%\begin{equation} \label{eq:3.5}
+%	A = 
+%	\begin{pmatrix}
+%		\Phi & 0 \\
+%		0 & \Phi
+%	\end{pmatrix},
+%\end{equation}
+%where $\Phi \in \mathbb{R}^{n\times k}$ is an orthonormal matrix. By constructing the \emph{combined snapshot matrix}
+%\begin{equation} \label{eq:3.6}
+%	S_{\text{combined}} = [q_1,\dots,q_N,p_1,\dots,p_N], \quad (q_i^T,p_i^T)^T = z(t_i),
+%\end{equation}
+%we define $\Phi=[u_1,\dots,u_k]$, where $u_i$ is the $i$-th left singular vector of $S_{\text{combined}}$.
 The symplectic basis obtained this way is referred to as the \emph{Cotangent Lift}.
@@
 Note that the basis $A$ constructed using the cotangent lift is an ortho-symplectic basis.
-
+%
 \subsection{The Reduced Dissipative Hamiltonian Method} \label{sec:3.3}
 
-Since the theory of symplectic model reduction in section \ref{sec:3.2} is based on the conservation law in theorem \ref{theorem:2.1}, it can no longer be applied to dissipative Hamiltonian systems.
-We expect that for dissipative systems, the evolution of energy is not correctly translated into the reduced system.
-The Reduced Dissipative Hamiltonian method considers a Hamiltonian extension to a dissipative Hamiltonian system to construct a closed system.
+Since the symplectic model reduction in Section \ref{sec:3.2} is based on the conservation law in Theorem \ref{theorem:2.1}, it can no longer be applied to dissipative Hamiltonian systems.
+Instead in the Reduced Dissipative Hamiltonian method, we considers a Hamiltonian extension to a dissipative Hamiltonian system to construct a closed system.
 A symplectic model reduction can then be naturally applied to conserve the total energy.
@@
 \end{equation}
-Since $A^T L^T L A$ is a symmetric and positive definite matrix, it has a unique Cholesky factorization $\tilde L^T \tilde L$ where $\tilde L = A^T L A$ is an upper triangular matrix of size $2k \times 2k$.
+Since $A^T L^T L A$ is symmetric and positive definite, it too has a unique Cholesky factorization $\tilde L^T \tilde L$ where $\tilde L = A^T L A$ is an upper triangular matrix of size $2k \times 2k$.
 Writing (\ref{eq:2.4}) in terms of the reduced coordinates $y$ and the Hamiltonian (\ref{eq:3.7}) reads
@@
 	& \dot y(t) = \mathbb J_{2k} A^T L^T f(t), \\
-	& A^T f(t) + \sqrt{2} A^T \sqrt{\chi} \phi(t,0) = A^T L A y
+	& A^T f(t) + \sqrt{2} A^T \sqrt{\chi} \phi(t,0) = A^T L A y,
 \end{align}
-Here we used the fact that $A^+\mathbb J_{2n} = \mathbb{J}_{2k} A^T$.
+where we use the fact that $A^+\mathbb J_{2n} = \mathbb{J}_{2k} A^T$.
 If we define $f = A \tilde f$, $\phi = A \tilde \phi$, $\theta = A\tilde \theta$ and the \emph{reduced susceptibility} as $\tilde \chi = A^T \chi A$ we recover the reduced Hamiltonian system
@@
 		\label{eq:3.12.a} & \dot{y}(t) = \mathbb J_{2k} {\tilde L}^T \tilde f(t), \\
-		\label{eq:3.12.b} & \partial_t \tilde \theta(t,x) = \partial_x^2 \tilde \phi(t,x) + \sqrt 2 \delta_0(x) \cdot \sqrt{\tilde \chi}  \tilde f(t), \\
-		\label{eq:3.12.c} & \partial_t \tilde \phi(t,x) = \tilde \theta(t,x),
+		\label{eq:3.12.c} & \partial_t \tilde \phi(t,x) = \tilde \theta(t,x),\\
+		\label{eq:3.12.b} & \partial_t \tilde \theta(t,x) = \partial_x^2 \tilde \phi(t,x) + \sqrt 2 \delta_0(x) \cdot \sqrt{\tilde \chi}  \tilde f(t),
 \end{align}
@@
 \end{equation}
-Equations (\ref{eq:3.12.a}) to (\ref{eq:3.12.c}) is a Hamiltonian system on the symplectic linear vector space $\mathbb R^{2k} \times \mathcal H^{2k}$ and contributes to the \emph{reduced TDD system}
+Equations (\ref{eq:3.12.a})-(\ref{eq:3.12.b}) is a Hamiltonian system on the symplectic linear vector space $\mathbb R^{2k} \times \mathcal H^{2k}$ and contributes to the \emph{reduced TDD system}
 \begin{equation}
@@
 \end{equation}
-Therefore, the system energy will be conserved along integral curves of (\ref{eq:3.12.a}-\ref{eq:3.12.c}).
+Therefore, the system energy will be conserved along integral curves of (\ref{eq:3.12.a})-(\ref{eq:3.12.c}).
 
-We point out that the transformation that transforms (\ref{eq:2.10.a}-\ref{eq:2.10.c}) to (\ref{eq:3.12.a}-\ref{eq:3.12.c}) is given by
+We point out that the transformation that connects (\ref{eq:2.10.a})-(\ref{eq:2.10.b}) to (\ref{eq:3.12.a})-(\ref{eq:3.12.b}) is given by
 \begin{equation}
@@
 \end{equation}
-This is a symplectic transformation, since $\mathcal A^T J_{2n} \mathcal A = J_{2k}$.
-Further, the dissipation of energy in the reduced system only depends on the reduced susceptibility.
+This is a symplectic transformation, since $\mathcal A^T \mathcal J_{2n} \mathcal A = \mathcal J_{2k}$.
+Furthermore, the dissipation of energy in the reduced system only depends on the reduced susceptibility.
 Thus, the choice of $A$ should be independent of the hidden strings $(\phi, \theta)$.
@@
 \begin{enumerate}
-	\item Construct the Hamiltonian extension (\ref{eq:2.10.a}-\ref{eq:2.10.c}) to the original TDD system (\ref{eq:2.4}).
+	\item Construct the Hamiltonian extension (\ref{eq:2.10.a})-(\ref{eq:2.10.b}) to the original TDD system (\ref{eq:2.4}).
 	\item Collect the snapshots $z(t_i)$, $i=1,\dots,N$ through time integration of the extended Hamiltonian.
 	\item Construct an ortho-symplectic basis $A$.
-	\item Define $\tilde L = A^T L A$, $\tilde \chi = A^T \chi A$ and construct the reduced dissipative Hamiltonian system (\ref{eq:3.12.a}-\ref{eq:3.12.c})
+	\item Define $\tilde L = A^T L A$, $\tilde \chi = A^T \chi A$ and construct the reduced dissipative Hamiltonian system (\ref{eq:3.12.a})-(\ref{eq:3.12.b})
 \end{enumerate}
@@
 Note that Algorithm \ref{alg:3.1} does not depend on the choice of the method to construct an ortho-symplectic basis $A$.
-Thus, an SVD-based method, such as the cotangent lift, or a greedy-natured method can be applied.
+Thus, for basis generation, the greedy approach introduced in section \ref{sec:3.2} or an SVD-based method, e.g.
+the cotangent lift \cite{Peng:2014di}, can be applied.
+
+The main advantage of the RDH method compared to the existing method is that it enables the reduced system to be integrated using a symplectic integrator.
+The reduced system constructed using the RDH is a closed Hamiltonian system, therefore the conservation law in Theorem \ref{theorem:2.1} holds and a symplectic integrator guarantees that the total energy is preserved in the reduced system.
+Alternative methods, e.g.
+\cite{peng2016geometric,Polyuga:2010gj,beattie2011structure}, either integrate the reduced system with a non-symplectic integrator, or do not construct a closed reduced system which result in accumulation of local errors or unstable solution during long time integration respectively.

### 2017-03-25 dissipative_hamiltonian/chapters/4.numer.tex
first edit of the paper

@@
 \section{Numerical Results} \label{sec:4}
-In this section we illustrate the performance of the discussed metthod through numerical solution of the dissipative wave equation and a port-Hamiltonian model for a dissipative circuit.
+In the following we illustrate the performance of the method through the reduced order model of the dissipative wave equation and a port-Hamiltonian model for a dissipative circuit.
 
@@
 \begin{equation} \label{eq:4.1}
+	\left\{
 	\begin{aligned}
@@
 	\end{aligned}
+	\right.
 \end{equation}
@@
 The discretization of $r$ corresponds to a diagonal and semi-positive definite matrix $r_\Delta$.
-Further, we discretize (\ref{eq:4.1}) using a standard centeral finite differences schemes to obtain
+Furthermore, we discretize (\ref{eq:4.1}) using a standard central finite differences schemes to obtain
 \begin{equation} \label{eq:4.2}
@@
 \end{equation}
-with $D_x^TD_x = D_{xx}$ the central finite differences matrix operator. Writing (\ref{eq:4.2}) in a TDD formulation gives
+with $D_x^TD_x = D_{xx}$ as the central finite differences matrix operator. Writing (\ref{eq:4.2}) in a TDD formulation yields
 \begin{equation} \label{eq:4.4}
-	\dot z = \mathbb J_{2n} K^T f(t), \quad f(t) + R \int_0^t f(t) = K z.
+	\dot z = \mathbb J_{2n} K^T f(t), \quad f(t) + R \int_0^t f(s) \ ds = K z.
 \end{equation}
-Here, since $R$ is not time dependent, it commutes with the integration operator. The Hamiltonian extension of (\ref{eq:4.4}), then takes the form (\ref{eq:2.10.a})-(\ref{eq:2.10.c}).
+Since $R$ is not time dependent, it commutes with the integration operator. The Hamiltonian extension of (\ref{eq:4.4}), then takes the form (\ref{eq:2.10.a})-(\ref{eq:2.10.b}).
 
@@
 \end{equation}
-
 For the numerical time integration of the extended Hamiltonian system, the Str\"omer-Verlet time stepping scheme (\ref{eq:2.3}) is used.
-Further in each time step, the system of linear equations (\ref{eq:2.11}) is solved to recover $z$.
+In each time step, the system of linear equations (\ref{eq:2.11}) is solved to recover $z$.
 System parameters are summarized below.
@@
 \vspace{0.5cm}
+The first numerical experiment corresponds to an inhomogeneous dissipative media.
+Here, $r_{\Delta}$ is a diagonal matrix with diagonal elements $r_i := 0.1 + 0.9(i/N_{\Delta x})$, for $i=1,\dots,N_{\Delta x}$.
 
-The first numerical experiment, corresponds to an inhomogeneous dissipative media.
-Here, $r_{\Delta} = \text{diag}(r_1,\dots,r_{N_{\Delta x}})$, with $r_i = 0.1 + 0.9(i/N_{\Delta x})$, and diag is the MATLAB notation for constructing a diagonal matrix.
-
-Figure \ref{fig:4.1}.a shows the solution of the original dissipative wave equation (\ref{eq:4.1}) at $t \in \{0,2.5,5,7.5\}$.
+Figure \ref{fig:4.1}.(a) shows the solution of the original dissipative wave equation (\ref{eq:4.1}) at $t \in \{0,2.5,5,7.5\}$.
 For a nonzero $r_\Delta$ the solution will converge to $(q(t=\infty,x),p(t=\infty,x)) = (\rho,0)$ where $\rho$ is the center of mass of $q_0$.
 
-To construct the reduced basis for different methods, we gather the snapshots of (\ref{eq:4.1}).
 We construct the RDH reduced system according to the Algorithm \ref{alg:3.1}.
-The cotangent lift is used to construct an ortho-symplectic reduced basis $A$ (\ref{eq:3.5}).
-The performance of the RDH method is then compared to the POD, and the method proposed in \cite{peng2016geometric} referred to as the PSD.
+The performance of the method is then compared to the POD and the method proposed in \cite{peng2016geometric}, referred to as the PSD.
 
-Figure \ref{fig:4.1}.b Shows the decay of the singular values of the snapshot matrix for the POD, PSD and the RDH methods.
+Figure \ref{fig:4.1}.(b) illustrates the decay of the singular values of the snapshot matrix \cite{hesthaven2015certified}, for the POD, PSD, and the RDH methods.
 Note that the snapshots for the PSD and the RDH are different since they have different canonical representations.
@@
 
-The $L^2$ error between the full systems and the RDH, the PSD and the POD methods are presented in Figure \ref{fig:4.1}.c.
-We notice that symplectic methods provide a more accurate solution with respect to the POD method.
+The $L^2$-error between the full systems and the RDH, the PSD, and the POD methods are presented in Figure \ref{fig:4.1}.(c).
+We notice that the symplectic methods provide a more accurate solution when compared to the POD method.
 In fact, the POD method does not yield a stable reduced system.
-Further it is seen that enriching the PSD reduced basis does not yield a significant enhancement in the accuracy of the reduced system.
-This is because in the PSD method, a non-conservative system is being numerically integrated with a symplectic integrator.
-This results in an incorrect evolution of energy and eventually, in a qualitatively wrong numerical solution.
+Furthermore, it is seen that enriching the PSD reduced basis does not yield a significant improvement in the accuracy of the reduced system.
+This happens as the PSD method, numerically integrate a non-conservative system with a symplectic integrator.
+This results in an incorrect evolution of the energy and eventually, in a qualitatively wrong numerical solution.
 
-On the other hand, We notice that the RDH method with 40 modes provides a significantly more accurate solution compared to the PSD method with 60 modes.
-The RDH method provides a conservative reduced system where the dissipated energy is absorbed by the hidden strings.
-The conservation of energy is then guaranteed using a symplectic integrator.
-Due to this reason, we observe remarkable increase in the accuracy by enriching the RDH reduced basis.
+On the other hand, we notice that the RDH method with 40 modes provides a significantly more accurate solution compared to the PSD method with 60 modes.
+The RDH method provides a conservative reduced system where the dissipated energy is absorbed by the hidden strings and the conservation of the energy is then guaranteed by using a symplectic integrator.
+Therefore, we observe remarkable increase in the accuracy by enriching the RDH reduced basis.
 
-Figure \ref{fig:4.1}.d shows the preservation of the energy in different methods.
-The conservation law expressed in theorem \ref{theorem:2.1} is destroyed through the POD model reduction.
-That is why we observe the blow-up of the system energy.
-The symplectic methods however, preserves the energy significantly better.
-Due to the reason discussed above, enriching the PSD basis does not significantly affect the preservation of energy.
-On the contrary, The RDH not only provides a noteworthy accuracy in preserving the energy, but also provides significant improvement under enrichment.
+Figure \ref{fig:4.1}.(d) shows the conservation of the energy in the different methods.
+The conservation law expressed in Theorem \ref{theorem:2.1} is destroyed through the POD model reduction and as a consequence we observe blow-up of the system energy.
+The symplectic methods preserves the energy significantly better.
+As discussed above, enriching the PSD basis does not significantly improve the preservation of energy.
+On the contrary, the RDH provides a substantial improvement in the accuracy of the energy.
 
-In Figure \ref{fig:4.1}.e we show the transfer of the energy from the TDD system to the hidden strings, for the full system and the RDH reduced system.
+In Figure \ref{fig:4.1}.(e) we show the transfer of the energy from the TDD system to the hidden strings, for the full system and the RDH reduced system.
 We notice that the RDH method preserves the total energy of the extended Hamiltonian system.
-Further, the transfer of the energy to the hidden strings in the full model is correctly translated in the reduced system.
+Furthermore, the transfer of energy to the hidden strings in the full model is correctly translated in the reduced system.
 
@@
 \end{tabular}
-\caption{(a) The solution to the original dissipative wave equation (\ref{eq:4.1}), (b) The decay of the singular values for the POD, the PSD and the RDH methods, (c) The $L^2$ error for the different methods, (d) Energy preservation for different methods, (e) Energy preservation of the Hamiltonian extension for the original and the reduced system.
-``FM'' and ``RM'' stand for the full model and the reduced model respectively.
-(f) The $L^2$ error between the solution to the reduced system and the full system in a near-zero dissipation regime.} \label{fig:4.1}
+\caption{(a) The solution to the original dissipative wave equation (\ref{eq:4.1}), (b) The decay of the singular values for the POD, the PSD, and the RDH methods, (c) The $L^2$-error for the different methods, (d) Evolution of error in the Hamiltonian for different methods, (e) Energy preservation of the Hamiltonian extension for the original and the reduced system.
+``FM'' and ``RM'' refer to the full model and the reduced model, respectively.
+(f) The $L^2$-error between the solution to the reduced system and the full system in a near-zero dissipation regime.} \label{fig:4.1}
 \end{figure}
@@
 The numerical setting is taken to be identical to the previous numerical experiment, but with the difference that $r_i = 10^{-5}$, for $i=1,\dots,N_{\Delta x}$.
-Figure \ref{fig:4.1}.f shows the $L^2$ error between the solution to the reduced system and the full system, for the POD, the PSD and the RDH methods.
-We notice that the POD does not yield a reduced system as the symplectic structure is lost via model reduction.
-Further we notice that error for the PSD and the RDH coincide since the two methods become identical as $\| \chi \|_{\infty}\to 0$.
+Figure \ref{fig:4.1}.(f) shows the $L^2$-error between the solution to the reduced system and the full system, for the POD, the PSD, and the RDH methods.
+We notice that the POD does not yield a stable reduced system as the symplectic structure is lost via model reduction.
+Furthermore we notice that error for the PSD and the RDH coincide as the two methods become identical as $\| \chi \|_{\infty}\to 0$.
 
@@
 Port-Hamiltonian systems are popular in network modeling and electrical engineering.
-In the framework of port-Hamiltonian modelling, Energy conservation and Hamiltonian structure is the fundamental principle of the dynamics of the system.
-Ports in the system network then allows the exchange of energy with the environment in the form of input forces, capacitors and dissipations \cite{vanderSchaft:2014:PST:2693645.2693646}.
+In the framework of port-Hamiltonian modelling, energy conservation and Hamiltonian structure is the fundamental principle of the dynamics of the system.
+Ports in the system network then allows the exchange of energy with the environment in the form of sources, capacitors, and dissipations \cite{vanderSchaft:2014:PST:2693645.2693646}.
 Port-Hamiltonian systems can be viewed as a forced and dissipative Hamiltonian system.
@@
 Consider the $n$-dimensional linear ladder network in Figure \ref{fig:4.2}.
-Here $C_i$, $L_i$ and $R_i$, $i=1,\dots,n$, are the capacitance, inductance and resistance of the corresponding capacitors, inductors and resistors respectively and $R_{n+1}$ is the load capacitor.
+Here $C_i$, $L_i$ and $R_i$, $i=1,\dots,n$, are the capacitance, inductance and resistance of the corresponding capacitors, inductors, and resistors, respectively, and $R_{n+1}$ is the load capacitor.
 The port-Hamiltonian model of the linear ladder network takes the form
@@
 \end{equation}
-$u=(1,0,\dots,0)^T$ is a constant input current and $J_{2n}$ is a skew-symmetric $2n\times 2n$ matrix with -1 and 1 on the superdiagonal and subdiagonal respectively.
+$u=(1,0,\dots,0)^T$ is a constant input current and $J_{2n}$ is a skew-symmetric $2n\times 2n$ matrix with -1 and 1 on the superdiagonal and subdiagonal, respectively.
 
-The energy associated to a port-Hamiltonian system of the form (\ref{eq:4.7}) at time $t$, is given as $H(x(t)) = \frac 1 2 x^T Q^T Q x$.
-Since $J_{2n}$ is skew symmetric it implies that $\frac d {dt} H(x) = u^T Q^T Q x - x^T Q^T Q R Q^T Q x \leq u^T Q^T Q x$ which is referred to as the \emph{passivity} of the system (\ref{eq:4.7}) \cite{vanderSchaft:1996es,Willems:1972ek}.
+The energy associated with a port-Hamiltonian system of the form (\ref{eq:4.7}) at time $t$, is given as $H(x(t)) = \frac 1 2 x^T Q^T Q x$.
+Since $J_{2n}$ is skew symmetric we have that that $\frac d {dt} H(x) = u^T Q^T Q x - x^T Q^T Q R Q^T Q x \leq u^T Q^T Q x$ which is referred to as the \emph{passivity} of the system (\ref{eq:4.7}) \cite{vanderSchaft:1996es,Willems:1972ek}.
 
-Since $J_{2n}$ is full ranked, one can always find a coordinate transformation $ x= T \tilde x$ such that $T^{-1} J_{2n} T^{-T} = \mathbb J_{2n}$.
-The dissipative Hamiltonian formulation for (\ref{eq:4.7}) takes the form
+Since $J_{2n}$ is full rank, one can always find a coordinate transformation $ x= T \tilde x$ such that $T^{-1} J_{2n} T^{-T} = \mathbb J_{2n}$.
+The dissipative Hamiltonian formulation of (\ref{eq:4.7}) takes the form
 \begin{equation} \label{eq:4.8}
@@
 where $\tilde Q = QT$, $\tilde R = T^{-1}RT^{-T}Q^TQ$ and $\tilde u = T^{-1} u$.
-Note that in this case, since $T$ is orthogonal and $R$ is diagonal, then $\tilde R$ is symmetric and semi-positive definite.
+Note that in this case, $\tilde R$ is symmetric and semi-positive definite since $T$ is orthogonal and $R$ is diagonal.
 The TDD formulation of (\ref{eq:4.8}) takes the form
@@
 \end{equation}
-The extended Hamiltonian formulation (\ref{eq:2.10.a}-\ref{eq:2.10.c}) with a quadratic Hamiltonian $H_{\text{ex}}$ can be then carried out according to section \ref{sec:2.2}.
+The extended Hamiltonian formulation (\ref{eq:2.10.a})-(\ref{eq:2.10.b}) with a quadratic Hamiltonian $H_{\text{ex}}$ can be carried out following Section \ref{sec:2.2}.
 We note that due to the input $\tilde u$, the Hamiltonian $H_{\text{ex}}$ is time dependent.
-As the matter of fact $\frac{d}{dt} H_{\text{ex}} = \tilde u^T\tilde Q^T \tilde Q \tilde x$.
+In fact $\frac{d}{dt} H_{\text{ex}} = \tilde u^T\tilde Q^T \tilde Q \tilde x$.
 If we define $\overset{\circ}{H}_{\text{ex}} : \mathbb R^{2n}\times \mathcal H^{2n}\times \mathbb R^{2}\to \mathbb R$ as
@@
 it is easily checked that $\frac d {dt} \overset{\circ}{H}_{\text{ex}} =0$ \cite{Hairer:1250576}.
-However for the time integration of the Hamiltonian system related to $\overset{\circ}{H}_{\text{ex}}$, since the evolution of $\tilde x$, $\phi$ and $\theta$ does not explicitly depend on $e$, we can apply a symplectic integrator directly on (\ref{eq:4.10}).
+However for the time integration of the Hamiltonian system related to $\overset{\circ}{H}_{\text{ex}}$ we can apply a symplectic integrator directly on (\ref{eq:4.10}), since the evolution of $\tilde x$, $\phi$ and $\theta$ does not explicitly depend on $e$.
 Thus, the passivity of (\ref{eq:4.7}) will be preserved through a symplectic time integration of (\ref{eq:4.9}).
 
-Using an ortho-symplectic reduced basis $A$ The Reduced Dissipative Hamiltonian method can be applied to (\ref{eq:4.9}) to construct a reduced system of the form (\ref{eq:3.12.a}-\ref{eq:3.12.c}) together with the extended Hamiltonian $\tilde H_{\text{ex}}$.
-We immediately notice that $\frac{d}{dt} \tilde H_{\text{ex}} = (A^+ \tilde u)^T A^T \tilde Q^T \tilde Q A y$, meaning that the reduced system is passive.
-Further the dissipative Hamiltonian structure of the reduced system indicates that the reduced system also carries a port-Hamiltonian structure.
+Using an ortho-symplectic reduced basis $A$, the Reduced Dissipative Hamiltonian method can be applied to (\ref{eq:4.9}) to construct a reduced system of the form (\ref{eq:3.12.a})-(\ref{eq:3.12.b}) together with the extended Hamiltonian $\tilde H_{\text{ex}}$.
+We note that $\frac{d}{dt} \tilde H_{\text{ex}} = (A^+ \tilde u)^T A^T \tilde Q^T \tilde Q A y$, ensuring that the reduced system is passive.
+Furthermore, the dissipative Hamiltonian structure of the reduced system indicates that the reduced system also carries a port-Hamiltonian structure.
 
@@
 We take $C_i=1$, $L_i = 1$, $R_i=0.2$ for $i=1,\dots,50$, and $R_{51} = 0.4$.
-The cotangent lift is used to construct a reduced basis and the RDH method is used to construct a reduced system according to Algorithm \ref{alg:3.1}.
+We construct the RDH reduced system following Algorithm \ref{alg:3.1}.
 
@@
 Reduced bases of size $2k = 10$, $2k = 20$ and $2k = 30$ are constructed with the RDH and the moment matching method.
-Figure \ref{fig:4.3}.c shows the error in the charge in $C_1$ for the two methods.
-We observe that although the moment matching method is bounded, over long-time integration, the RDH method provides a significantly more accurate solution.
-This is because in the moment matching method, the passivity of the reduced system implies that the energy of the system will be bounded by the input energy.
+Figure \ref{fig:4.3}.(c) shows the error in the charge in $C_1$ for the two methods.
+We observe that although the moment matching method is bounded over long-time integration, the RDH method provides a significantly more accurate solution.
+In the moment matching method, the passivity of the reduced system implies that the energy of the system will be bounded by the input energy.
 However, there is no guarantee that the dissipation of energy in the reduced system mimics the one of the original system.
-On the other hand, the RDH method allows correct dissipation of energy through the hidden strings.
-Also the symplectic time integration in the RDH method guarantees that the total energy will be preserved.
+On the other hand, the RDH method allows a correct dissipation of energy through the hidden strings and the symplectic time integration in the RDH method guarantees that the total energy is preserved.
 
 Over short-time integration, we notice that the moment matching method with 10 modes provides a more accurate solution than the RDH with 10 modes.
-Further, the moment matching method with 20 and 30 modes provide a comparable accuracy to the RDH method with 20 and 30 modes.
-But, the RDH method maintains the high accuracy over the long-time integration, while the moment matching method loses up to 3 orders of magnitude in the accuracy independent of the number of modes.
+Furthermore, the moment matching method with 20 and 30 modes provide a comparable accuracy to the RDH method with 20 and 30 modes.
+However, the RDH method maintains the high accuracy during long-time integration, while the moment matching method loses up to 3 orders of magnitude in the accuracy, independent of the number of modes.
 
-Figure \ref{fig:4.3}.a and Figure \ref{fig:4.3}.b show the average error in the charge and flux of the capacitors and inductors respectively.
+Figure \ref{fig:4.3}.(a) and Figure \ref{fig:4.3}.(b) show the average temporal error in the charge and flux of the capacitors and inductors, respectively.
 The RDH method provides a significantly better accuracy compared to the moment matching method.
 This is because the charge of $C_1$ is specified as the output of interest in the moment matching method and so it is expected that that method provides low accuracy for computing other outputs.
-On the other hand, the RDH method not only provides high accuracy in computing the charge for $C_1$ but also it has high accuracy for computing other outputs of interest.
+On the other hand, the RDH method not only provides high accuracy in computing the charge for $C_1$ but also high accuracy for all components of the system.

### 2017-03-25 dissipative_hamiltonian/paper.tex
first edit of the paper

@@
 \usepackage{amsmath}
+%\usepackage{amsthm}
 \usepackage{amssymb}
@@
 Reduced basis methods are popular for approximately solving large and complex systems of differential equations.
-However, conventional methods do not generally preserve conservation laws and symmetries of the full order model.
-Here, we present The Reduced Dissipative Hamiltonian Method, that preserves the symplectic symmetry of dissipative Hamiltonian systems over the course of model reduction.
-This method constructs a closed reduced Hamiltonian system by coupling the full model with a canonical heat bath.
-This results in a correct dissipation of energy, preservation of total energy and ultimately in the stability of the solution.
+However, conventional reduced basis methods do not generally preserve conservation laws and symmetries of the full order model.
+Here, we present an approach for reduced model construction, that preserves the symplectic symmetry of dissipative Hamiltonian systems.
+The method constructs a closed reduced Hamiltonian system by coupling the full model with a canonical heat bath.
+This results in a correct dissipation of energy, preservation of total energy and, ultimately, in the stability of the solution.
 Accuracy and stability of the method are illustrated through the numerical simulation of the dissipative wave equation and a port-Hamiltonian model of an electric circuit.

### 2017-03-27 dissipative_hamiltonian/chapters/3.model.tex
second edit finished

@@
 Here $\Omega$ is a closed and bounded set of parameters for the original Hamiltonian system.
-Note that since the Hamiltonian function is constant in time, then $\omega^{k+1}$ can be identified prior to time integration, i.e.
+Note that $\omega^{k+1}$ can be identified prior to time integration since the Hamiltonian function is constant in time, i.e.
 $H(z) = H(z_0)$ and $H(AA^+z) = H(AA^+z_0)$.
@@
 
-The main advantage of the RDH method compared to the existing method is that it enables the reduced system to be integrated using a symplectic integrator.
+The main advantage of the RDH method compared to the existing methods is that it enables the reduced system to be integrated using a symplectic integrator.
 The reduced system constructed using the RDH is a closed Hamiltonian system, therefore the conservation law in Theorem \ref{theorem:2.1} holds and a symplectic integrator guarantees that the total energy is preserved in the reduced system.
 Alternative methods, e.g.
-\cite{peng2016geometric,Polyuga:2010gj,beattie2011structure}, either integrate the reduced system with a non-symplectic integrator, or do not construct a closed reduced system which result in accumulation of local errors or unstable solution during long time integration respectively.
+\cite{peng2016geometric,Polyuga:2010gj,beattie2011structure}, either integrate the reduced system with a non-symplectic integrator, or do not construct a closed reduced system which result in accumulation of local errors or unstable solution during long time integration respectively \cite{Hairer:1250576}.

### 2017-03-27 dissipative_hamiltonian/chapters/4.numer.tex
second edit finished

@@
 
-\subsection{Dissipative wave equation}
+\subsection{Dissipative wave equation} \label{sec:4.1}
 
@@
 
-The $L^2$-error between the full systems and the RDH, the PSD, and the POD methods are presented in Figure \ref{fig:4.1}.(c).
+The $L^2$-error between the full system and the RDH, the PSD, and the POD methods are presented in Figure \ref{fig:4.1}.(c).
 We notice that the symplectic methods provide a more accurate solution when compared to the POD method.
@@
 Furthermore we notice that error for the PSD and the RDH coincide as the two methods become identical as $\| \chi \|_{\infty}\to 0$.
+
+\subsection{The sine-Gordon equation}
+Consider the one-dimensional dissipative nonlinear wave equation
+\begin{equation} \label{eq:added.4.1}
+	\left\{
+	\begin{aligned}
+		q_{t}(t,x) &= p(t,x), \\
+		p_{t}(t,x) &= q_{xx}(t,x) - \sin(q) - r(x)  p(t,x), \\
+		q(0,x) &= q_0(x), \quad q(t,0) = a, \quad q(t,L) = b,\\
+		p(0,x) &= p_0(x).
+	\end{aligned}
+	\right.
+\end{equation}
+defined on a domain of length $L$, which is known as the sine-Gordon equation. In the absence of dissipation, $r(x) = 0$, the \emph{kink} solution to (\ref{eq:added.4.1}) is given as
+\begin{equation} \label{eq:added.4.2}
+	q(t,x) = 4\arctan\left( \exp \left( \frac{(x-x_0 - vt)}{\sqrt{1-v^2}} \right) \right),
+\end{equation}
+where $|v|< 1$ is the wave speed. In the presence of dissipation, where $r(x)\geq 0$, the traveling wave de-accelerates and stops. The TDD formulation for (\ref{eq:added.4.1}) takes the form
+\begin{equation} \label{eq:added.4.3}
+	\dot z = \mathbb J_{2n} K^T f(t) + \mathbb J_{2n} g(z) - \mathbb J_{2n} z_{\text{bd}}, \quad f(t) + R \int_0^t f(s) \ ds = K z.	
+\end{equation}
+where $z$, $K$, $R$ and $f$ are defined similar to (\ref{eq:4.4}), with $r_{\Delta} = rI_{n}$, and 
+\begin{equation}
+g(z) = (\sin(q_1),\dots,\sin(q_{N_{\Delta x}}),0,\dots,0)^T,
+\end{equation}
+and $z_{\text{bd}}$ is the term corresponding to the Dirichlet boundary condition. Note that the extended Hamiltonian $H_{ex}$ takes the form
+\begin{equation}
+	H_\text{ex}(z,\phi,\theta) = \frac 1 2 \left( \| Kz - \phi(t,0) \|_2^2 + \| G(z) \|_2^2 + \| \theta(t) \|^2_{\mathcal H^{2n} } + \| \partial_x\phi(t)\|^2_{\mathcal H^{2n} }\right),
+\end{equation}
+where $G(z)$ is a potential for $g(z)$ given as $G(q_i) = 1 - \sin(q_i)$, for $i=1,\dots,N_{\Delta_x}$. System parameters are summarized below
+
+\vspace{0.5cm}
+\begin{center}
+\begin{tabular}{|l|l|}
+\hline
+Domain length & $L = 50$ \\
+No. grid points & $N = 500$ \\
+Space discretization size & $\Delta x = L/N$ \\
+Time discretization size & $\Delta t = 0.02$ \\
+Wave speed & $v = 0.5$ \\
+Boundary conditions & $a = 0$, $b=1$\\
+Dissipation coefficient & $r = 0.1$ \\
+\hline
+\end{tabular}
+\end{center}
+\vspace{0.5cm}
+
+The RDH reduced system is constructed following Algorithm \ref{alg:3.1}.
+To reduce the complexity of the nonlinear term, we used the symplectic discrete empirical interpolation method (SDEIM) \cite{Maboudi:2016}.
+The performance of the method is then compared to the PSD and the POD where the SDEIM proposed in \cite{Peng:2014di} and the classical DEIM \cite{Chaturantabut:2010cz} is applied to reduce the complexity of the nonlinear term, respectively.
+
+\begin{figure}[t]
+\begin{tabular}{cc}
+\includegraphics[width=0.5\textwidth]{./figs/wave_nonlin/error} & 
+\includegraphics[width=0.5\textwidth]{./figs/wave_nonlin/energy} \\
+(a) & (b)
+\end{tabular}
+\caption{(a) The $L^2$-error for the different methods, (a) Evolution of error in the kinetic energy for different methods.} \label{fig:added4.1}
+\end{figure}
+
+Figure \ref{fig:added4.1}.(a) shows the The $L^2$-error between the full system and the RDH, the PSD, and the POD methods.
+Although the Hamiltonian system of the sine-Gordon equation is nonlinear, the errors for the different methods show a similar behavior as those in Section (\ref{sec:4.1}).
+We observe that the POD does not yield a stable reduce system while the symplectic methods provide a high accuracy.
+Furthermore, we notice that enriching the PSD basis does not significantly enhance the accuracy of the method.
+
+The evolution of error in the kinetic energy $K(p) = \|p\|_2^2/2$ is illustrated in Figure \ref{fig:added4.1}.(b).
+We see that the POD does not conserve the evolution of the kinetic energy.
+The RDH method conserves the kinetic energy of the system with higher accuracy than the PSD method.
+Furthermore, the accuracy of the RDH method is better scaled under enrichment of the reduced basis, compared to the PSD method.
+
+It is observed in Figure 1 that the symplectic treatment of the nonlinear terms is essential in correct model reduction of Hamiltonian systems.
+In addition, the SDEIM can be combined with the RDH method to construct a reduced Hamiltonian system that can be integrated using a symplectic integrator.
+Thus, the combination preserves the system energy and the symplectic symmetry of Hamiltonian systems.
 
@@
 \end{tabular}
-\caption{Error between the full model and the reduced model obtained by the reduced dissipating Hamiltonian method ``RDH'' and the moment matching method ``MM''.
-(a) The average error of charge in capacitors in time.
-(b) the average error of the flux in inductors in time.
-(c) Error in $C_1$.} \label{fig:4.3}
+\caption{Error between the full model and the reduced model obtained by the reduced dissipative Hamiltonian method ``RDH'' and the moment matching method ``MM''.
+(a) The average temporal error of charge in capacitors.
+(b) the average temporal error of the flux in inductors.
+(c) The error in $C_1$.} \label{fig:4.3}
 \end{figure}

### 2017-03-27 dissipative_hamiltonian/paper.tex
second edit finished

@@
 The method constructs a closed reduced Hamiltonian system by coupling the full model with a canonical heat bath.
-This results in a correct dissipation of energy, preservation of total energy and, ultimately, in the stability of the solution.
+This allows the reduced system to be integrated with a symplectic integrator, resulting in a correct dissipation of energy, preservation of total energy and, ultimately, in the stability of the solution.
 Accuracy and stability of the method are illustrated through the numerical simulation of the dissipative wave equation and a port-Hamiltonian model of an electric circuit.

### 2017-03-31 dissipative_hamiltonian/chapters/2.Hamil.tex
paper finished

@@
 where $\Delta t$ is the time step.
-The recursive relation in (\ref{eq:2.12}) suggests that storing the value of the integral term together with the state of $f$ in the previous time step suffices to evaluate the integral for the new time step.
+The recursive relation in (\ref{eq:2.14}) suggests that storing the value of the integral term together with the state of $f$ in the previous time step suffices to evaluate the integral for the new time step.
 For other interpolation based quadrature rules, we can construct similar recursive rules of the form

### 2017-03-31 dissipative_hamiltonian/chapters/3.model.tex
paper finished

@@
 \end{equation}
-Therefore, the system energy will be conserved along integral curves of (\ref{eq:3.12.a})-(\ref{eq:3.12.c}).
+Therefore, the system energy will be conserved along integral curves of (\ref{eq:3.12.a})-(\ref{eq:3.12.b}).
 
@@
 Alternative methods, e.g.
-\cite{peng2016geometric,Polyuga:2010gj,beattie2011structure}, either integrate the reduced system with a non-symplectic integrator, or do not construct a closed reduced system which result in accumulation of local errors or unstable solution during long time integration respectively \cite{Hairer:1250576}.
+\cite{peng2016geometric,Polyuga:2010gj,beattie2011structure}, either integrate the reduced system with a non-symplectic integrator, or do not construct a closed reduced system which result in accumulation of local errors or unstable solution during long time integration, respectively \cite{Hairer:1250576}.

### 2017-03-31 dissipative_hamiltonian/chapters/4.numer.tex
paper finished

@@
 \begin{tabular}{cc}
-\includegraphics[width=0.5\textwidth]{./figs/wave_nonlin/error} & 
-\includegraphics[width=0.5\textwidth]{./figs/wave_nonlin/energy} \\
+\includegraphics[width=0.5\textwidth]{./figs/wave_nonlin/error_nonlin} & 
+\includegraphics[width=0.5\textwidth]{./figs/wave_nonlin/energy_nonlin} \\
 (a) & (b)
@@
 We see that the POD does not conserve the evolution of the kinetic energy.
-The RDH method conserves the kinetic energy of the system with higher accuracy than the PSD method.
+The RDH method conserves the kinetic energy of the system with a higher accuracy than the PSD method.
 Furthermore, the accuracy of the RDH method is better scaled under enrichment of the reduced basis, compared to the PSD method.

### 2017-03-31 dissipative_hamiltonian/paper.tex
paper finished

--- 2017-03-25 dissipative_hamiltonian/paper.tex
@@
 The method constructs a closed reduced Hamiltonian system by coupling the full model with a canonical heat bath.
-This results in a correct dissipation of energy, preservation of total energy and, ultimately, in the stability of the solution.
+This allows the reduced system to be integrated with a symplectic integrator, resulting in a correct dissipation of energy, preservation of the total energy and, ultimately, in the stability of the solution.
 Accuracy and stability of the method are illustrated through the numerical simulation of the dissipative wave equation and a port-Hamiltonian model of an electric circuit.

### 2017-07-11 greedy_hamiltonian_second_revision/chapters/1.introduction.tex
second revision added

--- 2016-12-09 greedy_hamiltonian_revised/chapters/1.introduction.tex
@@
 
-Convectional reduced basis techniques, such as proper orthogonal decomposition (POD) \cite{Kunisch:2002er,Atwell:2001ja,Ravindran:2002hn}, require the exploration of the entire parameter space.
+Convectional reduced basis techniques, such as {\blue the} Proper Orthogonal Decomposition (POD) \cite{Kunisch:2002er,Atwell:2001ja,Ravindran:2002hn}, require the exploration of the entire parameter space.
 This leads to a very expensive and often impractical offline {\edit stage} when dealing with multi-dimensional parameter domains.

### 2017-07-11 greedy_hamiltonian_second_revision/chapters/2.mor.tex
second revision added

--- 2016-12-09 greedy_hamiltonian_revised/chapters/2.mor.tex
@@
 Here $S$ is the snapshot matrix, containing snapshots $\mathbf x(t_i,\omega_j)$ in its columns, $\|\cdot \|_F$ is the Frobenius norm and $I_k$ is the identity matrix of size $k$.
-According to {\edit Schmidt-Mirsky-Eckart-Young theorem \cite{Markovsky:2011:LRA:2103589}}, the solution to (\ref{eq:MoOr:6}) is equivalent to the truncated singular value decomposition (SVD) of the snapshot matrix $S$ given by
+According to {\blue the} {\edit Schmidt-Mirsky-Eckart-Young theorem \cite{Markovsky:2011:LRA:2103589}}, the solution to (\ref{eq:MoOr:6}) is equivalent to the truncated singular value decomposition (SVD) of the snapshot matrix $S$ given by
 \begin{equation} \label{eq:MoOr:7}
@@
 \begin{equation} \label{eq:MoOr:8}
-	\frac{d}{dt} \mathbf y = \underbrace{(WV)^{-1} L V}_{\tilde L} \mathbf{y} + \underbrace{(WV)^{-1} \mathbf g(t,V\mathbf y,\omega)}_{\tilde N (\mathbf y)}.
+	\frac{d}{dt} \mathbf y = \underbrace{(W^TV)^{-1} L V}_{\tilde L} \mathbf{y} + \underbrace{(W^TV)^{-1} \mathbf g(t,V\mathbf y,\omega)}_{\tilde N (\mathbf y)}.
 \end{equation}

### 2017-07-11 greedy_hamiltonian_second_revision/chapters/3.ham.tex
second revision added

--- 2016-12-09 greedy_hamiltonian_revised/chapters/3.ham.tex
@@
 The differential of $H$, denoted by $\mathbf dH$, defines a 1-form on $\mathcal M$.
-The nondegeneracy of $\Omega$ implies that there is a unique vector field $X_H$, \emph{Hamiltonian vector field }\cite{da2003introduction,Marsden:1999ck}, on $\mathcal M$ such that}
+{\blue The nondegeneracy of $\Omega$ implies that there is a unique vector field $X_H$, \emph{the Hamiltonian vector field }\cite{da2003introduction,Marsden:1999ck}, on $\mathcal M$ such that}
 \begin{equation} \label{eq:Hasy:1}
-	\edit i_{X_H} \Omega = \mathbf dH, 
+	\edit i_{X_H} \Omega = \mathbf dH. 
 \end{equation}
-{\edit where $i_{X_H} \Omega$ is the interior product of $X_H$ with $\Omega$, i.e., that requiring}
+}
+{\blue Here $i_{X_H} \Omega$ is the interior product of $X_H$ with $\Omega$, i.e.,}
 \begin{equation}
@@
 For the case where the symplectic manifold is also a linear vector space, the pair $({\edit \mathcal M},\Omega)$ is also referred to as a \emph{symplectic vector space}.
-We {\edit need} the following theorems regarding symplectic vector spaces and refer the reader to \cite{de2006symplectic,Marsden:1999ck,de2006symplectic,Silva01lectureson} for detailed proofs.
+We {\edit need} the following theorems regarding symplectic vector spaces and refer the reader to \cite{de2006symplectic,Marsden:1999ck,Silva01lectureson} for detailed proofs.
 
@@
 \end{equation}
-where $\delta$ is the Kronecker's delta function. Moreover if $V = \mathbb{R}^{2n}$ we can choose basis vectors $\{e_i,f_i\}_{i=1}^n$ such that
+where $\delta$ is the Kronecker's delta function. {\blue Moreover, if $V = \mathbb{R}^{2n}$ then} we can choose basis vectors $\{e_i,f_i\}_{i=1}^n$ such that
 \begin{equation} \label{eq:Hasy:5}
@@
 \end{equation}
-with $\mathbb J_{2n}$ being the symplectic matrix, defined as
+with $\mathbb J_{2n}$ being the {\blue standard} symplectic matrix, defined as
 \begin{equation} \label{eq:Hasy:6}
@@
 \end{definition}
-\begin{theorem} \label{theorem:Hasy:1.6}  \cite{da2003introduction}
+\begin{theorem} \label{theorem:Hasy:1.6}  \cite{mehl2009perturbation,da2003introduction}
 Suppose $(V,\Omega)$ is a $2n$ dimensional symplectic vector space and $E\subset V$ is a Lagrangian subspace. Then there is an orthosymplectic basis for $V$.
@@
 \begin{proof}
-Starting from a Lagrangain subspace in $E \subset V$ an orthosymplectic basis can be easily constructed.
+{\blue A detailed proof can be found in \cite{mehl2009perturbation}.
+Since the proof contains an insight to the construction of symplectic bases, we are going to restate the proof.} Starting from a Lagrangain subspace in $E \subset V$ an orthosymplectic basis can be easily constructed.
 By Theorem \ref{theorem:Hasy:1.5} $E$ is $n$ dimensional.
@@
 \begin{equation}
-	\langle f_i, f_j \rangle = e_i^T \mathbb J_{2n} {\mathbb J_{2n}}^T e_i = \delta_{ij}, \quad \langle f_i, e_j \rangle = e_i^T \mathbb J_{2n} e_j = 0, \quad i,j=1,\dots,n,
+	\langle f_i, f_j \rangle = e_i^T \mathbb J_{2n} {\mathbb J_{2n}}^T e_j = \delta_{ij}, \quad \langle f_i, e_j \rangle = e_i^T \mathbb J_{2n} e_j = 0, \quad i,j=1,\dots,n,
 \end{equation}

### 2017-07-11 greedy_hamiltonian_second_revision/chapters/4.sympmor.tex
second revision added

--- 2016-12-09 greedy_hamiltonian_revised/chapters/4.sympmor.tex
@@
 It avoids the evaluation of the full snapshot matrix, hence substantially reduces the computational cost in the offline stage of the symplectic model reduction.
-Also, by construction, it yields a orthosymplectic basis and therefore a well-conditioned basis.
+Also, by construction, it yields an orthosymplectic basis and therefore a well-conditioned basis.
 
@@
 \item $e_1 \leftarrow \mathbf z_0(\omega^*)$
-\item $A \leftarrow [e_1,-\mathbb J^T_{2n}e_1]$
+\item $A \leftarrow [e_1,\mathbb J^T_{2n}e_1]$
 \item $k \leftarrow 1$
-\item \textbf{while} $\Delta H(\omega) > \delta$ for all $\omega \in \Omega_N$
-\item \hspace{0.5cm} $w^* \leftarrow$ $\underset{\omega\in \Omega_N}{\text{argmax }}\Delta H(\omega)$
+\item \textbf{while} $\Delta H(\omega) > \delta$ for all $\omega \in \Gamma_N$
+\item \hspace{0.5cm} $\omega^* \leftarrow$ $\underset{\omega\in \Gamma_N}{\text{argmax }}\Delta H(\omega)$
 \item \hspace{0.5cm} Compute trajectory snapshots $S=\{ \mathbf z(t_i,\omega^*) | i = 1,\dots,M \}$
@@
 \begin{equation} \label{eq:new3}
-	e_{k+1} := \underset{s\in S}{\text{argmax }}\sigma_{2k}(s), \quad f_{k+1} := \mathbb{J}^T e_{k+1}.
+	e_{k+1} := \underset{s\in S}{\text{argmax }}\sigma_{2k}(s), \quad f_{k+1} := \mathbb{J}_{2n}^T e_{k+1}.
 \end{equation}
@@
 
-We start by symplectifying the vectors provided by the greedy algorithm as
+We start by $\edit \mathbb J_{2n}$-orthogonalizing the vectors provided by the greedy algorithm as
 \begin{equation} \label{eq:new6}
@@
 \end{equation}
-for any $s\in S$. Since $\bar{\xi}_i$ is symplectic to span$(A_{2(k-1)})$ we have
+for any $s\in S$. Since $\bar{\xi}_i$ is {\edit $\mathbb J_{2n}$-orthogonal} to the span$(A_{2(k-1)})$ we have
 \begin{equation} \label{eq:new9}
@@
 \end{equation}
-Define $c = \kappa_j^{-1} \sum_{i=1,i\neq j}^{k+1} \kappa_i \xi_i + \kappa_j^{-1}\sum_{i=1}^{k+1} \kappa_{i+k+1} \bar{\xi}_i$ and note that $\mathbb{J}_{2n}^T c$ is symplectic to $\xi_j$. We recover
+Define $c = \kappa_j^{-1} \sum_{i=1,i\neq j}^{k+1} \kappa_i \xi_i + \kappa_j^{-1}\sum_{i=1}^{k+1} \kappa_{i+k+1} \bar{\xi}_i$.
+Using that $\mathbb{J}_{2n}^T c$ is {\edit $\mathbb J_{2n}$-orthogonal} to $\xi_j$ we recover
 \begin{equation} \label{eq:new19}
@@
 \subsection{Symplectic Discrete Empirical Interpolation Method (SDEIM)} Consider the Hamiltonian system (\ref{eq:Hasy:7}) and its reduced system (\ref{eq:SyMo:4}) equipped with a symplectic transformation $A$.
-One can split the Hamiltonian function $H = H_1 + H_2$ such that $\nabla H_1 = L\mathbf z$ and $\nabla H_2 = \mathbf g(\mathbf z)$, where $L$ is a constant matrix in $\mathbb R^{n\times n}$ and $\mathbf g$ is a nonlinear function.
+One can split the Hamiltonian function $H = H_1 + H_2$ such that $\nabla H_1 = L\mathbf z$ and $\nabla H_2 = \mathbf g(\mathbf z)$, where $L$ is a constant matrix in $\edit \mathbb R^{2n\times 2n}$ and $\mathbf g$ is a nonlinear function.
 The reduced system takes the form
@@
 To overcome this, when index $\mathfrak p_i$ with $\mathfrak p_i\leq n$ or $\mathfrak p_i>n$ is chosen in Algorithm \ref{alg:MoOr:1} we also include $\mathfrak p_i + n$ or $\mathfrak p_i-n$, respectively.
-{\edit Simple calculations verifies that $\mathbf g$ and $P$ commute.}
+{\edit Simple calculations verifies that $\mathbf g$ and $P$ then commute.}

### 2017-07-11 greedy_hamiltonian_second_revision/chapters/5.numres.tex
second revision added

--- 2016-12-09 greedy_hamiltonian_revised/chapters/5.numres.tex
@@
 The parametric linear wave equation is considered to compare SVD based methods with the greedy method.
-The nonlinear model order reduction using the combination of DIEM and the symplectic basis is then illustrated by considering the parametric nonlinear Schr\"odinger equation.
+{\blue The symplectic model reduction of nonlinear systems} is then illustrated by considering the parametric nonlinear Schr\"odinger equation.
 {\edit Finally we discuss the numerical convergence of the greedy method introduced in Algorithm \ref{alg:SyMo:3}.}

### 2017-07-11 greedy_hamiltonian_second_revision/chapters/6.conclusion.tex
second revision added

--- 2016-11-15 greedy_hamiltonian_new_template/chapters/6.conclusion.tex
@@
 An iteration of the greedy method comprises searching the parameter space using the error in the Hamiltonian, to find the best basis vectors that increase the overall accuracy of the reduced basis.
-We argued that for a compact subset with exponentially small Kolmogorov $n$-width we recover exponentially fast convergence of the greedy algorithm.
-For fast approximation of nonlinear terms, the basis obtained by the greedy was combined with the symplectic DEIM to construct a Hamiltonian reduced system.
+We argue that for a compact subset with exponentially small Kolmogorov $n$-width we recover exponentially fast convergence of the greedy algorithm.
+For fast approximation of nonlinear terms, the basis obtained by the greedy was combined with a symplectic DEIM to construct {\edit a reduced system with a Hamiltonian that is arbitrary close to the Hamiltonian of the original system.}
 
@@
 
-The numerical results demonstrated that the greedy method can save substantial computational cost in the offline stage as compared to alternative SVD-based techniques.
+The numerical results demonstrate that the greedy method can save substantial computational cost in the offline stage as compared to alternative SVD-based techniques.
 Also since the reduced system obtained by the greedy method is Hamiltonian, the greedy method yields a stable reduced system.
-Symplectic DEIM also effectively reduced computational cost of approximating nonlinear terms while preserving stability and symplectic structure.
+Symplectic DEIM effectively reduces computational cost of approximating nonlinear terms while preserving stability and symplectic structure.
 Hence, the greedy method is an efficient model reduction technique that provides an accurate and stable reduced system for large-scale parametric Hamiltonian systems.

### 2017-07-11 greedy_hamiltonian_second_revision/ex_shared.tex
second revision added

% SIAM Shared Information Template
% This is information that is shared between the main document and any
% supplement. If no supplement is required, then this information can
% be included directly in the main document.

% Packages and macros go here
\usepackage{lipsum}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{algorithmic}
\ifpdf
  \DeclareGraphicsExtensions{.eps,.pdf,.png,.jpg}
\else
  \DeclareGraphicsExtensions{.eps}
\fi

% Declare title and authors, without \thanks
\newcommand{\TheTitle}{Structure Preserving Model Reduction of Parametric Hamiltonian Systems} 
\newcommand{\TheAuthors}{B. Maboudi Afkham, and J. S. Hesthaven}

% Sets running headers as well as PDF title and authors
\headers{Symplectic Model Reduction of Hamiltonian Systems}{\TheAuthors}

% Title. If the supplement option is on, then "Supplementary Material"
% is automatically inserted before the title.
\title{{\TheTitle}}

% Authors: full names plus addresses.
\author{
	Babak Maboudi Afkham\thanks{Department of Mathematics, Chair of Computational Mathematics and Simulation Science (MCSS), \'Ecole Polytechnique F\'ed\'erale de Lausanne, Switzerland (\email{babak.maboudi@epfl.ch})} 
	\and 
	Jan S. Hesthaven\thanks{Department of Mathematics, Chair of Computational Mathematics and Simulation Schience (MCSS), \'Ecole Polytechnique F\'ed\'erale de Lausanne, Switzerland (\email{jan.hesthaven@epfl.ch})}}

\usepackage{amsopn}
\DeclareMathOperator{\diag}{diag}

%%% Local Variables: 
%%% mode:latex
%%% TeX-master: "ex_article"
%%% End:

### 2017-07-11 greedy_hamiltonian_second_revision/paper.tex
second revision added

--- 2016-12-09 greedy_hamiltonian_revised/paper.tex
@@
 \newcommand{\edit}{\color{black}}
+\newcommand{\blue}{\color{blue}}
 
@@
 While reduced-order models (ROMs) {\edit have been} popular for {\edit efficiently} solving large systems of differential equations, the stability of reduced models over long-time integration {\edit is of present challenges}.
-We present a greedy approach for ROM generation of parametric Hamiltonian systems {\edit that} captures the symplectic structure of Hamiltonian systems to ensure stability of the reduced model.
+We present a greedy approach for {\blue a} ROM generation of parametric Hamiltonian systems {\edit that} captures the symplectic structure of Hamiltonian systems to ensure stability of the reduced model.
 Through the greedy selection of basis vectors, {\edit two new vectors are added at each iteration to the linear vector space} to increase the accuracy of the {\edit reduced} basis.

### 2017-07-11 greedy_hamiltonian_second_revision/chapters/1.introduction.tex
new comments added

@@
 As a result, the reduced system preserves the symplectic structure.
-Although these methods preserve {\edit the} geometric structure, they use a POD-like approach for constructing the reduced basis and are not well {\edit suited} for problems with a high-dimensional parameter domain.
+Although these methods preserve {\edit the} geometric structure, they use a POD-like approach for constructing the reduced basis.
+{\blue If the numerical evaluation of the original model is computationally demanding, performing POD can be excessively expensive \cite{Quarteroni:2016wi}.}
 

### 2017-07-11 greedy_hamiltonian_second_revision/chapters/3.ham.tex
new comments added

@@
 \begin{proof}
-{\blue A detailed proof can be found in \cite{mehl2009perturbation}.
-Since the proof contains an insight to the construction of symplectic bases, we are going to restate the proof.} Starting from a Lagrangain subspace in $E \subset V$ an orthosymplectic basis can be easily constructed.
+{\blue We are going to summarize the proof given in \cite{mehl2009perturbation}.} Starting from a Lagrangain subspace in $E \subset V$ an orthosymplectic basis can be easily constructed.
 By Theorem \ref{theorem:Hasy:1.5} $E$ is $n$ dimensional.
@@
 A matrix of size $2n\times 2k$ satisfying (\ref{eq:Hasy:11}) is called a \emph{symplectic matrix}.
+{\blue We remind the reader that a symplectic matrix is conventionally referred to a square matrix, however, here we may allow symplectic matrices to be also rectangular.}
 

### 2017-07-11 greedy_hamiltonian_second_revision/chapters/3.ham.tex
second revision

@@
 A matrix of size $2n\times 2k$ satisfying (\ref{eq:Hasy:11}) is called a \emph{symplectic matrix}.
-{\blue We remind the reader that a symplectic matrix is conventionally referred to a square matrix, however, here we may allow symplectic matrices to be also rectangular.}
+{\blue We emphasize that a symplectic matrix is conventionally referred to a square matrix, however, here we may allow symplectic matrices to be also rectangular.}
 

### 2017-07-11 greedy_hamiltonian_second_revision/chapters/4.sympmor.tex
second revision

@@
 {\edit Finding eigen-spaces of Hamiltonian and symplectic matrices is studied in the context of optimal control problems \cite{Benner:2000ww,Benner:1997ef,Watkins:2004kz,BunseGerstner:1986dg} and model reduction of Riccati equations \cite{Benner:1997ef}, where also an SVD-like decomposition for Hamiltonian and symplectic matrices has been proposed \cite{Xu:2003kx}.
-However, the computation of a large snapshot matrix and use of the mentioned methods to compute its eigen-spaces, is usually computationally demanding.
+{\blue Specially computation of Lagrangian subspaces of a large scale Hamiltonian matrices using a CS-decomposition is presented in \cite{mehrmann2016inverse,mehrmann2012doubling}.} However, the computation of a large snapshot matrix and use of the mentioned methods to compute its eigen-spaces, is usually computationally demanding.
 Also, these methods generally do not guarantee the construction of a well-conditioned symplectic basis.
@@
 
-Note that the $SR$ method can be replaced with backward stable routines such as the isotropic Arnoldi or the isotropic Lanczos methods \cite{Mehrmann:2000dv}.
+{\blue Note that the $SR$ method is chosen due to its simplicity and it can be replaced with backward stable routines such as the isotropic Arnoldi or the isotropic Lanczos methods \cite{Mehrmann:2000dv}.}
 }
@@
 \end{equation}
-We emphasisze that the sequence of basis vectors generated by the greedy is generally not unique.
+We emphasisze that the sequence of basis vectors generated by the greedy is generally not unique \cite{Quarteroni:2016wi,Anonymous:2016wl}.
 
@@
 Nonlinear terms in Hamiltonian systems often take the from
-\begin{equation}
+\begin{equation} \label{eq:new29}
 	\mathbf g (\mathbf z) = \mathbf g (\mathbf q,\mathbf p) = 
@@
 {\edit Simple calculations verifies that $\mathbf g$ and $P$ then commute.}
+
+{\blue In case $\mathbf g$ is not of the form (\ref{eq:new29}) one can use MDEIM \cite{carlberg2012efficient,negri2015efficient} to accelerate the assembling of the Jacobian matrix.}

### 2017-07-11 response2/response.tex
second response added

\documentclass[a4paper]{article}
\usepackage{url}
\usepackage[margin=1in]{geometry}
\newcommand{\breview}{\begin{quotation}\begin{bf}\noindent}
\newcommand{\ereview}{\end{bf}\end{quotation}}
\newcommand{\reviewbullet}[1]{\breview \begin{itemize}\item #1 \end{itemize}\ereview}

\usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}

\begin{document}
Dear Editor, \\[1cm]

Attached you will find the revised version of manuscript \#M111206, titled ``Structure Preserving Model Reduction of Parametric Hamiltonian Systems".

Yours Sincerely,\\[1.0cm]
Babak Maboudi Afkham and
Jan S. Hesthaven
\\[1cm]

\section*{Response to Reviewer 1}

In our paper we present a greedy method for generation of a structure preserving reduced systems for Hamiltonian systems. The main goal of the method is to save computational costs in the generation of a reduced system. This is done by avoiding possibly a large scale structured eigenvalue problem.

It seems that the main concern of the reviewer is the comparison of the greedy method with the CS-decomposition proposed by Mehrmann and Poloni. The CS-decomposition method is used to construct eigen-spaces (sometimes Lagrangian subspaces) for Hamiltonian and other structured matrices. However, the eigen-decomposition of the snapshot matrix is not the only step in a model reduction routine. Although the CS-decomposition is a nice idea for structured model reduction, but it is not a model reduction technique by itself and it needs further investigation and analysis. For example in most of the references mentioned by the reviewer the CS-decomposition method are applied on a Hamiltonian matrix in the context of the Riccati equation while the snapshot matrix of a parametric Hamiltonian system is not of this form.

Furthermore, the CS-decomposition is generally expensive when dealing with high dimensional parameter domains since the evaluation of the entire snapshot matrix is required. This is precisely what the greedy method is avoiding. Thus, comparison between the two methods is not meaningful in the context of model reduction.

Further reponses to the reviewers comments are below:
\breview
Unfortunately still the revised version of the paper is prepared in a very sloppy way, this concerns a lot of typos, the lack of explanation of terminology and still a missing reflection of the existing literature. While a number of references have been added to the previous work, in several parts the paper is still not given full credit to previous work.
\ereview

Some references regarding the CS-decomposition is added specially regarding computing the eigen-spaces of a structured matrix. The review does not reflect further absences of the existing literature.

\breview
I recall my previous comments: There are also some mathematical questions that need to be addressed. It is well-known that symplectic bases are not norm bounded, so special precau- tions have to be taken to guarantee that they do not become unbounded. This can be done by the isotropic Arnoldi/Gram Schmidt and needs to be done here. Unfortunately this procedure requires the existence of Lagrange subspaces which is not the case here. A way out of this dilemma is the symplectic CS decomposition of the isotropic basis and a different representation of the basis as suggested in recent work by Mehrmann/Poloni.
I do not see the reflection of this unboundedness in the convergence proof, I guess it is reflected in the constant, but this should be analyzed.
\ereview

The existence of Lagrangian subspaces and subsequently the boundedness of the reduced basis and the stability of the reduced system was analysed excessively in the first revision of the paper. Theorem 10 guarantees the boundedness of a symplectic basis, and further in section 4.1.2 it is shown that the basis obtained by the greedy method contains a Lagrangian subspace and hence, norm-bounded by construction. We understand the reviewer's concern about backward instability of symplectic ortho-normalization of the basis. However we believe that the choice of an ortho-normalization routine can be done by the user upon application. This was explicitly pointed out in the first revision of the paper in lines 462-463 (468-469 in the new revision). As the matter of fact some of the reviewer's suggestions, e.g. the isotropic Arnoldi, were introduced as a backward stable replacement for the SR method.

\breview
line 61: this is a biased point of view that should be elaborated.
\ereview

Further comments and proper references to support the argument is added.

\breview
line 247: Proof of thm 10 not needed, it has been shown in work by C. Mehl (SIMAX)
\ereview

The proof of this theorem contains some intuition about the constructive approach of the greedy method. Therefore we decided to restate the proof while citing the original proof.

\breview
line 279: this is not a good definition, symplectic matrices should be square. This is an isotropic matric
line 280: This is a generalized inverse not an inverse.
\ereview

Indeed symplectic matrices are conventionally square matrices. But referring to linear symplectic transformations in a matrix form as isotropic matrices is misleading since these matrices contain a symplectic basis in their column span and not an isotropic basis. Therefore we decided to keep this notation for a better understanding and we added a comment here to avoid any confusion.

\breview
line 462: SR is bad, here is exactly where isotropic Arnoldi is used, and this is exactly what the authors do.
\ereview

As mentioned before different ortho-normalization routines can be used in the greedy method. The choice of the SR method is due to the simplicity of the mthod. However, there are suggestions to backward stable routines such as the isotropic Arnoldi in lines 468-470.

\breview
Line 513: It can be made unique via the CS decomposition
\ereview

Greedy approaches to reduce basis generation do not usually result a unique reduced basis. This is because the uniqueness is compromised with affordability and lower computation cost of the reduced basis. Therefore uniqueness is not really a concerning issue.

\breview
Line 621: here is the critical step of the new method, what happens if the function is not of this form?
\ereview

Further comments are added to remove the concern.

\breview
line 669: if POD is done with a orthosymplectic basis then it is keeping the structure. Here is where the paper gets a bit off the track. POD can be done structure preserving as well, see work by Beattie/Gugercin, Polyuga, Van der Schaft. This should be the method to compare with. Unstructured POD compared with structured reduced basis is not fair.

Line 698: here is where a well elaborated comparison should be done.

line 124: Also here Petrov Galerkin could be used if $W^T V = I$, in particular $W = V^T J$ could be used.
\ereview

As the matter fact the greedy method is compared with the Cotangent Lift method and the Complex SVD method which are structure preserving POD-based model reduction methods. Extensive comparison between the methods are presented in Section 5.

The references mentioned by the reviewer are mostly concerning model reduction of port-Hamiltonian systems. These systems are actually not a Hamiltonian system since the energy is not preserved and consequently there is no symplectic structure. Therefore, applying these methods to a Hamiltonian system is not straight forward and requires further investigation.
\end{document}

### 2017-09-11 response2/response.tex
final version of the greedy hamiltonian uploaded

@@
 However, the eigen-decomposition of the snapshot matrix is not the only step in a model reduction routine.
-Although the CS-decomposition is a nice idea for structured model reduction, but it is not a model reduction technique by itself and it needs further investigation and analysis.
-For example in most of the references mentioned by the reviewer the CS-decomposition method are applied on a Hamiltonian matrix in the context of the Riccati equation while the snapshot matrix of a parametric Hamiltonian system is not of this form.
+Although the CS-decomposition is a nice idea for structured model reduction, it is not a model reduction technique by itself and it needs further investigation and analysis.
+For example in most of the references mentioned by the reviewer the CS-decomposition method is applied on a Hamiltonian matrix in the context of the Riccati equation while the snapshot matrix of a parametric Hamiltonian system is not of this form.
 
@@
 
-Some references regarding the CS-decomposition is added specially regarding computing the eigen-spaces of a structured matrix. The review does not reflect further absences of the existing literature.
+We have again revised the paper and corrected some typos and added further explanations to the terminology.
+Some references regarding the CS-decomposition is added specially regarding computing the eigen-spaces of a structured matrix.
 
@@
 
-The existence of Lagrangian subspaces and subsequently the boundedness of the reduced basis and the stability of the reduced system was analysed excessively in the first revision of the paper.
-Theorem 10 guarantees the boundedness of a symplectic basis, and further in section 4.1.2 it is shown that the basis obtained by the greedy method contains a Lagrangian subspace and hence, norm-bounded by construction.
-We understand the reviewer's concern about backward instability of symplectic ortho-normalization of the basis.
-However we believe that the choice of an ortho-normalization routine can be done by the user upon application.
+The existence of Lagrangian subspaces and subsequently the boundedness of the reduced basis and the stability of the reduced system was analysed extensively in the paper.
+Theorem 10 guarantees the boundedness of a symplectic basis, and further in section 4.1.2 it is shown that the basis obtained by the greedy method contains a Lagrangian subspace and hence, is norm-bounded by construction.
+We appreciate the reviewer's concern about backward instability of symplectic ortho-normalization of the basis.
+However, the choice of an ortho-normalization routine can be done by the user upon application.
 This was explicitly pointed out in the first revision of the paper in lines 462-463 (468-469 in the new revision).
-As the matter of fact some of the reviewer's suggestions, e.g.
+Some of the reviewer's suggestions, e.g.
 the isotropic Arnoldi, were introduced as a backward stable replacement for the SR method.
@@
 
-The proof of this theorem contains some intuition about the constructive approach of the greedy method. Therefore we decided to restate the proof while citing the original proof.
+The proof of this theorem yields some intuition about the constructive approach of the greedy method. Therefore we decided to restate the proof while citing the original proof.
 
@@
 As mentioned before different ortho-normalization routines can be used in the greedy method.
-The choice of the SR method is due to the simplicity of the mthod.
-However, there are suggestions to backward stable routines such as the isotropic Arnoldi in lines 468-470.
+The choice of the SR method is due to the simplicity of the method.
+However, suggestions to backward stable routines such as the isotropic Arnoldi are included, see lines 468-470.
 
@@
 Greedy approaches to reduce basis generation do not usually result a unique reduced basis.
-This is because the uniqueness is compromised with affordability and lower computation cost of the reduced basis.
-Therefore uniqueness is not really a concerning issue.
+This is because the uniqueness is balanced with affordability and lower computation cost of the reduced basis.
+Uniqueness is not really an issue of concern in this context.
 
@@
 
-Further comments are added to remove the concern.
+To accelerate the assembling of the Jacobian matrix for the reduced system, we assumed that the interpolating index matrix $P$ can commute with the nonlinear function $\mathbf g$.
+The assumption holds for many Hamiltonian system.
+This leads to a simple expression for the Jacobian matrix and also a great acceleration in the assembly of the Jacobian matrix.
+However, the assembling of the Jacobian matrix (used in Newton's iterations) does not affect the symplectic structure of the reduced system.
+Therefore, one can use other existing methods to assemble the Jacobian matrix, such as the MDEIM method.
+This is addressed in the revision of the paper, see line 632 and 633.
 
@@
 
-As the matter fact the greedy method is compared with the Cotangent Lift method and the Complex SVD method which are structure preserving POD-based model reduction methods.
+The greedy method is compared with the Cotangent Lift method and the Complex SVD method which are structure preserving POD-based model reduction methods.
 Extensive comparison between the methods are presented in Section 5.
@@
 The references mentioned by the reviewer are mostly concerning model reduction of port-Hamiltonian systems.
-These systems are actually not a Hamiltonian system since the energy is not preserved and consequently there is no symplectic structure.
+These systems are not Hamiltonian systems since the energy is not preserved and consequently there is no symplectic structure.
 Therefore, applying these methods to a Hamiltonian system is not straight forward and requires further investigation.

### 2017-09-11 greedy_hamiltonian_final/paper.tex
final version of the greedy hamiltonian uploaded

% SIAM Article Template
\documentclass[final]{siamart}
\usepackage{mwe}
\usepackage{subfig}

% Information that is shared between the article and the supplement
% (title and author information, macros, packages, etc.) goes into
% ex_shared.tex. If there is no supplement, this file can be included
% directly.

\input{ex_shared}

% Optional PDF information
\ifpdf
\hypersetup{
  pdftitle={\TheTitle},
  pdfauthor={\TheAuthors}
}
\fi

% The next statement enables references to information in the
% supplement. See the xr-hyperref package for details.

\externaldocument{ex_supplement}

% FundRef data to be entered by SIAM
%<funding-group>
%<award-group>
%<funding-source>
%<named-content content-type="funder-name"> 
%</named-content> 
%<named-content content-type="funder-identifier"> 
%</named-content>
%</funding-source>
%<award-id> </award-id>
%</award-group>
%</funding-group>

\newcommand{\edit}{\color{black}}
\newcommand{\blue}{\color{black}}

\begin{document}

\maketitle

% REQUIRED
\begin{abstract}
While reduced-order models (ROMs) {\edit have been} popular for {\edit efficiently} solving large systems of differential equations, the stability of reduced models over long-time integration {\edit is of present challenges}. We present a greedy approach for {\blue a} ROM generation of parametric Hamiltonian systems {\edit that} captures the symplectic structure of Hamiltonian systems to ensure stability of the reduced model. Through the greedy selection of basis vectors, {\edit two new vectors are added at each iteration to the linear vector space} to increase the accuracy of the {\edit reduced} basis. We {\edit use} the error in the Hamiltonian due to model reduction as an error indicator to search the parameter space and identify the next best basis vectors. Under natural assumptions on the set of all solutions of the Hamiltonian system under variation of the parameters, we show that the greedy algorithm converges with exponential rate. Moreover, we demonstrate that combining the greedy basis with the discrete empirical interpolation method also preserves the symplectic structure. This enables the reduction of the computational cost for nonlinear Hamiltonian systems. The efficiency, accuracy, and stability of this model reduction technique is illustrated through simulations of the parametric wave equation and the parametric Schr\"odinger equation.
\end{abstract}

% REQUIRED
\begin{keywords}
  Symplectic model reduction, Hamiltonian system, Greedy basis generation, Symplectic Discrete Empirical Interpolation (SDEIM)
\end{keywords}

% REQUIRED
\begin{AMS}
78M34,34C20,35B30,37K05,65P10,37J25
\end{AMS}

\section{Introduction}
{\edit Parameterized partial differential equations often arise as a model in many problems in engineering and the applied sciences}. While the need for more accuracy has led to the development of exceedingly complex models, the {\edit limitations in computational cost and storage} often make direct approaches {\edit impractical}. Hence, we must seek alternative methods that allow us to approximate the desired output under variation of the input parameters while keeping the computational costs to a minimum.

Reduced basis methods have emerged as a powerful approach for the reduction of the intrinsic complexity of such models \cite{Ito:1998up,Ito:1998ch,Ito:2001ev,Peterson:1989ki}. These methods contain two stages: {\edit the offline stage and the online stage}. In the offline stage, one explores the parameter space to construct a low-dimensional basis that accurately represents the parametrized solution to the partial differential equation. In this stage, the evaluation of the solution of the original model for multiple parameter values is required. The online stage comprises a Galerkin projection onto the span of the reduced basis, which allows exploration of the parameter space at a significantly reduced complexity \cite{Antoulas:2005:ALD:1088857,Anonymous:2016wl}.

Convectional reduced basis techniques, such as {\blue the} Proper Orthogonal Decomposition (POD) \cite{Kunisch:2002er,Atwell:2001ja,Ravindran:2002hn}, require the exploration of the entire parameter space. This leads to a very expensive and often impractical offline {\edit stage} when dealing with multi-dimensional parameter domains. On the other hand, sampling techniques, usually of a greedy nature, search through the parameter space selectively, guided by an error estimate to certify the accuracy of the basis. This approach, accompanied with an efficient sampling procedure, {\edit balances the cost} of computation with the overall accuracy of the reduced-basis \cite{Cuong:2005gd,Rozza:2005ie,Anonymous:2016wl}.

{\edit Besides} computational complexity, another aspect of reduced order modeling is the preservation of structure and, in particular, {\edit the} stability of the original model. In general, reduced order models do not guarantee that such properties are preserved \cite{Anonymous:pMn0O0Q4}.

In the context of Hamiltonian and Lagrangian systems, recent work suggests modifications of POD to preserve {\edit some} geometric structures. Lall et al. \cite{Lall:2003iy} and Carlberg et al. \cite{Carlberg:2014ky} suggests that the reduced-order system should be identified by a Lagrangian function on a low-dimensional configuration space. In this way, the geometric structure of the original system is inherited by the reduced system. {\edit Model reduction for port-Hamiltonian systems can be found in the works of Beattie et al. \cite{Chaturantabut:2016he}, Polyuga et al. \cite{Polyuga:2010gj} and references therein. These works construct a reduced port-Hamiltonian system using Krylov or POD methods that inherit the passivity and stability of the original system.} For Hamiltonian systems, Peng et al. \cite{Peng:2014di}, using a symplectic transformation, constructs a reduced Hamiltonian, as an approximation to the Hamiltonian of the original system. As a result, the reduced system preserves the symplectic structure. Although these methods preserve {\edit the} geometric structure, they use a POD-like approach for constructing the reduced basis. {\blue If the numerical evaluation of the original model is computationally demanding, performing POD can be excessively expensive \cite{Quarteroni:2016wi}.}

In this paper, we present a greedy approach for the construction of a reduced system that preserves the geometric structure of Hamiltonian systems. This technique results in a reduced Hamiltonian system that mimics the symplectic properties of the original system and preserves the Hamiltonian structure and its stability over the course of time. On the other hand, since time integration of the original system is only required once per iteration, the proposed method saves substantial computational cost during the offline stage when compared to alternative POD-like approaches. {\edit It is well known that structured matrices, e.g. symplectic matrices, generally are not well-conditioned \cite{Karow:2006cf}. The greedy update of the symplectic basis presented here, yields a orthosymplectic basis and, therefore, a norm bounded basis.} Moreover, we demonstrate that assumptions, natural for the set of all solutions of the original Hamiltonian system under variation of parameters, lead to exponentially fast convergence of the greedy algorithm. For nonlinear Hamiltonian systems, we show how the basis can be combined with the discrete empirical interpolation method (DEIM) {\edit \cite{Chaturantabut:2010cz,Barrault:2004kz}} to enable a fast evaluation of nonlinear terms while maintaining the symplectic structure.

This paper is organized as follows. Section \ref{chap:MoOr:1} presents a brief overview of model order reduction, POD and DEIM. In Section \ref{chap:Hasy:1} we cover the required topics from symplectic geometry and Hamiltonian systems. Section \ref{chap:SyMo:1} discusses the greedy generation of a symplectic reduced basis as well as other SVD-based symplectic model reduction techniques. Accuracy, stability, and efficiency of the greedy method compared to other SVD-based methods are discussed in Section \ref{chap:NuRe:1}. {\edit Finally we offer some conclusive remarks in Section \ref{chap:Con:1}}.

\section{Model Order Reduction} \label{chap:MoOr:1}
Consider a parameterized, finite dimensional dynamical system described by a set of first order ordinary differential equations
\begin{equation} \label{eq:MoOr:1}
\left\{
\begin{split}
& \frac{d}{dt} \mathbf{x}(t,\omega) = \mathbf f (t,\mathbf x,\omega), \\
& \mathbf x(0,\omega) = \mathbf x_0(\omega).
\end{split}
\right.
\end{equation}
Here $\mathbf x \in \mathbb R^n$ is the state vector, $\omega \in \Gamma$ is a vector containing all the parameters of the system {\edit belonging} to a compact set $\Gamma$ ($\subset \mathbb R^d$) and $\mathbf f : \mathbb R \times \mathbb R^n \times \Gamma \to \mathbb R^n$ is a general {\edit vector valued} function of the state variables and parameters.

We define the solution manifold as the set of all solutions to (\ref{eq:MoOr:1}) under variation of {\edit the parameters in $\Gamma$}
\begin{equation} \label{eq:MoOr:2}
	\edit \mathcal M = \{ \mathbf x(t,\omega) | \omega \in \Gamma,\ t\geq 0 \} \subset \mathbb R^n.
\end{equation}
Note that the exact solution and solution manifold is {\edit often not} available; we assume that we have a numerical integrator that can approximate the solution to (\ref{eq:MoOr:1}) for any realization of $\omega$ {\edit with a given accuracy}. By abuse of notation, we refer to $\mathbf x$ and $\mathcal M$ as the exact solution and the exact solution manifold, respectively, rather than the discrete solution and discrete solution manifold.

Model order reduction is based on the assumption that $\mathcal M$ is of low dimension \cite{Anonymous:2016wl,Antoulas:2005:ALD:1088857} and that the span of appropriately chosen basis vectors $\{v_i\}_{i=1}^k$ covers most of the solution manifold {\edit to within} a small error. The set $\{v_i\}_{i=1}^k$ is denoted as the \emph{reduced basis} and its span as the \emph{reduced space}. Assuming that a $k$-dimensional $(k\ll n)$ reduced basis is given, the approximated solution can be represented as
\begin{equation} \label{eq:MoOr:3}
	\mathbf x \approx V \mathbf y,
\end{equation}
where $V$ is a matrix containing the reduced basis vectors as its columns and $\mathbf y$ {\edit contains} the coordinates of the approximation in this basis. By substituting (\ref{eq:MoOr:3}) into (\ref{eq:MoOr:1}) we obtain the overdetermined system
\begin{equation} \label{eq:MoOr:4}
	V \frac{d}{dt} \mathbf y = \mathbf f (t , V \mathbf y , \omega) + \mathbf r(t,\omega).
\end{equation}
Here we added the residual $\mathbf r$ to emphasize that (\ref{eq:MoOr:4}) is an approximation of (\ref{eq:MoOr:1}). Taking the Petrov-Galerkin projection \cite{Antoulas:2005:ALD:1088857} we construct a basis $W$ of size $n-k$ that is orthogonal to the residual $\mathbf r$ and {\edit requires that} $W^T V$ is invertible. This yields
\begin{equation} \label{eq:MoOr:5}
	\frac{d}{dt} \mathbf y = (W^TV)^{-1} \mathbf f(t,V\mathbf y,\omega).
\end{equation}
Equation (\ref{eq:MoOr:5}) consists of $k$ equations and is called the reduced system. Solving the reduced system instead of the original system can reduce the computational costs {\edit provided} $k$ is significantly smaller than $n$. For nonlinear systems, the evaluation of $\mathbf f$ may still have computational complexity that depends on $n$. We return to this question in detail in Section \ref{chap:MoOr.DEIM:1}.

\subsection{Proper Orthogonal Decomposition} \label{chap:MoOr.PrOr:1}
Let $\mathbf x (t_i,\omega_j)$ with $i=1,\dots,m$ and $j=1,\dots,n$ be a finite number of samples, referred to as \emph{snapshots}, from the solution manifold (\ref{eq:MoOr:2}). If we {\edit assume} that a reduced basis $V$ is provided, the projection operator from $\mathbb R^n$ onto the reduced space can be constructed as $VV^T$. The proper orthogonal decomposition (POD) requires the total error of projecting all the snapshots onto the reduced space to be {\edit minimized}. The POD basis of size $k$ is thus the solution to the optimization problem
\begin{equation} \label{eq:MoOr:6}
\begin{aligned}
& \underset{V\in \mathbb R^{n\times k}}{\text{minimize}}
& & \| S - VV^TS\|_F \\
& \text{subject to}
& & V^TV = I_k
\end{aligned}
\end{equation}
Here $S$ is the snapshot matrix, containing snapshots $\mathbf x(t_i,\omega_j)$ in its columns, $\|\cdot \|_F$ is the Frobenius norm and $I_k$ is the identity matrix of size $k$. According to {\blue the} {\edit Schmidt-Mirsky-Eckart-Young theorem \cite{Markovsky:2011:LRA:2103589}}, the solution to (\ref{eq:MoOr:6}) is equivalent to the truncated singular value decomposition (SVD) of the snapshot matrix $S$ given by
\begin{equation} \label{eq:MoOr:7}
	V = \sigma_1 u_1 v^T_1 + \cdots + \sigma_k u_k v^T_k.
\end{equation}
Here $\sigma_i, u_i$ and $v_i$ are the singular values, the left singular vectors, and the right singular vectors of $S$, respectively {\edit \cite{Markovsky:2011:LRA:2103589} }.

\subsection{Discrete Empirical Interpolation Method (DEIM)} \label{chap:MoOr.DEIM:1} \nocite{Chaturantabut:2010cz}
In this section we {\edit discuss the efficiency} of evaluating nonlinearities {\edit in the context of} projection based reduced models. Suppose that the right hand side in (\ref{eq:MoOr:1}) is of the form $\mathbf f(t,\mathbf x , \omega) = L\mathbf x + \mathbf g(t,\mathbf x ,\omega)$, where $L\in \mathbb R^{n\times n}$ reflects the linear part, and $\mathbf g$ is a nonlinear function. Now {\edit assume that} a $k$-dimensional reduced basis $V$ is provided. The reduced system takes the form
\begin{equation} \label{eq:MoOr:8}
	\frac{d}{dt} \mathbf y = \underbrace{(W^TV)^{-1} L V}_{\tilde L} \mathbf{y} + \underbrace{(W^TV)^{-1} \mathbf g(t,V\mathbf y,\omega)}_{\tilde N (\mathbf y)}.
\end{equation}
Here, $\tilde L$ is a $k\times k$ matrix which can be computed before time integration of the reduced system. However, {\edit the} evaluation of $\tilde N (\mathbf y)$ has a complexity that depends on $n$, the size of the original system. Suppose that the evaluation of $\mathbf g$ with $n$ components has the complexity $\alpha(n)$, for some function $\alpha$. Then the complexity of evaluating $\tilde N(\mathbf y)$ is $\mathcal{O}(\alpha(n) + 4nk)$ which consists of 2 matrix-vector operations and the {\edit evaluation of the nonlinear function, i.e. the evaluation of the nonlinear terms can be as expensive as solving the original system.}

{\edit To overcome this bottleneck we take an approach similar to that of Section \ref{chap:MoOr.PrOr:1} \cite{Chaturantabut:2010cz,Barrault:2004kz}}. Assume that the manifold $\mathcal M_{\mathbf g} = \{ \mathbf g(t,\mathbf x , \omega)| t\in \mathbb R, \mathbf x \in \mathbb R , \omega \in \Gamma\}$ is of a low dimension and that $\mathbf g$ can be approximated by a linear subspace of dimension $m\ll n$, spanned by the basis $\{ u_1 , \dots , u_m \}$, i.e.
\begin{equation} \label{eq:MoOr:10}
	\mathbf g(t,\mathbf x,\omega) \approx U \mathbf c(t,\mathbf x,\omega).
\end{equation}
Here $U$ contains basis vectors $u_i$ and $\mathbf c$ is the vector of coefficients. Now suppose $p_1,\dots,p_m$ are $m$ indices from $\{1,\dots,n\}$ and define an $n\times m$ matrix
\begin{equation} \label{eq:MoOr:11}
	P = [e_{p_1},\dots,e_{p_m}],
\end{equation}
where $e_{p_i}$ is the $p_i$-th column of the identity matrix $I_n$. Multiplying $P$ with $\mathbf g$ selects components $p_1,\dots,p_m$ of $\mathbf g$. If we assume that $P^TU$ is non-singular, the coefficient vector $\mathbf c$ can be uniquely determined from
\begin{equation} \label{eq:MoOr:12}
	P^T \mathbf g = (P^TU)\mathbf c.
\end{equation}
Finally the approximation of $\mathbf g$ is determined by
\begin{equation} \label{eq:MoOr:13}
	\mathbf g(t,\mathbf x,\omega) \approx U \mathbf c(t,\mathbf x,\omega) = U (P^TU)^{-1} P^T \mathbf g(t,\mathbf x,\omega),
\end{equation}
which is referred to as the \emph{Discrete Empirical Interpolation} (DEIM) approximation \cite{Chaturantabut:2010cz}. Applying DEIM to the reduced system (\ref{eq:MoOr:5}) yields
\begin{equation} \label{eq:MoOr:14}
	\frac{d}{dt} \mathbf y = \tilde L \mathbf y + (W^TV)^{-1} U(P^TU)^{-1}P^T \mathbf g(t,V\mathbf y , \omega).
\end{equation}
Note that the matrix $(WV)^{-1} U(P^TU)^{-1}$ can be computed offline and since $\mathbf g$ is evaluated only at $m$ of its components, the evaluation of the nonlinear term in (\ref{eq:MoOr:14}) does not depend on $n$.

{\edit To obtain the projection basis $U$, the POD can be applied to the ensemble of samples of the nonlinear term $\mathbf g(t_i,\mathbf x, \omega_j)$ with $i=1,\dots,m$ and $j=1,\dots,n$}. There is no additional cost {\edit associated with} computing the nonlinear snapshots, since they are generated when computing the trajectory snapshot matrix $S$. The interpolating indices $p_1,\dots,p_m$ can be constructed as follows. Given the projection basis $U = \{u_1,\dots,u_m\}$, the first interpolation index $p_1$ is chosen according to the component of $u_1$ with the largest magnitude. The rest of the interpolation indices, $p_2,\dots,p_m$ correspond to the component of the largest magnitude of the residual vector $\mathbf r = u_l - U \mathbf c$. It is shown in \cite{Chaturantabut:2010cz} that if the residual vector is a nonzero vector in each iteration then $P^TU$ is non-singular and (\ref{eq:MoOr:13}) is well defined.

\begin{algorithm} 
\caption{Discrete Empirical Interpolation Method} \label{alg:MoOr:1}
{\bf Input:}  Basis vectors $\{u_1,\dots , u_m\}\subset \mathbb R^n$
\begin{enumerate}
\item pick $p_1$ to be the index of the largest component of $u_1$.
\item $U \leftarrow [u_1]$
\item $P \leftarrow [p_1]$
\item \textbf{for} $i\leftarrow 2$ \textbf{to} $m$
\item \hspace{0.5cm} solve $(P^TU)\mathbf c = P^T u_i$ for $\mathbf c$
\item \hspace{0.5cm} $\mathbf r \leftarrow u_i - U\mathbf c$
\item \hspace{0.5cm} pick $p_i$ to be the index of the largest component of $\mathbf r$
\item \hspace{0.5cm} $U \leftarrow [u_1,\dots,u_i]$
\item \hspace{0.5cm} $P \leftarrow [p_1,\dots,p_i]$
\item \textbf{end for}
\end{enumerate}
\vspace{0.5cm}
{\bf Output:} Interpolating indices $\{p_1,\dots,p_m\}$
\end{algorithm}

{\edit The numerical solution of (\ref{eq:MoOr:8}) may involve the computation of the Jacobian of the nonlinear function $\mathbf g(t,\mathbf x, \omega)$ with respect to the reduced state variable $\mathbf y$}
\begin{equation} \label{eq:MoOr:9}
	\edit \mathbf J_{\mathbf y}(\mathbf g) = (W^TV)^{-1} \mathbf J_{\mathbf x}(\mathbf g) V,
\end{equation}
{\edit where $\mathbf J_\alpha(\mathbf g)$ is the Jacobian matrix of $\mathbf g$ with respect to the variable $\alpha$.} The complexity of (\ref{eq:MoOr:9}) is $\mathcal{O}(\alpha(n) +2n^2k+2nk^2+2nk)$, comprising several matrix-vector multiplications and an evaluation of the Jacobian which depends on the size of the original system. Approximating the Jacobian in (\ref{eq:MoOr:9}) is usually both problem and discretization dependent. Often the nonlinear function $\mathbf g$ is evaluated component-wise i.e.
\begin{equation} \label{eq:MoOr:15}
	\mathbf g(\mathbf x) =
	\begin{pmatrix}
		g_1(x_1,\dots,x_n) \\
		g_2(x_1,\dots,x_n) \\
		\vdots \\
		g_n(x_1,\dots,x_n)
	\end{pmatrix}
	=
	\begin{pmatrix}
		g_1(x_1) \\
		g_2(x_2) \\
		\vdots \\
		g_n(x_n)
	\end{pmatrix}.
\end{equation}
In such cases the interpolating index matrix $P$ and the nonlinear function $\mathbf g$ commute, i.e.,
\begin{equation} \label{eq:MoOr:16}
	\tilde N(\mathbf y) \approx (W^TV)^{-1} U(P^TU)^{-1}P^T \mathbf g(V\mathbf y) = (W^TV)^{-1} U(P^TU)^{-1}\mathbf g(P^TV\mathbf y)
\end{equation}
If we now take the Jacobian of the approximate function we recover
\begin{equation}
	\edit \mathbf J_{\mathbf y}(\mathbf g) = \underbrace{ (W^TV)^{-1} U(P^TU)^{-1} }_{k\times m} \underbrace{ \mathbf J_{\mathbf x}(\mathbf g(P^T V \mathbf y) ) }_{m\times m} \underbrace{P^T V}_{m\times k}.
\end{equation}
The matrix $(WV)^{-1} U(P^TU)^{-1}$ can be computed offline and the Jacobian is evaluated only for $m\times m$ components. Hence the overall complexity of computing the Jacobian is now independent of $n$. {\edit We refer the reader to \cite{Barrault:2004kz,Chaturantabut:2010cz} for more detail.}

\section{Hamiltonian Systems and Symplectic Geometry} \label{chap:Hasy:1}
Let $\edit \mathcal M$ be a manifold and $\edit \Omega:\mathcal M \times \mathcal M \to \mathbb R$ be a closed, nondegenerate {\edit and skew-symmetric} 2-form on $\edit \mathcal M$. The pair $\edit (\mathcal M,\Omega)$ is called a \emph{symplectic manifold} {\edit \cite{Marsden:1999ck}}.

{\edit Let $(\mathcal M,\Omega)$ be a symplectic manifold and suppose that $H:\mathcal M \to \mathbb R$ is a smooth scalar function. The differential of $H$, denoted by $\mathbf dH$, defines a 1-form on $\mathcal M$. {\blue The nondegeneracy of $\Omega$ implies that there is a unique vector field $X_H$, \emph{the Hamiltonian vector field }\cite{da2003introduction,Marsden:1999ck}, on $\mathcal M$ such that}
\begin{equation} \label{eq:Hasy:1}
	\edit i_{X_H} \Omega = \mathbf dH. 
\end{equation}
}
{\blue Here $i_{X_H} \Omega$ is the interior product of $X_H$ with $\Omega$, i.e.,}
\begin{equation}
	\edit \Omega(X_H,Y) = \mathbf dH(Y),
\end{equation}
{\edit for any vector field $Y$ on $\mathcal M$.} Note that when $\edit \mathcal M$ belongs to a Euclidean space then $\mathbf d H = \nabla_z H$. The equations of evolution are then defined by
\begin{equation} \label{eq:Hasy:2}
	\dot z = X_H(z)
\end{equation}
and known as \emph{Hamilton's equation} \cite{Marsden:1999ck}. A fundamental feature of Hamiltonian systems is the conservation of the Hamiltonian along integral curves on $\edit \mathcal M$. To emphasize the importance of this property we recall {\edit \cite{Marsden:1999ck}}

\begin{theorem} \label{theorem:Hasy:1}
Suppose that $X_H$ is a Hamiltonian vector field with the flow $\phi_t$ on a symplectic manifold $\mathcal M$. Then $H\circ \phi_t = H$.
\end{theorem}

\begin{proof}
$H$ is constant along integral curves since
\begin{equation} \label{eq:Hasy:3}
\begin{aligned}
	\frac{d}{dt}(H\circ \phi_t)(z) &= \mathbf d H(\phi_t(z)) \cdot( \frac{d}{dt} \phi_t(z) ) \\
	&= \mathbf d H (\phi_t(z))\cdot X_H(\phi_t(z)) \\
	&= \Omega_z( X_H(\phi_t(z)), X_H(\phi_t(z)) ) = 0,
\end{aligned}
\end{equation}
{\edit by using} the chain rule and bilinearity of $\Omega$ in the argument.
\end{proof}

For the case where the symplectic manifold is also a linear vector space, the pair $({\edit \mathcal M},\Omega)$ is also referred to as a \emph{symplectic vector space}. We {\edit need} the following theorems regarding symplectic vector spaces and refer the reader to \cite{de2006symplectic,Marsden:1999ck,Silva01lectureson} for detailed proofs.

{\edit
\begin{theorem} \label{theorem:Hasy:1.1} \cite{Marsden:1999ck}
If $(V,\Omega)$ is a symplectic vector space then $\Omega$ is a constant form, that is $\Omega_z$ is independent of $z\in V$. 
\end{theorem}
\begin{theorem} \label{theorem:Hasy:1.2} \cite{Marsden:1999ck}
If $(V,\Omega)$ is a finite-dimensional symplectic manifold then $V$ is even dimensional.
\end{theorem}
\begin{theorem} \label{theorem:Hasy:1.3} \cite{de2006symplectic}
(The Symplectic Gram-Schmidt) If $(V,\Omega)$ is a $2n$-dimensional symplectic vector space, then there is a basis $e_1,\dots e_n,f_1, \dots , f_n$ of $V$ such that
\begin{equation} \label{eq:Hasy:4}
\begin{aligned}
	& \Omega(e_i,e_j) = 0 = \Omega(f_i,f_j), \quad & i\neq j,\\
	& \Omega(e_i,f_j) = \delta_{ij}, & i\leq i,j \leq n.
\end{aligned}
\end{equation}
where $\delta$ is the Kronecker's delta function. {\blue Moreover, if $V = \mathbb{R}^{2n}$ then} we can choose basis vectors $\{e_i,f_i\}_{i=1}^n$ such that
\begin{equation} \label{eq:Hasy:5}
	\Omega(v_1,v_2) = v_1^T \mathbb J_{2n} v_2, \qquad v_1,v_2\in \mathbb R^n,
\end{equation}
with $\mathbb J_{2n}$ being the {\blue standard} symplectic matrix, defined as
\begin{equation} \label{eq:Hasy:6}
	\mathbb{J}_{2n} = 
	\begin{pmatrix}
		0_n & I_n \\
		-I_n & 0_n
	\end{pmatrix}.
\end{equation}
Here $I_n$ and $0_n$ is the identity matrix and the zero square matrix of size $n$, respectively.
\end{theorem}
\begin{theorem} \label{theorem:Hasy:1.4} \cite{Marsden:1999ck}
The classical inner product $\langle \cdot,\cdot \rangle:\mathbb R^{2n}\times \mathbb R^{2n}\to \mathbb R$ can be written in terms of the 2-form as
\begin{equation}
	\langle v,u \rangle = \Omega(\mathbb J_{2n}v,u),\quad \forall u,v \in \mathbb R^{2n}.
\end{equation}
\end{theorem}
}

{\edit 
\begin{definition}  \cite{de2006symplectic}
Suppose $(V,\Omega)$ is a finite dimensional symplectic vector space and $E\subset V$ is a subspace. Then the symplectic complement of $E$ inside $V$ is defined as
\[
	E^{\perp} := \{ v\in V |\ \Omega(v,e) = 0,\ \forall e\in E \}
\]
\end{definition}
Note that $E \cap E^{\perp}$ is not empty in general. 
\begin{definition} \cite{de2006symplectic}
Suppose $(V,\Omega)$ is a finite dimensional symplectic vector space. A subspace $E\subset V$ is called a Lagrangian subspace inside $V$ if $E = E^\perp$.
\end{definition}
\begin{theorem} \label{theorem:Hasy:1.5} \cite{abraham1978foundations}
Suppose $(V,\Omega)$ is a finite dimensional symplectic vector space. If $E\subset V$ is a Lagrangian subspace then $dim(E)=\frac 1 2dim(V)$. Here $dim$ denotes the dimension of the subspace.
\end{theorem}
\begin{definition}
A basis of $(V,\Omega)$ is called orthosymplectic if it is both a symplectic basis and an orthogonal basis with respect to the classical scalar product.
\end{definition}
\begin{theorem} \label{theorem:Hasy:1.6}  \cite{mehl2009perturbation,da2003introduction}
Suppose $(V,\Omega)$ is a $2n$ dimensional symplectic vector space and $E\subset V$ is a Lagrangian subspace. Then there is an orthosymplectic basis for $V$.
\end{theorem}
\begin{proof}
{\blue We are going to summarize the proof given in \cite{mehl2009perturbation}.} Starting from a Lagrangain subspace in $E \subset V$ an orthosymplectic basis can be easily constructed. By Theorem \ref{theorem:Hasy:1.5} $E$ is $n$ dimensional. Suppose that $\{ e'_1,\dots, e'_n \}$ is a basis for $E$, using the classical Gram-Schmidt orthogonalization process we can construct an orthonormal basis $\{ e_1,\dots,e_n \}$. Define a new set of vectors $f_1 = \mathbb J_{2n}^Te_1$, $f_2 =\mathbb J_{2n}^T e_2$, $\dots$, $f_n= \mathbb J_{2n}^Te_n$. We have
\begin{equation}
	\langle f_i, f_j \rangle = e_i^T \mathbb J_{2n} {\mathbb J_{2n}}^T e_j = \delta_{ij}, \quad \langle f_i, e_j \rangle = e_i^T \mathbb J_{2n} e_j = 0, \quad i,j=1,\dots,n,
\end{equation}
where we used the fact that $\mathbb J_{2n} {\mathbb J_{2n}}^T = I_{2n}$ in the first identity and the second identity is due to the fact that the basis $\{ e_1,\dots,e_n \}$ forms a Lagrangian subspace. This shows that the set $\{ e_1,\dots,e_n \}\cup \{ f_1,\dots,f_n \}$ forms an orthonormal basis. Also, it can be easily verified that this is a symplectic basis. Thus $\{ e_1,\dots,e_n \}\cup \{ f_1,\dots,f_n \}$ is an orthosymplectic basis.
\end{proof}
\begin{theorem} \label{theorem:Hasy:1.7} \cite{Marsden:1999ck}
On a finite-dimensional symplectic vector space the relationship (\ref{eq:Hasy:1}) becomes 
\begin{equation} \label{eq:Hasy:7}
\left\{
\begin{aligned}
	&\dot {\mathbf z} = \mathbb{J}_{2n} \nabla_{\mathbf z} H(\mathbf z), \\
	& \mathbf z(0) = \mathbf z_0.
\end{aligned}
\right.
\end{equation}
or, by introducing the canonical coordinates $\mathbf z = (\mathbf q^T, \mathbf p^T)^T$,
\begin{equation} \label{eq:Hasy:8}
\left\{
\begin{aligned}
	&\dot {\mathbf q} = \nabla_{\mathbf p} H(\mathbf q,\mathbf p),\\
	&\dot {\mathbf p} = - \nabla_{\mathbf q} H(\mathbf q,\mathbf p).
\end{aligned}
\right.
\end{equation}
\end{theorem}
}

{\edit Let} us now introduce \emph{symplectic transformations}, i.e., mappings between symplectic manifolds which preserve the 2-form $\Omega$. The accurate numerical treatment of Hamiltonian systems often requires preservation of the symmetry expressed in Theorem \ref{theorem:Hasy:1}. Symplectic transformations can be used to construct such symmetry preserving numerical methods.

{\edit
\begin{definition}
Let $(V,\Omega)$ and $(W,\Pi)$ be two linear symplectic vector spaces of dimensions $2n$ and $2k$, respectively. A linear mapping $\phi:V \to W$ is called \emph{symplectic} or \emph{canonical} if
\begin{equation} \label{eq:Hasy:9}
	\Omega = \phi^* \Pi
\end{equation}
where $\phi^* \Pi$ is the pullback of $\Pi$ by $\phi$, i.e. for all $\mathbf{z}_1, \mathbf{z}_2\in V$
\begin{equation}
	\Omega(\mathbf{z}_1,\mathbf{z}_2) = \Pi(\phi(\mathbf{z}_1),\phi(\mathbf{z}_2)).
\end{equation}
\end{definition}

Note that if we represent the transformation $\phi$ as a matrix $A\in \mathbb R^{2n\times 2k}$ condition (\ref{eq:Hasy:9}) is equivalent to \cite{Marsden:1999ck}}

\begin{equation} \label{eq:Hasy:11}
	A^T \mathbb{J}_{2n}A = \mathbb{J}_{2k}.
\end{equation}
A matrix of size $2n\times 2k$ satisfying (\ref{eq:Hasy:11}) is called a \emph{symplectic matrix}. {\blue We emphasize that a symplectic matrix is conventionally referred to a square matrix, however, here we may allow symplectic matrices to be also rectangular.}

\begin{definition}
	The \emph{symplectic inverse} of a matrix $A\in \mathbb{R}^{2n\times 2k}$ is denoted by $A^+$ and defined by {\edit \cite{Peng:2014di}}
\begin{equation}\label{eq:Hasy:12}
	A^+ := \mathbb{J}_{2k}^T A^T \mathbb{J}_{2n}.
\end{equation}
\end{definition}
We point out the properties of the symplectic inverse and refer the reader to \cite{Peng:2014di} for detailed proof.
\begin{lemma} \label{lemma:Hasy:1}
Let $A\in \mathbb{R}^{2n\times 2k}$ be a symplectic matrix and $A^+$ its symplectic inverse as defined in (\ref{eq:Hasy:12}). Then ${(A^+)}^T$ is a symplectic matrix and $A^+A = I_{2k}$.
\end{lemma}

{\edit A straight-forward calculation verifies} that $AA^+$ is idempotent, i.e., a symplectic projection onto the column span of $A$.

{\edit It is natural to expect a numerical integrator that solves (\ref{eq:Hasy:7}) to also satisfy the conservation law in Theorem \ref{theorem:Hasy:1}. Common numerical integrators e.g., Runge-Kutta methods, do not generally preserve the Hamiltonian which results in a qualitative wrong behavior of the solution \cite{Hairer:1250576}. Symplectic integrators are a class of numerical integrators for Hamiltonian systems that preserve the symplectic structure and ensure stability in long-time integration.} The St\"ormer-Verlet time stepping scheme is an example of symplectic integrators and is given by
\begin{equation} \label{eq:Hasy:13}
\begin{aligned}
	q_{n+1/2} &= q_n + \frac{\Delta t}{2} \nabla_pH(q_{n+1/2},p_n), \\
	p_{n+1} &= p_n - \frac{\Delta t}{2} \left( \nabla_qH(q_{n+1/2},p_n) + \nabla_qH(q_{n+1/2},p_{n+1}) \right),\\
	q_{n+1} &= q_{n+1/2} + \frac{\Delta t}{2} \nabla_pH(q_{n+1/2},p_{n+1}),
\end{aligned}
\end{equation}
and
\begin{equation} \label{eq:Hasy:14}
\begin{aligned}
	p_{n+1/2} &= p_n - \frac{\Delta t}{2} \nabla_qH(q_{n},p_{n+1/2}), \\
	q_{n+1} &= q_n + \frac{\Delta t}{2} \left( \nabla_pH(q_{n},p_{n+1/2}) + \nabla_pH(q_{n+1},p_{n+1/2}) \right),\\
	p_{n+1} &= p_{n+1/2} - \frac{\Delta t}{2} \nabla_qH(q_{n+1},p_{n+1/2}).
\end{aligned}
\end{equation}
For a general Hamiltonian system, the St\"ormer-Verlet scheme is implicit. However, for separable Hamiltonians, i.e. $H(q,p) = K(p) + U(q)$, this {\edit scheme becomes} explicit. We refer the reader to \cite{Hairer:1250576} for more information about the construction and applications of symplectic and geometric numerical integrators.

\section{Symplectic Model Reduction} \label{chap:SyMo:1}
We now {\edit discuss} how to modify reduced order modeling to ensure that {\edit the resulting scheme preserves} the symplectic structure of the Hamiltonian system.

Consider a Hamiltonian system (\ref{eq:Hasy:7}) on a $2n$-dimensional symplectic vector space $\edit (V,\Omega)$. Suppose that the solution manifold $\mathcal M_H$ is well approximated by a low dimensional symplectic subspace $\edit (W,\Omega)$ of dimension $2k$ $(k\ll n)$. We can {\edit then} construct a symplectic basis $A$ for $\edit W$ and approximate the solution to (\ref{eq:Hasy:7}) as
\begin{equation} \label{eq:SyMo:1}
	\mathbf z \approx A\mathbf y.
\end{equation}
Substituting this into (\ref{eq:Hasy:7}) we obtain
\begin{equation} \label{eq:SyMo:2}
	A\mathbf y = \mathbb{J}_{2n} \nabla_{\mathbf z} H(A \mathbf y). 
\end{equation}
Multiplying both sides with the symplectic inverse of $A$ and using the chain rule we have
\begin{equation} \label{eq:SyMo:3}
	\mathbf y = A^+ \mathbb J_{2n} (A^+)^T \nabla_{\mathbf y} H(A\mathbf y).
\end{equation}
Since $A$ is a symplectic basis, Lemma \ref{lemma:Hasy:1} ensures that $(A^+)^T$ is a symplectic matrix i.e., $A^+ \mathbb J_{2n} (A^+)^T = \mathbb{J}_{2k}$. By defining the reduced Hamiltonian $\tilde H:\mathbb R^{2k} \to \mathbb R$ as $\tilde H (y) = H(Ay)$ we obtain the reduced system
\begin{equation} \label{eq:SyMo:4}
\left\{
\begin{aligned}
	 \frac{d}{dt} \mathbf y &= \mathbb J_{2k} \nabla_{\mathbf y} \tilde H(\mathbf y), \\
	 \mathbf y_0 &= A^+ \mathbf z_0.
\end{aligned}
\right.
\end{equation}
The system obtained from the Petrov-Galerkin projection in (\ref{eq:MoOr:5}) is not a Hamiltonian system and does not guarantee conservation of the symplectic structure. On the other hand, we observe that the reduced system in (\ref{eq:SyMo:4}) is of the form (\ref{eq:Hasy:7}) and, hence, is a Hamiltonian system, i.e. the symplectic structure will be conserved along integral curves of (\ref{eq:SyMo:4}). Note that the original and the reduced systems are {\edit endowed with} different Hamiltonians. In the next proposition we show that the error in the Hamiltonian is constant in time.

\begin{proposition}
Let $\mathbf{z} (t)$ be the solution of (\ref{eq:Hasy:7}) at time $t$. Further suppose that $\tilde{\mathbf{z}} (t)$ is the approximate solution of the reduced system (\ref{eq:SyMo:4}) in the original coordinate system. Then the error in the Hamiltonian defined by
\begin{equation} \label{eq:SyMo:5}
	\Delta H(t)  = |H(\mathbf z(t)) - H(\tilde{\mathbf z}(t))|,
\end{equation}
is constant for all $t\in \mathbb R$.
\end{proposition}

\begin{proof}
	Let $\phi_t$ and $\psi_t$ be the Hamiltonian flow of the original and the reduced system respectively. By definition $\mathbf z(t) = \phi_t(\mathbf z_0)$ and $\mathbf y(t) = \psi_t(\mathbf y_0)$. Using the definition of the reduced Hamiltonian and Theorem \ref{theorem:Hasy:1} we have
\begin{equation} \label{eq:SyMo:6}
\begin{aligned}
	H(\tilde{\mathbf{z}} (t)) = H( A\mathbf y (t) ) = \tilde H(\mathbf y (t)) = \tilde H(\psi_t(\mathbf y_0)) = \tilde H(\mathbf y_0) = \tilde H(A^+ \mathbf z_0) = H(AA^+\mathbf z_0).
\end{aligned}
\end{equation}
The error in the Hamiltonian can then be written in terms of $\mathbf z_0$ and the symplectic basis $A$ as
\begin{equation} \label{eq:SyMo:7}
	\Delta H(t) = |H(\mathbf z_0) - H(AA^+\mathbf z_0)|
\end{equation}
\end{proof}

{\edit 
The following theorems provide a strong indication of the stability of the reduced system.

\begin{definition} \label{definition:SyMo:1} \cite{bhatia2002stability}
Consider a dynamical system of the form $\dot{\mathbf z} = \mathbf f(\mathbf z)$ and suppose that $\mathbf z_e$ is an equilibrium point for the system so that $\mathbf f(\mathbf z_e) = 0$. $\mathbf z_e$ is called nonlinearly stable or Lyapunov stable if, for any $\epsilon > 0$, we can find $\delta > 0$ such that for any trajectory $\phi_t$, if $\| \phi_0 - \mathbf z_e \|_2 \leq \delta$, then for all $0 \leq t < \infty$, we have $\| \phi_t - \mathbf z_e \|_2 < \epsilon$, where $\| \cdot \|_2$ is the Euclidean norm.
\end{definition}	
The following proposition, also known as Dirichlet's theorem \cite{bhatia2002stability}, states the sufficient condition for an equilibrium point to be Lyapunov stable. We refer the reader to \cite{bhatia2002stability} for the proof.
\begin{proposition} \label{proposition:SyMo:1} \cite{bhatia2002stability}
An equilibrium point $\mathbf z_e$ is Lyapunov stable if there exists a scalar function $W : \mathbb R^{n} \to  \mathbb R$ such that $\nabla W(\mathbf z_e) = 0$, $\nabla^2 W(\mathbf z_e)$ is positive definite, and that for any trajectory $\phi_t$ defined in the neighborhood of $\mathbf z_e$, we have $\frac{d}{dt} W(\phi_t) \leq 0$. Here $\nabla^2W$ is the Hessian matrix of $W$.
\end{proposition}
The scalar function $W$ is referred to as the \emph{Lyapunov function}. In the context of the Hamiltonian systems, a suitable candidate for the Lyapunov function is the Hamiltonian function $H$. The following theorem shows that when $H$ (or $-H$) is a Lyapunov function, then the equilibrium points of the original and the reduced system are Lyapunov stable \cite{abraham1978foundations}. 
\begin{theorem} \label{theorem:SyMo:1}
Consider a Hamiltonian system of the form (\ref{eq:Hasy:7}) together with the reduced system (\ref{eq:SyMo:4}). Suppose $\mathbf z_e$ is an equilibrium point for (\ref{eq:Hasy:7}) and that $\mathbf y_e = A^+\mathbf z_e$. If $H$ (or $-H$) is a Lyapunov function satisfying Proposition \ref{proposition:SyMo:1}, then $\mathbf z_e$ and $\mathbf y_e$ are Lyapunov stable equilibrium points for (\ref{eq:Hasy:7}) and (\ref{eq:SyMo:4}), respectively. 
\end{theorem}
\begin{proof}
	It is a direct consequence of Proposition \ref{proposition:SyMo:1} that $\mathbf z_e$ is a local minimum or maximum of (\ref{eq:Hasy:7}) and also a Lyapunov stable point. It can be easily checked that if $\mathbf z_e$ is a local minimum of $H$ then $\mathbf y_e$ is a local minimum for $\tilde H$ and an equilibrium point for (\ref{eq:SyMo:4}). Also from the chain rule we have
\[
	\nabla^2_{\mathbf y} \tilde H = A^T \nabla^2_{\mathbf z} H A.
\]
So for any $\xi\in \mathbb R^{2k}$
\[
	\xi^T \nabla^2_{\mathbf y} \tilde H \xi = (A\xi)^T \nabla^2_{\mathbf z} H (A\xi) \geq 0.
\]
Here the last inequality is due to the positive definiteness of $H$. Therefore $\tilde H$ is also positive definite. By Proposition \ref{proposition:SyMo:1} we conclude that $\mathbf y_e$ is a Lyapunov stable point.
\end{proof}
}

While the symplectic structure is not guaranteed to be preserved in the reduced systems obtained by the Petrov-Galerkin projection, the reduced system obtained by the symplectic projection guarantees the preservation of the energy up to the error in the Hamiltonian (\ref{eq:SyMo:5}). In the next section we discuss  different methods for obtaining a symplectic basis.

\subsection{Proper Symplectic Decomposition (PSD)} \label{chap:SyMo.PrSy:1}

Similar to Section \ref{chap:MoOr.PrOr:1} we gather snapshots $\mathbf z_i = [q_i^T , p_i^T]^T$ in the snapshot matrix $S$. Suppose that a symplectic basis $A$ of size $2n\times2k$ and its symplectic inverse $A^+$ is provided. {\edit The Proper Symplectic Decomposition} requires that the error of the symplectic projection onto the symplectic subspace {\edit be minimized}. Hence, the PSD symplectic basis of size $2k$ is the solution to the optimization problem

\begin{equation} \label{eq:SyMo:8}
\begin{aligned}
& \underset{V\in \mathbb R^{2n\times 2k}}{\text{minimize}}
& & \| S - AA^+S\|_F \\
& \text{subject to}
& & A^T \mathbb{J}_{2n}A = \mathbb{J}_{2k}
\end{aligned}
\end{equation}
Compared to POD, in (\ref{eq:SyMo:8}) the orthogonal projection is replaced with a symplectic projection $AA^+$. At first, the minimization looks similar to the one obtained by POD. {\edit However, it is well known that symplectic bases are not generally orthogonal, and therefore not norm bounded. This means that numerical errors may become dominant in the symplectic projection \cite{Karow:2006cf} which makes the minimization (\ref{eq:SyMo:8}) a harder problem than (\ref{eq:MoOr:6}).}

As the optimization problem (\ref{eq:SyMo:8}) is nonlinear, the direct solution is usually expensive. A simplified version of the optimization (\ref{eq:SyMo:8}) can be found in \cite{Peng:2014di}, but there is no guarantee that the method provides a near optimal basis.

{\edit Finding eigen-spaces of Hamiltonian and symplectic matrices is studied in the context of optimal control problems \cite{Benner:2000ww,Benner:1997ef,Watkins:2004kz,BunseGerstner:1986dg} and model reduction of Riccati equations \cite{Benner:1997ef}, where also an SVD-like decomposition for Hamiltonian and symplectic matrices has been proposed \cite{Xu:2003kx}. {\blue Specially computation of Lagrangian subspaces of a large scale Hamiltonian matrices using a CS-decomposition is presented in \cite{mehrmann2016inverse,mehrmann2012doubling}.} However, the computation of a large snapshot matrix and use of the mentioned methods to compute its eigen-spaces, is usually computationally demanding. Also, these methods generally do not guarantee the construction of a well-conditioned symplectic basis.

The greedy approach presented in Section \ref{Chap:Symo.PrSy:3} is an iterative method for construction of a symplectic basis. It avoids the evaluation of the full snapshot matrix, hence substantially reduces the computational cost in the offline stage of the symplectic model reduction. Also, by construction, it yields an orthosymplectic basis and therefore a well-conditioned basis.

In Section \ref{chap:SyMo.PrSy:2} we briefly outline non-direct methods for finding solutions to (\ref{eq:SyMo:8}), proposed by \cite{Peng:2014di}, and assuming a specific structure for $A$. In Section \ref{Chap:Symo.PrSy:3} we introduce a greedy approach for the symplectic basis generation.}

\subsubsection{SVD Based Methods for Symplectic Basis Generation} \label{chap:SyMo.PrSy:2}

\paragraph{\bf Cotangent lift} Suppose that $A$ is of the form
\begin{equation} \label{eq:SyMo:9}
	A = 
	\begin{pmatrix}
		\Phi & 0 \\
		0 & \Phi
	\end{pmatrix},
\end{equation}
where $\Phi \in \mathbb{R}^{n\times k}$ is an orthonormal matrix. It is easy to check that $A$ is a symplectic matrix, i.e., $A^T \mathbb J_{2n} A = \mathbb J_{2k}$. The construction of $A$ suggests that the range of $\Phi$ should cover both the potential and the momentum spaces. Hence, we can construct $A$ by forming the combined snapshot matrix
\begin{equation} \label{eq:SyMo:10}
	S_{\text{combined}} = [q_1,\dots,q_n,p_1,\dots,p_n], \qquad \mathbf z_i = (q_i^T,p_i^T)^T,
\end{equation}
and define $\Phi=[u_1,\dots,u_k]$, where $u_i$ is the $i$-th left singular vector of $S_{\text{combined}}$. It is shown in \cite{Peng:2014di} that among all symplectic bases of the form (\ref{eq:SyMo:9}) cotangent lift minimizes the projection error.

\paragraph{\bf Complex SVD} Suppose instead that $A$ takes the form \cite{Peng:2014di}
\begin{equation} \label{eq:SyMo:11}
	A = 
	\begin{pmatrix}
		\Phi & -\Psi \\
		\Psi & \Phi
	\end{pmatrix},
\end{equation}
while $\Phi$ and $\Psi$ are real matrices of size $n\times k$ satisfying conditions
\begin{equation} \label{eq:SyMo:12}
\Phi^T \Phi + \Psi^T \Psi = I_k,\quad \Phi^T \Psi = \Psi^T \Phi.
\end{equation}
It can be checked that $A$ forms a symplectic matrix. To construct $A$ we first define the complex snapshot matrix
\begin{equation} \label{eq:SyMo:13}
	S_{\text{complex}} = [ q_1 + i p_1, \dots , q_N + i p_N ].
\end{equation}
Each left singular vector of $S_{\text{complex}}$ now takes the form $u_m = r_m + i s_m$. We define
\begin{equation} \label{eq:SyMo:14}
	 \Phi = [r_1,\dots, r_k], \quad \Psi = [s_1,\dots, s_k].
\end{equation}
One can easily check that (\ref{eq:SyMo:12}) is satisfied {\edit since the matrix of singular vectors is unitary}. It is shown in \cite{Peng:2014di} that among all symplectic bases of the form (\ref{eq:SyMo:11}) the complex SVD minimizes the projection error.

\subsubsection{The Greedy Approach to Symplectic Basis Generation} \label{Chap:Symo.PrSy:3} Greedy generation of the reduced basis is an iterative procedure which, in each iteration, adds the two best possible basis vectors to the symplectic basis to enhance overall accuracy. In contrast to the cotangent lift and the complex SVD methods, the greedy approach does not require the symplectic basis to have a specific structure. This typically results in a more compact basis and/or more accurate reduced systems. For parametric problems, the greedy approach only requires one numerical solution to be computed per iteration hence saving substantial computational cost in the offline stage.

{\edit The orthonormalization step is an essential step in most greedy approaches for basis generation in the context of model reduction \cite{Anonymous:2016wl,Quarteroni:2016wi}. However common orthonormalization processes, e.g. the QR method, destroy the symplectic structure of the original system \cite{BunseGerstner:1986dg}. Here we use a variation of the QR method known as the SR \cite{Salam2014} method which is based on the symplectic Gram-Schmidt method and yields a symplectic basis. 
}

{\edit
As discussed in Section \ref{chap:Hasy:1}, any finite dimensional symplectic linear vector space has a symplectic basis that satisfies conditions (\ref{eq:Hasy:4}). Further, Theorem \ref{theorem:Hasy:1.6} provides an iterative process for constructing an orthosymplectic basis \cite{Matsuo:2014wl,Salam2014}. To briefly describe the SR method, suppose that an orthosymplectic basis
\begin{equation} \label{eq:SyMo:14.1}
	A_{2k}=\{ e_1 , \dots , e_k \} \cup \{ \mathbb J_{2n}^T e_1 , \dots , \mathbb J_{2n}^T e_k \},
\end{equation}
and a vector $z\not \in \text{span}(A_{2k})$ is provided. We aim to symplectically orthogonalize ($\mathbb J_{2n}$-orthogonalize) $z$ with respect to $A_{2k}$ and seek $\alpha_1,\dots,\alpha_k,\beta_1,\dots,\beta_k \in \mathbb R$ such that
\begin{equation} \label{eq:SyMo:14.2}
	\Omega\left(z + \sum_{i=1}^k \alpha_i e_i + \sum_{i=1}^k \beta_i \mathbb J_{2n}^Te_i , \sum_{i=1}^k \bar{\alpha}_i e_i + \sum_{i=1}^k \bar{\beta}_i \mathbb J_{2n}^Te_i \right) = 0,
\end{equation}
for all possible $\bar{\alpha}_1,\dots,\bar{\alpha}_k,\bar{\beta}_1,\dots,\bar{\beta}_k \in \mathbb R$. It is easily seen that the unique solution is 
\begin{equation} \label{eq:SyMo:14.3}
	\alpha_i = - \Omega(z,\mathbb J_{2n}^Te_i), \quad \beta_i = \Omega(z,e_i),
\end{equation}
for $i=1,\dots,k$. Now define the modified vectors as
\begin{equation} \label{eq:SyMo:14.4}
	\tilde z = z - \sum_{i=1}^k \Omega(z,\mathbb J_{2n}^Te_i) e_i + \sum_{i=1}^k \Omega(z,e_i) \mathbb J_{2n}^Te_i.
\end{equation}
If we introduce $e_{k+1} = \tilde z / \| \tilde z \|_2$, it is easily checked that $e_{k+1}$ is also orthogonal to $A_{2k}$ with respect to the classical inner product. Therefore span$\{e_1,\dots,e_{k+1}\}$ forms a Lagrangian subspace and according to Theorem \ref{theorem:Hasy:1.6} the basis $A_{2k+2}= A_{2k}\cup \{ e_{k+1} , \mathbb J_{2n}^T e_{k+1} \}$ forms an orthosymplectic basis.

{\blue Note that the $SR$ method is chosen due to its simplicity and it can be replaced with backward stable routines such as the isotropic Arnoldi or the isotropic Lanczos methods \cite{Mehrmann:2000dv}.}
}

The key element of the greedy algorithm is the availability of an error function which evaluates the error associated with the model reduction \cite{Anonymous:2016wl}. In the framework of symplectic model reduction, one possible candidate is the error in the Hamiltonian (\ref{eq:SyMo:5}). Correctly approximating symplectic systems relies on preservation of the Hamiltonian, hence the error in the Hamiltonian {\edit arises as a} a natural choice. Moreover, since the error in the Hamiltonian depends on the initial condition and the reduced symplectic basis, evaluation of the error does not require the time integration of the full system.

Suppose that a $2k$-dimensional {\edit orthosymplectic basis (\ref{eq:SyMo:14.1})} is generated at the $k$-th step of the greedy method and we seek to enrich it by two additional vectors. Using the error in the Hamiltonian (\ref{eq:SyMo:7}) we search the parameter space to identify the value that maximizes the error in the Hamiltonian
\begin{equation} \label{eq:SyMo:14.5}
	\omega_{k+1} := \underset{\omega\in \Gamma}{\text{argmax }}\Delta H(\omega).
\end{equation}
The goal is to approximate the Hamiltonian function as well as possible.

We then propagate (\ref{eq:Hasy:7}) in time to produce trajectory snapshots 
\begin{equation}
	S=\{ \mathbf z(t_i,\omega_{k+1}) | i = 1,\dots,M \}.
\end{equation} 
The next basis vector is the snapshot that maximises the projection error (\ref{eq:SyMo:8})
{\edit 
\begin{equation} \label{eq:SyMo:14.6}
	z := \underset{s\in S}{\text{argmax }} \| s - A_{2k}{A_{2k}}^+s \|.
\end{equation}
}
Finally, we update the basis as
{\edit
\begin{equation} \label{eq:SyMo:14.7}
	e_{k+1} = \tilde z, \quad A_{2k+1} = A_{2k}\cup \{ e_{k+1} , \mathbb J_{2n}^Te_{k+1} \},
\end{equation}
}
where $\tilde z$ is the vector obtained {\edit after} applying the symplectic Gram-Schmidt process to $z$.

Since the maximization over the entire parameter space $\Gamma$ is impossible, we discretize the parameter set into a grid with $N$ points: $\Gamma_N = \{ \omega_1,\dots,\omega_N\}$. However, since the selection of parameters only require the evaluation of the error in the Hamiltonian and not time integration of the original system, then $\Gamma_N$ can be chosen {\edit to be} very rich.

We summarize the greedy algorithm for the generation of a symplectic basis in Algorithm \ref{alg:SyMo:3}.

\begin{algorithm} 
\caption{The greedy algorithm for generation of a symplectic basis} \label{alg:SyMo:3}
{\bf Input:} Tolerated loss in the Hamiltonian $\delta$, parameter set $\Gamma_N = \{\omega_1,\dots,\omega_N\}$, initial condition $\mathbf z_0(\omega)$
\begin{enumerate}
\item $\omega^* \leftarrow \omega_1$
\item $e_1 \leftarrow \mathbf z_0(\omega^*)$
\item $A \leftarrow [e_1,\mathbb J^T_{2n}e_1]$
\item $k \leftarrow 1$
\item \textbf{while} $\Delta H(\omega) > \delta$ for all $\omega \in \Gamma_N$
\item \hspace{0.5cm} $\omega^* \leftarrow$ $\underset{\omega\in \Gamma_N}{\text{argmax }}\Delta H(\omega)$
\item \hspace{0.5cm} Compute trajectory snapshots $S=\{ \mathbf z(t_i,\omega^*) | i = 1,\dots,M \}$
\item \hspace{0.5cm} $\mathbf z^* \leftarrow$ $\underset{s\in S}{\text{argmax }} \| s - AA^+s \|$
\item \hspace{0.5cm} Apply symplectic Gram-Schmidt on $\mathbf z^*$
\item \hspace{0.5cm} $e_{k+1} \leftarrow \mathbf z^*/ \| \mathbf  z^*\|$
\item \hspace{0.5cm} $A \leftarrow [e_1,\dots ,e_{k+1} , \mathbb J^T_{2n}e_1,\dots,\mathbb J^T_{2n}e_{k+1}]$
\item \hspace{0.5cm} $k \leftarrow k+1$
\item \textbf{end while}
\end{enumerate}
\vspace{0.5cm}
{\bf Output:} Symplectic basis $A$.
\end{algorithm}

%The convergence and convergence rate of the greedy algorithm with respect to Kolmogorov $n$-width is investigated in Section \ref{chap:SyMo.PrSy:3}. We finish this section by stating the advantage of using the greedy algorithm over the POD-based methods:
%\begin{itemize}
%\item Lack of a symplectic linear algebra toolbox for solving the minimization in (\ref{eq:SyMo:8}) requires a possibly very expensive nonlinear optimizations to recover near optimal symplectic bases. Alternative approaches for finding solutions to (\ref{eq:SyMo:8}) requires assumptions on the structure of the symplectic basis. The greedy algorithm on the other hand, is a cheap alternative to the nonlinear optimization and it allows flexibility in terms of the structure.
%\item The greedy method reduces the total cost of the offline stage by solving the original system (\ref{eq:Hasy:7}) only $k$ times to obtain a reduced system of size $2k$, unlike POD-based methods that require time integration of (\ref{eq:Hasy:7}) for the entire parameter space.
%\item If enrichment of the symplectic basis is required, the greedy algorithm can add new basis vectors. The POD-based methods, are bound to re-compute the basis for the refined parameter set.
%\item POD-based methods, e.g., the cotangent lift and the complex SVD, require performing SVD decompositions on possibly very large matrices. The greedy algorithm avoids this.
%\end{itemize}

\subsubsection{Convergence of the Greedy Method} \label{chap:SyMo.PrSy:3}

To show convergence of the greedy method we {\edit consider} a slightly different version based on the projection error. The error in the Hamiltonian is then introduced as a cheap surrogate to the projection error to accelerate the parameter selection.

Suppose that we are given a compact subset $S$ of $\mathbb R^{2n}$. Our intention is to find a set of vectors $A=\{e_1,\dots,e_k,f_1,\dots,f_k\}$ such that $A$ forms {\edit an orthosymplectic} basis and any $s\in S$ is well approximated by elements of the subspace span$(A)$. The modified greedy method for generating basis vectors $e_i$ and $f_i$ is as follows. In the initial step we pick $e_1$ such that $\edit \|e_1\|_2 = \max_{s\in S} \|s\|_2$. Then define $f_1 = \mathbb{J}_{2n}^T e_1$. It is easy to check that the span of $A_2 = \{e_1,f_1\}$ is {\edit orthosymplectic}, so $A_2$ is the first subspace that approximates elements of $S$. In the $k$-th step of the greedy method, suppose we have a basis $A_{2k} = \{ e_1,\dots, e_k , f_1,\dots ,f_k \}$. We define $P_{2k}$ to be a symplectic projection operator that projects elements of $S$ onto span$(A_{2k})$ and define
\begin{equation} \label{eq:new1}
	\sigma_{2k}(s) := \|s-P_{2k}(s)\|_2,
\end{equation}
as the projection error. Moreover we denote by $\sigma_{2k}$ the maximum approximation error of $S$ using elements in span$(A_{2k})$ as
\begin{equation} \label{eq:new2}
	\sigma_{2k} := \max_{s\in S} \sigma_{2k}(s).
\end{equation}
The next set of basis vectors in the greedy selection are
\begin{equation} \label{eq:new3}
	e_{k+1} := \underset{s\in S}{\text{argmax }}\sigma_{2k}(s), \quad f_{k+1} := \mathbb{J}_{2n}^T e_{k+1}.
\end{equation}
We emphasisze that the sequence of basis vectors generated by the greedy is generally not unique \cite{Quarteroni:2016wi,Anonymous:2016wl}.

To estimate the quality of the reduced subspace, it is natural to compare it with the best possible $2k$-dimensional subspace in the sense of the minimum projection (not necessary symplectic) error. For this we introduce the Kolmogorov $n$-width \cite{Kolmogoroff:1936fj,Pinkus:1985vy}.

\begin{definition}
Let $S$ be a subset of $\mathbb R^{m}$ and $Y_n$, $n\leq m$, be a general $n$-dimensional subspace of $\mathbb R^{m}$. The angle between $S$ and $Y_n$ is given by
\begin{equation} \label{eq:new4}
	E(S,Y_n) := \sup_{s\in S} \inf_{y\in Y_n} \|s-y\|_2.
\end{equation}
The Kolmogorov $n$-width of $S$ in $\mathbb R^m$ is given by
\begin{equation} \label{eq:new5}
	d_{n}(S,\mathbb{R}^m) := \inf_{Y_n} E(S,Y_n) = \inf_{Y_n} \sup_{s\in S} \inf_{y\in Y_n} \|s-y\|_2
\end{equation}
\end{definition}

For a given subspace $Y_n$, the angle between $S$ and $Y_n$ measures the worst possible projection error of elements in $S$ onto $Y_n$. Hence the Kolmogorov $n$-width quantifies how well $S$ can be approximated by {\edit an} $n$-dimensional subspace.

We seek to show that the decay of $\sigma_{2k}$, obtained by the greedy algorithm, has the same rate as of $d_{2k}(S)$, i.e., the greedy method provides the best possible accuracy attained by a $2k$-dimensional subspace.

We start by $\edit \mathbb J_{2n}$-orthogonalizing the vectors provided by the greedy algorithm as
\begin{equation} \label{eq:new6}
\begin{aligned}
	& \xi_1 = e_i, & \bar{\xi}_1 = \mathbb{J}_{2n}^T \xi_1, &\\
	& \xi_i = e_i - P_{2(i-1)} (e_i), & \bar{\xi}_i = \mathbb{J}_{2n}^T, \xi_i &\quad i = 2,3,\dots
\end{aligned}
\end{equation}
The projection of a vector $s\in S$ onto span$(A_{2k})$ can be written using the symplectic basis as
\begin{equation} \label{eq:new7}
	P_{2k}(s) = \sum_{i=1}^k \left( \alpha_i(s) \xi_i + \bar{\alpha}_i(s) \bar{\xi}_i \right),
\end{equation}
where $\alpha_i(s)$ and $\bar{\alpha}_i(s)$ for $i=1,\dots,k$ are the expansion coefficients
\begin{equation} \label{eq:new8}
	\alpha_i(s) = - \frac{\Omega(\bar{\xi}_i,s)}{\Omega(\xi_i,\bar{\xi}_i)}, \quad \bar{\alpha}_i(s) = \frac{\Omega(\xi_i,s)}{\Omega(\xi_i,\bar{\xi}_i)},
\end{equation}
for any $s\in S$. Since $\bar{\xi}_i$ is {\edit $\mathbb J_{2n}$-orthogonal} to the span$(A_{2(k-1)})$ we have
\begin{equation} \label{eq:new9}
\begin{aligned}
	|\alpha_i(s)| = \frac{|\Omega(\bar{\xi}_i,s)|}{|\Omega(\xi_i,\bar{\xi}_i)|} = \frac{|\Omega( \bar{\xi}_i, s - P_{2(k-1)}(s))|}{|\Omega(\xi_i,\bar{\xi}_i)|}  &\leq \frac{\|\bar{\xi}_i\|_2 \| s - P_{2(k-1)}(s) \|_2}{ \|\xi_i\|_2 \|\bar{\xi}_i\|_2 } \\
	&= \frac{\| s - P_{2(k-1)}(s) \|_2}{\| e_i - P_{2(k-1)}(e_i) \|_2} \leq 1.
\end{aligned}
\end{equation}
Here, we use the fact that $\edit |\Omega(\xi_i,\bar{\xi}_i)| = \| \xi_i \|^2_2 = \|\bar{\xi}_i\|^2_2$ with the last inequality following from the greedy algorithm which maximizes $e_i$. Similarly we deduce that $|\bar{\alpha}_i(s)|\leq 1$.

We write
\begin{equation} \label{eq:new10}
\begin{aligned}
	\xi_j = \sum_{i=1}^j \left( \mu_i^j e_i + \gamma_i^j f_i \right), \quad \bar{\xi}_j = \sum_{i=1}^j \left( \lambda_i^j e_i + \eta_i^j f_i,  \right), \quad j=1,2,\dots
\end{aligned}
\end{equation}
with
\begin{equation} \label{eq:new11}
\begin{aligned}
	&\mu^j_j = 1, \quad \gamma^j_j = 0, \\
	&\mu_i^j = \sum_{l=i}^{j-1}\left( - \alpha_l(f_j) \mu_i^l  + \bar{\alpha}_l(f_j) \gamma_i^l \right), \quad\gamma_i^j = \sum_{l=i}^{j-1}\left( - \alpha_l(f_j) \gamma_i^l  + \bar{\alpha}_l(f_j) \mu_i^l \right), \\
	&\lambda^j_i = - \gamma ^j_i, \quad \eta^j_i = \mu^j_i,
\end{aligned}
\end{equation}
for $j=2,3,\dots$. By induction and using the bound in (\ref{eq:new9}) we deduce that
\begin{equation} \label{eq:new12}
	\mu^j_i,\gamma^j_i,\lambda^j_i,\eta^j_i \leq 3^{j-i}, \quad \text{for } j\geq i.
\end{equation}
Now let $2k$ be the dimension of the desired reduced space. Looking at the definition of Kolmogorov $n$-width we observe that for any $\theta > 1$ we can find a subspace $Y_{2k}$ such that $E(S,Y_{2k}) \leq \theta d_{2k}(S,\mathbb R^n)$. Hence we can find vectors $v_1,\dots,v_k,u_1,\dots,u_k\in Y_{2k}$ such that
\begin{equation} \label{eq:new13}
\begin{aligned}
	& \|e_i - v_i\|_2 \leq \theta d_{2k}(S,\mathbb R^n), \\
	& \|f_i - u_i\|_2 \leq \theta d_{2k}(S,\mathbb R^n).
\end{aligned}
\end{equation}
Now we construct a set of $2(k+1)$ new vectors
\begin{equation} \label{eq:new14}
\begin{aligned}
	& \zeta_j = \sum_{i=1}^{k+1} \mu_i^j v_i + \gamma^j_i u_i,\quad \bar{\zeta}_j = \sum_{i=1}^{k+1} \lambda_i^j v_i + \eta^j_i u_i.
\end{aligned}
\end{equation}
for $j = 1,\dots,k+1$. Note that since $u_i$ and $v_i$ belong to $Y_{2k}$ so does their linear combination including all $\zeta_j$ and $\bar{\zeta}_j$. We can use the inequality (\ref{eq:new12}) to write
\begin{equation}
	\| \xi_i - \zeta_i \|_2 \leq 3^i \theta d_{2k}(S,\mathbb R^n),\quad \| \bar{\xi}_i - \bar{\zeta}_i \|_2 \leq 3^i \theta d_{2k}(S,\mathbb R^n).
\end{equation}
Moreover since $Y_{2k}$ is of dimension $2k$ we find $\kappa_i$, $i=1,\dots,2(k+1)$ such that
\begin{equation} \label{eq:new15}
	\sum_{i=1}^{2(k+1)} \kappa_i^2 = 1, \quad\sum_{i=1}^{k+1} \kappa_i \zeta_i + \sum_{i=1}^{k+1} \kappa_{i+k+1} \bar{\zeta}_i = 0.
\end{equation}
We have
\begin{equation} \label{eq:new17}
\begin{aligned}
	\left\| \sum_{i=1}^{k+1} \kappa_i \xi_i + \sum_{i=1}^{k+1} \kappa_{i+k+1} \bar{\xi}_i \right\|_2 &= \left\| \sum_{i=1}^{k+1} \kappa_i (\xi_i - \zeta_i) + \sum_{i=1}^{k+1} \kappa_{i+k+1} (\bar{\xi}_i-\bar{\zeta}_i) \right\|_2 \\
	&\leq 2\cdot 3^{k+1} \sqrt{2(k+1)} \theta d_{2k}(S,\mathbb R^n).
\end{aligned}
\end{equation}
We know there exists $1 \leq j\leq 2k+2$ such that $\kappa_j > 1/\sqrt{2(k+1)}$. Without loss of generality let us assume that $j\leq k+1$. This yields
\begin{equation} \label{eq:new18}
	\left\| \xi_j +  \kappa_j^{-1} \sum_{i=1,i\neq j}^{k+1} \kappa_i \xi_i + \kappa_j^{-1}\sum_{i=1}^{k+1} \kappa_{i+k+1} \bar{\xi}_i \right\|_2 \leq 4\cdot 3^{k+1} (k+1) \theta d_{2k}(S,\mathbb R^n).
\end{equation}
Define $c = \kappa_j^{-1} \sum_{i=1,i\neq j}^{k+1} \kappa_i \xi_i + \kappa_j^{-1}\sum_{i=1}^{k+1} \kappa_{i+k+1} \bar{\xi}_i$. Using that $\mathbb{J}_{2n}^T c$ is {\edit $\mathbb J_{2n}$-orthogonal} to $\xi_j$ we recover
\begin{equation} \label{eq:new19}
\begin{aligned}
	\| \xi_j \|_2 &\leq \| \xi_j \|_2 + \| c \|_2 = \Omega(\xi_j,\mathbb{J}_{2n}^T \xi_j) + \Omega(c,\mathbb{J}_{2n}^T c) \\
	       &= \Omega(\xi_j,\mathbb{J}_{2n}^T \xi_j) + \Omega(c,\mathbb{J}_{2n}^T c) + \Omega(\xi_j,\mathbb{J}_{2n}^T c) + \Omega(c,\mathbb{J}_{2n}^T \xi_j) \\
       	&= \Omega(\xi_j + c, \mathbb{J}^T_{2n} (\xi_j + c)) = \| \xi_j + c \|_2	
\end{aligned}
\end{equation}
Combining this with (\ref{eq:new18}) yields
\begin{equation} \label{eq:new20}
	\| \xi_j \|_2 \leq 4\cdot 3^{k+1} (k+1) \theta d_{2k}(S,\mathbb R^n).
\end{equation}
Finally using the definition of $\xi_j$ for all $s\in S$ we have
\begin{equation} \label{eq:new21}
	\| s - P_{2(j-1)}(s) \|_2 \leq \| f_j - P_{2(j-1)}(f_j) \|_2 = \|\xi_j \|_2 \leq 4\cdot 3^{k+1} (k+1) \theta d_{2k}(S,\mathbb R^n)
\end{equation}
Hence, for any given $\lambda > 1$
\begin{equation} \label{eq:new22}
	\| s - P_{2k}(s) \|_2 \leq \| s - P_{2(j-1)}(s) \|_2 \leq 4\cdot 3^{k+1} (k+1) \theta d_{2k}(S,\mathbb R^n).
\end{equation}
This establishes the following theorem.
\begin{theorem} \label{theorem:SyMo:2}
	Let $S$ be a compact subset of $\mathbb{R}^{2n}$ with exponentially small Kolmogorov $n$-width $\edit d_{k}\leq c\exp(-\alpha k)$ with $\alpha > \log3$. Then there exists $\beta>0$ such that the symplectic subspaces $A_{2k}$ generated by the greedy algorithm provide exponential approximation properties such that
\begin{equation} \label{eq:new23}
	\| s - P_{2k}(s) \|_2 \leq C \exp(-\beta k)
\end{equation}
for all $s\in S$ and some $C>0$.
\end{theorem}

%{\edit
%\begin{theorem} \label{theorem:SyMo:3}
%	The basis $A_{2k}$ generated by the greedy algorithm is orthosymplectic.
%\end{theorem}
%\begin{proof}
%	Direct result of theorem \ref{theorem:Hasy:1.6}.
%\end{proof}
%}

\subsection{Symplectic Discrete Empirical Interpolation Method (SDEIM)} Consider the Hamiltonian system (\ref{eq:Hasy:7}) and its reduced system (\ref{eq:SyMo:4}) equipped with a symplectic transformation $A$. One can split the Hamiltonian function $H = H_1 + H_2$ such that $\nabla H_1 = L\mathbf z$ and $\nabla H_2 = \mathbf g(\mathbf z)$, where $L$ is a constant matrix in $\edit \mathbb R^{2n\times 2n}$ and $\mathbf g$ is a nonlinear function. The reduced system takes the form
\begin{equation} \label{eq:new24}
	\frac{d}{dt} \mathbf y = \underbrace{A^+ \mathbb J_{2n} L A}_{\tilde L} \mathbf y + A^+ \mathbb J_{2n} \mathbf g(A\mathbf y)
\end{equation}
As discussed in Section \ref{chap:MoOr.DEIM:1}, the complexity of evaluating the nonlinear term still depends on $n$, the size of the original system. To overcome this computational bottleneck we use the DEIM approximation for evaluating the nonlinear function $\mathbf g$ as
\begin{equation} \label{eq:new25}
	\frac{d}{dt} \mathbf y = \tilde L \mathbf y + \underbrace{ A^+ \mathbb J_{2n} V (P^TV)^{-1} P^T \mathbf g(A\mathbf y) }_{\tilde N(\mathbf y)}
\end{equation}
For a general choice of $V$ the system (\ref{eq:new25}) is not guaranteed to be a Hamiltonian system, impacting long time accuracy and stability. However, we can guarantee that (\ref{eq:new25}) is a Hamiltonian system by choosing $V=(A^+)^T$. To {\edit see this}, we note that the system (\ref{eq:new25}) is a Hamiltonian system if and only if $\tilde N(\mathbf y) = \mathbb J_{2k} \nabla_{\mathbf y} \mathbf g(\mathbf y)$. Also we have 
\begin{equation} \label{eq:new26}
	\mathbf g(A\mathbf y) = \nabla_{\mathbf z} H_2(\mathbf z) = (A^+)^T \nabla_{\mathbf y} H_2(A \mathbf y),
\end{equation}
where the chain rule is used for the second equality. Substituting this into $\tilde N$ we obtain
\begin{equation} \label{eq:new27}
	\tilde N(\mathbf y)= A^+ \mathbb J_{2n} V (P^TV)^{-1} P^T  (A^+)^T \nabla_{\mathbf y} H_2(A \mathbf y).
\end{equation}
Taking $V = (A^+)^T$ yields
\begin{equation} \label{eq:new28}
	\tilde N(\mathbf y) = A^+ \mathbb J_{2n}(A^+)^T \nabla_{\mathbf y} H_2(A \mathbf y) = \mathbb J_{2k} \nabla_{\mathbf y} H_2(A \mathbf y),
\end{equation}
since $(A^+)^T$ is a symplectic matrix. Hence, $V = (A^+)^T$ is a sufficient condition for (\ref{eq:new25}) to {\edit be Hamiltonian}.

Regarding the construction of the projection space, suppose that we have already constructed a symplectic basis $A=\{ e_1,\dots , e_k,f_1,\dots f_k \}$ using the greedy algorithm. Note that $(A^+)^T$ is a symplectic basis and $(A^+)^+=A$. Thus, we can move between these two symplectic bases by simply using the transpose operator and the symplectic inverse operator. Let $S_{\mathbf g} = \{ \mathbf g (\mathbf x(t_i,\omega_j)) \}$ with $i = 1,\dots,M$ and $ j = 1 ,\dots,N$ be the nonlinear snapshots that were gathered in the greedy algorithm. We then form $(A^+)^T = \{ e'_1,\dots, e'_k,f'_1,\dots,f'_k\}$ and use a greedy approach to add new basis vectors to $(A^+)^T$. At the $i$-th iteration of the symplectic DEIM, we use $(A^+)^T$ to approximate elements in $S_{\mathbf g}$ and choose the vector that maximizes the error as the next basis vector 
\begin{equation}
	s^* := \underset{s \in S_{\mathbf g}}{\text{argmax }}\| s - (A^+)^T A^+ s \|_2.	
\end{equation}
After applying the symplectic Gram-Schmidt on $s^*$, we update $(A^+)^T$ as
\begin{equation}
\begin{aligned}
	e'_{k+i+1} = \frac{s^*}{\| s^* \|_2},\quad f'_{k+i+1} = \mathbb J_{2n}^T e'_{k+i+1}.
\end{aligned}
\end{equation}
Finally when $(A^+)^T$ approximates elements $S_{\mathbf g}$ with the desired accuracy, we transpose and symplectically invert $(A^+)^T$ to obtain $A$. We summarize the symplectic DEIM algorithm in Algorithm \ref{alg:SyMo:4}.

\begin{algorithm} 
\caption{Symplectic Discrete Empirical Interpolation Method} \label{alg:SyMo:4}
{\bf Input:} Symplectic basis $A=\{ e_1,\dots,e_k,f_1,\dots,f_k \}$, nonlinear snapshots $S_{\mathbf g} = \{ \mathbf g(\mathbf x(t_i,\omega_j)) \}$ and tolerance $\delta$
\begin{enumerate}
\item Compute $(A^+)^T = \{ e'_1,\dots,e'_k,f'_1,\dots,f'_k \}$
\item $i \leftarrow 1$
\item \textbf{while} max$\| s - (A^+)^T A^+s \| > \delta$ for all $s\in S_{\mathbf g}$
\item \hspace{0.5cm} $s^* \leftarrow \underset{s \in S_{\mathbf g}}{\text{argmax }}\| s - (A^+)^T A^+ s \|$
\item \hspace{0.5cm} Apply symplectic Gram-Schmidt on $s^*$
\item \hspace{0.5cm} $e'_{k+i} = s^* / \| s^* \|$
\item \hspace{0.5cm} $f'_{k+i} = \mathbb J_{2n} e'_{k+i}$
\item \hspace{0.5cm} $(A^+)^T \leftarrow [e'_1,\dots,e'_{k+i},f'_1,\dots,f'_{k+i}]$
\item \hspace{0.5cm} $i\leftarrow i+1$
\item \textbf{end while}
\item take transpose and symplectic inverse of $(A^+)^T$
\end{enumerate}
\vspace{0.5cm}
{\bf Output:} Symplectic basis $A$ that guarantees a Hamiltonian reduced system.
\end{algorithm}

When using an implicit time integration scheme we face inefficiencies when evaluating the Jacobian of nonlinear terms, as discussed in Section \ref{chap:MoOr.DEIM:1}. We recall that the key to fast approximation of the Jacobian is that the interpolating index matrix $P$, obtained in the DEIM approximation, commutes with the nonlinear function. Nonlinear terms in Hamiltonian systems often take the from
\begin{equation} \label{eq:new29}
	\mathbf g (\mathbf z) = \mathbf g (\mathbf q,\mathbf p) = 
	\begin{pmatrix}
		g_1(q_1,p_1) \\
		g_2(q_2,p_2) \\
		\vdots \\
		g_{2n}(q_{n},p_{n})
	\end{pmatrix}.
\end{equation}
Thus, the interpolating index matrix, obtained by Algorithm \ref{alg:MoOr:1} does not necessarily commute with the function $\mathbf g$. To overcome this, when index $\mathfrak p_i$ with $\mathfrak p_i\leq n$ or $\mathfrak p_i>n$ is chosen in Algorithm \ref{alg:MoOr:1} we also include $\mathfrak p_i + n$ or $\mathfrak p_i-n$, respectively. {\edit Simple calculations verifies that $\mathbf g$ and $P$ then commute.}

{\blue In case $\mathbf g$ is not of the form (\ref{eq:new29}) one can use MDEIM \cite{carlberg2012efficient,negri2015efficient} to accelerate the assembling of the Jacobian matrix.}

\section{Numerical Results} \label{chap:NuRe:1}
In this section, we illustrate the performance of the greedy generation of a symplectic basis. The parametric linear wave equation is considered to compare SVD based methods with the greedy method. {\blue The symplectic model reduction of nonlinear systems} is then illustrated by considering the parametric nonlinear Schr\"odinger equation. {\edit Finally we discuss the numerical convergence of the greedy method introduced in Algorithm \ref{alg:SyMo:3}.}

\subsection{Parametric Linear Wave equation} \label{chap:NuRe:1.1} Consider the {\edit parametric} linear wave equation
\begin{equation} \label{eq:NuRe:1}
\left\{
\begin{aligned}
& u_{tt}(x,t,\omega) = \kappa(\omega) u_{xx}(x,t,\omega), \\
& u(x,0) = u^0(x),
\end{aligned}
\right.
\end{equation}
where $x$ belongs to a one-dimensional torus of length $L$, $\omega = (\omega_1,\dots,\omega_4)$ and
\begin{equation} \label{eq:NuRe:2}
	\kappa(\omega) = c^2\left( \sum_{l=1}^4 \frac{1}{l^2} \omega_l \right).
\end{equation}
{\edit Here $\omega_l \in [0,1]$ for $l=1,\dots,4$} and $c\in \mathbb{R}$ is a constant number. By rewriting (\ref{eq:NuRe:1}) in canonical form, using the change of variable $q = u$ and $\partial q/ \partial t= p$, we obtain the symplectic form
\begin{equation} \label{eq:NuRe:3}
\left\{
\begin{aligned}
& q_t(x,t,\omega) = p(x,t,\omega), \\
& p_t(x,t,\omega) = \kappa(\omega) q_{xx}(x,t,\omega),
\end{aligned}
\right.
\end{equation}
with the associated Hamiltonian
\begin{equation} \label{eq:NuRe:4}
	H(q,p,\omega) = \frac 1 2 \int_0^L p^2 + \kappa(\omega) q_x^2 \ dx.
\end{equation}
We discretize the torus into $N$ equidistant points and define $\Delta x = L/N$, $x_i = i\Delta x$, $q_i=q(t,x_i,\omega)$ and $p_i=p(t,x_i,\omega)$ for $i = 1, \dots, N$. Furthermore, we discretize (\ref{eq:NuRe:3}) using a standard central finite differences scheme to obtain
\begin{equation} \label{eq:NuRe:5}
	\frac{d}{dt} \mathbf z = \mathbb{J}_{2N} L\mathbf z,
\end{equation}
where $\mathbf z=(q,\dots,q_N,p_q,\dots,p_n)^T$ and
\begin{equation} \label{eq:NuRe:6}
L = 
\begin{pmatrix}
	I_n & 0_N \\
	0_N & \kappa(\omega)D_{xx}
\end{pmatrix},\quad 
\end{equation}
with $D_{xx}$ the central finite differences matrix operator. The discrete Hamiltonian can finally be written as
\begin{equation} \label{eq:NuRe:7}
	H_{\Delta x}(\mathbf z) = \frac{\Delta x}2 \sum_{i=1}^{N} \left( p_i^2 + \kappa(\omega) \frac{(q_{i+1} - q_i)^2}{2\Delta x ^ 2} + \kappa(\omega) \frac{(q_{i} - q_{i-1})^2}{2\Delta x ^ 2} \right).
\end{equation}
The initial condition is given by
\begin{equation} \label{eq:NuRe:8}
	q_i(0) = h( 10\times|x_i - \frac{1}{2}| ), \quad p_i = 0, \quad i=1,\dots,N
\end{equation}
where $h(s)$ is the cubic spline function
\begin{equation} \label{eq:NuRe:9}
h(s) = 
\left\{
\begin{aligned}
& 1 - \frac{3}{2}s^2 + \frac{3}{4}s^3, \quad & 0\leq s \leq 1, \\
& \frac{1}{4}(2-s)^3, & 1< s \leq 2, \\
& 0, & s > 2.
\end{aligned}
\right.
\end{equation}
This will result in waves propagating in both directions on the torus.

For numerical time integration we {\edit use} the St\"ormer-Verlet (\ref{eq:Hasy:13}) scheme, {\edit which is explicit since} the Hamiltonian is separable for the linear wave-equation. The full model uses the following parameter set \vspace{0.5cm}
\begin{center}
\begin{tabular}{|l|l|}
\hline
Domain length & $L = 1$ \\
No. grid points & $N = 500$ \\
Space discretization size & $\Delta x = 0.002$ \\
Time discretization size & $\Delta t = 0.01$ \\
Wave speed & $c^2 = 0.1$ \\
\hline
\end{tabular}
\end{center}
\vspace{0.5cm}
We compare the reduced system obtained by the greedy algorithm with the methods based on SVD. To generate snapshots, we discretize the parameter space $[0,1]^4$ into in total of $5^4$ equidistant grid points. For the SVD based methods and POD, snapshots are gathered in the snapshot matrices $S$, $S_{\text{combined}}$ and $S_{\text{complex}}$, respectively, and the SVD is performed to construct the reduced basis. The greedy method is applied following Algorithm \ref{alg:SyMo:3}; as input, the tolerance for the error in the Hamiltonian is set to $\delta = 5 \times 10^{-3}$. All reduced systems are taken to have an identical size ($k=80$ for POD and $k=40$ for the symplectic methods). We use the {\edit St\"ormer-Verlet} scheme for symplectic methods and a second order Runge-Kutta method for the POD. {\edit The choice of different time integration routines is due to the fact that the POD destroys the canonical form of the original equations and a symplectic integrator cannot be applied. One can alternatively use separate reduced subspaces for the potential and the momentum spaces, which however is not a standard model reduction approach and requires further analysis.} Finally we use transformation (\ref{eq:SyMo:1}) to transfer the solution of the reduced systems into the high-dimensional space for illustration purposes.

We reduced the cost by 50\% in the offline stage when using the greedy method as compared to SVD-based methods (cotangent lift and complex SVD method). This happens because the SVD-based methods require time integration of the full system for all discrete parameter points, while the greedy method picks a number of parameters from the parameter space.

%We summarized the duration of offline and online stages of different model reduction methods in the table below.
%
%\vspace{0.5cm}
%\begin{center}
%\begin{tabular}{c|cc}
% & offline (seconds) & online (seconds) \\
%\hline
%full system & - & 0.19 \\
%cotangent lift & 61.4 & 0.11 \\
%complex SVD & 61.4 & 0.11 \\
%greedy & 33.7 & 0.11 \\
%\end{tabular}
%\end{center}
%\vspace{0.5cm}
%
%In our implementations the offline stage of cotangent lift and the complex SVD are substantially longer compared to the greedy method. This is because SVD based methods need to explore the entire parameter space while the greedy method only hand pick a number of parameters from the parameter space. We see that the online stage of all reduced systems are faster than solving the full model. Of course efficiency of reduced systems depend on the implementation.

\begin{figure}

\begin{minipage}{.5\linewidth}
\centering
\subfloat[$t=0$]{\label{fig:NuRe:1a}\includegraphics[width=1\textwidth]{./figs/wave/solution/solution_t0}}
\end{minipage}%
\begin{minipage}{.5\linewidth}
\centering
\subfloat[$t=1$]{\label{fig:NuRe:1b}\includegraphics[width=\textwidth]{./figs/wave/solution/solution_t1}}
\end{minipage}\par\medskip
\centering
\subfloat[$t=2$]{\label{fig:NuRe:1c}\includegraphics[width=0.5\textwidth]{./figs/wave/solution/solution_t2}}
\caption{The solution $q$ at $t=0$, $t=1$ and $t=2$ of the linear wave equation for parameter value $c= 0.1019$ different from training parameters. Here, the solution of the full system together with the solution of the POD, cotangent lift, complex SVD and the greedy reduced system is shown.}
\label{fig:NuRe:1}
\end{figure}

\begin{figure}

\begin{minipage}{.5\linewidth}
\centering
\subfloat[]{\label{fig:NuRe:2c}\includegraphics[width=\textwidth]{./figs/wave/error}}
\end{minipage}%
\begin{minipage}{.5\linewidth}
\centering
\subfloat[]{\label{fig:NuRe:2b}\includegraphics[width=\textwidth]{./figs/wave/hamiltonian}}
\end{minipage}\par\medskip
\centering
%\subfloat[]{\label{fig:NuRe:2c}\includegraphics[width=0.5\textwidth]{./figs/wave/error}}

\caption{{\edit (a) The $L^2$-error between the solution of the full system and the reduced system for different model reduction methods for $t \in [0,30]$.} (b) Plot of the Hamiltonian function for $t \in [0,30]$. }
\label{fig:NuRe:2}
\end{figure}

Figure \ref{fig:NuRe:1a} shows the solution of the linear wave equation for parameter values {\edit $(\omega_1,\omega_2,\omega_3,\omega_4) = (0.8456,0.1320,0.9328,0.5809)$ or $\kappa(\omega) = 0.1019$}, chosen to be different from training parameters, at $t=0$, $t = 1$ and $t=2$. While we see instability and divergence from the exact solution for the POD reduced system, the symplectic methods provide a good approximation of the full model.

{\edit The decay of the singular values for the POD are shown in Figure \ref{fig:NuRe:5a}. The decay} of the singular values suggests that a low dimensional solution manifold indeed exists. However, since the linear subspace, constructed by the POD, is not symplectic, we observe blow up of the Hamiltonian function in Figure \ref{fig:NuRe:2b} and the instability of the solution in Figure \ref{fig:NuRe:1}. The symplectic methods (using a reduced basis of the same size as POD) preserve the Hamiltonian function as shown in Figure \ref{fig:NuRe:2b}.

Figure \ref{fig:NuRe:2c} shows the $L^2$-error between the solution of the full model and the reduced systems constructed by different methods. We note that the error for the POD reduced system rapidly increases, confirming that {\edit the} projection based reduced system does not yield a stable solution. Furthermore, the symplectic methods provide a better approximation since the geometric structure of the original system is preserved. Although the greedy method is almost {\edit twice faster than the SVD-based methods} in the offline stage, its accuracy is comparable. {\edit The cotangent lift method provides a more accurate solution, on the other hand the cotangent lift basis (\ref{eq:SyMo:9}) takes a less general form and usually computationally more demanding than the greedy method.}

{\edit For complex systems were the solution of the full system is expensive and for high dimensional parameter domains, POD-based methods become impractical \cite{Anonymous:2016wl,Quarteroni:2016wi}. However, the greedy method requires substantially fewer (proportional to the size of the reduced basis) evaluation of the time integration of the original system.}

%\begin{figure*}[t]
%\begin{center}
%\begin{tabular}{cc}
%	\includegraphics[width=0.4\textwidth]{./figs/wave/solution/solution_t0} &
%	\includegraphics[width=0.4\textwidth]{./figs/wave/solution/solution_t1} \\
%	$t=0$ & $t = 1$
%\end{tabular}
%\end{center}
%\end{figure*}

%\begin{figure}[t]
%\centering
%\begin{tabular}{cc}
%\includegraphics[width=0.4\textwidth]{./figs/wave/solution/solution_t2} \\
%	$t=2$
%\end{tabular}
%\caption{The solution $q$ at $t=0$, $t=1$ and $t=2$ of the linear wave equation for parameter value $c= 0.1019$ different from training parameters. Here the solution of full system together with the solution of the POD, cotangent lift, complex SVD and the greedy reduced system is presented.}\label{fig:NuRe:3}
%\end{figure}

%\begin{figure*}[t]
%\begin{center}
%\begin{tabular}{cc}
%	\includegraphics[width=0.5\textwidth]{./figs/wave/singular} &
%	\includegraphics[width=0.5\textwidth]{./figs/wave/hamiltonian} \\
%	(a) & (b)
%\end{tabular}
%\end{center}
%\end{figure*}

%\begin{figure}[t] 
%\begin{center}
%\begin{tabular}{c}
%	\includegraphics[width=0.5\textwidth]{./figs/wave/error} \\
%	(c)
%\end{tabular}
%\end{center}
%\caption{(a) The decay of singular values for POD, cotangent lift and complex SVD methods. (b) The $L^2$ error between the solution of th full system and the reduced system for different model reduction methods for $t \in [0,30]$. (c) Plot of the Hamiltonian function for $t \in [0,30]$.}\label{fig:NuRe:4}
%\end{figure}

%\begin{figure*}[t]
%  \centering
%  \includegraphics[width=0.5\textwidth]{./figs/wave/hamiltonian}
%  \label{fig:testfig}
%\end{figure*}

\subsection{Nonlinear Schr\"odinger equation} \label{chap:NuRe:1.2} Let us consider the one-dimensional parametric Schr\"odinger equation
\begin{equation} \label{eq:NuRe:10}
\left\{
\begin{aligned}
	& i u_t(t,x,\epsilon) = - u_{xx}(t,x,\epsilon) - \epsilon |u(t,x,\epsilon)|^2 u(t,x,\epsilon),\\
	& u(0,x) = u_0(x),
\end{aligned}
\right.
\end{equation}
where $u$ is a complex valued wave function, $i$ is the imaginary unit, $|\cdot|$ is the modulus operator and $\epsilon$ is a parameter that belongs to the interval $\Gamma = [0.9,1.1]$. We consider periodic boundary conditions, i.e., $x$ belongs to a one-dimensional torus of length $L$. We consider the initial condition
\begin{equation} \label{eq:NuRe:11}
	u_0(x) = \frac{\sqrt 2}{\cosh(x - x_0)} \exp(i\frac{c(x-x_0)}{2}),
\end{equation}
for a positive constant $c$. In quantum mechanics, the quantity $|u(t,x)|^2$ represents the probability of finding the system in state $x$ at time $t$. For the choice of $\epsilon = 1$, $|u(x,t)|$ becomes a solitary wave, and the initial condition will be transported in the positive $x$ direction with a constant speed. For other choices of $\epsilon$, the solution comprises an ensemble of solitary waves, moving in either direction \cite{Faou:2012vh}.

By introducing the real and imaginary variables $u = p + iq$, we can rewrite (\ref{eq:NuRe:10}) in canonical form as
\begin{equation} \label{eq:NuRe:12}
\left\{
\begin{aligned}
 q_t &= p_{xx} + \epsilon (q^2+p^2)p, \\
 p_t &= -q_{xx} - \epsilon (q^2 + p^2)q,
\end{aligned}
\right.
\end{equation}
with the Hamiltonian function
\begin{equation} \label{eq:NuRe:13}
	H(q,p) = \int_{0}^{L} (q_x^2 + p_x^2) + \frac \epsilon 2 (q^2 + p^2)^2\ dx.
\end{equation}
We discretize the torus into $N$ equidistant points and take $\Delta x = L/N$, $x_i = i\Delta x$, $q_i=q(t,x_i,\epsilon)$ and $p_i = p(t,x_i,\omega)$ for $i = 1 ,\dots,N$. A central finite differences scheme is used to discretize (\ref{eq:NuRe:12}) as
\begin{equation}  \label{eq:NuRe:14}
	\frac{d}{dt} \mathbf z = \mathbb J_{2N} L\mathbf z + \mathbb J_{2N} \mathbf g(\mathbf z).
\end{equation}
Here $\mathbf z = (q_1,\dots,q_N,p_1,\dots,p_n)^T$ and
\begin{equation}  \label{eq:NuRe:15}
	L = 
	\begin{pmatrix}
		D_{xx} & 0_N \\
		0_N & D_{xx}
	\end{pmatrix}.
\end{equation}
Here $\mathbf g$ is a vector valued nonlinear function defined as
\begin{equation}  \label{eq:NuRe:16}
	\mathbf g(\mathbf z) =
	\begin{pmatrix}
	(q_1^2 + p_1^2)q_1 \\
	\vdots \\
	(q_N^2 + p_N^2)q_N \\
	(q_1^2 + p_1^2)p_1 \\
	\vdots \\
	(q_N^2 + p_N^2)p_N
	\end{pmatrix}.
\end{equation}
We discretize the Hamiltonian to obtain
\begin{equation}  \label{eq:NuRe:17}
	H_{\Delta x}(\mathbf z) = {\Delta x}\sum_{i=1}^{N} \left( \frac{q_i q_{i-1} - q_i^2}{\Delta x ^2} + \frac{p_i p_{i-1} - p_i^2}{\Delta x ^2} + \frac \epsilon 4 (p_i^2 + q_i^2)^2  \right),
\end{equation}
and use a St\"ormer-Verlet (\ref{eq:Hasy:13}) scheme for time integration. Since the Hamiltonian function (\ref{eq:NuRe:17}) is non-separable, this scheme becomes implicit so in each time iteration, a system of nonlinear equations is solved using Newton's iteration. We summarize the physical and numerical parameters for the full model in the following table

\vspace{0.5cm}
\begin{center}
\begin{tabular}{|l|l|}
\hline
Domain length & $L = 2\pi /l$ \\
Domain scaling factor & $l = 0.11$ \\
wave speed & $c =1$\\
No. grid points & $N = 256$ \\
Space discretization size & $\Delta x = 0.2231$ \\
Time discretization size & $\Delta t = 0.01$ \\
\hline
\end{tabular}
\end{center}
\vspace{0.5cm}
Regarding computation of the nonlinear terms of reduced systems, {\edit we} compare the DEIM with the symplectic DEIM. For generation of the DEIM reduced basis we apply Algorithm \ref{alg:MoOr:1} to the set of nonlinear snapshots. Algorithm \ref{alg:SyMo:4} is used to construct a reduced basis appropriate for the symplectic DEIM. As input, we provide the symplectic basis generated by Algorithm \ref{alg:SyMo:3} with the set of nonlinear snapshots and a tolerance for the error $\delta = 10^{-4}$.

We compare the reduced system obtained using the greedy algorithm with the cotangent lift, the complex SVD, DEIM, the symplectic DEIM and also the POD. For the SVD-based methods, we discretize the parameter space $[0.9,1.1]$ into $M=500$ equidistant grid points {\edit across} the discrete parameter space $\Gamma_M = \{\epsilon_1,\dots,\epsilon_M \}$, and gather trajectory snapshots for each $\epsilon_i$ for $i = 1,\dots,M$ in the snapshots matrix $S$. All reduced systems are taken to have identical sizes ($k=90$ for the symplectic methods and $k=180$ for the POD method). Following Algorithm \ref{alg:SyMo:3} we construct the reduced system using the same discrete parameter space $\Gamma_M$. The tolerance for the error in the Hamiltonian is set to $\delta = 10^{-3}$. Moreover, for DEIM and symplectic DEIM, we construct bases of size $k'=80$. Note that the reduced system, generated in the symplectic DEIM, will be of size $k+k'=170$.

The cost of the offline stage is reduced to 20\% when using the greedy method for constructing a symplectic basis of size $k=90$, as compared to the SVD-based methods. The online stage, i.e., time integration for a new parameter in $\Gamma$, is generally more than 3 times faster than {\edit for} the original system. We point out that the efficiency of reduced systems are implementation and platform dependent {\edit and we expect further reduction as the size of the problem increases.}

%The duration of the online and offline stages for different model reduction techniques are summarized in table below. The offline time for the cotangent lift and the complex SVD consist of time integration of the full model for all the parameters in the discrete parameter space $\Omega_M$. This is while the greedy method requires substantially shorter offline stage to construct a reduced basis of size $k = 90$. For a reduced system of size $k=90$, all methods almost have the same online duration. The efficiency of reduced systems compared to the full system are heavily implementation dependent, nevertheless solving the reduced systems with our implementation is more than 3 times faster than solving the original system.

%\vspace{0.5cm}
%\begin{center}
%\begin{tabular}{c|cc}
% & offline (hours) & online (seconds) \\
%\hline
%full system & - & 187 \\
%cotangent lift & 25.8 & 59$\pm 1$ \\
%complex SVD & 25.8 & 59$\pm 1$ \\
%greedy & 4.6 & 59$\pm 1$ \\
%greedy + SDEIM & 4.7 & ???
%\end{tabular}
%\end{center}
%\vspace{0.5cm}

%\begin{figure*}[t]
%\begin{center}
%\begin{tabular}{cc}
%	\includegraphics[width=0.5\textwidth]{./figs/schrodinger/solution/solution_t0} &
%	\includegraphics[width=0.5\textwidth]{./figs/schrodinger/solution/solution_t10} \\
%	$t = 0$ & $t = 10$
%\end{tabular}
%\end{center}
%\end{figure*}
%
%\begin{figure}[t] 
%\begin{center}
%\begin{tabular}{c}
%	\includegraphics[width=0.5\textwidth]{./figs/schrodinger/solution/solution_t20}\\
%	$$t = 20$$
%\end{tabular}
%\end{center}
%\caption{The solution $|u(t,x)| = \sqrt{q^2 + p^2}$ at $t=0$, $t=10$ and $t=20$ of the Nonlinear Schr\"odinger equation for parameter value $\epsilon = 1.0932$. Here the solution of full system together with the solution of the POD, cotangent lift, complex SVD and the greedy reduced system is presented.}\label{fig:NuRe:1}
%\end{figure}
%
%\begin{figure*}[t]
%\begin{center}
%\begin{tabular}{cc}
%	\includegraphics[width=0.5\textwidth]{./figs/schrodinger/singular} &
%	\includegraphics[width=0.5\textwidth]{./figs/schrodinger/hamiltonian} \\
%	(a) & (b)
%\end{tabular}
%\end{center}
%\end{figure*}
%
%\begin{figure}[t] 
%\begin{center}
%\begin{tabular}{c}
%	\includegraphics[width=0.5\textwidth]{./figs/schrodinger/error} \\
%	(c)
%\end{tabular}
%\end{center}
%\caption{(a) The decay of singular values for POD, cotangent lift and complex SVD methods. (b) The $L^2$ error between the solution of th full system and the reduced system for different model reduction methods for $t \in [0,30]$. (c) Plot of the Hamiltonian function for $t \in [0,30]$.}\label{fig:NuRe:2}
%\end{figure}

\begin{figure}

\begin{minipage}{.5\linewidth}
\centering
\subfloat[$t=0$]{\label{fig:NuRe:3a}\includegraphics[width=1\textwidth]{./figs/schrodinger/solution/solution_t0}}
\end{minipage}%
\begin{minipage}{.5\linewidth}
\centering
\subfloat[$t=10$]{\label{fig:NuRe:3b}\includegraphics[width=\textwidth]{./figs/schrodinger/solution/solution_t10}}
\end{minipage}\par\medskip
\centering
\subfloat[$t=20$]{\label{fig:NuRe:3c}\includegraphics[width=0.5\textwidth]{./figs/schrodinger/solution/solution_t20}}
\caption{The solution $|u(t,x)| = \sqrt{q^2 + p^2}$ at $t=0$, $t=10$ and $t=20$ of the Nonlinear Schr\"odinger equation for parameter value $\epsilon = 1.0932$. Here the solution of {\edit the} full system, together with the solution of the POD, cotangent lift, complex SVD and the greedy reduced system, is shown.}
\label{fig:NuRe:3}
\end{figure}

\begin{figure}

\begin{minipage}{.5\linewidth}
\centering
\subfloat[]{\label{fig:NuRe:4c}\includegraphics[width=\textwidth]{./figs/schrodinger/error}}
\end{minipage}%
\begin{minipage}{.5\linewidth}
\centering
\subfloat[]{\label{fig:NuRe:4b}\includegraphics[width=\textwidth]{./figs/schrodinger/hamiltonian}}
\end{minipage}\par\medskip
\centering
%\subfloat[]{\label{fig:NuRe:4c}\includegraphics[width=0.5\textwidth]{./figs/schrodinger/error}}

\caption{{\edit (a) Plot of the Hamiltonian function for $t \in [0,30]$.} (b) The $L^2$ error between the solution of the full system and the reduced system for different model reduction methods for $t \in [0,30]$. }
\label{fig:NuRe:4}
\end{figure}

Figure \ref{fig:NuRe:3} shows the solution of the Schr\"odinger equation for parameter value $\epsilon = 1.0932$ at $t=0$, $t=10$ and $t=20$. We first compare the reduced system obtained by the greedy algorithm with the POD, the cotangent lift, and the complex SVD method. The size of the reduced systems are taken identical for all methods ($k=180$ for POD and $k=90$ for the rest). Although the decay of the singular values in Figure \ref{fig:NuRe:5b} suggests that the accuracy of the POD reduced system should be comparable to that of the other methods, we observe instabilities in the solution at $t=10$. The greedy, the cotangent lift and the complex SVD method, on the other hand, generate a stable reduced system that accurately approximates the solution of the full model.

{\edit In Figure \ref{fig:NuRe:4b} we observe that the symplectic methods preserve the Hamiltonian function, unlike the POD and the DEIM methods. We emphasise that using the reduced basis, obtained by the greedy, together with the DEIM (purple line) does not preserve the symplectic structure as suggested in this figure.}

Figure \ref{fig:NuRe:4c} illustrates the $L^2$-error between the solution of the full model with the reduced systems, generated by different methods. We first observe that symplectic methods yield a lower computational error {\edit when} compared to non-symplectic methods. Secondly, we observe that although the reduced systems from the cotangent lift and the complex SVD are of the same size, their accuracy is different by an order of magnitude. We notice that the greedy algorithm is slightly less accurate than the cotangent lift method while its offline computational cost is reduced to 20\% {\edit when compared to the cotangent lift}. Lastly we notice that the combination of the greedy reduced basis and DEIM yields large errors in the solution while the solution using the symplectic DEIM is very accurate. We note that the symplectic DEIM is even more accurate than the greedy itself since it has been enriched by the nonlinear snapshots.

\subsection{Numerical Convergence} {\edit In this section we discuss the numerical convergence of the symplectic greedy method introduced in Section \ref{chap:SyMo:1}. The exponential convergence properties of the conventional greedy \cite{Quarteroni:2016wi} is presented in \cite{Buffa:2012iz,Binev:2011fj}. Theorem \ref{theorem:SyMo:2} suggests that the symplectic greedy method has similar properties. To illustrate this we compare the convergence of the conventional greedy with the convergence of the symplectic greedy method through the numerical simulations in Sections \ref{chap:NuRe:1.1} and \ref{chap:NuRe:1.2}.

The decay of the singular values of the snapshot matrix for the parametric wave equation and the nonlinear Schr\"odinger equation are given in Figure \ref{fig:NuRe:5}. The decay rate of the singular values is a strong indicator for the decay rate of the Kolmogorov $n$-width of the solution manifold. We expect that the conventional greedy method and the symplectic greedy method provide a similar rate in the decay of the error.

Figure \ref{fig:NuRe:5} shows the maximum $L^2$ error between the original system and the reduced system at each iteration of different greedy methods. In this figure we find the conventional greedy with orthogonal projection error as a basis selection criterion (orange), the symplectic greedy method with a symplectic projection error as a basis selection criterion (green), and the symplectic greedy method with energy loss $\Delta H$ as a basis selection criterion (red).

It is observed that the decay rate of the error for greedy with the orthogonal projection and the greedy with the symplectic projection is similar to the decay of the singular values. This matches our expectation from Theorem \ref{theorem:SyMo:2}. We also notice that the greedy method with the loss in Hamiltonian provides an excellent error indication as a basis selection criterion.
}

\begin{figure}

\begin{minipage}{.5\linewidth}
\centering
\subfloat[]{\label{fig:NuRe:5a}\includegraphics[width=\textwidth]{./figs/wave/convergence}}
\end{minipage}%
\begin{minipage}{.5\linewidth}
\centering
\subfloat[]{\label{fig:NuRe:5b}\includegraphics[width=\textwidth]{./figs/schrodinger/convergence}}
\end{minipage}\par\medskip
\centering
%\subfloat[]{\label{fig:NuRe:4c}\includegraphics[width=0.5\textwidth]{./figs/schrodinger/error}}

\caption{ (a) Convergence of the greedy method for the wave equation. (b) Convergence of the greedy method for the nonlinear Schr\"odinger equation equation. }
\label{fig:NuRe:5}
\end{figure}

\section{Conclusion} \label{chap:Con:1}

In this paper, we present a greedy approach for the construction of a reduced system that preserves the geometric structure of Hamiltonian systems. An iteration of the greedy method comprises searching the parameter space using the error in the Hamiltonian, to find the best basis vectors that increase the overall accuracy of the reduced basis. We argue that for a compact subset with exponentially small Kolmogorov $n$-width we recover exponentially fast convergence of the greedy algorithm. For fast approximation of nonlinear terms, the basis obtained by the greedy was combined with a symplectic DEIM to construct {\edit a reduced system with a Hamiltonian that is arbitrary close to the Hamiltonian of the original system.}

The numerical results demonstrate that the greedy method can save substantial computational cost in the offline stage as compared to alternative SVD-based techniques. Also since the reduced system obtained by the greedy method is Hamiltonian, the greedy method yields a stable reduced system. Symplectic DEIM effectively reduces computational cost of approximating nonlinear terms while preserving stability and symplectic structure. Hence, the greedy method is an efficient model reduction technique that provides an accurate and stable reduced system for large-scale parametric Hamiltonian systems.

\section*{Acknowledgments}
{\edit We would like to thank the referees for providing us with very useful comments which served to improve the paper.}

\bibliographystyle{siamplain}
\bibliography{ref}
\end{document}

### 2017-09-11 greedy_hamiltonian_second_revision/chapters/3.ham.tex
final version of the greedy hamiltonian uploaded

@@
 Common numerical integrators e.g., Runge-Kutta methods, do not generally preserve the Hamiltonian which results in a qualitative wrong behavior of the solution \cite{Hairer:1250576}.
-Symplectic integrators are a class of numerical integrators for Hamiltonian systems that preserve the symplectic structure and ensure stability in long-time integration.} The Str\"omer-Verlet time stepping scheme is an example of symplectic integrators and is given by
+Symplectic integrators are a class of numerical integrators for Hamiltonian systems that preserve the symplectic structure and ensure stability in long-time integration.} The St\"ormer-Verlet time stepping scheme is an example of symplectic integrators and is given by
 \begin{equation} \label{eq:Hasy:13}
@@
 \end{equation}
-For a general Hamiltonian system, the Str\"omer-Verlet scheme is implicit.
+For a general Hamiltonian system, the St\"ormer-Verlet scheme is implicit.
 However, for separable Hamiltonians, i.e.

### 2017-09-11 greedy_hamiltonian_second_revision/chapters/5.numres.tex
final version of the greedy hamiltonian uploaded

@@
 
-For numerical time integration we {\edit use} the Str\"omer-Verlet (\ref{eq:Hasy:13}) scheme, {\edit which is explicit since} the Hamiltonian is separable for the linear wave-equation.
+For numerical time integration we {\edit use} the St\"ormer-Verlet (\ref{eq:Hasy:13}) scheme, {\edit which is explicit since} the Hamiltonian is separable for the linear wave-equation.
 The full model uses the following parameter set \vspace{0.5cm}
@@
 All reduced systems are taken to have an identical size ($k=80$ for POD and $k=40$ for the symplectic methods).
-We use the {\edit Str\"omer-Verlet} scheme for symplectic methods and a second order Runge-Kutta method for the POD.
+We use the {\edit St\"ormer-Verlet} scheme for symplectic methods and a second order Runge-Kutta method for the POD.
 {\edit The choice of different time integration routines is due to the fact that the POD destroys the canonical form of the original equations and a symplectic integrator cannot be applied.
@@
 \end{equation}
-and use a Str\"omer-Verlet (\ref{eq:Hasy:13}) scheme for time integration.
+and use a St\"ormer-Verlet (\ref{eq:Hasy:13}) scheme for time integration.
 Since the Hamiltonian function (\ref{eq:NuRe:17}) is non-separable, this scheme becomes implicit so in each time iteration, a system of nonlinear equations is solved using Newton's iteration.

### 2017-12-05 response_dissipative_hamiltonian/response.tex
dissipative Hmiltonian revision submitted

\documentclass[a4paper]{article}
\usepackage{url}
\usepackage[margin=1in]{geometry}
\newcommand{\breview}{\begin{quotation}\begin{bf}\noindent}
\newcommand{\ereview}{\end{bf}\end{quotation}}
\newcommand{\reviewbullet}[1]{\breview \begin{itemize}\item #1 \end{itemize}\ereview}

\usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}

\begin{document}

Dear Editor, \\[1cm]

Bellow you will find the response to the review of the manuscript \#JOMP-D-17-00165, titled ``Structure-Preserving Model-Reduction of Dissipative Hamiltonian Systems".

Yours Sincerely,\\[1.0cm]
Babak Maboudi Afkham and
Jan S. Hesthaven
\\[1cm]

\section*{Response to the Review}

\breview
1- In the abstract, the authors emphasize that this method ``allows the reduced system to be integrated with a symplectic integrator, resulting in a correct dissipation of energy, preservation of the total energy and, ultimately, in the stability of the solution.''

I agree that by applying a symplectic integrator, the reduced system preserves Hex (the Hamiltonian of the TDD system) as time evolution. However, I am not convinced that the reduced system has stability guarantee. The preservation of Hex (defined in Equation 18) is merely a necessary condition of stability, not an efficient one. In general, a  Hamiltonian system does not necessarily to be stable unless certain conditions can be satisfied (such as conditions in Theorem 3). Without a clarification of the merits of this method with more strict language, I can not recommend the publication of the material in its current form. 
\ereview

Stability is a very important aspect, not only in the context of model reduction, but also in time integration of Hamiltonian systems. Therefore the concern of the reviewer with stability is understandable and appreciated.

Although model reduction of unstable Hamiltonian system is very interesting, e.g. to study Hamiltonian systems containing bifurcation or chaotic behaviour, this paper assumes stability in the high fidelity system prior to model reduction. This step is an important step in the model reduction of Hamiltonian systems since standard techniques such as POD do not guarantee a stable reduced system, even if the high fidelity system is stable as seen in [20,22,23].

To address the concern of the reviewer, we have added Theorem 4 and Theorem 5 that discuss the stability of the reduced system starting from a stable and bounded high fidelity system. These two theorems show that under proper assumptions on the high fidelity system, the solution of the reduced system remains bounded and the stable equilibrium points of the extended system remains locally stable. This is a strong indication of the preservation of the stability in the reduced system and is supported by the numerical experiments.

Finally, the claims in the abstract are corrected to remove any confusion of stability and to resemble the evidence presented in this article.

\breview

2- The proof of theorem 2 has already been shown in reference [8]. It should be cited without a proof.

\ereview

The suggestion of the reviewer is applied with the proper citation.

\breview
3- In model reduction community, not everyone is very familiar with the (TDD) system. Thus, in example 1, it would be nice if the authors can clarify how K, f, and $\chi$ in TDD systems be represented in the parameters of the damped harmonic oscillator. If the f in the damped harmonic equation does not correspond the f of the TDD system, it might cause  confusions. 
\ereview

This is corrected in the revision of the article and hopefully the confusion is removed.

\breview
4- In the numerical examples with the RHD framework, which method is used to construct basis? Cotangent lift or the greedy approach method 1 in section 3.2? I would suggest authors test both methods and make a comparison. 
\ereview

The symplectic greedy method and the cotangent lift method has been compared extensively in [20]. The comparison of the two methods, other than restating the findings in [20] (suggested by the numerical experiments), does not add a significantly novel insight into the performance of the RDH method. However, usage of the greedy and the cotangent lift is mentioned clearly in the revision of the article and proper citations has been added to remove possible confusions.

\breview
5- The authors mention a particular form of Hex for the sine-Gordon equation. Is it possible to prove that the conservation of Hex related to the stability for this particular example?
\ereview

The Sine-Gordon equation is a proof of concept, and is brought as an example for model reduction of nonlinear systems with the RDH method. However, model reduction of general nonlinear dissipative systems requires a deeper study and is left for future works. The suggestion of the reviewer, although very interesting, is difficult to address and is beyond the scope of the paper.

\end{document}

### 2017-12-05 dissipative_hamiltonian_revision/chapters/2.Hamil.tex
dissipative Hmiltonian revision submitted

--- 2017-03-31 dissipative_hamiltonian/chapters/2.Hamil.tex
@@
 Note that without a damping term, (\ref{eq:2.5}) is a Hamiltonian system.
-The TDD formulation for the damped harmonic oscillator takes the form
-\begin{equation} \label{eq:2.6}
-	\dot q(t) = f(t), \quad \dot p(t) = - k q(t), \quad f(t) + \int_0^t r f(s) \ ds = p(t).
+{\edit The TDD formulation for the damped harmonic oscillator takes the form (\ref{eq:2.4}) with
+\begin{equation}
+	z = \begin{pmatrix}
+	q \\ \dot q
+	\end{pmatrix},
+       	\quad K = 
+	\begin{pmatrix}
+	1 & 0 \\
+	0 & \sqrt{k}
+	\end{pmatrix}, \quad \chi(t) = r.
 \end{equation}
-Here $(q,p)$ are the canonical coordinates and the susceptibility is the constant function $r$.
+%\begin{equation} \label{eq:2.6}
+%	\dot q(t) = f(t), \quad \dot p(t) = - k q(t), \quad f(t) + \int_0^t r f(s) \ ds = p(t).
+%\end{equation}
+Here $\mathbb J_{2n} = \mathbb J_{2}$, $(q,\dot q)^T$ is the canonical coordinate and the susceptibility is the constant function $r$.
+}
 \end{example}
@@
 
-\begin{theorem}
+{\edit
+\begin{theorem} \cite{Figotin:2006jy}
 Suppose that $K$ is full rank and $\chi(t)$ is symmetric. Then there is a quadratic extension to (\ref{eq:2.4}) of the form (\ref{eq:2.7}), if
@@
 \end{theorem}
-\emph{Proof.} Here we prove the theorem for the case where $\chi$ is a constant symmetric matrix, where condition (\ref{eq:2.8}) corresponds to $\chi$ being positive semi-definite.
-We refer the reader to \cite{Figotin:2006jy} for the proof of the general case.
-Consider the Hamiltonian system
+When $\chi$ is a constant symmetric matrix, condition (\ref{eq:2.8}) corresponds to $\chi$ being positive semi-definite \cite{Figotin:2006jy}.
+In this case the Hamiltonian extension to (\ref{eq:2.4}) reads
 \begin{subequations}
@@
 		\label{eq:2.10.a} & \dot{z}(t) = \mathbb J_{2n} K^T f(t), \\
-		\label{eq:2.10.c} & \partial_t \phi(t,x) = \theta(t,x), \\
-		\label{eq:2.10.b} & \partial_t \theta(t,x) = \partial_x^2 \phi(t,x) + \sqrt 2 \delta_0(x) \cdot \sqrt{\chi}  f(t), 
+		\label{eq:2.10.b} & \partial_t \phi(t,x) = \theta(t,x), \\
+		\label{eq:2.10.c} & \partial_t \theta(t,x) = \partial_x^2 \phi(t,x) + \sqrt 2 \delta_0(x) \cdot \sqrt{\chi}  f(t), 
 \end{align}
@@
 \end{equation}
-To show that the Hamiltonian system (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) is an extension to (\ref{eq:2.4}) in the sense discussed above, it is enough to show that the solution $f$ to equation (\ref{eq:2.11}) also satisfies (\ref{eq:2.4.1}).
-Equations (\ref{eq:2.10.c}) and (\ref{eq:2.10.b}) are equations for a vibrating string, and can be solved analytically
+Equations (\ref{eq:2.10.b}) and (\ref{eq:2.10.c}) are equations for a vibrating string, and can be solved analytically
 \begin{equation} \label{eq:2.12}
@@
 \end{equation}
-Then, we recover (\ref{eq:2.4.1}) by substituting (\ref{eq:2.12}) into (\ref{eq:2.11}).
+}We can recover (\ref{eq:2.4.1}) by substituting (\ref{eq:2.12}) into (\ref{eq:2.11}).
 The extended Hamiltonian $H_\text{ex}$ for the system (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) takes the quadratic from
 \begin{equation} \label{eq:2.13}
-	H_\text{ex}(z,\phi,\theta) = \frac 1 2 \left( \| Kz - \phi(t,0) \|_2^2 + \| \theta(t) \|^2_{\mathcal H^{2n} } + \| \partial_x\phi(t)\|^2_{\mathcal H^{2n} }\right)
+	H_\text{ex}(z,\phi,\theta) = \frac 1 2 \left( \| Kz - \phi(t,0) \|_2^2 + \| \theta(t) \|^2_{\mathcal H^{2n} } + \| \partial_x\phi(t)\|^2_{\mathcal H^{2n} }\right),
 \end{equation}
@@
 
-Equations (\ref{eq:2.10.c}) and (\ref{eq:2.10.b}) are called the \emph{hidden strings}.
+Equations (\ref{eq:2.10.b}) and (\ref{eq:2.10.c}) are called the \emph{hidden strings}.
 The dissipation of energy in the original system (\ref{eq:2.4}) is carried away, as vibrations, along the added strings making the extended system conservative.

### 2017-12-05 dissipative_hamiltonian_revision/chapters/3.model.tex
dissipative Hmiltonian revision submitted

--- 2017-03-31 dissipative_hamiltonian/chapters/3.model.tex
@@
 	\dot {y}(t) &= \mathbb J_{2k} \nabla_{y} \tilde H, \\
-	y(0) &= A^+ z_0,
+	y(0) &= A^+ z_0.
 	\end{aligned}
@@
 Suppose that at the $k$-th generic step of the greedy basis selection, an ortho-symplectic basis $A_{2k} = \{ e_1,\dots,e_k,f_1,\dots,f_k\}$ of size $2k$ is provided.
-The first step in an iteration of the greedy basis selection comprises finding $t^{k+1}\in[0,T]$ such $z(t^{k+1})$ is worst approximated by the current reduced space.
+The first step in an iteration of the greedy basis selection comprises finding $t^{k+1}\in[0,T]$ such that $z(t^{k+1})$ is worst approximated by the current reduced space.
 In other words
 \begin{equation} \label{eq:3.4}
-	t^{k+1} := \underset{t\in [0,T]}{\text{argmax }} \| z(t) - A_{2k}{A_{2k}}^+z(t) \|.
+	t^{k+1} := \underset{t\in [0,T]}{\text{argmax }} \| z(t) - A_{2k}{A_{2k}}^+z(t) \|_2.
 \end{equation}
@@
 \item $k \leftarrow 1$
-\item \textbf{while} $\| z(t) - A_{2k}{A_{2k}}^+z(t) \| > \delta$ for all $t \in [0,T]$
-\item \hspace{0.5cm} $t^{k+1} := \underset{t\in [0,T]}{\text{argmax }} \| z(t) - A_{2k}{A_{2k}}^+z(t) \|$
+\item \textbf{while} $\| z(t) - A_{2k}{A_{2k}}^+z(t) \|_2 > \delta$ for all $t \in [0,T]$
+\item \hspace{0.5cm} $t^{k+1} := \underset{t\in [0,T]}{\text{argmax }} \| z(t) - A_{2k}{A_{2k}}^+z(t) \|_2$
 \item \hspace{0.5cm} $\mathbb J_{2n}$-orthogonalize $ z(t^{k+1})$ to obtain $e_{k+1}$
@@
 		\label{eq:3.12.a} & \dot{y}(t) = \mathbb J_{2k} {\tilde L}^T \tilde f(t), \\
-		\label{eq:3.12.c} & \partial_t \tilde \phi(t,x) = \tilde \theta(t,x),\\
-		\label{eq:3.12.b} & \partial_t \tilde \theta(t,x) = \partial_x^2 \tilde \phi(t,x) + \sqrt 2 \delta_0(x) \cdot \sqrt{\tilde \chi}  \tilde f(t),
+		\label{eq:3.12.b} & \partial_t \tilde \phi(t,x) = \tilde \theta(t,x),\\
+		\label{eq:3.12.c} & \partial_t \tilde \theta(t,x) = \partial_x^2 \tilde \phi(t,x) + \sqrt 2 \delta_0(x) \cdot \sqrt{\tilde \chi}  \tilde f(t),
 \end{align}
@@
 Equations (\ref{eq:3.12.a})-(\ref{eq:3.12.b}) is a Hamiltonian system on the symplectic linear vector space $\mathbb R^{2k} \times \mathcal H^{2k}$ and contributes to the \emph{reduced TDD system}
-\begin{equation}
+\begin{equation} \label{eq:3.14}
 	\dot {y} = \mathbb J_{2k} \tilde L^T \tilde f(t), \quad \tilde f(t) + \int_0^t \tilde \chi\cdot \tilde f(s)\ ds = \tilde L y.
@@
 
-We point out that the transformation that connects (\ref{eq:2.10.a})-(\ref{eq:2.10.b}) to (\ref{eq:3.12.a})-(\ref{eq:3.12.b}) is given by
+We point out that the transformation that connects (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) to (\ref{eq:3.12.a})-(\ref{eq:3.12.c}) is given by
 \begin{equation}
@@
 \begin{enumerate}
-	\item Construct the Hamiltonian extension (\ref{eq:2.10.a})-(\ref{eq:2.10.b}) to the original TDD system (\ref{eq:2.4}).
+	\item Construct the Hamiltonian extension (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) to the original TDD system (\ref{eq:2.4}).
 	\item Collect the snapshots $z(t_i)$, $i=1,\dots,N$ through time integration of the extended Hamiltonian.
 	\item Construct an ortho-symplectic basis $A$.
-	\item Define $\tilde L = A^T L A$, $\tilde \chi = A^T \chi A$ and construct the reduced dissipative Hamiltonian system (\ref{eq:3.12.a})-(\ref{eq:3.12.b})
+	\item Define $\tilde L = A^T L A$, $\tilde \chi = A^T \chi A$ and construct the reduced dissipative Hamiltonian system (\ref{eq:3.12.a})-(\ref{eq:3.12.c})
 \end{enumerate}
@@
 \cite{peng2016geometric,Polyuga:2010gj,beattie2011structure}, either integrate the reduced system with a non-symplectic integrator, or do not construct a closed reduced system which result in accumulation of local errors or unstable solution during long time integration, respectively \cite{Hairer:1250576}.
+
+{\edit The following theorem provides an indication for the boundedness of the reduced system.
+\begin{theorem}
+Consider the TDD system (\ref{eq:2.4}) together with the initial condition (\ref{eq:2.10.1}).
+If in the absence of dissipation $(\chi = 0)$, a bounded open neighborhood $U\subset \mathbb R^{2n}$ of $z_0$ exists such that $H(z)<c$, for some constant value $c > H(z_0)$, then the RDH method yields a bounded reduced system.
+\end{theorem}
+\begin{proof} For a Hamiltonian of the form (\ref{eq:3.7}), $H(z)\geq 0$ for all $z\in \mathbb R^{2n}$.
+It is shown in \cite{Figotin:2005,Figotin:2006jy} that for a symmetric and positive definite $\chi$, $\frac d {dt} H(z(t)) \leq 0$.
+Therefore, for all $0\leq t <\infty$, $H(z) < c$ which implies $z(t) \in U$.
+Now suppose that the symplectic basis $A$ is used to construct the RDH reduced system (\ref{eq:3.12.a})-(\ref{eq:3.12.c}) such that $y_0 = A^+z_0$.
+It is easily checked that $V = \text{Range}(A)\cap U$ is a bounded open neighborhood of $z_0$ and that $H(z) = H(Ay) < c$ for all $z \in V$.
+Since $\tilde H(y) = H(Ay)$ is a Hamiltonian function for the dissipative Hamiltonian system (\ref{eq:3.14}) and that $\tilde \chi$ is symmetric and positive definite, then $\frac d {dt} \tilde H(y) \leq0$.
+This implies that $\tilde H(y)<c$ and $y(t)\in U$ for all $0\leq t <\infty$.
+Also it is easy to check that the boundedness of the solution of the original TDD system implies boundedness of the extended system in $\mathbb R^{2n}\times \mathcal H^{2n}$ with respect to the norm $\| (x,f) \|_{\text{ex}} = \|x\|_2 + \| f \|_{\mathcal H ^{2n}}$.
+\end{proof}
+
+Note that in practice, ensuring $\frac d {dt} H(z(t)) \leq 0$, requires a careful time integration of the dissipative system.
+The symplectic time integration of the extended system guarantees the correct evolution of the Hamiltonian.
+
+We conclude this section by showing that the RDH method preserves the stability of the equilibrium points of the extended system.
+This provides a strong indication that the overall dynamics and stability of the original system system is translated into the RDH system.
+
+\begin{definition}
+Consider a dynamical system of the form $\dot z = f( z)$ and suppose that $z_e$ is an equilibrium point for the system so that $ f( z_e) = 0$.
+$z_e$ is called Lyapunov stable if, for any $\epsilon > 0$, we can find $\delta > 0$ such that for any trajectory $\psi_t$, if $\| \psi_0 - z_e \|_{\text{ex}} \leq \delta$, then for all $0 \leq t < \infty$, we have $\| \psi_t - z_e \|_{\text{ex}} < \epsilon$.
+\end{definition}
+
+The Dirichlet's theorem \cite{bhatia2002stability}, states that the sufficient condition for an equilibrium point to be stable is the existence of a Lyapunov function \cite{bhatia2002stability}.
+In the context of the Hamiltonian systems, the Hamiltonian is a candidate for the Lyapunov function.
+
+\begin{theorem}
+Consider the Hamiltonian system (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) together with the reduced system (\ref{eq:3.12.a})-(\ref{eq:3.12.c}).
+Suppose that $(z_e,\phi_e,\theta_e)$ is an equilibrium point of (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) and that $z_e$ is in the range of the symplectic basis $A$.
+Further, suppose that $H_{\text{ex}}$ is a Lyapunov function ($\nabla^2_{(z,\phi,\theta)}H_{\text{ex}}$ is a positive definite operator).
+Then $(z_e,\phi_e,\theta_e)$ and $(A^+z_e,A^+\phi_e,A^+\theta_e)$ are lyapunov stable points of (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) and (\ref{eq:2.10.a})-(\ref{eq:2.10.c}), respectively.
+\end{theorem}
+\begin{proof} It is easily checked that $(z_e,\phi_e,\theta_e)$ is a local minimum of $H_{\text{ex}}(z,\phi,\theta)$ and a lypunov stable point for (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) and also $(A^+z_e,A^+\phi_e,A^+\theta_e)$ is a local minimum of $\tilde H(y,\tilde \phi,\tilde \theta) = H_{\text{ex}}(Ay,A\tilde \phi,A\tilde \theta)$.
+We notice that
+\begin{equation}
+	\nabla^2_{(y,\tilde \phi,\tilde \theta)}\tilde H_{\text{ex}} = \mathcal A^T \nabla^2_{(z,\phi,\theta)}H_{\text{ex}} \mathcal A,
+\end{equation}
+is positive definite.
+Therefore, $\tilde H$ is a lyapunov function for (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) and $(A^+z_e,A^+\phi_e,A^+\theta_e)$ is a Lyapunov stable point of  (\ref{eq:2.10.a})-(\ref{eq:2.10.c}).
+\end{proof}
+}

### 2017-12-05 dissipative_hamiltonian_revision/chapters/4.numer.tex
dissipative Hmiltonian revision submitted

--- 2017-03-31 dissipative_hamiltonian/chapters/4.numer.tex
@@
 
-We construct the RDH reduced system according to the Algorithm \ref{alg:3.1}.
-The performance of the method is then compared to the POD and the method proposed in \cite{peng2016geometric}, referred to as the PSD.
+{\edit
+We construct the RDH reduced system according to the Algorithm \ref{alg:3.1} using a basis constructed by the greedy basis selection described in Algorithm \ref{alg:3.1}.
+The performance of the method is then compared to the POD and the PSD proposed in \cite{peng2016geometric}.
+The PSD method constructs a symplectic basis using the cotangent lift method \cite{peng2016geometric}.
+Note the cotangent lift method can also be used to construct a basis for the RDH method.
+However, the greedy method and the cotangent lift yield very similar results, inline with the results in previous works \cite{Maboudi:2016}.
+}
 
@@
 Furthermore we notice that error for the PSD and the RDH coincide as the two methods become identical as $\| \chi \|_{\infty}\to 0$.
+{\edit Note that in this case the basis for the RDH method and PSD method are both generated using the cotangent lift method in order to show the convergence}.
 

### 2017-12-05 dissipative_hamiltonian_revision/chapters/5.conc.tex
dissipative Hmiltonian revision submitted

\section{Conclusion} \label{sec:5}

In this paper we present the Reduced Dissipative Hamiltonian method. The method preserves the symplectic structure of dissipative Hamiltonian systems and guarantees the correct dissipation of energy through time integration. The RDH method couples the reduced system with a canonical heat bath such that the reduced system forms a closed system.

The main advantage of the RDH method compared to the existing methods is that it enables the reduced system to be integrated using a symplectic integrator which naturally preserves the Hamiltonian structure and the symplectic symmetry of the Hamiltonian systems. Applying a symplectic integrator to a non-conservative system or using a non-symplectic integrator for the reduced system can cause accumulation of local errors or wrong qualitative solution over long-time integration, respectively.

The numerical simulations illustrate that the RDH method preserves the system energy with significantly higher accuracy than other methods. Furthermore, it is shown that the hidden strings assure that the dissipation of energy in the reduce system mimics the dissipation of energy in the full system. This ensures that the local error do not accumulate over long-time integration.

### 2017-12-05 dissipative_hamiltonian_revision/paper.tex
dissipative Hmiltonian revision submitted

--- 2017-03-25 dissipative_hamiltonian/paper.tex
@@
 \usepackage{amsmath}
-%\usepackage{amsthm}
 \usepackage{amssymb}
+\let\proof\relax
+\let\endproof\relax
+\usepackage{amsthm}
 \usepackage{color}
@@
 \begin{document}
+
+\newcommand{\edit}{\color{black}}
 
@@
 The method constructs a closed reduced Hamiltonian system by coupling the full model with a canonical heat bath.
-This results in a correct dissipation of energy, preservation of total energy and, ultimately, in the stability of the solution.
+This allows the reduced system to be integrated with a symplectic integrator, resulting in a correct dissipation of energy, preservation of the total energy and, ultimately, {\edit it helps conserving the} stability of the solution.
 Accuracy and stability of the method are illustrated through the numerical simulation of the dissipative wave equation and a port-Hamiltonian model of an electric circuit.

### 2018-07-12 conservative_mor/paper.tex
new paper added

%%%%%%%%%%%%%%%%%%%% author.tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% sample root file for your "contribution" to a contributed volume
%
% Use this file as a template for your own input.
%
%%%%%%%%%%%%%%%% Springer %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% RECOMMENDED %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[graybox]{svmult}

% choose options for [] as required from the list
% in the Reference Guide

\usepackage{mathptmx}       % selects Times Roman as basic font
\usepackage{helvet}         % selects Helvetica as sans-serif font
\usepackage{courier}        % selects Courier as typewriter font
\usepackage{type1cm}        % activate if the above 3 fonts are
                            % not available on your system
%
\usepackage{makeidx}         % allows index generation
\usepackage{graphicx}        % standard LaTeX graphics tool
                             % when including figure files
\usepackage{subfigure}
\usepackage{epstopdf}

\usepackage{multicol}        % used for the two-column index
\usepackage[bottom]{footmisc}% places footnotes at page bottom
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
%\usepackage[linesnumbered, ruled, vlined]{algorithm2e}
%\usepackage{algorithm}% http://ctan.org/pkg/algorithms
%\usepackage{algorithmicx} 
%\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
%\renewcommand{\algorithmicrequire}{\textbf{Input:}}
%\renewcommand{\algorithmicensure}{\textbf{Output:}}
\usepackage{enumerate}
\usepackage{multirow}
\usepackage{cleveref}

\usepackage[hang, small,labelfont=bf,up,textfont=it,up]{caption} % Custom captions under/above floats in tables or figures
\usepackage{booktabs} % Horizontal rules in tables
\usepackage{float} % Required for tables and figures in the multi-column environment - they need to be placed in specific locations with the [H] (e.g. \begin{table}[H])
%\usepackage{hyperref} % For hyperlinks in the PDF

% see the list of further useful packages
% in the Reference Guide

\makeindex             % used for the subject index
                       % please use the style svind.ist with
                       % your makeindex program

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\title*{Model Order Redcution of Fluid Flow While Preserving Quadratic Invariants}
% Use \titlerunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
\author{Bababk Maboudi Afkham, Nicol\`o Ripamonti, Qian Wang, Jan S. Hesthaven and Karen E. Willcox}
% Use \authorrunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
\institute{Name of First Author \at Name, Address of Institute, \email{name@email.address}
\and Name of Second Author \at Name, Address of Institute \email{name@email.address}}
%
% Use the package "url.sty" to avoid
% problems with special characters
% used in your e-mail or web address
%
\maketitle

%\abstract*{Each chapter should be preceded by an abstract (10--15 lines long) that summarizes the content. The abstract will appear \textit{online} at \url{www.SpringerLink.com} and be available with unrestricted access. This allows unregistered users to read the abstract as a teaser for the complete chapter. As a general rule the abstracts will not appear in the printed version of your book unless it is the style of your particular book or that of the series to which your book belongs.
%Please use the 'starred' version of the new Springer \texttt{abstract} command for typesetting the text of the online abstracts (cf. source file of this chapter template \texttt{abstract}) and include them with the source files of your manuscript. Use the plain \texttt{abstract} command if the abstract is also to appear in the printed version of the book.}

\abstract{In the context of model order reduction, conservation of stability in the reduced order systems of partial differential equation, especially with a strong advection term, remains a challenge. Recently preserving structures, invariants and conservation laws, in order to help with stability, and ensure robustness in long-time integrations, has been an area of active research. Energy conservation is a fundamental feature of fluid flows which is often violated in the course of model reduction. In this paper we discuss a model reduction routine, that exploits skew-symmetry of conservative and centered discretization schemes, to recover conservation of energy at the level of reduced system. This results in an, overall, correct evolution of the fluid and ensures robustness of the reduced system. We evaluate the performance of the proposed method through numerical simulation of various fluid flows, and also in a numerical simulation of continuous variable resonance combustor model.}

% sections are added here

\input{./sections/1.intro.tex}
\input{./sections/2.mor.tex}
\input{./sections/3.skew.tex}
\input{./sections/4.mor_skew.tex}
\input{./sections/5.res.tex}
\input{./sections/6.con.tex}

\bibliographystyle{plain}
\bibliography{references}

\end{document}

### 2018-07-12 conservative_mor/sections/1.intro.tex
new paper added

\section{Introduction} \label{sec:intro}

Model order reduction (MOR), and in particular reduced basis (RB) methods, has emerged as a powerful approach to cope with the complex and computationally intensive models in engineering and science. Such techniques, construct a reduced ordered representation for the state of a model which accurately approximates the configuration of the system. The evaluation of the this representation is, then, possible with considerable acceleration.

Although RB methods are successful in reducing the computational complexity of models with elliptic and parabolic partial differential equations (PDEs), MOR of systems of hyperbolic equations, or models with strong advective terms, remains a challenge. Often, such models arise from a set of invariants and conservation laws. Such symmetries are violated over the course of MOR which result in qualitatively wrong and sometimes unstable solution.

Constructing MOR techniques and RB methods that preserve intrinsic structures has recently attracted attention \cite{doi:10.1137/17M1111991,1705.00498,kalashnikova2014stabilization,farhat2015structure,doi:10.1137/110836742,beattie2011structure,doi:10.1137/140978922}. Structure preservation can recover a physically meaningful reduced model, rather than a pure algebraic coupling of equations. This enforces robustness and can help with the stability of reduced model. Preserving time symmetries of Lagrangian, Hamiltonian, and port-Hamiltonian systems can be found in the works of \cite{Carlberg:2014ky,doi:10.1137/140978922,doi:10.1137/17M1111991,1705.00498,beattie2011structure,chaturantabut2016structure,gugercin2012structure}. Conserving the inf-sup stability, in the context of finite element methods, can be found in the works of \cite{farhat2015structure,ballarin2015supremizer}. Furthermore, a flux preserving model reduction for finite volume methods is presented in \cite{carlberg2018conservative}.

Large scale simulation of fluid flows appear in a wide range of disciplines and industries. Therefore, MOR of fluid flows, specially when advective terms are dominant, is of a high importance. It is well known that conservation of the energy, specially kinetic energy, is essential for a qualitatively correct numerical integration of fluid flows. Conventional model reduction techniques often violates conservation of mass, momentum or energy in fluid flows which result in an unstable reduced system, specially over long time integration. The method discussed in \cite{carlberg2018conservative} conserves the mass and momentum for a finite volume discretization scheme. However, a method that also conserves the energy of the fluid flow is not known to the authors.

Skew-symmetric forms for fluid flows, are centered time and space discretization scheme, typically using a finite differences discretization method, that recovers time-symmetries of a fluid at the discrete level. Such discretization schemes are studies comprehensively over the past few decades and can be found in the works of \cite{morinishi2010skew,morinishi1998fully,desjardins2008high,reiss2014conservative,reiss2014conservative} and the references therein. These schemes construct a skew-symmetric differential operator, acting on the momentum vector field, that ensures conservation quadratic invariants, such as energy.

In this paper we discuss how to preserve skew-symmetry of differential operators at the level of reduced system. This results in conservation of quadratic invariants. Furthermore, we show that the reduced system, as a system of coupled differential equations, contains quadratic invariants and an associated energy which approximates the energy of the high-fidelity system. Therefore, a proper time stepping scheme can preserve the reduced energy, and, the loss in energy due to model reduction remains constant in time.

The rest of this paper is organized as follows. In \Cref{sec:mor} we summarize the theory on MOR and introduce proper orthogonal decomposition (POD) and discrete empirical interpolation method (DEIM) as conventional RB methods. We discuss skew-symmetric and conservatives methods for compressible and incompressible fluid flows in \Cref{sec:skew}. Conservative and energy-preserving model reduction of fluid flows is discussed in \Cref{sec:mor_skew}. We evaluate the performance of the method in numerical simulation of incompressible and compressible fluid flow in \Cref{sec:res}. We also apply the method to construct a reduced system for the continuous variable resonance combustor, a one dimensional reaction-diffusion model for a rocket engine. Finally, we present conclusive remarks in \Cref{sec:con}.

### 2018-07-12 conservative_mor/sections/2.mor.tex
new paper added

\section{Model Order Reduction for Time Dependent Problems} \label{sec:mor}

Consider a dynamical system of the form
\begin{equation} \label{eq:1}
	\left\{
	\begin{aligned}
		\frac{d}{dt} u(t) &= f(t,u(t)),\\
		u(0) &= u_0.
	\end{aligned}
	\right.
\end{equation}
Here, $u(t),u_0\in \mathbb R^{n}$ and $f: [0,T]\times \mathbb R^{n} \to \mathbb R^{n}$, for some $T<\infty$, is a Lipschitz function. We may apply the \emph{method of lines} \cite{Edsberg:2008:ICM:1477735} to a system of partial differential equations to obtain a dynamical system of the form \eqref{eq:1}. The \emph{solution manifold} for \eqref{eq:1} is defined as
\begin{equation} \label{eq:2}
	\mathcal M_u := \{ u(t) | t \in [0,T] \}.
\end{equation}
When $\mathcal M_u$ has a low-dimensional representation, it is referred to as \emph{reducible}. Assume that $\mathcal M_u$ can be well approximated by $k$-dimensional linear subspace $\mathcal V_k$, with $k\ll n$ and let $E_k = \{ v_1,\dots,v_k \}$ be the basis vectors for $\mathcal V_k$ and $V_k$ the basis matrix that contains these vectors in its columns. A reduced basis (RB) method assumes that $u \approx \tilde u = V_k v$, where $v$ is the expansion coefficients of $\tilde u$ in the basis $E_k$. Substituting this into \eqref{eq:1} yields
\begin{equation} \label{eq:3}
	V_k \frac{d}{dt} v(t) = f(t,V_{k}v) + r(t,u).	
\end{equation}
Here, $r$ is the error vector in this approximation. The Petrov-Galerkin projection of \eqref{eq:1} onto $\mathcal V_k$ requires $r$ to be orthogonal to a $k$-dimensional subspace $\mathcal W_k$. One can construct a projection operator $P_{\mathcal V_{k},\mathcal W_k}$ that projects elements of $\mathbb R^{n}$ onto $\mathcal V_k$, orthogonal to $\mathcal W_k$ as $P_{\mathcal V_{k},\mathcal W_k} = V_k(W_k^TV_k)^{-1}W_k^T$, where $W_k$ is the basis matrix that contains the basis vectors of $\mathcal W_k$ in its columns and $W_k^TV_k$ is assumed to be invertible. With this projection, \eqref{eq:1} reduces to
\begin{equation} \label{eq:4}
	\left\{
	\begin{aligned}
		\frac{d}{dt} v(t) &= (W_k^TV_k)^{-1} f(t,V_{k}v),\\
		v(0) &= (W_k^TV_k)^{-1}u_0.
	\end{aligned}
	\right.
\end{equation}
When we require $W_k=V_k$, then \eqref{eq:4} is referred to as the \emph{Galerkin} projection of \eqref{eq:1} onto $\mathcal V_k$. Since \eqref{eq:4} has a smaller size, as compared to \eqref{eq:1}, one can expect accelerated evaluation. To numerically identify the best possible subspace $\mathcal V_{k}$ we first discretize the solution manifold to obtain
\begin{equation} \label{eq:5}
	\mathcal M_{u}^{\Delta} = \{ u(t_i) | i\in \{ 1,\dots,N_t \} \}.
\end{equation}
Members of $M_{u}^{\Delta}$ are referred to as \emph{snapshots} of \eqref{eq:1}. One can obtain these snapshots by applying a time-integration scheme, e.g. the Runge-Kutta methods, to \eqref{eq:1} to obtain $\tilde {\mathcal M}_{u}^{\Delta}$ an approximation to $\mathcal M_{u}^{\Delta}$. Throughout this paper, we assume that we can choose $\tilde{\mathcal M}_{u}^{\Delta}$ arbitrary close to $\mathcal M_{u}^{\Delta}$, therefore, by an abuse of notation, we may drop the overscript ``\textasciitilde''. For a Galerkin projection, the best possible basis $V_k$ is the one that minimizes the collective projection error \cite{hesthaven2015certified}, i.e., the solution to the minimization problem
\begin{equation} \label{eq:6}
\begin{aligned}
&  \underset{V_k\in\mathbb R^{n\times k}}{\text{minimize}}
& &  \| S - V_kV_k^TS\|_F, \\
& \text{subject to}
& & V_k^TV_k=I_k.
\end{aligned}
\end{equation}
Here $S$ collects vectors in $\mathcal M_{u}^{\Delta}$ in its columns, referred to as the \emph{snapshot matrix}, $\|\cdot \|_F$ is the Frobenius norm \cite{trefethen97}, and $I_{k}$ is the identity matrix of size $k$. Note that the constraint in \eqref{eq:6} requires $V_k$ to be orthonormal. The basis matrix $V_k$ that solves the minimization problem \eqref{eq:6} is referred to as the \emph{proper orthogonal decomposition} (POD) of $S$ of size $k$ \cite{hesthaven2015certified}, and, according to the Schmidt-Mirsky theorem, can be constructed using the left singular vectors of $S$ as
\begin{equation}
	V_{k} = [u_i]_{i=1}^{k}.
\end{equation}
Here $u_i$, for $i=1,\dots k$, are the first $k$ singular vectors of $S$.

Often MOR is studied in the parametric setting, where the vector $u,u_0$, and the right hand side $f$ of \eqref{eq:1} are of the form $u(t;\mu)$, $u_0(\mu)$, and $f(t,u;\mu)$, respectively. Here $\mu$ belongs to $\mathbb P$ a closed subset of $\mathbb R^d$. In this case, the reduced system can approximate the quantities of interest at an accelerated rate. Since the nature of time, as a parameter, is different from other spacial and physical parameters, in this paper we solely focus on $t$ as the parameter. Nevertheless, it is straight forward to extend the results of this paper to the parameter setting by using POD in time and parameter space, or by using the POD-greedy \cite{haasdonk2013convergence,hesthaven2015certified,quarteroni2015reduced} method to generate a basis $V_k$.

Since the approximated solution $\tilde u$ is a linear combination of the POD basis vectors, $\tilde u$ inherits linear properties of these basis vectors. However, when the solution $u$ to \eqref{eq:1} satisfies some nonlinear invariants, there is no guarantee that, in general, $\tilde u$ also satisfy such invariants \cite{doi:10.1137/140959602,doi:10.1137/140978922,doi:10.1137/17M1111991,MaboudiAfkham2018}. This results in a qualitatively wrong and often unstable solution. In the later sections, we discuss how the skew-symmetric formulation of the fluid flow allows conservation of quadratic invariants, e.g. the kinetic energy, at the level of the reduced system.

### 2018-07-12 conservative_mor/sections/3.skew.tex
new paper added

\section{Skew Symmetric and Centered Schemes for Fluid Flows} \label{sec:skew}

\subsection{Conservation Laws} \label{sec:skew.1}
Transport of conserved quantities, in the context of fluid flows, can be formulated as
\begin{equation} \label{eq:3.1}
	\frac{\partial }{\partial t} \rho \varphi + \nabla \cdot ( \rho u \varphi  ) = \nabla \cdot F_{\varphi}\quad \text{defined in} \quad \Omega \subset \mathbb R^{d}.
\end{equation}
Here, $d = 1,2$ or $3$, $\rho:\Omega\to \mathbb R$ is the density, $u\in \Omega \to \mathbb R^{d}$ is the velocity vector field, $\varphi$ is a measured scalar quantity of the flow, and $F_{\varphi}$ is the flux function associated to $\varphi$. Integration of \eqref{eq:3.1} over $ \Omega$ yields
\begin{equation} \label{eq:3.2}
	\frac{d}{dt} \int_{\Omega} \rho \varphi \ dx = \int_{\partial \Omega} (F_{\varphi} - \rho u \varphi) \cdot \hat n\ ds,
\end{equation}
where $\partial \Omega$ is the boundary of $\Omega$, and $\hat n$ is the unit outward normal vector to $\partial \Omega$. When the right hand side of \eqref{eq:3.2} vanishes, e.g. in case where $\Omega$ is periodic and $\varphi$ is flux-free, the quantity $(\rho \varphi)$ is conserved. Therefore, the form \eqref{eq:3.1} is referred to the \emph{conservative form} and the convective term in \eqref{eq:3.1} is referred to as the \emph{divergence form}. However, using the \emph{continuity equation}
\begin{equation} \label{eq:3.3}
	\frac{\partial }{\partial t} \rho + \nabla \cdot (\rho u) = 0,
\end{equation}
we can rewrite \eqref{eq:3.1} as
\begin{equation} \label{eq:3.4}
	\rho \frac{\partial }{\partial t} \varphi + (\rho u)\cdot \nabla \varphi = \nabla \cdot F_{\varphi}.
\end{equation}
The convective term in this formulation is referred to as the \emph{advective form}. The \emph{skew-symmetric} form for the convective term is obtained by the arithmetic average of the divergent and the advective form:
\begin{equation} \label{eq:3.5}
	\frac{1}{2} \left( \rho \frac{\partial }{\partial t} \varphi + \frac{\partial }{\partial t} (\rho \varphi) \right) + \frac 1 2 \left( (\rho u)\cdot \nabla \varphi + \nabla \cdot (\rho u \varphi) \right) = \nabla \cdot F_{\varphi}.
\end{equation}
Multiplying \eqref{eq:3.5} with $\varphi$ yields
\begin{equation}
	\frac{\partial }{\partial t} \rho \varphi^2 + \nabla \cdot ( \rho u \varphi^2  ) = \varphi \nabla \cdot F_{\varphi}.
\end{equation}
Therefore, $\varphi^2$ is a conserved quantity for a flux-free $\varphi$. Since the divergence, the advective and the skew-symmetric forms are identical at the continuous level, then $\varphi^2$ is a conserved quantity for all forms. However, equivalence of these forms is not preserved through a general discretization scheme and, therefore, we do not expect $\varphi^2$ to be a conserved quantity in discrete level. Assuming proper boundary conditions, the operator 
\begin{equation}
	S_{\rho u}(\cdot) = \frac 1 2 ( [ \nabla \cdot \rho u ] + (\rho u)\cdot \nabla )(\cdot),
\end{equation}
is a skew-adjoint operator with respect to the $L^2$ inner-product. Here, $[\cdot]$ indicates that the inside of brackets act as a differential operator. This skew-adjoint property is used in later sections to show the conservation of some quadratic quantities in \eqref{eq:3.1}. Similarly, we can define the skew-adjoint operator in time variable as
\begin{equation}
	S_{\rho,\partial_t} = \frac{1}{2} \left( \rho \frac{\partial}{\partial t} + [ \frac{\partial}{\partial t} \rho] \right).
\end{equation}
Here, the subscript $\partial t$ is to emphasize that $S_{\rho,\partial_t}$ is a differential operator with respect to $t$. Note that numerical time integration of \eqref{eq:3.5} can be challenging since the time differentiation of different variable is present. However, following derivations in \cite{morinishi2010skew}, one can rewrite \eqref{eq:3.5} as
\begin{equation} \label{eq:3.6}
	\sqrt{\rho } \frac{\partial }{\partial t} (\sqrt \rho \varphi ) + S_{\rho u}(\varphi) = \nabla \cdot F_{\varphi}.
\end{equation}
Time integration of this form is presented in the works of \cite{?}. Note that one can also generate a quasi-skew-symmetric form \cite{blaisdell1991numerical,morinishi2003dns} of \eqref{eq:3.1} as
\begin{equation} \label{eq:3.7}
	\frac{\partial }{\partial t} (\rho \varphi) + S_{\rho u}(\varphi) = \nabla \cdot F_{\varphi}.
\end{equation}
Even though this is not a fully skew-symmetric form (skew-symmetric only in space), the numerical stabilities of this form is significantly better than the divergence and advective form \cite{morinishi2010skew,blaisdell1991numerical,morinishi2003dns}. Note that the quasi-skew-symmetric form is identical to the skew-symmetric form in the incompressible limit.

\subsection{Incompressible Fluid} \label{sec:skew.2}
Consider the governing equations of an incompressible fluid with skew-symmetric convective term:
\begin{equation} \label{eq:3.8}
	\left\{
	\begin{aligned}
	&\nabla \cdot u = 0, \\
	&\frac{\partial}{\partial t} u + S_{u}(u) + \nabla p = \nabla \cdot \tau,
	\end{aligned}
	\right.
\end{equation}
defined on $\Omega$. Here, $p: \Omega \to \mathbb R^+$ is the pressure, $\tau: \Omega \to \mathbb R^{d\times d}$ is the viscous stress tensor, and $S_u = \frac 1 2 ([ \nabla \cdot u] + u\cdot \nabla)$. It is straight forward to check
\begin{equation} \label{eq:3.9}
	\frac{d}{dt} K + \nabla \cdot (Ku) + \nabla \cdot (pu)= \nabla \cdot (\tau u) - (\tau \nabla)\cdot u,
\end{equation}
where $K = \frac 1 2 \sum_{i=1}^d u_i^2 $ is the kinetic energy. Here we used 
\begin{equation} \label{eq:3.10}
	\begin{aligned}
	u\cdot \nabla p &= \nabla \cdot (up) - p\nabla \cdot u, \\
	u \cdot (\nabla \cdot \tau) &= \nabla \cdot (\tau u) - (\tau \nabla)\cdot u,
	\end{aligned} 
\end{equation}
and
\begin{equation} \label{eq:3.11}
	u\cdot S_{u}(u) = \nabla \cdot(Ku).
\end{equation}
The only non-conservative term in \eqref{eq:3.9} is $-(\tau \nabla)\cdot u$, which corresponds to dissipation of kinetic energy. Therefore, in the absence of the viscous terms, $K$ is a conserved quantity of the system, and $\frac d {dt} \int_{\Omega} K \ dx <0$ when $\tau\neq 0$. Note that as long as $\nabla \cdot u = 0$, as discussed in \Cref{sec:skew.1}, the divergence, the convective, and the skew-symmetric forms are identical for the incompressible fluid equation. Thus, kinetic energy is conserved for all forms. However, for a general discretization scheme, these forms are not identical, and often the conservation of kinetic energy, in the discrete sense, is violated.

The skew-symmetric discretization of \eqref{eq:3.8} is a centered scheme that exploits the skew-adjoint property of $S_u$, and ensures conservation of kinetic energy at the discrete level. We uniformly discretize $\Omega$ into $N$ points and denote by $\mathbf u \in \mathbb R^{N\times d}$, $\mathbf p \in \mathbb R^N$, and $T \in R^{N\times d\times d} $ the discrete representation of $u$, $p$, and $\tau$, respectively. Let $D_j$ be the centered finite difference matrix operator for $\partial / \partial x_j$, for $j = 1,\dots,d$. The momentum equation in \eqref{eq:3.8} is then discretized as
\begin{equation} \label{eq:3.13}
	\frac{d}{dt}{\mathbf u}_i + S_{\mathbf u} \mathbf u_i + D_i \mathbf p = \sum_{j=1}^d D_j T_{ij}, \quad i=1,\dots,d.
\end{equation}
where $S_{\mathbf u}$ is the discretization of $S_{u}$ given by
\begin{equation} \label{eq:3.14}
	S_{\mathbf u} = \sum_{j=1}^d D_j U_j + U_j D_j,
\end{equation}
and $U_i$ contains components of $\mathbf u_i$ on its diagonal. We require $D_i$ to satisfy
\begin{enumerate}
	\item $D_i = -D_i^T$
	\item $D_i \mathbf 1 = \mathbf 0$, where $\mathbf 1$ and $\mathbf 0$ are vectors of ones and zeros, respectively.
\end{enumerate}
Condition 1 and 2 yield
\begin{equation} \label{eq:3.15}
	S_{\mathbf u} = -S_{\mathbf u}^T, \quad \mathbf 1^T S_{\mathbf u} \mathbf u_i = 0, \quad i=1,\dots,d.
\end{equation}
Conservation of momentum in the discrete sense reads
\begin{equation} \label{eq:3.16}
	\frac{d}{dt} \sum_{i=1}^d  \mathbf 1^T \mathbf u_i = \sum_{i=1}^d \left( - \mathbf 1^T S_{\mathbf u} \mathbf u_i - \mathbf 1^T D_i \mathbf p + \sum_{j=1}^d \mathbf 1^T D_j T_{ij}  \right) = 0.
\end{equation}
Similarly, using \eqref{eq:3.10}, it is verified that
\begin{equation} \label{eq:3.17}
\frac{d}{dt} \sum_{i=1}^d \left( \frac 1 2 \mathbf u_i^T \mathbf u_i \right) = - \sum_{i,j=1}^d T_{ij}D_j \mathbf u_i \leq 0.
\end{equation}

Conditions 1 and 2 for $D_i$ can be easily checked for a centered finite differences scheme on a periodic domain. For other types of boundaries, e.g., wall boundary and inflow/outflow, we refer the reader to \cite{morinishi1998fully,desjardins2008high} for construction of the proper discrete centered differentiation operator.

\subsection{Compressible Fluid} \label{sec:skew.3}
Consider the equation governing the evolution of a compressible fluid in a skew-symmetric form in one spacial dimension
\begin{equation} \label{eq:3.18}
\left\{
\begin{aligned}
	&\frac{\partial}{\partial t} \rho + \frac{\partial }{\partial x}(\rho u) = 0, \\ 
	& S_{\rho,\partial_t}(u)+ S_{\rho u}(u) + \frac{\partial }{\partial x} p = \frac{\partial }{\partial x} \tau, \\
	&\frac{\partial}{\partial t} \rho E + \frac{\partial}{\partial x}(u E + up) = \frac{\partial }{\partial x}(u\tau - \phi).
\end{aligned}
\right.
\end{equation}
Here $E= e + u^2/2$ is the total energy per unit mass, with $e = \frac{p}{\rho(\gamma - 1)}$ the internal energy, $\gamma$ the adiabatic gas index, and $\phi = -\lambda \frac{\partial T}{\partial x}$ is the heat flux, with $\lambda$ the heat conductivity. The rest of the variables are the same as those discussed in \Cref{sec:skew.2}. Following \cite{reiss2014conservative}, the evolution of the momentum equation reads
\begin{equation} \label{eq:3.19}
	\begin{aligned}
	\frac{\partial}{\partial t}(\frac{\rho u^2}{2}) + \frac{\partial }{\partial x}(\rho u \frac{u^2}{2}) &= \frac 1 2 u( \frac{d}{dt} \rho u + \rho \frac{d}{dt} u ) + \frac 1 2 u ( [ \frac{\partial}{\partial x} \rho u ] u + \rho u \frac{\partial }{\partial x} u ) \\
	& = - u \frac{\partial}{\partial x} p + u \frac{\partial}{\partial x} \tau.
	\end{aligned}
\end{equation}
Substituting this into the energy equation in \eqref{eq:3.18} while assuming a constant adiabatic index yields
\begin{equation} \label{eq:3.20}
	\frac{1}{\gamma -1} \frac{d}{dt} p + \frac{\gamma}{\gamma -1} \frac{\partial }{\partial x} up - u \frac{\partial }{\partial x}(p) = - u \frac{\partial}{\partial x} \tau + \frac{\partial }{\partial x}(u\tau - \phi).
\end{equation}

We discretize the real line, uniformly, into $N$ grid points and denote by $\mathbf r, \mathbf u, \mathbf p \in \mathbb R^{N}$, the discrete representations of $\rho$, $u$, and $p$, respectively. Using the matrix differentiation operator $D\in \mathbb R^{N\times N}$ (we omit the subscript ``$i$'' for the one dimensional case), introduced in \Cref{sec:skew.2}, we define the skew-symmetric matrix operator $S_{\mathbf r \mathbf u} = \frac 1 2 (DUR + RUD)$, where $R$ is the matrix that contains $r$ in its diagonal.Now, the semi-discretization of \eqref{eq:3.18} and \eqref{eq:3.20} takes now takes the form
\begin{equation} \label{eq:3.21}
\left\{
\begin{aligned}
	& \frac{d}{dt} \mathbf r + DU\mathbf r = 0, \\
	& S_{\mathbf r,\partial_t} (\mathbf u) + S_{\mathbf r \mathbf u} \mathbf u + D \mathbf p = D T, \\
	&\frac{1}{\gamma -1} \frac{d}{dt} \mathbf p + \frac{\gamma}{\gamma -1} D U \mathbf p - UD\mathbf p = - UDT + D(UT - \mathbf \phi).
\end{aligned}
\right.
\end{equation}

Considering conditions 1 and 2 for $D$, discussed in \Cref{sec:skew.2}, it is easily verified that
\begin{equation} \label{eq:3.22}
	S_{\mathbf r \mathbf u}^T = - S_{\mathbf r \mathbf u}, \quad \mathbf 1^T S_{\mathbf r \mathbf u} \mathbf u = - \mathbf u^T DU \mathbf r.
\end{equation}
Conservation of mass yields
\begin{equation} \label{eq:3.23}
	\frac{d}{dt} (\mathbf 1^T \mathbf r) = - \mathbf 1^T DR\mathbf u = 0. 
\end{equation}
Furthermore, we recover conservation of momentum in the discrete sense as
\begin{equation} \label{eq:3.24}
\begin{aligned}
	\frac{d}{dt}(\mathbf r^T \mathbf u) &= \frac{1}{2} \frac{d}{dt}(\mathbf r^T \mathbf u) + \frac{1}{2} \left( \mathbf r^T \frac d{dt} \mathbf u +\mathbf u^T \frac{d}{dt} \mathbf r \right)\\
	&= \frac{1}{2}u^T \frac d{dt} \mathbf r + \mathbf 1^T S_{\mathbf r,\partial_t} (\mathbf u) \\
	&= -\frac 1 2 \mathbf u^T DU \mathbf r  - \frac 1 2 \mathbf 1^T S_{\mathbf r \mathbf u} \mathbf u - \mathbf 1^T D\mathbf p +  \mathbf 1^T D T = 0.
\end{aligned}
\end{equation}
Here we used \eqref{eq:3.22} and the mass and the momentum equation in \eqref{eq:3.21}. Similarly, for the conservation of the total energy, we have
\begin{equation} \label{eq:3.25}
\begin{aligned}
	\frac{d}{dt} \left( \frac{1}{\gamma - 1} \mathbf 1^T \mathbf p + \frac 1 2 \mathbf (R\mathbf u)^T \mathbf u  \right) &= \frac{d}{dt} \left( \frac{1}{\gamma - 1} \mathbf 1^T \mathbf p \right) + \frac 1 2 \mathbf u^T  S_{\mathbf r,\partial_t} (\mathbf u) = 0.
\end{aligned}
\end{equation}
In addition to the conservation of the total energy, skew-symmetric form of \eqref{eq:3.21} also conserves the evolutions of the kinetic energy:
\begin{equation} \label{eq:3.26}
\begin{aligned}
	\frac{d}{dt} ( \frac 1 2 \mathbf u^T R\mathbf u) = \frac 1 2 \mathbf u^T S_{\mathbf r,\partial_t} (\mathbf u) &= -\mathbf u ^T S_{\mathbf r \mathbf u} \mathbf u + \mathbf u^T D p + \mathbf u^T DT \\
	&= \mathbf u^T D p + \mathbf u^T DT.
\end{aligned}
\end{equation}
Here, we used the skew-symmetry of $S_{\mathbf r \mathbf u}$. Therefore, only the pressure and viscous terms contribute to the change in the level of kinetic energy.

We point out that there are other methods to obtain a skew-symmetric form for \eqref{sec:skew.3}, that result in the conservation of other quantities. An entropy preserving skew-symmetric form can be found in \cite{sjogreen2010skew}. Furthermore, a fully quasi-skew-symmetric form for \eqref{sec:skew.3}, where all quadratic fluxes are in a skew-symmetric form, is shown to have lowest aliasing error \cite{honein2004higher,honein2005numerical}

\subsection{Time integration}
Following steps in \cite{reiss2014conservative,morinishi2010skew} we can construct a fully discrete scheme for \eqref{sec:skew.3} as
\begin{equation} \label{eq:3.27}
	\left\{
	\begin{aligned}
	&\frac 1 2 \sqrt{\mathbf r} ^{n+1/2} \frac{\sqrt{\mathbf r}^{n+1} - \sqrt{\mathbf r}^{n}}{\Delta t} + DU^{n+1/2} \mathbf r^{n} = 0, \\
	& \sqrt{\mathbf r} ^{n+1/2}  \frac{\sqrt{ \mathbf R}^{n+1} u^{n+1} - \sqrt{\mathbf R}^{n}u^n}{\Delta t} + S_{\mathbf r^{n} \mathbf u^n} \mathbf u^{n+1/2}_\alpha + D \mathbf p^{n} = DT^{n}, \\
	& \frac 1 {\gamma -1} \frac{\mathbf p^{n+1} - \mathbf p^n}{\Delta t} + \frac{\gamma}{\gamma -1} D U^{n} \mathbf p - U^{n} D \mathbf p = - U^{n}D T^{n} + D (U^nT^n - \phi^n).
	\end{aligned}
	\right.
\end{equation}
Here, $\Delta t$ is the time discretization unit, superscript $n$ denotes evaluating the variable at $t = n\Delta t$, superscript $n+1/2$ denotes the arithmetic average of the variable at $t=n\Delta t$, the square root sign denotes element-wise application of square root,  and $t=(n+1)\Delta t$ and 
\begin{equation}
	\mathbf u_{\alpha}^{n+1/2} = \frac{\sqrt{\mathbf R}^{n+1} \mathbf u^{n+1} + \sqrt{\mathbf R}^{n} \mathbf u^{n}}{2\sqrt{\mathbf r}^{n+1/2} }
\end{equation}
It is discussed in \cite{reiss2014conservative} that this time discretization scheme preserves the symmetries expressed in \eqref{eq:3.17}, \eqref{eq:3.24}, \eqref{eq:3.25}, and \eqref{eq:3.26}. In the incompressible case, the method reduces to the implicit mid-point scheme \cite{hairer2006geometric}. For further information see \cite{reiss2014conservative,morinishi2010skew}.

### 2018-07-12 conservative_mor/sections/4.mor_skew.tex
new paper added

\section{Model Reduction of Fluid Flow} \label{sec:mor_skew}
Straight forward model reduction of \eqref{eq:3.8} and \eqref{eq:3.18} does not generally preserves symmetries and conservation laws that were presented in \Cref{sec:skew}. In this section we discuss how to exploit the discrete skew-symmetric structure of \eqref{eq:3.13} and \eqref{eq:3.21} to recover conservation of mass, momentum, and energy at the level of reduced system.

Let $V_{\mathbf r}$, $V_{\mathbf r \mathbf u}$, and $V_{\mathbf u_i}$ be the reduced bases for the snapshots of $\mathbf r$, $R \mathbf u$, and $\mathbf u$, respectively. For the purpose of simplicity, we assume that all bases have the size $k$. In case of an incompressible fluid, $V_{ \mathbf r}$ and $V_{\mathbf r \mathbf u}$ are omitted. We seek to project $S_{\mathbf u}$ and $S_{\mathbf r \mathbf u}$ onto the reduced space, such that the projection preserves the skew-symmetry property. The projected operators, using a Galerkin projection, read
\begin{equation} \label{eq:4.1}
	S^r _{\mathbf u} = V_{ \mathbf u_i}^T S _{\mathbf u} V_{ \mathbf u_i}, \quad i=1,\dots,d,
\end{equation}
and
\begin{equation} \label{eq:4.2}
	S^r_{\mathbf r ,\partial_t} =V_{\mathbf r \mathbf u}^T  S_{\mathbf r ,\partial_t} V_{\mathbf u}, \quad S^r _{\mathbf r \mathbf u} = V_{\mathbf r \mathbf u}^T  S _{\mathbf r \mathbf u} V_{\mathbf u}.
\end{equation}
Note that $S^r_{\mathbf r ,\partial_t}$ is not computed explicitly. It is seen that the $S^r _{\mathbf u}$ is already in a skew-symmetric form. On the other hand, $S^r_{\mathbf r ,\partial_t}$ and $S^r _{\mathbf r \mathbf u}$ are not, in general, skew-adjoint and skew-symmetric, respectively. This can be ensured, however, by requiring $V_{\mathbf r \mathbf u} = V_{\mathbf u}$. We denote such a basis by $V_{\mathbf r \mathbf u, \mathbf u}$.Using \eqref{eq:4.1} and \eqref{eq:4.2}, the Galerkin projection of momentum equation in \eqref{eq:3.13} and equations governing the compressible fluid in \eqref{eq:3.21} take the form
\begin{equation} \label{eq:4.3}
	\frac{d}{dt} {\mathbf u^r}_i + S^r_{\mathbf u} \mathbf u^r_i + V_{\mathbf u_i} ^T D_i \mathbf p = \sum_{j=1}^d V_{k_3, \mathbf u_i}^T D_j T_{ij}(V_{ \mathbf u_i} \mathbf u^r_i), \quad i=1,\dots,d,
\end{equation}
and
\begin{equation} \label{eq:4.4}
\left\{
\begin{aligned}
	& \frac{d}{dt} \mathbf r^r + \sum_{i=1}^k V^T_{\mathbf r}DU_iV_{\mathbf r}\mathbf r^r = 0, \\
	& S^r_{\mathbf r ,\partial_t} + S^r _{\mathbf r \mathbf u} \mathbf u^r + V_{\mathbf r \mathbf u, \mathbf u}^T D \mathbf V_{\mathbf p} \mathbf p^r = V_{\mathbf r \mathbf u, \mathbf u}^T D T, \\
	&\frac{1}{\gamma -1} \frac{d}{dt} \mathbf p^r + \frac{\gamma}{\gamma -1} \mathbf V_{\mathbf p}^T D U V_{\mathbf p} \mathbf p^r - \mathbf V_{\mathbf p}^T UD V_{\mathbf p} \mathbf p^r = - V_{\mathbf p}^T UDT + V_{\mathbf p}^T D(UT - \mathbf \phi).
\end{aligned}
\right.
\end{equation}
Note that in \eqref{eq:4.4}, dependency of $T$ on $V_{\mathbf r \mathbf u , \mathbf u}$ is not shown for abbreviation. In \eqref{eq:4.3} and \eqref{eq:4.4}, $D_i$ is always multiplied from left with a basis matrix or a diagonal matrix. Therefore, the telescoping sum discussed in condition 2 in \Cref{sec:skew.1} cannot be used to show conservation of mass and momentum. However, POD preserves linear properties of snapshots. To demonstrate, let the overscript ``\textasciitilde'' denote the representation of a reduced variable in the high fidelity space. A reduced variable, e.g. density, can be represented as a linear combination of some snapshots as $\mathbf r \approx \tilde{\mathbf r} = \sum_{i=1}^k c_i \mathbf r_i$, for some snapshots $\mathbf r_i$ and some coefficients $c_i \in \mathbb R$, for $i=1,\dots,k$. The conservation mass for the reduced system reads
\begin{equation} \label{eq:4.5}
	\frac{d}{dt} \mathbf 1^T \tilde {\mathbf r} = \sum_{i=1}^k c_i  \left( \mathbf 1^T \frac{d}{dt} \mathbf r_i \right) = - \sum_{i=1}^k c_i  \left( \mathbf 1^T DR_i\mathbf u_i \right) = 0.
\end{equation}
Here we used the fact that $\mathbf 1^T D = \mathbf 0^T$. Similarly, we recover conservation of momentum
\begin{equation} \label{eq:4.6}
\begin{aligned}
	\frac{d}{dt}(\tilde {\mathbf r}^T \tilde{\mathbf u}) &= \frac{1}{2} \frac{d}{dt}(\tilde{\mathbf r}^T \tilde{\mathbf u}) + \frac{1}{2} \left( \tilde{ \mathbf r }^T \frac d{dt} \tilde{ \mathbf u } + \tilde {\mathbf u}^T \frac{d}{dt} \tilde {\mathbf r} \right)\\
	&= \sum_{i,j=1}^k d_i c_j \left( \mathbf u_i^T \frac{d}{dt} \mathbf r_j + \left( \mathbf r_j^T \frac d{dt} \mathbf u_i +\mathbf u_i^T \frac{d}{dt} \mathbf r_j \right)\right) = 0.\\
\end{aligned}
\end{equation}
Here, $\tilde{\mathbf u} = \sum_{i=1}^k d_i \mathbf u_i$, for some snapshot $\mathbf u_i$ and coefficients $d_i \in \mathbb R$. Denoting by $\{ R \mathbf u\}^r$ the reduced representation of $R\mathbf u$ in basis $V_{\mathbf r \mathbf u , \mathbf u}$, the evolution of kinetic energy reads
\begin{equation} \label{eq:4.7}
	\begin{aligned}
	\frac{d}{dt}\left( \frac 1 2 \tilde{\mathbf u}^T \tilde{ R } \tilde {\mathbf u} \right) &= \frac{d}{dt}\left( \frac 1 2 {\mathbf u^r}^T V_{\mathbf r \mathbf u , \mathbf u}^T V_{\mathbf r \mathbf u , \mathbf u} \{ R \mathbf u\}^r \right) = \frac{d}{dt}\left( \frac 1 2 {\mathbf u^r}^T \{ R \mathbf u\}^r \right) \\
	&= \frac 1 2 \left( {\mathbf u^r}^T \frac{d}{dt} \{ R \mathbf u \}^r + \{ R \mathbf u \}^r \frac{d}{dt} \mathbf u^r \right) \\
	&= \frac 1 2 \left( {\mathbf u^r}^T V_{\mathbf r \mathbf u , \mathbf u}^T V_{\mathbf r \mathbf u , \mathbf u} \frac{d}{dt} \{ R \mathbf u \}^r + \{ R \mathbf u \}^r \frac{d}{dt} V_{\mathbf r \mathbf u , \mathbf u}^T V_{\mathbf r \mathbf u , \mathbf u} \mathbf u^r \right) \\
	&= {\mathbf u^r}^T S^r_{\mathbf r, \partial_t} \mathbf u^r = { \mathbf u^r }^T V_{\mathbf r \mathbf u , \mathbf u} D V_{\mathbf p} \mathbf P^r + { \mathbf u^r }^T V_{\mathbf r \mathbf u , \mathbf u}^T D T.
	\end{aligned}
\end{equation}
In the missing steps in the last line, the skew-symmetry of $S_{\mathbf r \mathbf u}^r$ is used. Note, that only the reduced pressure and the viscous term contribute to the evolution of kinetic energy. Furthermore, the quantity $ \frac 1 2 {\mathbf u^r}^T \{ R \mathbf u\}^r$ is the kinetic energy associated with the reduced system \eqref{eq:4.4} that approximates the kinetic energy of the high-fidelity system \eqref{eq:3.21} and is a quadratic form with respect to the reduced variables. Conservation of kinetic energy for \eqref{eq:4.3} follows similarly. With what discussed above, it is  straight forward to check that
\begin{equation} \label{eq:4.8}
	\frac{d}{dt} \left( \frac{1}{\gamma - 1} \mathbf 1^T \tilde{\mathbf p} + \frac{1}{2} \tilde{\mathbf u}^T \tilde R \tilde{\mathbf u} \right) = 0,
\end{equation}
i.e., the total energy is conserved. We can immediately recognize that $ \mathbf p^r /(\gamma - 1)$ is the total internal energy of the reduced system. However, the internal energy of the system is a weighted sum, $ b^T\mathbf p^r /(\gamma - 1)$, with $b = V_{\mathbf p}^T \mathbf 1$ which is an approximation of the internal energy of \eqref{eq:3.21}. From \eqref{eq:4.5}, \eqref{eq:4.6}, \eqref{eq:4.7}, and \eqref{eq:4.8} we conclude the following proposition.
\begin{proposition}
	The loss in the mass, momentum and energy associated with model reduction in \eqref{eq:4.4} is constant in time, and therefore, bounded.
\end{proposition}

\subsection{Assembling Nonlinear Terms and Time Integration}
Nonlinear terms that appear in \eqref{eq:4.3} and \eqref{eq:4.4} are of quadratic nature. These terms can be evaluated exactly using a set of precomputed matrices as proposed in \cite{Benner2018}. As an example, consider, for simplicity, in one dimension
\begin{equation}
	S_{\mathbf u}^r = V^T_{\mathbf u} ( DU + UD ) V^T_{\mathbf u}.
\end{equation}
We write $U$ as a linear combination of matrices as $U = \sum_{j=1}^k \mathbf u ^r_j U_j$, where $\mathbf u ^r_j$ is the $j$th component of $\mathbf u ^r$, and $U_j$ contains the $j$th column of $V_{\mathbf u}$ on its diagonal. It follows
\begin{equation}
	S_{\mathbf u}^r = \sum_{j=1}^k \mathbf u ^r_j \left( V^T_{\mathbf u} ( DU_j + U_jD ) V^T_{\mathbf u} \right),
\end{equation}
and the matrices $V^T_{\mathbf u} ( DU_j + U_jD ) V^T_{\mathbf u}$ can be computed prior to the time integration of the reduced system. However, the form of the fully discrete system in \eqref{eq:3.27} introduces cubic and even quartic terms. Although, in principle, the same method can be applied to assemble the nonlinear terms, however, the number of precomputed matrices can grow proportional to the order of nonlinear term.

To accelerate the assembly of the nonlinear terms that require a large number of precomputed matrices, we may approximately evaluate them using the discrete empirical interpolation method (DEIM). This approximation can affect the accuracy of conserved quantities in \eqref{eq:4.4}. Therefore, the accuracy of DEIM approximation should be chosen considerably higher than the one from POD.

To integrate \eqref{eq:4.4} in time, the fully discrete system \eqref{eq:3.27} is modified prior to model reduction, by multiplying the mass and momentum equation with $\sqrt{\mathbf r}^{n+1}$. Note that since the new form is identical to \eqref{eq:3.27}, it does not affect the conserved quantities. Subsequently, a basis for $\sqrt{\mathbf r}$, denoted by $V_{\sqrt{\mathbf r}}$, is constructed. The nonlinear terms, are evaluated exactly using the quadratic expansion or approximated using the DEIM.

### 2018-07-12 conservative_mor/sections/5.res.tex
new paper added

\section{Numerical Experiments} \label{sec:res}

\subsection{The incompressible Euler Equation}

\subsection{The Compressible Euler Equation}

\subsection{Continuous Variable Resonance Combustor}

CVRC is a model rocket combustor designed and operated at Purdue University (Indiana, U.S.) to investigate combustion instabilities \cite{yu2008combustion}. This setup is called the Continuously Variable Resonance Combustor (CVRC) because the length of the oxidizer injector can be varied continuously, allowing for a detailed investigation of the coupling between acoustics and combustion in the chamber \cite{garby2013simulations}. The 2D/3D high-fidelity simulations of CVRC are expensive. Thus to get a fast analysis tool, a quasi-1D model has been proposed by Smith et al. \cite{smith2008computational} and further developed by Frezzotti et al. \cite{frezzotti2015determination,frezzotti2017numerical,frezzotti2018quasi}.

The CVRC consists of three parts: oxidizer post, combustion chamber and exit nozzle, as shown in Fig. \ref{fig:radius}. The oxidizer is injected from the left end of the oxidizer post and meets the fuel that is injected through an annular ring around the oxidizer injector, at the back-step. The combustion happens in a region around the back-step. The combustion products flow through the chamber and exit the system from the nozzle.  Both the injector and the nozzle are operated at choked condition during the experiment. The length of the oxidizer post $L_{op}$ of the CVRC can be varied continuously, leading to different behavior of the combustion stability. In this paper, we will focus on the case with $L_{op}= 14.0$ cm, in which the combustion is unstable.

The geometry parameters of the quasi-1D CVRC with a oxidizer post length  $L_{op}= 14.0$ cm are shown in Table \ref{tab:geometry_parameters}.  The back-step and the converging part of the nozzle are sinusoidally contoured to avoid discontinuity of the radius that will invalidate the quasi-1D governing equations presented in the next subsection.

\begin{figure}
	\centering
	\includegraphics[width=0.8\linewidth]{pic/radius}
	\caption{Geometry of quasi-1D CVRC model.}
	\label{fig:radius}
\end{figure}

\begin{table} [h]
	\centering
	\caption{Geometry parameters of the quasi-1D CVRC with an oxidizer post length $L_{op}=14$ cm.}
	\centering
	\begin{tabular}{c c c c c c }
		\toprule
		\centering
		\multirow{2}{*}{Section} &
		\multicolumn{2}{c}{Oxidizer post} &
		\multirow{2}{*}{Chamber} &
		\multicolumn{2}{c}{Nozzle} \\
		\cmidrule(lr){2-3} \cmidrule(lr){5-6}
		& injector & back-step & & converging part & diverging part\\
		\midrule
		Length (cm) & 12.99 & 1.01 & 38.1 & 1.27 & 3.4 \\
		Radius (cm) & 1.02  & $1.02 \sim 2.25$ & 2.25 & $2.25 \sim 1.04$ & $1.04 \sim 1.95$ \\
		\bottomrule
		\label{tab:geometry_parameters}
	\end{tabular} 
\end{table}

The fuel is pure gaseous methane. The oxidizer is a mixture of 42\% oxygen and 58\% water (per unit mass). The oxidizer is injected in the oxidizer post at a temperature $T_{ox}=1030$ K so that both water and oxygen are in the gaseous phase. The operating conditions are listed in Table \ref{tab:operating-conditions}.

\begin{table} [h]
	\centering
	\caption{CVRC operating conditions.}
	\centering
	\begin{tabular}{l l l}
		\toprule
		\centering
		Parameter & Unit & Value \\
		\midrule
		Fuel mass flow rate, $\dot{m}_{f}$ & kg/s & 0.027   \\
		Fuel temperature, $T_{f}$ & K & 300   \\
		Oxidizer mass flow rate, $\dot{m}_{ox}$ & kg/s & 0.32   \\
		Oxidizer temperature, $T_{ox}$ & K & 1030   \\
		$O_2$ mass fraction in oxidizer, $Y_{O_2}$ & -- & 42.4\%   \\
		$H_2O$ mass fraction in oxidizer, $Y_{H_2O}$ & -- & 57.6\%   \\
		Mean chamber pressure & MPa & 1.34 \\
		Equivalence ratio, $E_r$ & -- & 0.8 \\
		\bottomrule
		\label{tab:operating-conditions}
	\end{tabular} 
\end{table}

For the combustion, we consider the one-step reaction model
\begin{equation*}\label{eq:combustion}
CH_4 + 2O_2 \rightarrow CO_2 + 2H_2O
\end{equation*}
We assume that the fuel reacts instantaneously to form products, allowing us to neglect intermediate species and finite reaction rates. As the equivalence ratio is less than one, there is oxidizer left after the combustion. Therefore, only two species need to be considered: oxidizer and combustion products.

The governing equations that describe the conservation of mass, momentum, and energy of the quasi-1D CVRC flow, are the quasi-1D unsteady Euler equations for multiple species, expressed in conservative form as
\begin{equation*}\label{eq:quasi-1D-equation}
\frac{\partial \mathbf{u}}{\partial t} + \frac{\partial \mathbf{f}}{\partial x} = \mathbf{s}_A + \mathbf{s}_f + \mathbf{s}_q.
\end{equation*}
The conserved variable vector $\mathbf{u}$ and the convective flux vector $\mathbf{f}$ are
\begin{equation*}\label{eq:CVRC-u-f}
\mathbf{u}= \left( \begin{gathered}
\rho A  \\
\rho uA  \\
\rho EA  \\
\rho Y_{ox} A \\
\end{gathered} \right), 
\mathbf{f} = \left( \begin{gathered}
\rho uA  \\
\left(\rho u^2 + p\right)A  \\
\left(\rho E + p\right)uA  \\
\rho uY_{ox} A \\
\end{gathered} \right),
\end{equation*}
where $\rho$ is the density, $u$ is the velocity, $p$ is the pressure, $E$ is the total energy, $Y_{ox}$ is the mass fraction of oxidizer, and $A=A(x)$ is the cross section area of the duct. The pressure $p$ can be computed using the conserved variables as
\begin{equation*}\label{eq:total-engery}
E = \frac{p}{\rho (\gamma - 1)} + \frac{u^2}{2} - C_p T_{ref},
\end{equation*}
where $T_{ref}$ is the reference temperature and is set as 298.15 K in this paper. The temperature $T$ is recovered from the equation of state $p = \rho R T$. The gas properties  $C_p$, $R$ and $\gamma$ are computed as $C_p= \sum C_{pi}Y_i$, $R=\sum R_iY_i$ and $ \gamma= C_p/(C_p-R)$, respectively.

The source terms are
\begin{equation*}\label{eq:source-terms}
\mathbf{s}_A = \left( \begin{gathered}
0  \\
p \frac{dA}{dx}  \\
0  \\
0 \\
\end{gathered} \right), 
\mathbf{s}_f = \left( \begin{gathered}
{\dot \omega}_f  \\
{\dot \omega}_f u  \\
{\dot \omega}_f \left(h_{0}^{f} + \Delta h_{0}^{rel} \right)  \\
{\dot \omega}_{ox} \\
\end{gathered} \right), 
{\mathbf{s}_q} = \left( \begin{gathered}
0  \\
0  \\
q'  \\
0 \\
\end{gathered} \right),
\end{equation*}
where $\dot{\omega}_f$ is the depletion rate of the fuel, $\dot{\omega}_{ox}$ is the depletion rate of the oxidizer, $h_0^f$ is the total enthalpy of the fuel, $\Delta h_{0}^{rel}$ is the heat of reaction per unit mass of fuel and $q'$ is the unsteady heat release term. $\mathbf{s}_A$ accounts for area variations, $\mathbf{s}_f$ and $\mathbf{s}_q$ are related to the combustion. $\mathbf{s}_f$ represents the addition of the fuel and its combustion with the oxidizer, which in turn results in the creation of the combustion products. The depletion rate of the fuel is
\begin{equation}\label{eq:wf}
\dot{\omega}_{f}= \frac{k_f \dot{m}_f Y_{ox} \left(1+sin\xi\right)}{l_f-l_s},
\end{equation} 
where
\begin{equation*}\label{eq:xi}
\xi= -\frac{\pi}{2} + 2\pi\frac{x-l_s}{l_f-l_s}, \hspace{0.5cm} \forall \hspace{0.2cm} l_s < x < l_f.
\end{equation*}
The setting of the fuel injection restricts the combustion to the region $l_s < x < l_f$. The reaction constant $k_f$ is selected to insure that the fuel is consumed within the specified combustion zone. The depletion rate of the oxidizer is computed by 
\begin{equation*}\label{eq:wox}
\dot{\omega}_{ox} = C_{o/f} \dot{\omega}_f,
\end{equation*}
where $C_{o/f}$ is the oxidizer-to-fuel ratio.

The unsteady heat release term $q'$, also called the combustion response function, models the coupling between acoustics and combustion. In this paper, we use the combustion response function designed by Frezzotti et al. \cite{frezzotti2017numerical,frezzotti2018quasi}, which is a function of the velocity, sampled at specific abscissa $\hat{x}$ that is almost coincident with the antinode of the first longitudinal modal shape, with a certain time lag $\tau$, i.e.,
\begin{equation}\label{eq:response-function}
q'\left( x,t\right) = \alpha g\left(x\right)  A\left(x\right) \left[ u\left( \hat{x},t-\tau \right) - \bar{u}\left( \hat{x} \right) \right].
\end{equation}
Here $\bar{u}$ is the time averaged velocity, estimated with the steady-state quasi-1D model assuming $q'=0$, and $g(x)$ is a Gaussian distribution  
\begin{equation*}\label{eq:gx}
g\left(x\right)= \frac{e^{-\frac{\left(x-\mu\right)^2}{2\sigma^2}}}{\sqrt{2\pi\sigma^2}},
\end{equation*}
where $\mu$ is the mean and $\sigma$ is the standard deviation. The amount of heat release due to velocity oscillations is controlled by the parameter $\alpha$.

The boundary conditions for the quasi-1D CVRC flow include the fixed mass flow rate and the stagnation temperature at the head-end of the oxidizer injector, and the supersonic outflow at the exit of the nozzle.

Prior to unsteady simulation, the quasi-1D CVRC needs to be excited, which can be achieved by adding a perturbation to the steady-state solution. The perturbation is added by forcing the mass flow rate with a multi-sine signal
\begin{equation}\label{eq:multisine}
\dot{m}_{ox} \left(t\right)= \dot{m}_{ox,0} \left[1 + \delta\sum_{k=1}^{K}  sin\left(2\pi k\Delta f t\right) \right],
\end{equation}
where $\dot{m}_{ox,0}$ is the oxidizer mass flow rate in Table \ref{tab:operating-conditions}, $\Delta f$ is the frequency resolution and $K$ is the number of frequencies. In this paper, $\Delta f = 50 $ Hz and $K=140$, resulting in a minimal frequency of 50 Hz and a maximal frequency of 7000 Hz. $\delta$ is required to be small to control the amplitude of the perturbation and is set as 0.1\%.

The procedure of the unsteady simulation of the quasi-1D CVRC flow includes three steps:
\begin{enumerate}[(1)]
	\item  Compute the steady-state solution by setting $\dot{m}_{ox}=\dot{m}_{ox,0} $ and $q'=0$.
	\item  Excite the system by adding a perturbation to the oxidizer mass flow rate according to (\ref{eq:multisine}) and setting $q'=0$.
	\item  Perform the unsteady simulation by turning on the combustion response function $q'$ in (\ref{eq:response-function}) and turning off the oxidizer mass flow rate perturbation by setting $\dot{m}_{ox}=\dot{m}_{ox,0} $.	
\end{enumerate}

A high-fidelity quasi-1D CVRC flow solver is built by employing the MUSCL reconstruction \cite{delis1998tvd}, the Lax-Friedrichs flux \cite{shu1988efficient} and the strong stability preserving, three-stage Runge-Kutta (SSP RK3) time stepping \cite{jiang1996efficient}.
